# Descriptive Statistics & Exploratory Data Analysis {#descriptive-statistics}

```{r setup}
#| include: false
#| eval: true
source(here::here("R/functions.R"))
source(here::here("R/load-packages.R"))
``` 

This session teaches practical descriptive statistics and exploratory
data analysis (EDA) with a focus on interpretation for clinical
research. Each section shows when to use the method, a short code demo,
how to read the output, and a short exercise for practice.

## When to use EDA

Use EDA as the first analytic step to understand data quality, detect
missingness and outliers, and choose reasonable transformations or
models. Good EDA prevents common modeling mistakes and grounds
interpretation in the actual data.

## Short code demo (dplyr-style)

```{r demo-load, echo=TRUE}
# load and clean a canonical clinical dataset
pbc <- survival::pbc %>%
  as_tibble() %>%
  janitor::clean_names() %>%
  mutate(stage = factor(stage, levels = 1:4))

# counts by stage
pbc %>% count(stage)
```

```{r demo-summary, echo=TRUE}
# descriptive summary by stage (handle missingness)
pbc %>%
  filter(!is.na(bili)) %>%
  group_by(stage) %>%
  summarize(
    n = n(),
    mean_bili = mean(bili, na.rm = TRUE),
    sd_bili = sd(bili, na.rm = TRUE),
    median_bili = median(bili, na.rm = TRUE),
    iqr_bili = IQR(bili, na.rm = TRUE)
  )
```

## How to explain the output (teacher notes)

- **What to report:** show counts, an estimate of central tendency
  (median if skewed), and a measure of spread (IQR) for each group.
- **What it means:** for example, higher median bilirubin in later
  stages suggests worsening liver function; emphasize clinical units
  and whether differences are clinically meaningful, not just
  statistically significant.
- **Implications:** use these findings to guide choice of transformations
  (e.g., log-transform skewed labs), adjust sample-size expectations,
  and plan covariate adjustment in regression.

## Short exercise

- Compute mean, median, IQR and the proportion missing for `bili` by
  `stage`. Write one brief interpretation (2–3 sentences) focusing on
  clinical relevance.

**Hint:** use `group_by(stage)` and `summarize(..., na.rm = TRUE)` to
compute summaries; `sum(is.na(bili))/n()` gives proportion missing.

## Packages used

- `dplyr`, `janitor`, `survival`, `ggplot2` (for plotting in follow-up
  exercises)
