# Linear Regression {#linear-regression}

```{r setup}
#| include: false
#| eval: true
source(here::here("R/functions.R"))
source(here::here("R/load-packages.R"))
```

This short session introduces simple linear regression for continuous outcomes.

## When to use linear regression

Use linear regression to model a continuous outcome as a function of one or
more predictors when the outcome is approximately continuous and residuals are
reasonably well-behaved.

## Short code demo (dplyr-style)

```{r demo-lm, echo=TRUE}
# load example data and fit a simple model
mtcars <- datasets::mtcars %>% as_tibble(rownames = "model")
lm_fit <- lm(mpg ~ wt, data = mtcars)

# tidy coefficient table and model summary
broom::tidy(lm_fit)
broom::glance(lm_fit)
```

## How to interpret the output (teacher notes)

- **Slope:** expected change in the outcome per one-unit increase in the predictor.
- **Intercept:** expected outcome when predictors equal zero (context dependent).
- **R-squared:** fraction of variance explained by the model; interpret with caution.

## Short exercise

- Fit `mpg ~ wt + hp` and report the coefficient for `wt` with a one-sentence interpretation.

**Hint:** use `lm()` and `broom::tidy()`; consider whether effect sizes are clinically meaningful.

## Packages used

- `dplyr`, `broom`, `ggplot2`

**Exercises**

- See `dev_code_data/excercise3.R` for extended regression examples (scaling, centering, scatterplots, and diagnostic plots).

::: {.callout-tip collapse="true" title="Exercise 3 — Solution"}
- Solution (PDF): `dev_code_data/Solution3.pdf`  
- Solution script: `dev_code_data/Solution3.R`  
Run: `Rscript dev_code_data/Solution3.R` or open the PDF in your browser/RStudio.
:::

## Further reading

- Practical regression and diagnostics: @Henderson-Velleman-1981; @Matloff-2011.
- Applied examples and tidy workflows: @Wickham-Grolemund-2017.

## Resources

- Example script: `dev_code_data/excercise3.R` (scatterplots, fitted lines, diagnostics).
