---
title: "Simple Linear Regression"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(survival)
library(broom)
```

## When Do You Use This? {#sec-linear-regression .callout-tip}

> You have a continuous outcome and a continuous predictor, and you want to quantify how much the outcome changes per unit increase in the predictor, and test whether that relationship is statistically meaningful. Simple linear regression gives you a straight-line model with an estimated slope and its confidence interval. For example: predicting serum albumin from bilirubin, or estimating how bill length changes with flipper length in penguins.

## Learning Objectives

After completing this session you will be able to:

- Fit a simple linear regression model with `lm()` in R
- Interpret the intercept, slope, standard error, and $R^2$
- Check the four key regression assumptions using diagnostic plots
- Extract and present results with `broom::tidy()` and `broom::glance()`
- Predict new values with `predict()` and plot the regression line with uncertainty


## Background: The Linear Model

A simple linear regression models the relationship between a continuous outcome $Y$ and a single predictor $X$:

$$Y_i = \beta_0 + \beta_1 X_i + \varepsilon_i \quad \varepsilon_i \sim N(0, \sigma^2)$$

- $\beta_0$: **intercept**: the expected value of $Y$ when $X = 0$
- $\beta_1$: **slope**: the expected change in $Y$ for a one-unit increase in $X$
- $\varepsilon_i$: **residual**: what the model does not explain

The model is fitted by minimising the sum of squared residuals (ordinary least squares, OLS).

**Assumptions (LINE):**

| Assumption | How to check |
|---|---|
| **L**inearity | Residuals vs fitted plot → no pattern |
| **I**ndependence | Study design → cannot check from data alone |
| **N**ormality of residuals | QQ-plot of residuals |
| **E**qual variance (homoscedasticity) | Scale-location plot → flat red line |


## Example 1: Clinical Data (PBC)

### Predict Albumin from Log(Bilirubin)

Bilirubin is right-skewed; we log-transform it to achieve a more linear relationship.

```{r pbc-setup}
pbc <- survival::pbc %>%
  as_tibble() %>%
  janitor::clean_names() %>%
  filter(!is.na(albumin), !is.na(bili)) %>%
  mutate(log_bili = log(bili))
```

Always plot first:

```{r pbc-plot}
pbc %>%
  ggplot(aes(x = log_bili, y = albumin)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "lm", color = "tomato", se = TRUE) +
  labs(title = "Albumin vs log(bilirubin) in PBC",
       x = "log(Bilirubin)", y = "Albumin (g/dL)") +
  theme_bw()
```

```{r fit-lm}
fit <- lm(albumin ~ log_bili, data = pbc)
summary(fit)
```

::: {.callout-note}
### Reading the `summary()` Output: Line by Line

When you run `summary(fit)`, R prints several blocks. Here is what each one means.

**`Call:`**
Just echoes back the formula you used: `albumin ~ log_bili`. Confirms R fitted the right model.

**`Residuals:`**
```
Min      1Q  Median      3Q     Max
-1.52   -0.29   0.01    0.31    1.18
```
These are the differences between each patient's observed albumin and the model's prediction. Ideally:
- The **Median** should be close to 0 (no systematic over- or under-prediction)
- The **Min** and **Max** should be roughly symmetric  -  if the biggest error is `+1.18` and smallest is `-1.52`, that is reasonably balanced.
- Very large Max or very negative Min suggests outliers.

**`Coefficients:` table**
```
              Estimate Std. Error t value Pr(>|t|)
(Intercept)   4.15      0.06      69.2    <2e-16 ***
log_bili     -0.34      0.03     -11.3    <2e-16 ***
```

| Column | What it means |
|---|---|
| **Estimate** | The fitted value of the coefficient: the slope or intercept |
| **Std. Error** | How uncertain the estimate is: smaller is more precise |
| **t value** | Estimate / Std. Error: how many standard errors away from zero |
| **Pr(>\|t\|)** | p-value: probability of seeing a t-value this large *if the true slope were zero* |
| **Stars (\*\*\*)** | Just a shorthand: `***` = p < 0.001, `**` = p < 0.01, `*` = p < 0.05, `.` = p < 0.1 |

**`Residual standard error: 0.43 on 416 degrees of freedom`**
This is the typical size of a prediction error. On average, the model's albumin prediction is off by about 0.43 g/dL. Degrees of freedom = number of observations minus number of parameters estimated (here, 418 - 2 = 416).

**`Multiple R-squared: 0.22`**
Log(bilirubin) explains **22% of the variability** in albumin across patients. The remaining 78% is explained by other factors not in this model.

**`Adjusted R-squared: 0.22`**
Almost the same as R² here because we only have one predictor. It would drop relative to R² if we had many predictors but most were uninformative. Prefer adjusted R² when comparing models with different numbers of predictors.

**`F-statistic: 128 on 1 and 416 DF, p-value: < 2.2e-16`**
Tests whether the *whole model* explains significantly more variance than just predicting the mean for everyone. With one predictor, this is equivalent to the t-test on the slope. If p < 0.05, at least one predictor is useful.
:::

### Interpreting the Output

```{r tidy-output}
# Coefficients table
tidy(fit, conf.int = TRUE)

# Model-level statistics
glance(fit)
```

**Reading the coefficients:**
- **Intercept (4.15):** Expected albumin when log(bilirubin) = 0 (i.e., bilirubin = 1 mg/dL) is 4.15 g/dL.
- **Slope (-0.34):** For each one-unit increase in log(bilirubin), albumin decreases by 0.34 g/dL on average.
- **$R^2$ (0.22):** Log(bilirubin) explains 22% of the variability in albumin.

::: {.callout-tip}
### What Does This Actually Tell Us?

Translate the numbers into a scientific sentence before you report anything:

> *"In PBC patients, higher bilirubin is associated with lower albumin. For every approximate 2.7-fold increase in bilirubin (one unit on the log scale), albumin is on average 0.34 g/dL lower (95% CI: -0.40 to -0.28, p < 0.001). Log(bilirubin) explains 22% of the variability in albumin across patients."*

That is what you would write in a Results section. Notice:

- You state the **direction** (lower albumin)
- You give the **magnitude** (0.34 g/dL)
- You give the **uncertainty** (95% CI)
- You give the **p-value**  -  but note it only tells you the effect is unlikely to be zero, not that it is large or clinically important
- You give **R²** to indicate how much of the outcome this one predictor explains

**Is 22% a good R²?** In clinical and biological data, a single biomarker rarely explains more than 20-30% of an outcome, there are too many other factors involved. R² = 0.22 is actually quite respectable here. R² of 0.90+ is typical only in physics or engineering, not in medicine.
:::

### Checking Assumptions

```{r diagnostics}
par(mfrow = c(2, 2))
plot(fit)
par(mfrow = c(1, 1))
```

**Residuals vs Fitted:** Should show no pattern. A curve suggests non-linearity.

**QQ-plot:** Residuals should follow the diagonal line. Heavy tails indicate non-normality.

**Scale-Location:** Should be flat. An upward slope indicates heteroscedasticity (variance increases with fitted values).

**Cook's Distance:** Identifies influential observations. Points labeled by row number warrant investigation.

::: {.callout-note}
### How to Read Diagnostic Plots: What Good and Bad Look Like

**Plot 1: Residuals vs Fitted**
You are looking for a flat, horizontal cloud of points with no pattern.
- ✓ **Good:** Points scattered randomly around the horizontal line at zero.
- ✗ **Bad (U-shape or arc):** The relationship is non-linear; your straight line is the wrong shape. Try log-transforming X.
- ✗ **Bad (funnel shape):** Variance increases with fitted values (heteroscedasticity). Try log-transforming Y.

**Plot 2: Normal Q-Q**
You are checking whether the residuals follow a normal distribution.
- ✓ **Good:** Points follow the diagonal line closely.
- ✗ **Bad (S-curve):** Residuals are skewed.
- ✗ **Bad (heavy tails):** More extreme values than expected; the model underestimates rare large errors. Mild deviation is usually acceptable for large samples.

**Plot 3: Scale-Location**
Same as Plot 1 but using the square root of the absolute residuals, making it easier to spot unequal variance.
- ✓ **Good:** Red line is flat, points evenly spread.
- ✗ **Bad (upward slope):** Variance increases; patients with high predicted albumin also have more unpredictable albumin.

**Plot 4: Cook's Distance / Residuals vs Leverage**
Identifies observations that have a disproportionate effect on the fitted line.
- **Leverage** = how unusual the X value is (extreme bilirubin, for example)
- **Cook's D** = combined measure: unusual X *and* large residual
- **Rule of thumb:** Cook's D > 4/n (here, 4/418 ≈ 0.01) or > 1 flags cases to investigate
- These are not automatically errors; they may be biologically interesting patients
:::

### Predictions

```{r predictions}
# Predict albumin for bilirubin = 2, 5, 10 mg/dL
new_data <- tibble(log_bili = log(c(2, 5, 10)))
predict(fit, newdata = new_data, interval = "confidence")
```

```{r prediction-plot}
# Plot with prediction band
pred_df <- tibble(
  log_bili = seq(min(pbc$log_bili), max(pbc$log_bili), length.out = 100)
)
pred_df <- bind_cols(pred_df, as_tibble(predict(fit, newdata = pred_df, interval = "prediction")))

pbc %>%
  ggplot(aes(x = log_bili, y = albumin)) +
  geom_ribbon(data = pred_df, aes(y = fit, ymin = lwr, ymax = upr),
              alpha = 0.15, fill = "steelblue") +
  geom_line(data = pred_df, aes(y = fit), color = "steelblue", linewidth = 1) +
  geom_point(alpha = 0.3, size = 1.5) +
  labs(title = "Linear regression with 95% prediction interval",
       x = "log(Bilirubin)", y = "Albumin (g/dL)") +
  theme_bw()
```


## Example 2: Animal Data (palmerpenguins)

We predict body mass from flipper length, a practical calibration question in ecology and animal research.

```{r penguin-lm}
library(palmerpenguins)

peng <- penguins %>%
  drop_na(body_mass_g, flipper_length_mm)

fit_p <- lm(body_mass_g ~ flipper_length_mm, data = peng)
tidy(fit_p, conf.int = TRUE)
glance(fit_p)
```

```{r penguin-plot}
peng %>%
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species), alpha = 0.6) +
  geom_smooth(method = "lm", color = "black", se = TRUE) +
  scale_color_manual(values = c("steelblue", "tomato", "seagreen")) +
  labs(title = "Body mass predicted by flipper length",
       x = "Flipper length (mm)", y = "Body mass (g)") +
  theme_bw()
```

```{r penguin-diagnostics}
par(mfrow = c(2, 2))
plot(fit_p)
par(mfrow = c(1, 1))
```

Notice that the residuals may cluster by species, a sign that species confounds the relationship. This motivates multiple regression (see next session).


## Where to Find Data Like This {.callout-note}

**PBC dataset:** `survival::pbc` - multiple continuous variables with clinically interpretable relationships.

**palmerpenguins:** `palmerpenguins::penguins` - clean morphological measurements, ideal for regression with a natural grouping structure.

**MASS::Boston:** Housing data commonly used to teach regression; however, contains a problematic racial variable; examine the data carefully before use.


## What Can Go Wrong {.callout-warning}

**Extrapolation**
A regression line is only valid within the range of the observed data. Predicting outside this range (extrapolation) can produce nonsensical values - e.g., negative albumin.

**Ignoring non-linearity**
If the Residuals vs Fitted plot shows a U-shape or arc, the linear model is misspecified. Try a log-transformation of X, or consider polynomial or spline terms.

**Heteroscedasticity**
If variance increases with fitted values (fan-shaped residuals), OLS standard errors are incorrect. Consider transforming Y, or use heteroscedasticity-consistent (robust) standard errors.

**Confounding**
A simple regression coefficient is not causal. A third variable may drive both X and Y (see [Multiple Regression](06-multiple-regression.qmd) and [Correlation and Association](correlation-and-association.qmd)).

**Outliers and influential points**
Check Cook's distance. Points with Cook's D > 1 (or > 4/n) exert disproportionate influence on the fitted line. Investigate them; they may be data errors or genuinely unusual cases.
::: {.callout-warning}
### Common Misinterpretations: Don't Make These Mistakes

**"p < 0.05 means the effect is large"**
No. p < 0.05 means the effect is unlikely to be exactly zero given your sample size. A slope of -0.001 can be highly significant with 10,000 patients. Always report the *magnitude* (the estimate and CI), not just the p-value.

**"p = 0.06 means there is no effect"**
No. It means you do not have enough evidence to rule out chance with this sample size. A study with 20 patients and p = 0.06 may simply be underpowered. Never say "no association" based on p > 0.05 alone.

**"R² = 0.05 means the model is useless"**
Not necessarily. In medicine and biology, outcomes are driven by hundreds of factors. A single biomarker explaining 5% of variance may still be scientifically meaningful. Judge R² relative to what is realistic in your field.

**"The intercept is always meaningful"**
Only if X = 0 is a realistic value. In `lm(albumin ~ log_bili)`, the intercept is albumin when bilirubin = 1 mg/dL. If few patients have bilirubin near 1 mg/dL, the intercept is a mathematical anchor, not a clinically interpretable number.

**"A significant p-value means my model assumptions are met"**
The p-value tells you nothing about model assumptions. Check the diagnostic plots regardless of the p-value.
:::

## Exercises

### Exercise 1 (Guided): Prothrombin Time Predicts Albumin

1. Plot albumin (y) vs prothrombin time (x) in `pbc`. Log-transform prothrombin time.
2. Fit `lm(albumin ~ log(protime), data = pbc)`. Interpret the slope.
3. Check all four diagnostic plots.
4. Report the slope, 95% CI, and $R^2$.

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r}
pbc2 <- pbc %>% filter(!is.na(protime))

# 1. Plot
pbc2 %>%
  ggplot(aes(x = log(protime), y = albumin)) +
  geom_point(alpha = 0.4) + geom_smooth(method = "lm", color = "tomato") +
  labs(x = "log(Prothrombin time)", y = "Albumin (g/dL)") + theme_bw()

# 2. Fit and interpret
fit2 <- lm(albumin ~ I(log(protime)), data = pbc2)
tidy(fit2, conf.int = TRUE)
glance(fit2)

# 3. Diagnostics
par(mfrow = c(2, 2)); plot(fit2); par(mfrow = c(1, 1))
```
:::

### Exercise 2 (Semi-guided): Bill Length Predicts Body Mass (Penguins)

1. Fit `lm(body_mass_g ~ bill_length_mm, data = penguins)`.
2. Does the model meet the assumptions? Check the diagnostic plots.
3. Now fit the model separately for each species. Compare the slopes. How do they differ from the overall model?

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r}
peng2 <- penguins %>% drop_na()

# 1. Overall model
fit_all <- lm(body_mass_g ~ bill_length_mm, data = peng2)
tidy(fit_all, conf.int = TRUE)

# 2. Diagnostics
par(mfrow = c(2, 2)); plot(fit_all); par(mfrow = c(1, 1))

# 3. By species
peng2 %>%
  group_by(species) %>%
  do(tidy(lm(body_mass_g ~ bill_length_mm, data = .), conf.int = TRUE)) %>%
  filter(term == "bill_length_mm") %>%
  select(species, estimate, conf.low, conf.high, p.value)
```
:::

### Exercise 3 (Open-ended)

In your own research, identify a continuous outcome and a single continuous predictor you expect to be linearly related. Fit a simple linear regression, check all assumptions, and write a Results paragraph including: the slope (95% CI), the $R^2$, and a sentence on whether the assumptions are satisfied.


## Comprehension Check

1. You fit `lm(albumin ~ log_bili)` and get slope = -0.34, SE = 0.03. Interpret the slope in plain language.
2. The QQ-plot of residuals shows a heavy right tail. Is this a problem? What would you do?
3. $R^2 = 0.05$. Does this mean the model is useless?
4. You predict albumin = 2.1 g/dL for a patient with bilirubin = 80 mg/dL. The maximum observed bilirubin in your data is 28 mg/dL. Should you trust this prediction?
5. What does the intercept represent in `lm(albumin ~ log_bili)` when many patients have bilirubin values far from 1 mg/dL?

::: {.callout-note collapse="true"}
### Answers
1. For every one-unit increase in log(bilirubin), roughly a 2.7-fold increase in bilirubin on the original scale, and albumin is expected to be 0.34 g/dL lower, on average, after accounting for the uncertainty in this estimate.
2. A heavy right tail in the QQ-plot indicates that large positive residuals occur more often than expected under normality. For inference (t-tests, CIs), mild non-normality is acceptable with large samples (CLT). For severe non-normality, try a log-transformation of the outcome or use robust standard errors.
3. $R^2 = 0.05$ means the predictor explains 5% of the outcome's variability. Whether this is "useful" depends on context. In complex biological systems, a single biomarker explaining even 5% of variance may be scientifically meaningful. The slope and CI are more informative than $R^2$ alone.
4. No. A bilirubin of 80 mg/dL is well outside the range of the data used to fit the model (max = 28). Extrapolation is unreliable; the linear relationship may not hold at extreme values.
5. The intercept is the predicted albumin when log(bilirubin) = 0, i.e., when bilirubin = 1 mg/dL. If few patients have bilirubin near 1 mg/dL, the intercept is an extrapolation with high uncertainty and limited clinical meaning. It is mainly a mathematical artefact of the model parameterisation.
:::


## How to Report

### Reporting a Linear Regression

::: {.callout-note}
**In a methods section:** "The association between [outcome] and [predictor] was estimated using linear regression. Potential confounders [list] were included in the model. Model assumptions (linearity, homoscedasticity, normality of residuals) were assessed using diagnostic plots."

**In results:** "Each additional [unit] of [predictor] was associated with a [\u03b2] [unit] change in [outcome] (\u03b2 = X, 95% CI [L, U], p = Y)."

**What to always include:**

- Regression coefficient (\u03b2) with 95% CI and p-value
- Adjusted R² (proportion of variance explained)
- Number of observations and any excluded due to missing data
- Whether the model is unadjusted or adjusted (and for what)

**Reporting R² alone is not enough** --- always report the coefficient and CI so readers know the direction and magnitude of the effect.
:::

## Further Reading

- @Dalgaard-2008 --- Chapter on linear models in *Introductory Statistics with R*
- @Bland-2015 --- Regression chapters in *An Introduction to Medical Statistics*
- `?lm`, `?predict.lm` in R: full documentation
- `broom` package vignette for tidy model output
