# Logistic Regression {#logistic-regression}

```{r setup}
#| include: false
#| eval: true
source(here::here("R/functions.R"))
source(here::here("R/load-packages.R"))
```

Logistic regression models binary outcomes; coefficients exponentiated are odds ratios.

## When to use logistic regression

Use logistic regression for binary outcomes (event / no event). Adjust for confounders when estimating associations.

## Data {.unnumbered}

- Clinical example: Pima Indians Diabetes (public CSV) — suitable for logistic regression, classification, and model calibration examples. See `data_sources.md` for additional clinical datasets.

## Short code demo

```{r demo-glm, echo=TRUE}
# load Pima diabetes dataset (public CSV)
pima <- readr::read_csv('https://raw.githubusercontent.com/selva86/datasets/master/PimaIndiansDiabetes.csv')
pima <- pima %>% mutate(diabetes = as.factor(diabetes))

fit_log <- glm(diabetes ~ age + bmi, family = binomial, data = pima)

# tidy coefficients and approximate odds ratios
broom::tidy(fit_log)
exp(coef(fit_log))
```

## How to explain the output (teacher notes)

- **Odds ratio:** multiplicative change in odds for a one-unit predictor increase.
- **Predicted probabilities** are usually easier to communicate than odds ratios.

## Short exercise

- Fit a model with `diabetes ~ bmi + age` and compute predicted probabilities for representative covariate values.

**Hint:** use `predict(..., type = 'response')` for probabilities.

## Packages used

- `dplyr`, `readr`, `broom`, `ggplot2`

**Exercises**

- Example code is inline above; see `data_sources.md` for public datasets and suggested exercise replacements.

::: {.callout-tip collapse="true" title="Exercise 6 — Solution"}
- Solution files previously stored in `dev_code_data` have been removed.
- Run the inline examples above or use one of the public datasets listed in `data_sources.md` for extended logistic exercises.
:::

## Further reading

- Logistic regression references and interpretation: @Matloff-2011; @Henderson-Velleman-1981.
- Practical modeling guidance: @Wickham-Grolemund-2017.

## Resources

- Example scripts moved out of `dev_code_data`; see `data_sources.md` and inline examples above.
