# Logistic Regression {#logistic-regression}

```{r setup}
#| include: false
#| eval: true
source(here::here("R/functions.R"))
source(here::here("R/load-packages.R"))
```

Logistic regression models binary outcomes; coefficients exponentiated are
odds ratios.

## When to use logistic regression

Use logistic regression for binary outcomes (event / no event). Adjust for
confounders when estimating associations.

## Short code demo (dplyr-style)

```{r demo-glm, echo=TRUE}
# example using mtcars: model probability of manual transmission (am)
mtcars <- datasets::mtcars %>% as_tibble(rownames = "model") %>%
  mutate(am = factor(am, labels = c("auto", "manual")))

fit_log <- glm(am ~ wt + hp, family = binomial, data = mtcars)

# tidy coefficients and convert to odds ratios
broom::tidy(fit_log)
exp(coef(fit_log))
```

## How to explain the output (teacher notes)

- **Odds ratio:** multiplicative change in odds for a one-unit predictor increase.
- **Predicted probabilities** are usually easier to communicate than odds ratios.

## Short exercise

- Fit a model with `am ~ wt` and compute predicted probability for two example weights.

**Hint:** use `predict(..., type = "response")` for probabilities.

## Packages used

- `dplyr`, `broom`, `ggplot2`
