# T-tests and Group Comparisons {#sec-t-tests}

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(survival)
library(broom)
```

## When Do You Use This? {.callout-tip}

> You measured haemoglobin before and after treatment in 30 patients. Or you measured body weight in 20 control mice and 20 treated mice. A t-test tells you whether the observed difference is larger than you would expect by chance if there were no real effect.

## Learning Objectives

After completing this session you will be able to:

- Choose the right t-test variant (one-sample, two-sample, paired)
- Decide whether to use Welch's or Student's t-test
- Check the normality assumption and know when the test is still valid
- Perform all three variants in R and interpret the output
- Report t-test results correctly in a manuscript


## Background: Which t-test?

| Situation | Test | R function |
|---|---|---|
| Compare sample mean to a known value | One-sample t-test | `t.test(x, mu = value)` |
| Compare means of two independent groups | Two-sample t-test | `t.test(y ~ group, data = df)` |
| Compare two measurements on the same subjects | Paired t-test | `t.test(x, y, paired = TRUE)` |

**Welch vs Student:** Always use Welch's t-test (default in R: `var.equal = FALSE`) unless you have a specific reason to assume equal variances. Welch's test is more robust and the cost in power is negligible.

**Normality:** The t-test assumes the *sampling distribution of the mean* is normal, not the raw data. By the CLT, this holds for $n \geq 30$ even with moderately skewed data. For small samples with heavily skewed data, use a nonparametric test (see [Nonparametric Tests](nonparametric-tests.qmd)).


## Worked Example 1: Clinical Data (PBC)

### One-Sample: Is Mean Albumin Below a Clinical Threshold?

Normal serum albumin is = 3.5 g/dL. Is the mean albumin in PBC patients at this threshold?

```{r pbc-setup}
pbc <- survival::pbc %>%
  as_tibble() %>%
  janitor::clean_names() %>%
  mutate(
    sex = factor(sex, levels = c("f", "m"), labels = c("Female", "Male")),
    trt = factor(trt, levels = c(1, 2), labels = c("D-penicillamine", "Placebo"))
  )
```

```{r one-sample}
# H0: mean albumin = 3.5 g/dL
# H1: mean albumin != 3.5 g/dL
t.test(pbc$albumin, mu = 3.5)
```

```{r one-sample-tidy}
tidy(t.test(pbc$albumin, mu = 3.5))
```

`broom::tidy()` extracts the key numbers into a data frame, useful for piping into tables or inline reporting.

### Two-Sample: Does Albumin Differ Between Treatment Groups?

```{r two-sample}
pbc_rand <- pbc %>% filter(!is.na(trt), !is.na(albumin))

# Visualise first
pbc_rand %>%
  ggplot(aes(x = trt, y = albumin, fill = trt)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1.5) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(title = "Serum albumin by treatment group",
       x = NULL, y = "Albumin (g/dL)") +
  theme_bw() + theme(legend.position = "none")
```

```{r two-sample-test}
t_res <- t.test(albumin ~ trt, data = pbc_rand, var.equal = FALSE)
tidy(t_res)
```

**Reporting:** "Mean albumin was `r round(mean(pbc_rand$albumin[pbc_rand$trt == "D-penicillamine"], na.rm=TRUE), 2)` g/dL in the D-penicillamine group and `r round(mean(pbc_rand$albumin[pbc_rand$trt == "Placebo"], na.rm=TRUE), 2)` g/dL in the placebo group. There was no statistically significant difference (Welch's two-sample t-test: t = [value], df = [df], p = [p]; 95% CI for difference: [CI])."

### Paired: Change in Albumin Before and After (Hypothetical)

In a paired design, each subject provides two measurements (e.g., baseline and follow-up). We need the difference per subject, not the group means.

```{r paired-example}
# Simulate: 50 patients with paired albumin measurements
set.seed(2024)
paired_dat <- tibble(
  patient    = 1:50,
  albumin_t0 = rnorm(50, mean = 3.5, sd = 0.5),
  albumin_t1 = albumin_t0 + rnorm(50, mean = -0.3, sd = 0.3)  # slight decline
)

# Visualise the paired differences
paired_dat %>%
  mutate(diff = albumin_t1 - albumin_t0) %>%
  ggplot(aes(x = diff)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Change in albumin (t1 - t0)",
       x = "Change (g/dL)", y = "Count") +
  theme_bw()
```

```{r paired-test}
t.test(paired_dat$albumin_t1, paired_dat$albumin_t0, paired = TRUE)
```

The paired t-test reduces variability by removing between-patient differences, making it more powerful than a two-sample test for this design.


## Worked Example 2: Animal Data (ToothGrowth)

The `ToothGrowth` dataset compares two Vitamin C delivery methods in guinea pigs.

```{r toothgrowth}
tg <- ToothGrowth %>%
  as_tibble() %>%
  mutate(
    supp = factor(supp, levels = c("VC", "OJ"),
                  labels = c("Ascorbic acid", "Orange juice")),
    dose = factor(dose)
  )

# Compare at the 1.0 mg/day dose only
tg_1mg <- tg %>% filter(dose == "1")

tg_1mg %>%
  ggplot(aes(x = supp, y = len, fill = supp)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.1, alpha = 0.5, size = 2) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(title = "Tooth length by supplement type at 1.0 mg/day",
       x = NULL, y = "Odontoblast length (µm)") +
  theme_bw() + theme(legend.position = "none")
```

```{r toothgrowth-test}
tidy(t.test(len ~ supp, data = tg_1mg, var.equal = FALSE))
```


## Checking Assumptions: QQ-plots

For small samples ($n < 30$), check that data are approximately normal using a QQ-plot.

```{r qq-plot}
pbc_rand %>%
  ggplot(aes(sample = albumin)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  facet_wrap(~ trt) +
  labs(title = "QQ-plot of albumin by treatment group") +
  theme_bw()
```

Points falling close to the diagonal line indicate approximate normality. With $n \approx 150$ per group, slight deviations are not a concern for the t-test.


## Where to Find Data Like This {.callout-note}

**PBC dataset:** `survival::pbc` — built in, no download needed.

**Sleep dataset (paired):** `datasets::sleep` — built into base R. Contains the effect of two sleep drugs on 10 patients (classic paired t-test example).

**MASS::birthwt:** birth weight data with clinical predictors; good for comparing birth weight between groups.

```{r}
#| eval: false
data(birthwt, package = "MASS")
t.test(bwt ~ smoke, data = birthwt)
```


## What Can Go Wrong {.callout-warning}

**Using a paired test when data are independent (or vice versa)**
The paired t-test requires one-to-one matching. If you use it on independent group data, you will get wrong degrees of freedom and an incorrect p-value.

**Testing normality with a formal test**
Shapiro-Wilk (`shapiro.test()`) is not a reliable guide for choosing a t-test. With small samples it has low power; with large samples it rejects normality for trivially small deviations. Use a QQ-plot and think about the biology instead.

**Reporting only the p-value**
Always report the mean difference and 95% CI alongside the p-value. "p = 0.03" tells the reader almost nothing about the size or direction of the effect.

**Not accounting for unequal variances**
If the two groups have very different standard deviations, Welch's t-test (default in R) is the right choice. Using `var.equal = TRUE` when variances differ inflates the Type I error rate.


## Exercises

### Exercise 1 (Guided): One-Sample Test for Bilirubin

The upper limit of normal for bilirubin is about 1.2 mg/dL.

1. Plot the bilirubin distribution. Is it normal?
2. Log-transform bilirubin. Test whether mean log-bilirubin equals `log(1.2)`.
3. Report: mean, 95% CI, t-statistic, df, p-value, and a one-sentence interpretation.

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r}
# 1. Distribution
ggplot(pbc, aes(x = bili)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(x = "Bilirubin (mg/dL)", y = "Count") + theme_bw()

# 2. One-sample test on log scale
res <- t.test(log(pbc$bili), mu = log(1.2))
tidy(res)

# 3. Back-transform the CI
exp(res$conf.int)
```

The geometric mean bilirubin is well above 1.2 mg/dL (the upper limit of normal), and this difference is highly statistically significant (p << 0.001). This confirms that PBC patients have elevated bilirubin, consistent with cholestatic liver disease.
:::

### Exercise 2 (Semi-guided): Two-Sample Comparison

Using `ToothGrowth`, test whether tooth length differs by supplement type across **all** dose levels (not just 1 mg/day).

1. Run a two-sample Welch's t-test.
2. Is the test valid? Check sample sizes and a QQ-plot.
3. Run the test also on log-transformed tooth length. Do the conclusions differ?

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r}
tg_all <- ToothGrowth %>%
  as_tibble() %>%
  mutate(supp = factor(supp))

# Two-sample test
tidy(t.test(len ~ supp, data = tg_all))

# QQ-plot by group
tg_all %>%
  ggplot(aes(sample = len)) +
  stat_qq() + stat_qq_line(color = "red") +
  facet_wrap(~ supp) + theme_bw()

# Log scale
tidy(t.test(log(len) ~ supp, data = tg_all))
```

With 30 observations per group, the CLT applies. The QQ-plot shows approximate normality. The raw and log-transformed tests give similar conclusions.
:::

### Exercise 3 (Open-ended)

Use a dataset from your own research. Identify a comparison where a t-test is appropriate. Run the test, check assumptions, and write a Methods + Results paragraph as you would for a journal submission.


## Comprehension Check

1. What is the difference between a one-sample, two-sample, and paired t-test? Give a biomedical example for each.
2. When should you use Welch's t-test instead of Student's t-test?
3. A QQ-plot of your data shows points diverging from the line at the upper tail. The sample size is n = 12. What should you do?
4. You run a two-sample t-test and get p = 0.04, 95% CI for difference: (-0.01 g/dL to 2.1 g/dL). Is this a convincing result? What would you report?
5. A colleague says "I tested normality with Shapiro-Wilk (p = 0.06) so the data are normal." What is wrong with this reasoning?

::: {.callout-note collapse="true"}
### Answers
1. **One-sample:** compare a sample mean to a known or hypothetical value (e.g., "Is mean haemoglobin in my patients below 12 g/dL?"). **Two-sample:** compare means of two independent groups (e.g., "Does albumin differ between treatment and control mice?"). **Paired:** compare two measurements on the same subjects (e.g., "Did blood pressure change from baseline to 6 weeks?").
2. Use Welch's when the two groups may have different population variances. It is the default in R (`var.equal = FALSE`) and is almost always appropriate. Student's t-test (equal variance assumed) is rarely preferred.
3. With n = 12 and non-normality, use a nonparametric Wilcoxon test (see next session) instead of a t-test.
4. The result is statistically significant (p = 0.04), but the CI is very wide: the true difference could be essentially zero (-0.01) or moderately large (2.1 g/dL). Report both the p-value and the full CI, and discuss whether the upper end of the CI is clinically meaningful.
5. Shapiro-Wilk with p = 0.06 is not evidence that data are normal; it means you failed to detect non-normality at the 5% level. With n = 12, the test has low power and would miss moderate non-normality. Use a QQ-plot and domain knowledge instead.
:::

## Further Reading

- @Welch-1947 --- Original Welch t-test paper
- @Bland-2015 --- Chapter on t-tests in *An Introduction to Medical Statistics*
- `?t.test` in R — the help page has a detailed description of all options
