# T-tests & Group Comparisons {#t-tests}

```{r setup}
#| include: false
#| eval: true
source(here::here("R/functions.R"))
source(here::here("R/load-packages.R"))
```

This session covers one-sample, two-sample, and paired t-tests and when to
use Welch's correction.

## When to use t-tests

Use t-tests to compare means between groups (one-sample, two-sample) or
paired measurements on the same subjects (paired t-test). Use Welch's t-test
when variances are unequal.

## Short code demo

```{r demo-ttest, echo=TRUE}
# two-sample example from ToothGrowth
tg <- datasets::ToothGrowth %>% as_tibble() %>% mutate(dose = factor(dose))
tg_low <- tg %>% filter(dose == 0.5)

# two-sample t-test comparing supplements
t_test_res <- t.test(len ~ supp, data = tg_low)
broom::tidy(t_test_res)

# paired example using built-in sleep dataset
paired_res <- t.test(extra ~ group, data = datasets::sleep, paired = TRUE)
broom::tidy(paired_res)
```

## How to explain the output (teacher notes)

- **Report:** mean difference and 95% CI. State whether CI includes zero.
- **Context:** choose paired vs independent test based on study design.

## Short exercise

- Select two groups from `mtcars` (e.g., `am` levels) and run an appropriate t-test; write a one-line interpretation.

**Hint:** use `t.test(..., var.equal = TRUE)` for pooled variance, omit for Welch.

## Packages used

- `broom`, `dplyr`

**Exercises**

- See `dev_code_data/excercise2.R` for t-test examples and paired/independent comparisons.

::: {.callout-tip collapse="true" title="Exercise 2 — Solution"}
- Solution (PDF): `dev_code_data/Solution2.pdf`  
- Solution script: `dev_code_data/Solution2.R`  
Run: `Rscript dev_code_data/Solution2.R` or open the PDF in your browser/RStudio.
:::

## Further reading

- Testing and nonparametrics: @Teetor-2011; @Wickham-Grolemund-2017.

## Resources

- Example script: `dev_code_data/excercise2.R` (t-tests, paired tests, and Wilcoxon examples).
