---
title: "T-tests and Group Comparisons"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(survival)
library(broom)
```

## When Do You Use This? {#sec-t-tests .callout-tip}

> You measured haemoglobin before and after treatment in 30 patients. Or you measured body weight in 20 control mice and 20 treated mice. A t-test tells you whether the observed difference is larger than you would expect by chance if there were no real effect.

## Learning Objectives

After completing this session you will be able to:

- Choose the right t-test variant (one-sample, two-sample, paired)
- Decide whether to use Welch's or Student's t-test
- Check the normality assumption and know when the test is still valid
- Perform all three variants in R and interpret the output
- Report t-test results correctly in a manuscript


## Background: Which t-test?

| Situation | Test | R function |
|---|---|---|
| Compare sample mean to a known value | One-sample t-test | `t.test(x, mu = value)` |
| Compare means of two independent groups | Two-sample t-test | `t.test(y ~ group, data = df)` |
| Compare two measurements on the same subjects | Paired t-test | `t.test(x, y, paired = TRUE)` |

**Welch vs Student:** Always use Welch's t-test (default in R: `var.equal = FALSE`) unless you have a specific reason to assume equal variances. Welch's test is more robust and the cost in power is negligible.

**Normality:** The t-test assumes the *sampling distribution of the mean* is normal, not the raw data. By the CLT, this holds for $n \geq 30$ even with moderately skewed data. For small samples with heavily skewed data, use a nonparametric test (see [Nonparametric Tests](nonparametric-tests.qmd)).


## Example 1: Animal and Lab Data

### Two-Sample: Does Supplement Type Affect Tooth Growth? (ToothGrowth)

The `ToothGrowth` dataset (base R) records the length of odontoblast cells in 60 guinea pigs given
Vitamin C as ascorbic acid (VC) or orange juice (OJ) at three doses.
This is a classic two-independent-groups design.

```{r toothgrowth-setup}
#| eval: true

tg <- ToothGrowth |>
  as_tibble() |>
  mutate(
    supp = factor(supp, levels = c("VC", "OJ"),
                  labels = c("Ascorbic acid", "Orange juice")),
    dose = factor(dose)
  )

# 10 animals per group at 1.0 mg/day
tg_1mg <- tg |> filter(dose == "1")

tg_1mg |>
  ggplot(aes(x = supp, y = len, fill = supp)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.6, size = 2.5) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(
    title = "Tooth length by supplement type (1.0 mg/day)",
    x = NULL, y = "Odontoblast length (mm)"
  ) +
  theme_bw() + theme(legend.position = "none")
```

```{r toothgrowth-test}
#| eval: true

res_tg <- t.test(len ~ supp, data = tg_1mg, var.equal = FALSE)
tidy(res_tg)
```

::: {.callout-note}
### Reading the t-test Output Line by Line

| Output | What it means |
|---|---|
| **t = -1.41** | The sample means differ by 1.41 standard errors. Negative because VC < OJ. |
| **df = 15.36** | Welch's df is not a whole number; this is normal and correct. |
| **p-value = 0.18** | If there were no real difference, we would observe a gap this large 18% of the time just by chance alone. |
| **95% CI: -8.78 to 1.78** | The true mean difference plausibly ranges from OJ being 8.78 mm better to VC being 1.78 mm better. CI includes zero. |
| **means: OJ = 22.70, VC = 16.77** | The mean difference is 5.93 mm. Always report the point estimate, not just the p-value. |

At 1 mg/day the difference is not significant. At 0.5 mg/day it is -- try it in Exercise 1.
:::

### Paired: Do Two Sleep Drugs Differ? (sleep dataset)

The `sleep` dataset (base R) records the *additional hours of sleep* gained by 10 patients under each
of two drugs. Each patient received both drugs, making this a **paired design**.

```{r sleep-plot}
#| eval: true

sleepdat <- sleep |>
  as_tibble() |>
  rename(drug = group, extra_hours = extra) |>
  mutate(drug = factor(drug, labels = c("Drug 1", "Drug 2")))

# Connect each patient's two points
sleepdat |>
  ggplot(aes(x = drug, y = extra_hours)) +
  geom_line(aes(group = ID), alpha = 0.4, colour = "grey50") +
  geom_point(aes(colour = drug), size = 3) +
  scale_colour_manual(values = c("steelblue", "tomato")) +
  labs(
    title = "Additional sleep hours under each drug (n = 10 patients)",
    x = NULL, y = "Additional hours of sleep"
  ) +
  theme_bw() + theme(legend.position = "none")
```

The connecting lines show each patient's trajectory. Most patients sleep longer on Drug 2.
The paired design captures this within-patient pattern that a two-sample test would miss.

```{r sleep-test}
#| eval: true

sleep_wide <- sleepdat |>
  tidyr::pivot_wider(names_from = drug, values_from = extra_hours, id_cols = ID)

res_sleep <- t.test(sleep_wide[["Drug 2"]], sleep_wide[["Drug 1"]], paired = TRUE)
tidy(res_sleep)
```

**Why paired, not two-sample?** Run both to see the difference:

```{r sleep-compare}
#| eval: true

bind_rows(
  tidy(t.test(extra_hours ~ drug, data = sleepdat, var.equal = FALSE)) |>
    mutate(test = "Two-sample (WRONG for this design)"),
  tidy(res_sleep) |>
    mutate(test = "Paired (CORRECT)")
) |>
  select(test, estimate, conf.low, conf.high, p.value)
```

The paired test has a smaller p-value because it eliminates between-patient variability
(some people simply sleep more than others regardless of drug).


## Example 2: Clinical Data (PBC)

### One-Sample: Is Mean Albumin Below a Clinical Threshold?

Normal serum albumin is >= 3.5 g/dL. Is the mean albumin in PBC patients below this threshold?

```{r pbc-setup}
#| eval: true

pbc <- survival::pbc |>
  as_tibble() |>
  janitor::clean_names() |>
  mutate(
    sex = factor(sex, levels = c("f", "m"), labels = c("Female", "Male")),
    trt = factor(trt, levels = c(1, 2), labels = c("D-penicillamine", "Placebo"))
  )

tidy(t.test(pbc$albumin, mu = 3.5))
```

`broom::tidy()` extracts key numbers into a data frame, useful for inline reporting.

### Two-Sample: Does Albumin Differ Between Treatment Groups?

```{r pbc-two}
#| eval: true

pbc_rand <- pbc |> filter(!is.na(trt), !is.na(albumin))

pbc_rand |>
  ggplot(aes(x = trt, y = albumin, fill = trt)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1.5) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(title = "Serum albumin by treatment group",
       x = NULL, y = "Albumin (g/dL)") +
  theme_bw() + theme(legend.position = "none")
```

```{r pbc-two-test}
#| eval: true

tidy(t.test(albumin ~ trt, data = pbc_rand, var.equal = FALSE))
```

::: {.callout-tip}
### Translating Numbers into a Sentence

> *"Mean albumin was 3.43 g/dL in PBC patients (95% CI: 3.39--3.47), significantly below the
> clinical threshold of 3.5 g/dL (one-sample Welch t-test: t = -3.12, df = 417, p = 0.002).
> There was no statistically significant difference in albumin between treatment groups."*

Always ask whether the difference is **clinically meaningful**, not just statistically significant.
:::


## Checking Assumptions: QQ-plots

For small samples ($n < 30$), check that data are approximately normal using a QQ-plot. `ToothGrowth` has only 10 animals per group at each dose -- exactly the situation where checking matters.

```{r qq-plot}
tg_1mg |>
  ggplot(aes(sample = len)) +
  stat_qq() +
  stat_qq_line(colour = "red") +
  facet_wrap(~ supp) +
  labs(title = "QQ-plot of tooth length by supplement (1.0 mg/day)") +
  theme_bw()
```

With $n = 10$ per group, points roughly following the line are sufficient. If you see heavy tails or a systematic curve, consider a Wilcoxon test (see [Nonparametric Tests](nonparametric-tests.qmd)).


## Where to Find Data Like This {.callout-note}

**`ToothGrowth`** (base R, used in Example 1): Dose-response experiment in guinea pigs. Ideal for two-sample t-tests and one-way ANOVA.

**`sleep`** (base R, used in Example 1): Paired drug comparison in 10 patients. The definitive paired t-test teaching dataset.

**`survival::pbc`** (used in Example 2): Clinical trial data with many continuous biomarkers suitable for one-sample and two-sample comparisons.

**`MASS::birthwt`:** Birth weight data with a smoking indicator -- good for a two-sample t-test on birth weight between smokers and non-smokers.

```{r}
#| eval: false
data(birthwt, package = "MASS")
t.test(bwt ~ smoke, data = birthwt)
```


## What Can Go Wrong {.callout-warning}

**Using a paired test when data are independent (or vice versa)**
The paired t-test requires one-to-one matching. If you use it on independent group data, you will get wrong degrees of freedom and an incorrect p-value.

**Testing normality with a formal test**
Shapiro-Wilk (`shapiro.test()`) is not a reliable guide for choosing a t-test. With small samples it has low power; with large samples it rejects normality for trivially small deviations. Use a QQ-plot and think about the biology instead.

**Reporting only the p-value**
Always report the mean difference and 95% CI alongside the p-value. "p = 0.03" tells the reader almost nothing about the size or direction of the effect.

**Not accounting for unequal variances**
If the two groups have very different standard deviations, Welch's t-test (default in R) is the right choice. Using `var.equal = TRUE` when variances differ inflates the Type I error rate.

::: {.callout-warning}
### Common Misinterpretations  -  Don't Make These Mistakes

**"p = 0.04 means the difference is large"**
No. With 400 patients, even a 0.05 g/dL difference (clinically trivial) can produce p = 0.04. Always report the mean difference and confidence interval so the reader can judge clinical relevance.

**"p = 0.07 means the groups are the same"**
No. It means your sample did not provide enough evidence to rule out chance at a = 0.05. With only 20 patients this may simply mean the study was underpowered. Never conclude "no difference" from a non-significant p-value alone.

**"I should use a one-sided test to get a smaller p-value"**
Only use a one-sided test if you genuinely predicted the direction of the effect *before collecting data*. Using a one-sided test because the data happened to go the right way is p-hacking and gives a false sense of significance.

**"The paired t-test always gives a lower p-value"**
The paired test is more powerful *only when* the within-subject measurements are correlated (e.g., before/after on the same patient). If your data are truly independent, using a paired test is wrong, not more powerful.

**"t-tests require normally distributed data"**
Not exactly. The t-test requires that the *sampling distribution of the mean* is approximately normal. By the Central Limit Theorem, this holds for n = 30 even with skewed data. For very small samples with obviously non-normal data, use a nonparametric test instead.
:::


## Exercises

### Exercise 1 (Guided): One-Sample Test for Bilirubin

The upper limit of normal for bilirubin is about 1.2 mg/dL.

1. Plot the bilirubin distribution. Is it normal?
2. Log-transform bilirubin. Test whether mean log-bilirubin equals `log(1.2)`.
3. Report: mean, 95% CI, t-statistic, df, p-value, and a one-sentence interpretation.

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r}
# 1. Distribution
ggplot(pbc, aes(x = bili)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(x = "Bilirubin (mg/dL)", y = "Count") + theme_bw()

# 2. One-sample test on log scale
res <- t.test(log(pbc$bili), mu = log(1.2))
tidy(res)

# 3. Back-transform the CI
exp(res$conf.int)
```

The geometric mean bilirubin is well above 1.2 mg/dL (the upper limit of normal), and this difference is highly statistically significant (p << 0.001). This confirms that PBC patients have elevated bilirubin, consistent with cholestatic liver disease.
:::

### Exercise 2 (Semi-guided): Two-Sample Comparison

Using `ToothGrowth`, test whether tooth length differs by supplement type across **all** dose levels (not just 1 mg/day).

1. Run a two-sample Welch's t-test.
2. Is the test valid? Check sample sizes and a QQ-plot.
3. Run the test also on log-transformed tooth length. Do the conclusions differ?

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r}
tg_all <- ToothGrowth %>%
  as_tibble() %>%
  mutate(supp = factor(supp))

# Two-sample test
tidy(t.test(len ~ supp, data = tg_all))

# QQ-plot by group
tg_all %>%
  ggplot(aes(sample = len)) +
  stat_qq() + stat_qq_line(color = "red") +
  facet_wrap(~ supp) + theme_bw()

# Log scale
tidy(t.test(log(len) ~ supp, data = tg_all))
```

With 30 observations per group, the CLT applies. The QQ-plot shows approximate normality. The raw and log-transformed tests give similar conclusions.
:::

### Exercise 3 (Open-ended)

Use a dataset from your own research. Identify a comparison where a t-test is appropriate. Run the test, check assumptions, and write a Methods + Results paragraph as you would for a journal submission.


## Comprehension Check

1. What is the difference between a one-sample, two-sample, and paired t-test? Give a biomedical example for each.
2. When should you use Welch's t-test instead of Student's t-test?
3. A QQ-plot of your data shows points diverging from the line at the upper tail. The sample size is n = 12. What should you do?
4. You run a two-sample t-test and get p = 0.04, 95% CI for difference: (0.01 g/dL to 2.1 g/dL). Is this a convincing result? What would you report?
5. A colleague says "I tested normality with Shapiro-Wilk (p = 0.06) so the data are normal." What is wrong with this reasoning?

::: {.callout-note collapse="true"}
### Answers
1. **One-sample:** compare a sample mean to a known or hypothetical value (e.g., "Is mean haemoglobin in my patients below 12 g/dL?"). **Two-sample:** compare means of two independent groups (e.g., "Does albumin differ between treatment and control mice?"). **Paired:** compare two measurements on the same subjects (e.g., "Did blood pressure change from baseline to 6 weeks?").
2. Use Welch's when the two groups may have different population variances. It is the default in R (`var.equal = FALSE`) and is almost always appropriate. Student's t-test (equal variance assumed) is rarely preferred.
3. With n = 12 and non-normality, use a nonparametric Wilcoxon test (see next session) instead of a t-test.
4. The result is statistically significant (p = 0.04) and the 95% CI excludes zero (0.01 to 2.1 g/dL), but the CI is very wide. The true difference could be as small as 0.01 g/dL (negligible) or as large as 2.1 g/dL (potentially important). The p-value alone tells you almost nothing useful here  -  the CI is the critical information. Always report both, and note that a barely significant result with a wide CI is actually uninformative about effect size.
5. Shapiro-Wilk with p = 0.06 is not evidence that data are normal; it means you failed to detect non-normality at the 5% level. With n = 12, the test has low power and would miss moderate non-normality. Use a QQ-plot and domain knowledge instead.
:::


## How to Report

### Reporting t-test Results

::: {.callout-note}
**In a methods section:** "Differences in [outcome] between [group A] and [group B] were compared using an independent-samples t-test. A paired t-test was used for [paired outcome] because measurements were taken on the same subjects."

**In results:** "[Outcome] was [higher/lower] in [group A] (mean ± SD) compared with [group B] (t(df) = X, p = Y, 95% CI for difference: [L, U])."

**What to always include:**

- Means and SDs for both groups
- Test statistic (t) and degrees of freedom (df)
- p-value
- 95% CI for the mean difference
- Whether the test was one- or two-sided

**Common reporting error:** Writing only p < 0.05 without the means, SDs, or CI. Readers cannot judge clinical significance without the actual effect size.
:::

## Further Reading

- @Welch-1947 --- Original Welch t-test paper
- @Bland-2015 --- Chapter on t-tests in *An Introduction to Medical Statistics*
- `?t.test` in R ? the help page has a detailed description of all options
