# ANOVA: Comparing Means Across Groups {#sec-anova}

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(survival)
library(broom)
```

::: {.callout-important}
## Before You Start
ANOVA extends the t-test to three or more groups. If you haven't already, complete the **t-tests** session first.
:::

## When Do You Use This? {.callout-tip}

> You recorded albumin levels in 418 PBC patients at four stages of disease (Stage 1–4). You want to know whether albumin differs across these stages simultaneously — running six separate t-tests would inflate your Type I error. ANOVA tests all groups at once.

## Learning Objectives

After completing this session you will be able to:

- Explain what ANOVA tests and when it is the right choice
- Run a one-way ANOVA in R and interpret the F-statistic and p-value
- Conduct post-hoc pairwise comparisons with Tukey's HSD
- Run a two-way ANOVA and interpret a significant interaction term
- Check ANOVA assumptions: normality of residuals, homogeneity of variance

---

## Background: The Idea Behind ANOVA

A t-test compares two group means. ANOVA (**AN**alysis **O**f **VA**riance) generalises this to $k$ groups. Despite the name, ANOVA tests means, not variances.

**How it works:** ANOVA partitions total variability into:
- **Between-group** variability (signal): differences between group means
- **Within-group** variability (noise): natural variation within each group

The **F-statistic** is the ratio of these two quantities:

$$F = \frac{\text{Mean square between groups}}{\text{Mean square within groups}}$$

A large F means group differences are large relative to the within-group scatter. Under the null hypothesis that all group means are equal, F follows an F-distribution.

**ANOVA assumptions:**
1. Independence: observations are independent
2. Normality: residuals (within-group deviations) are approximately normal
3. Homogeneity of variance: all groups have similar variance (Levene's test)

---

## Worked Example 1: Clinical Data (PBC)

### One-Way ANOVA: Albumin by Disease Stage

Does albumin decline across the four disease stages?

```{r pbc-setup}
pbc <- survival::pbc |>
  as_tibble() |>
  janitor::clean_names() |>
  mutate(
    stage = factor(stage, levels = 1:4, labels = paste("Stage", 1:4)),
    trt   = factor(trt, levels = c(1, 2), labels = c("D-penicillamine", "Placebo"))
  )

pbc_stage <- pbc |> filter(!is.na(stage), !is.na(albumin))
```

Always visualise first:

```{r pbc-anova-plot}
pbc_stage |>
  ggplot(aes(x = stage, y = albumin, fill = stage)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1.5) +
  scale_fill_brewer(palette = "Blues") +
  labs(title = "Serum albumin by disease stage",
       x = NULL, y = "Albumin (g/dL)") +
  theme_bw() + theme(legend.position = "none")
```

```{r pbc-anova}
fit_aov <- aov(albumin ~ stage, data = pbc_stage)
summary(fit_aov)
```

**Interpreting the output:**
- `Df`: degrees of freedom (between groups = $k-1 = 3$; within groups = $n-k$)
- `Sum Sq`: sum of squares for between-group and within-group variation
- `Mean Sq`: sum of squares divided by degrees of freedom
- `F value`: ratio of the two mean squares
- `Pr(>F)`: p-value; if small, at least one pair of means differs

### Post-Hoc Comparisons: Tukey's HSD

A significant ANOVA only tells you *some* groups differ. Tukey's Honestly Significant Difference (HSD) test identifies *which* pairs:

```{r tukey}
TukeyHSD(fit_aov) |> tidy() |>
  select(contrast, estimate, conf.low, conf.high, adj.p.value) |>
  arrange(adj.p.value)
```

### Checking Assumptions

```{r anova-diagnostics}
# Residual plot and QQ-plot
par(mfrow = c(1, 2))
plot(fit_aov, which = c(1, 2))
par(mfrow = c(1, 1))
```

```{r levene}
# Levene's test for homogeneity of variance
pbc_stage |>
  group_by(stage) |>
  summarise(n = n(), mean = mean(albumin), sd = sd(albumin))
```

If the standard deviations differ substantially (rule of thumb: max/min > 2), consider Welch's ANOVA (`oneway.test()`) or a nonparametric alternative (Kruskal-Wallis).

### Welch's ANOVA (Relaxing Equal Variance Assumption)

```{r welch-anova}
oneway.test(albumin ~ stage, data = pbc_stage, var.equal = FALSE)
```

---

## Worked Example 2: Animal Data (Two-Way ANOVA)

The `ToothGrowth` dataset is a 2×3 factorial experiment: two Vitamin C delivery methods × three dose levels. Two-way ANOVA tests for the effect of each factor *and* their interaction.

```{r tg-setup}
tg <- ToothGrowth |>
  as_tibble() |>
  mutate(
    supp = factor(supp, levels = c("VC", "OJ"),
                  labels = c("Ascorbic acid", "Orange juice")),
    dose = factor(dose)
  )
```

```{r tg-interaction-plot}
tg |>
  group_by(supp, dose) |>
  summarise(mean_len = mean(len), se_len = sd(len) / sqrt(n()), .groups = "drop") |>
  ggplot(aes(x = dose, y = mean_len, color = supp, group = supp)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_len - se_len, ymax = mean_len + se_len), width = 0.1) +
  scale_color_manual(values = c("steelblue", "tomato")) +
  labs(title = "Interaction plot: dose × supplement",
       x = "Dose (mg/day)", y = "Mean tooth length (μm)", color = "Supplement") +
  theme_bw()
```

If the lines cross or have markedly different slopes, there is a **dose × supplement interaction**: the effect of dose depends on which supplement was used.

```{r two-way-anova}
fit2 <- aov(len ~ supp * dose, data = tg)
summary(fit2)
tidy(fit2)
```

**Reading the output:**
- `supp`: main effect of supplement type
- `dose`: main effect of dose
- `supp:dose`: interaction term — does the effect of dose differ between the two supplements?

When the interaction is significant, the main effects alone do not tell the full story. Report the interaction and plot it.

```{r two-way-posthoc}
# Post-hoc: all pairwise combinations
TukeyHSD(fit2) |> tidy() |>
  arrange(adj.p.value) |>
  head(10)
```

---

## Where to Find Data Like This {.callout-note}

**PBC dataset:** `survival::pbc` — four-stage disease data, ideal for one-way ANOVA.

**ToothGrowth:** `datasets::ToothGrowth` — built-in factorial experiment, ideal for two-way ANOVA.

**PlantGrowth:** `datasets::PlantGrowth` — simple three-group ANOVA.

**agridat package:** Many balanced factorial agricultural/animal experiments suitable for two-way ANOVA.

---

## What Can Go Wrong {.callout-warning}

**Running multiple t-tests instead of ANOVA**
If you test all pairs of groups with separate t-tests, you inflate the Type I error rate. With 4 groups, there are 6 comparisons. At α = 0.05, you expect 0.3 false positives even with no real effects. Use ANOVA followed by a single post-hoc procedure.

**Ignoring a significant interaction**
If the interaction term is significant in a two-way ANOVA, the main effects are misleading. For example, supplement A may be better only at low doses. Always plot the interaction.

**Not checking the equal variance assumption**
ANOVA is reasonably robust to mild variance inequality, but not severe inequality. Check residual plots and use Welch's ANOVA (`oneway.test()`) or Kruskal-Wallis if variances are very unequal.

**Using ANOVA on very small, non-normal groups**
With small samples (< 10 per group) and clearly non-normal residuals, use Kruskal-Wallis instead.

---

## Exercises

### Exercise 1 (Guided): One-Way ANOVA for Bilirubin by Stage

Using `survival::pbc`, test whether log(bilirubin) differs by stage.

1. Plot log(bilirubin) by stage.
2. Run `aov(log(bili) ~ stage, ...)`. What do you conclude?
3. Run Tukey's HSD. Which stage pairs differ?
4. Check residual normality with a QQ-plot.

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r}
pbc_s <- pbc |> filter(!is.na(stage), !is.na(bili))

pbc_s |>
  ggplot(aes(x = stage, y = log(bili), fill = stage)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_brewer(palette = "Blues") +
  labs(x = NULL, y = "log(Bilirubin)") + theme_bw() + theme(legend.position = "none")

fit_bili <- aov(log(bili) ~ stage, data = pbc_s)
summary(fit_bili)
TukeyHSD(fit_bili) |> tidy() |> arrange(adj.p.value)

qqnorm(residuals(fit_bili)); qqline(residuals(fit_bili), col = "red")
```
:::

### Exercise 2 (Semi-guided): Two-Way ANOVA for PlantGrowth

Using `datasets::PlantGrowth`, run a one-way ANOVA comparing weight across the three groups.

1. Visualise weight by group.
2. Run `aov(weight ~ group, ...)`.
3. Post-hoc with Tukey's HSD.
4. Is the assumption of equal variance plausible? Use `leveneTest()` from the `car` package or compare group SDs.

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r}
pg <- PlantGrowth |> as_tibble()

pg |>
  ggplot(aes(x = group, y = weight, fill = group)) +
  geom_boxplot(alpha = 0.7) + geom_jitter(width = 0.1, alpha = 0.5) +
  theme_bw() + theme(legend.position = "none")

fit_pg <- aov(weight ~ group, data = pg)
summary(fit_pg)
TukeyHSD(fit_pg) |> tidy()

pg |> group_by(group) |> summarise(sd = sd(weight))
```
:::

### Exercise 3 (Open-ended)

Using a dataset from your own research, identify a continuous outcome and a categorical grouping variable (≥ 3 groups). Run a one-way ANOVA. Check the assumptions. Write a Methods section (one paragraph) and a Results section (two sentences) suitable for a journal paper.

---

## Comprehension Check

1. What is the null hypothesis in a one-way ANOVA? What does a significant result tell you?
2. Why should you run Tukey's HSD after a significant ANOVA rather than multiple t-tests?
3. In a two-way ANOVA output, the `supp:dose` row is significant. What does this mean, and how should you proceed?
4. Your residual QQ-plot shows clear non-normality. Your groups have n = 8 each. What should you do?
5. ANOVA tests variances according to its name, so it assumes the group variances are equal. True or false?

::: {.callout-note collapse="true"}
### Answers
1. The null hypothesis is that all group population means are equal: $H_0: \mu_1 = \mu_2 = \ldots = \mu_k$. A significant result means at least one group mean differs from at least one other — but not which pairs differ.
2. Running $\binom{k}{2}$ t-tests independently inflates the family-wise error rate. With $k = 4$ groups there are 6 pairs; at α = 0.05 you expect false positives. Tukey's HSD controls the family-wise error rate by adjusting the p-value for the number of comparisons.
3. A significant interaction means the effect of supplement depends on dose (or vice versa). The main effects alone are misleading. You should plot the interaction (an interaction plot) and report group-specific means rather than marginal means.
4. With small non-normal groups, use Kruskal-Wallis (a nonparametric ANOVA alternative that is robust to non-normality).
5. **False.** Despite the name, ANOVA tests whether group *means* are equal, not variances. However, it does *assume* that the within-group variances are equal (homogeneity of variance). The name refers to the technique of partitioning total variance into components.
:::

## Further Reading

- @Dalgaard-2008 --- Chapter on ANOVA in *Introductory Statistics with R*
- @Bland-2015 --- One-way and two-way ANOVA in *An Introduction to Medical Statistics*
- `?aov`, `?TukeyHSD`, `?oneway.test` in R — built-in documentation
