# ANOVA & Multiple Comparisons {#anova}

```{r setup}
#| include: false
#| eval: true
source(here::here("R/functions.R"))
source(here::here("R/load-packages.R"))
```

One-way ANOVA compares means across three or more groups; use post-hoc
comparisons to identify which groups differ.

## When to use ANOVA

Use ANOVA for group mean comparisons when the outcome is continuous and you
have three or more groups. Follow with a multiplicity-corrected post-hoc test
if you need pairwise contrasts.

## Short code demo (dplyr-style)

```{r demo-anova, echo=TRUE}
pg <- datasets::PlantGrowth %>% as_tibble()
pg_aov <- aov(weight ~ group, data = pg)
summary(pg_aov)

# Tukey pairwise comparisons
TukeyHSD(pg_aov)
```

## How to explain the output (teacher notes)

- **Report:** global F-test result and pairwise comparisons (adjusted CIs).
- **Context:** emphasize effect sizes and which contrasts are clinically meaningful.

## Short exercise

- Run the ANOVA and report which groups differ using Tukey's test; write a one-line conclusion.

**Hint:** use `TukeyHSD()` on the fitted `aov` object.

## Packages used

- `dplyr`
