# ANOVA & Multiple Comparisons {#anova}

```{r setup}
#| include: false
#| eval: true
source(here::here("R/functions.R"))
source(here::here("R/load-packages.R"))
```

One-way ANOVA compares means across three or more groups; use post-hoc
comparisons to identify which groups differ.

## When to use ANOVA

Use ANOVA for group mean comparisons when the outcome is continuous and you
have three or more groups. Follow with a multiplicity-corrected post-hoc test
if you need pairwise contrasts.

## Short code demo

```{r demo-anova, echo=TRUE}
pg <- datasets::PlantGrowth %>% as_tibble()
pg_aov <- aov(weight ~ group, data = pg)
summary(pg_aov)

# Tukey pairwise comparisons
TukeyHSD(pg_aov)
```

## How to explain the output (teacher notes)

- **Report:** global F-test result and pairwise comparisons (adjusted CIs).
- **Context:** emphasize effect sizes and which contrasts are clinically meaningful.

## Short exercise

- Run the ANOVA and report which groups differ using Tukey's test; write a one-line conclusion.

**Hint:** use `TukeyHSD()` on the fitted `aov` object.

## Packages used

- `dplyr`

**Exercises**

- See `dev_code_data/excercise4.R` for worked examples on dose–response comparisons, pairwise tests, and model diagnostics.

::: {.callout-tip collapse="true" title="Exercise 4 — Solution"}
- Solution (PDF): `dev_code_data/Solution4.pdf`  
- Solution script: `dev_code_data/Solution4.R`  
Run: `Rscript dev_code_data/Solution4.R` or open the PDF in your browser/RStudio.
:::

## Further reading

- ANOVA and multiple comparisons: @Teetor-2011; @Matloff-2011.

## Resources

- Example script: `dev_code_data/excercise4.R` (dose–response comparisons and post-hoc tests).
