# Multiple Regression & Confounding {#multiple-regression}

```{r setup}
#| include: false
#| eval: true
source(here::here("R/functions.R"))
source(here::here("R/load-packages.R"))
```

Multiple regression models an outcome using several predictors and helps
assess confounding and adjusted associations.

## When to use multiple regression

Use multiple regression to estimate associations while adjusting for potential
confounders; check collinearity and model diagnostics.

## Short code demo (dplyr-style)

```{r demo-multilm, echo=TRUE}
mtcars <- datasets::mtcars %>% as_tibble(rownames = "model") %>%
  mutate(am = factor(am, labels = c("auto", "manual")))

multilm <- lm(mpg ~ wt + hp + am, data = mtcars)
broom::tidy(multilm)
broom::glance(multilm)
```

## How to interpret the output (teacher notes)

- **Adjusted coefficients:** each coefficient represents the expected change
  in the outcome holding other variables constant.
- **Diagnostics:** check residuals, leverage, and variance inflation where
  appropriate.

## Short exercise

- Fit `mpg ~ wt + hp` and compare the `wt` coefficient to the simple model `mpg ~ wt`.

**Hint:** use `broom::tidy()` and `broom::glance()` for concise summaries.

## Packages used

- `dplyr`, `broom`

**Exercises**

- See `dev_code_data/excercise7.Rmd` for a worked example on multiple regression with interaction terms, centering, and adjusted comparisons (birth-weight example).

::: {.callout-tip collapse="true" title="Exercise 7 — Solution"}
- Solution (PDF): `dev_code_data/Solution7.pdf`  
- Solution script: `dev_code_data/Solution7.R`  
Run: `Rscript dev_code_data/Solution7.R` or open the PDF in your browser/RStudio.
:::

## Further reading

- Confounding, model interpretation, and interactions: @Henderson-Velleman-1981; @Matloff-2011.
- Tidy modeling workflows: @Wickham-Grolemund-2017.

## Resources

- Example: `dev_code_data/excercise7.Rmd` (interaction terms and interpreting adjusted coefficients).
