# Advanced Visualization and Scientific Communication {#sec-advanced-viz}

```{r setup}
#| include: false
#| eval: true
source(here::here("R/functions.R"))
source(here::here("R/load-packages.R"))
pbc <- survival::pbc %>%
  as_tibble() %>%
  janitor::clean_names() %>%
  mutate(
    sex = factor(sex, levels = c("f", "m"), labels = c("Female", "Male")),
    stage = factor(stage, labels = c("Stage 1", "Stage 2", "Stage 3", "Stage 4"))
  )
```

## Introduction

This session shows how research figures support statistical interpretation.
A strong plot does more than display data: it answers a question, shows
uncertainty, and makes the result easier to explain.

## Learning outcomes

- Choose visual forms that fit a research question.
- Interpret trend lines, confidence bands, and subgroup comparisons.
- Create layered and faceted plots with `ggplot2`.
- Use annotation and labels to communicate the key message.
- Build figures that support evidence-based conclusions.

## Methods covered in this session

- Regression line visualization and interpretation
- Confidence and prediction interval plotting
- Group comparisons using color and facets
- Model diagnostics and residual visualization
- Communicating uncertainty with plots

## 1. Design principles for research plots

A useful research figure is:

- **Clear**: the main message is easy to read.
- **Accurate**: the visual reflects the underlying data.
- **Informative**: it answers a specific question.
- **Interpretable**: it helps readers draw the right conclusion.
- **Honest**: it shows uncertainty and avoids misleading design.

## 2. Use a consistent dataset for interpretation

We use the `pbc` dataset from the `survival` package in this session.
That allows us to connect visualization directly to clinical research
questions about bilirubin, age, and sex.

```{r pbc-summary}
pbc %>%
  summarize(
    n = n(),
    mean_age = mean(age, na.rm = TRUE),
    mean_bili = mean(bili, na.rm = TRUE),
    prop_female = mean(sex == "Female")
  )
```

This dataset is useful because it includes both continuous and categorical
variables, which lets us practice different plot types.

## 3. Layered plotting with ggplot2

Layered plots let us combine raw data, trend lines, and uncertainty in one
figure.

```{r layered-plot}
pbc %>%
  ggplot(aes(x = age, y = bili, color = sex)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = TRUE, linetype = "dashed") +
  labs(
    title = "Bilirubin and age by sex",
    subtitle = "Trend lines with 95% uncertainty bands",
    x = "Age (years)",
    y = "Bilirubin (mg/dL)",
    color = "Sex"
  ) +
  theme_minimal(base_size = 14)
```

### What this figure tells us

- The points show the raw relationship between age and bilirubin.
- The trend lines estimate the average pattern for each sex.
- The shaded bands show uncertainty around those estimates.
- If the bands overlap, the difference between groups is less certain.

## 4. Faceting to compare subgroups

Faceting makes it easy to compare the same relationship across groups.

```{r faceting}
pbc %>%
  ggplot(aes(x = age, y = bili)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  facet_wrap(~ sex) +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(
    title = "Bilirubin by age for female and male patients",
    x = "Age (years)",
    y = "Bilirubin (mg/dL)"
  ) +
  theme_minimal()
```

Faceting is useful when the research question asks whether a pattern is
consistent across groups. It highlights differences in slope, spread, and
uncertainty.

## 5. Visualizing uncertainty

A key strength of research graphics is showing how precise the estimates are.

```{r uncertainty}
pbc %>%
  ggplot(aes(x = age, y = bili)) +
  geom_point(color = "steelblue", alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, fill = "lightblue") +
  labs(
    title = "Age and bilirubin with uncertainty band",
    x = "Age (years)",
    y = "Bilirubin (mg/dL)"
  ) +
  theme_minimal()
```

The ribbon shows a confidence band around the trend line. Wider bands
indicate more uncertainty, which is especially important when the data are
noisy or sparse.

## 6. Annotation and narrative

The best figures include titles, subtitles, and axis labels that explain
the message. Good annotations help a reader see the conclusion without
needing to inspect every detail.

```{r theme-annotation}
pbc %>%
  ggplot(aes(x = age, y = bili, color = sex)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Bilirubin increases slightly with age",
    subtitle = "The trend is visible for both female and male patients",
    x = "Age (years)",
    y = "Bilirubin (mg/dL)",
    color = "Sex"
  ) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11)
  )
```

This figure is prepared for a report because it has a clear title and a
subtitle that state the main conclusion.

## 7. Exercise: refine an analysis plot

1. Use the `pbc` dataset from the `survival` package.
2. Plot `bili` versus `age`.
3. Add a smooth line and uncertainty band.
4. Separate by `sex` using color or facets.
5. Write a one-sentence summary of what the plot shows.

## 8. Model diagnostics and prediction visualization

Checking model assumptions and visualizing predictions are essential
steps before turning plots into clinical messages.

```{r diagnostics-predictions}
# Fit a simple linear model and inspect residuals
lm_bili_age <- lm(bili ~ age + sex, data = pbc)
pbc_diag <- pbc %>%
  mutate(fitted = fitted(lm_bili_age), resid = residuals(lm_bili_age))

# Residuals vs fitted values
ggplot(pbc_diag, aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted: bili ~ age + sex", x = "Fitted values", y = "Residuals") +
  theme_minimal()

# QQ-plot of residuals
ggplot(pbc_diag, aes(sample = resid)) +
  stat_qq() +
  stat_qq_line(color = "firebrick") +
  labs(title = "QQ-plot of model residuals") +
  theme_minimal()

# Prediction ribbon across age
new_age <- expand.grid(age = seq(min(pbc$age, na.rm = TRUE), max(pbc$age, na.rm = TRUE), length.out = 100),
                       sex = levels(pbc$sex)) %>%
  as_tibble()
preds_conf <- predict(lm_bili_age, newdata = new_age, interval = "confidence")
preds_df <- bind_cols(new_age, as_tibble(preds_conf))

ggplot(preds_df, aes(x = age, y = fit, color = sex, fill = sex)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.15, color = NA) +
  facet_wrap(~ sex) +
  labs(title = "Predicted mean bilirubin by age with 95% CI", y = "Predicted bilirubin (mg/dL)") +
  theme_minimal()
```

### How to explain diagnostics and predictions to a clinician

- **Diagnostics:** "I checked residuals and a QQ-plot to assess whether model assumptions (linearity, homoscedasticity, approximate normality of residuals) are reasonable. If residuals show systematic patterns, consider transformation or robust methods."  
- **Predictions:** "Report predicted mean values for representative ages with 95% confidence intervals, and give prediction intervals when discussing individual patient-level expectations. Confidence bands show uncertainty of the mean estimate; prediction intervals are wider and capture individual variability."

## 8. Key takeaways

Advanced visualization is not about making plots more complex. It is about
making research findings easier to understand, more accurate, and more
persuasive.

A good research figure answers a question, shows uncertainty, and helps a
reader draw the right conclusion.

## Further reading

- Advanced plotting and scientific communication: @Wickham-2016-ggplot2; @Chang-2018.
- Tidy workflows for visualization: @Wickham-Grolemund-2017.

## Resources

- Example demos moved from `dev_code_data`; use inline examples above or see `data_sources.md` for dataset choices and suggested demo scripts.
