# Advanced Visualization and Scientific Communication {#sec-advanced-viz}

```{r setup}
#| include: false
#| eval: true
source(here::here("R/functions.R"))
source(here::here("R/load-packages.R"))
```

## Introduction

This session moves beyond basic charts to research-quality figures.
We cover how to design plots that support clinical conclusions, show
uncertainty, and communicate results to non-technical audiences.

## Learning outcomes

- Create layered and faceted plots using `ggplot2`.
- Apply thoughtful themes, colors, and annotations for clarity.
- Visualize uncertainty with ribbons, error bars, and confidence bands.
- Build plots that support clinical interpretation and decision-making.

## 1. Design principles for research plots

A good research figure is:

- **Clear**: the message is easy to read.
- **Accurate**: the visual representation matches the data.
- **Informative**: it supports a specific research question.
- **Reusable**: it can be included in reports and presentations.

## 2. Layered plotting with ggplot2

```{r layered-plot}
mtcars <- datasets::mtcars %>%
  as_tibble(rownames = "model") %>%
  mutate(am = factor(am, labels = c("Automatic", "Manual")))

mtcars %>%
  ggplot(aes(x = wt, y = mpg, color = am)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, linetype = "dashed") +
  labs(
    title = "Fuel efficiency by weight and transmission type",
    subtitle = "A clear view of how weight and transmission influence mpg",
    x = "Weight (1000 lbs)",
    y = "Miles per gallon",
    color = "Transmission"
  ) +
  theme_minimal(base_size = 14)
```

### Explanation

- `geom_point()` displays the raw observations.
- `geom_smooth()` adds a trend line and uncertainty band.
- Color groups make it easy to compare categories.

## 3. Faceting to compare subgroups

Faceting creates a series of similar plots, one per subgroup.

```{r faceting}
mtcars %>%
  ggplot(aes(x = wt, y = mpg)) +
  geom_point(size = 2, alpha = 0.7) +
  facet_wrap(~ am) +
  labs(
    title = "MPG vs Weight by Transmission",
    x = "Weight (1000 lbs)",
    y = "Miles per gallon"
  ) +
  theme_minimal()
```

Faceting is useful when you want to compare the same relationship across
different clinical or experimental groups.

## 4. Visualizing uncertainty

Showing uncertainty is critical in research communication.

```{r uncertainty}
mtcars %>%
  ggplot(aes(x = wt, y = mpg)) +
  geom_point(color = "steelblue", alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, fill = "lightblue") +
  labs(
    title = "Regression with uncertainty band",
    x = "Weight (1000 lbs)",
    y = "Miles per gallon"
  ) +
  theme_minimal()
```

The shaded ribbon around the trend line shows the 95% confidence band,
which helps readers understand the precision of the estimate.

## 5. Publication-ready themes and annotation

A clean theme and well-chosen labels are essential.

```{r theme-annotation}
mtcars %>%
  ggplot(aes(x = wt, y = mpg, color = am)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Weight and fuel efficiency across transmission types",
    subtitle = "A publication-ready figure with clear annotations",
    x = "Weight (1000 lbs)",
    y = "Miles per gallon",
    color = "Transmission"
  ) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11)
  )
```

## 6. Exercise: refine an analysis plot

1. Use the `pbc` dataset from the `survival` package.
2. Plot `bili` versus `age`.
3. Add a smooth trend line and separate by `sex` using color or facets.
4. Add titles and axis labels that explain the plot to a clinical reader.

## 7. Key takeaways

Advanced visualization is not about making plots more complex. It is
about making research findings easier to understand, more accurate, and
more persuasive.
