---
title: "ANOVA: Comparing Means Across Groups"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(survival)
library(broom)
```

::: {.callout-note}
### Session at a Glance

**Total core time: about 120 minutes (~2 hours).**

| Section | Time | Type |
|---|---|---|
| When Do You Use This? + Learning Objectives | 5 min | Reading |
| Background: How ANOVA Works | 20 min | Reading |
| Example 1: One-Way ANOVA (PlantGrowth) | 25 min | Worked example |
| Example 2: Two-Way ANOVA (ToothGrowth) | 20 min | Worked example |
| What Can Go Wrong | 10 min | Reading |
| Exercises 1 & 2 | 30 min | Practice |
| Comprehension Check | 10 min | Self-test |

In a taught course, this session fits one teaching block. For
self-paced study, split it into two sittings, e.g. Background and
Example 1 first, Example 2 onward later. Exercise 3 is open-ended and
works best as a follow-up using a dataset from your own field.
:::

::: {.callout-tip}
### How to Use the Code in This Session

Every code chunk below is written so you can copy it into your R
console (or an R script) and run it yourself, in the order shown.
Where you see a "Run It Yourself" box, stop and run the code above it
before reading the explanation that follows.
:::

::: {.callout-important}
## Before You Start

ANOVA extends the t-test to three or more groups. Make sure you have
completed [T-tests and Group Comparisons](t-tests.qmd) first: the
logic here builds directly on the signal ÷ noise framework introduced
there.
:::

## When Do You Use This? {#sec-anova}

::: {.callout-tip}
You have a continuous outcome in three or more groups and you want to
know whether the group means differ. Maybe you are comparing a
biomarker across four disease stages, or a drug response across a
control and three dose levels. The temptation is to run all possible
pairs of t-tests. The problem: with four groups there are six pairs,
and at α = 0.05 the probability of at least one false positive from
six tests is 1 − 0.95⁶ = 26%. ANOVA tests all groups simultaneously
in a single model and keeps the false positive rate at 5%.
:::

::: {.callout-note collapse="true"}
### Teacher Note

Have the room verify 1 − 0.95⁶ = 26% on a calculator before moving on
— it is easy to nod along to "26%" without registering how large that
is. This is the same multiple-testing logic as Example 2 in
[Hypothesis Testing](hypothesis-testing.qmd); if you taught that
session, ask the room to connect the two before you explain ANOVA's
solution.
:::

## Learning Objectives

After completing this session you will be able to:

- Explain what the F-statistic is and how it relates to the t-statistic
- Run a one-way ANOVA in R and interpret the output table
- Run Tukey's HSD post-hoc test to identify which pairs differ
- Fit a two-way ANOVA and interpret a significant interaction term
- Check ANOVA assumptions using residual plots


## Background: How ANOVA Works

*Estimated time: ~20 minutes (reading)*

::: {.callout-note collapse="true"}
### Teacher Note

Connect this back to the t-statistic from [T-tests](t-tests.qmd):
for exactly two groups, ANOVA's F-statistic equals the square of the
two-sample t-statistic ($F = t^2$), and the ANOVA p-value equals the
two-sided t-test p-value. ANOVA is the generalisation of the t-test to
more than two groups, not a different idea.
:::

### The F-statistic is signal ÷ noise

In the t-tests session, the t-statistic was signal ÷ noise: the
difference between group means divided by the standard error of that
difference. The F-statistic follows the same logic but for multiple
groups simultaneously:

$$F = \frac{\text{variance between group means (signal)}}{\text{variance within groups (noise)}}$$

A large F means the group means spread out more than would be expected
from the natural variation within groups. Under the null hypothesis that
all group means are equal, F follows a known F-distribution. A small
p-value means "an F this large would be rare if the groups truly had
the same mean."

ANOVA works by **partitioning total variation** into two parts:

- **Between-group variation** (signal): how much the group means deviate
  from the overall mean
- **Within-group variation** (noise): how much individuals deviate from
  their own group mean

The ratio of these, after accounting for degrees of freedom, is the
F-statistic.

### The one thing ANOVA tells you, and the one thing it doesn't

A significant ANOVA tells you **at least one group mean differs from at
least one other**. It does not tell you which groups differ. That
requires a post-hoc test (Tukey's HSD is standard for ANOVA).

### ANOVA assumptions

| Assumption | How to check |
|---|---|
| **Independence** | Study design: each observation from a different subject |
| **Normality of residuals** | QQ-plot of model residuals (not raw data) |
| **Equal variances** | Check group SDs; use Welch's ANOVA if unequal |

ANOVA is reasonably robust to mild variance inequality, especially with
balanced group sizes. With small, clearly non-normal groups, use
Kruskal-Wallis instead (see [Nonparametric Tests](nonparametric-tests.qmd)).


## Example 1: One-Way ANOVA (PlantGrowth)

*Estimated time: ~25 minutes (worked example)*

::: {.callout-note collapse="true"}
### Teacher Note

After the boxplot (`pg-setup`), ask the room to predict, by eye, which
of the three pairwise comparisons (ctrl vs trt1, ctrl vs trt2, trt1 vs
trt2) will turn out significant in the Tukey output later. Most people
correctly guess trt1 vs trt2 because that gap looks largest relative
to the spread — a good opportunity to reinforce "signal relative to
noise," not just "signal."
:::

The `PlantGrowth` dataset records dried plant weight under three
conditions: a control and two experimental treatments, with 10 plants
per group.

```{r pg-setup}
#| eval: true

pg <- PlantGrowth %>% as_tibble()

pg %>%
  ggplot(aes(x = group, y = weight, fill = group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.6, size = 2.5) +
  scale_fill_manual(values = c("grey70", "steelblue", "tomato")) +
  labs(
    title = "Dried plant weight by treatment group",
    x     = NULL,
    y     = "Dry weight (g)"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

Treatment 2 (trt2) looks higher and treatment 1 (trt1) looks lower than
the control. Before running the test, notice this is the visual signal
ANOVA will try to detect relative to the within-group scatter.

```{r pg-anova}
#| eval: true

fit_pg <- aov(weight ~ group, data = pg)
summary(fit_pg)
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. You should get `F ≈ 4.85` on (2, 27) degrees of
freedom, with `Pr(>F) ≈ 0.016`. Below 0.05, so at least one group mean
differs. Which one(s)? The table alone cannot say — that is what
Tukey's HSD is for, next.
:::

::: {.callout-note}
### Reading the ANOVA Table

```
          Df  Sum Sq  Mean Sq  F value  Pr(>F)
group      2   3.77    1.88    4.85     0.016 *
Residuals 27  10.49    0.39
```

| Column | What it means |
|---|---|
| **Df (group = 2)** | Degrees of freedom for groups = number of groups − 1 = 3 − 1 = 2 |
| **Df (Residuals = 27)** | Degrees of freedom for error = total n − groups = 30 − 3 = 27 |
| **Sum Sq (group = 3.77)** | Total variation explained by group membership |
| **Sum Sq (Residuals = 10.49)** | Remaining variation within groups (noise) |
| **F = 4.85** | The between-group variance is 4.85× the within-group variance |
| **Pr(>F) = 0.016** | If all groups had the same true mean, an F this large would occur 1.6% of the time |

Conclusion: at least one group mean differs (p = 0.016). We do not yet
know which pairs. That requires a post-hoc test.
:::

### Post-hoc comparisons: Tukey's HSD

Tukey's Honestly Significant Difference test performs all pairwise
comparisons while controlling the family-wise error rate.

```{r pg-tukey}
#| eval: true

TukeyHSD(fit_pg) %>%
  tidy() %>%
  select(contrast, estimate, conf.low, conf.high, adj.p.value) %>%
  arrange(adj.p.value)
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. Only `trt2-trt1` has `adj.p.value ≈ 0.012`, below
0.05. The mean difference is about 0.865 g (95% CI: 0.17 to 1.56).
`trt2-ctrl` (adj.p ≈ 0.198) and `trt1-ctrl` (adj.p ≈ 0.391) are not
significant. Did this match your prediction?
:::

Read each row as a pairwise comparison: `estimate` is the mean
difference, `conf.low`/`conf.high` is the 95% CI for that difference,
and `adj.p.value` is the p-value adjusted for multiple comparisons.
The trt1 vs trt2 contrast drives the overall significance.

### Checking assumptions

Check two things: (1) are residuals approximately normal? (2) are
group variances similar?

```{r pg-diagnostics}
#| eval: true

aug_pg <- augment(fit_pg)

ggplot(aug_pg, aes(x = .fitted, y = .resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title    = "Residuals vs Fitted",
    subtitle = "No pattern = equal variance assumption met",
    x        = "Fitted values",
    y        = "Residuals"
  ) +
  theme_bw()

ggplot(aug_pg, aes(sample = .resid)) +
  stat_qq() +
  stat_qq_line(colour = "red") +
  labs(
    title    = "QQ-plot of residuals",
    subtitle = "Points near the line = normality assumption met"
  ) +
  theme_bw()
```

```{r pg-sd}
#| eval: true

pg %>%
  group_by(group) %>%
  summarise(n = n(), mean = mean(weight), sd = sd(weight))
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. Means are ctrl ≈ 5.03, trt1 ≈ 4.66, trt2 ≈ 5.53;
SDs are ctrl ≈ 0.58, trt1 ≈ 0.79, trt2 ≈ 0.44. The largest SD (0.79) is
less than twice the smallest (0.44), so the equal-variance assumption
is reasonable.
:::

The residual plot shows no pattern (equal variance holds). The QQ-plot
shows approximate normality. The group SDs are within a factor of 2 of
each other. The ANOVA assumptions are met.

::: {.callout-tip}
### What if variances are unequal?

If group SDs differ markedly (roughly, if the largest SD is more than
twice the smallest), use Welch's ANOVA:

```r
oneway.test(weight ~ group, data = pg, var.equal = FALSE)
```

Follow this with Games-Howell post-hoc tests rather than Tukey's HSD.
:::


## Example 2: Two-Way ANOVA (ToothGrowth)

*Estimated time: ~20 minutes (worked example)*

::: {.callout-note collapse="true"}
### Teacher Note

This dataset also appears in [Nonparametric Tests](nonparametric-tests.qmd)
(comparing the two supplements at dose 0.5 only). Here the design is
extended to all three doses, which lets the interaction term emerge.
Before showing the interaction plot, ask the room what a "no
interaction" plot would look like (two parallel lines), so the
converging lines below stand out as a meaningful pattern rather than
just "two lines that happen to be close."
:::

The `ToothGrowth` dataset is a 2 × 3 factorial experiment: two Vitamin
C delivery methods × three dose levels, 10 guinea pigs per cell. This
design lets us ask three questions simultaneously:

1. Does the supplement type affect tooth length? (main effect of `supp`)
2. Does the dose level affect tooth length? (main effect of `dose`)
3. Does the effect of dose *depend on* which supplement was used?
   (interaction between `supp` and `dose`)

The interaction question is the most interesting, and the most
commonly misunderstood.

```{r tg-setup}
#| eval: true

tg <- ToothGrowth %>%
  as_tibble() %>%
  mutate(
    supp = factor(supp, levels = c("VC", "OJ"),
                  labels = c("Ascorbic acid", "Orange juice")),
    dose = factor(dose)
  )
```

### Visualise before testing: the interaction plot

An **interaction plot** shows the mean response at each dose level for
each supplement. If the two lines are roughly parallel, the dose effect
is similar for both supplements (no interaction). If the lines cross or
converge, the dose effect differs by supplement, an interaction.

```{r tg-interaction-plot}
#| eval: true

tg %>%
  group_by(supp, dose) %>%
  summarise(
    mean_len = mean(len),
    se_len   = sd(len) / sqrt(n()),
    .groups  = "drop"
  ) %>%
  ggplot(aes(x = dose, y = mean_len, colour = supp, group = supp)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = mean_len - se_len, ymax = mean_len + se_len),
    width = 0.1
  ) +
  scale_colour_manual(values = c("steelblue", "tomato")) +
  labs(
    title    = "Interaction plot: mean tooth length by dose and supplement",
    subtitle = "Error bars = ±1 SE; non-parallel lines suggest an interaction",
    x        = "Dose (mg/day)",
    y        = "Mean tooth length (mm)",
    colour   = "Supplement"
  ) +
  theme_bw()
```

The lines are not parallel and converge at the highest dose. This
suggests the benefit of orange juice over ascorbic acid is larger at
lower doses, but the two supplements become similar at 2 mg/day. ANOVA
will test whether this interaction is statistically meaningful.

```{r two-way-anova}
#| eval: true

fit2 <- aov(len ~ supp * dose, data = tg)
summary(fit2)
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. All three rows are significant: `supp` (F ≈ 15.6,
p ≈ 0.0002), `dose` (F ≈ 92.0, p < 2e-16), and `supp:dose` (F ≈ 4.11,
p ≈ 0.022). Because the interaction term is significant, the two main
effects cannot be interpreted on their own — see below.
:::

::: {.callout-note}
### Reading the Two-Way ANOVA Table

| Row | What it tests |
|---|---|
| **supp** | Main effect: does supplement type affect length, averaged across doses? |
| **dose** | Main effect: does dose affect length, averaged across supplement types? |
| **supp:dose** | Interaction: does the *effect of dose* differ between supplements? |

When the **interaction is significant**, the main effects alone are
misleading. The story is more subtle: the effect of dose depends on
which supplement you use. Report the interaction and show the plot:
do not summarise by "supplement A is better" if it is only better at
certain doses.

When the interaction is **not** significant, the main effects can be
interpreted directly.
:::

```{r two-way-posthoc}
#| eval: true

TukeyHSD(fit2) %>%
  tidy() %>%
  arrange(adj.p.value) %>%
  head(10)
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. With 2 supplements × 3 doses there are 15 possible
pairwise contrasts; the ten shown here are the most significant. The
largest effect is `dose 2-0.5` (estimate ≈ 15.5 mm, adj.p ≈ 4e-13): the
dose effect dwarfs the supplement effect, consistent with `dose` having
by far the largest F-statistic above.
:::


## What Can Go Wrong

*Estimated time: ~10 minutes (reading)*

::: {.callout-note collapse="true"}
### Teacher Note

Quick true/false round on the "Common misinterpretations" callout below
— read each bold statement aloud and ask for a thumbs up/down before
revealing the explanation. The "F = 12.4 means a large effect" item
connects directly back to the η² discussion in "How to Report" later,
so it's worth flagging that effect size is still to come.
:::

**Running multiple t-tests instead of ANOVA**

With four groups and six pairwise t-tests at α = 0.05, the chance of
at least one false positive is 26%. This is not a minor concern: it
means roughly one in four such analyses produces a spurious finding by
chance. ANOVA + Tukey's HSD keeps the family-wise error rate at 5%.

**Ignoring a significant interaction in two-way ANOVA**

If `supp:dose` is significant, reporting only the main effects is
wrong. "Orange juice increases tooth length by 3.7 mm on average" is
misleading if orange juice is only better at low doses. Always check
the interaction term and plot the group means before summarising the
results.

**Claiming "all groups differ" after a significant ANOVA**

A significant F-test means *at least one* pair differs. The post-hoc
test identifies which ones. Stage 1 vs Stage 4 may differ significantly
while Stage 1 vs Stage 2 may not.

**Using ANOVA with very small, non-normal groups**

With fewer than 10 observations per group and visibly non-normal
residuals, use Kruskal-Wallis. ANOVA is robust to mild non-normality
with larger samples but not with small, skewed groups.

::: {.callout-warning}
### Common misinterpretations

**"F = 12.4 means a large effect"**
The F-statistic is not an effect size and depends on sample size. A
large F in a study of 500 subjects may reflect a trivially small
difference. Report η² (eta-squared = SS_between / SS_total) as the
effect size: η² < 0.06 is small, 0.06--0.14 is medium, > 0.14 is
large.

**"ANOVA tests variances (it says so in the name)"**
ANOVA tests whether group *means* are equal by partitioning variance.
It assumes equal within-group variances but does not test them. The
name refers to the technique, not the hypothesis.

**"A non-significant interaction means effects are purely additive"**
A non-significant interaction p-value may reflect low power rather
than true additivity. Always plot the group means to see whether the
pattern is consistent with your scientific interpretation.
:::


## Exercises

::: {.callout-note collapse="true"}
### Teacher Note

Pair students for Exercises 1 and 2: one runs the code while the other
predicts each result (ANOVA significance, then which Tukey pairs will
be significant) before it appears. Both exercises use the same `pbc`
stage variable as [Nonparametric Tests](nonparametric-tests.qmd), so if
you taught that session, ask the room to predict whether the ANOVA
post-hoc pattern will look like the Kruskal-Wallis post-hoc pattern
they already saw.
:::

### Exercise 1 (Guided): One-Way ANOVA for Bilirubin by Stage

*Estimated time: ~20 minutes*

Using `survival::pbc`, test whether log(bilirubin) differs across
the four disease stages.

1. Plot log(bilirubin) by stage (boxplot + jitter, using `scale_fill_brewer()`).
2. Run `aov(log(bili) ~ stage, ...)`. Report F, degrees of freedom,
   and p-value.
3. Run Tukey's HSD. Which stage pairs differ significantly?
4. Extract residuals with `augment()` and plot a QQ-plot. Is the
   normality assumption met?

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r ex1}
pbc <- survival::pbc %>%
  as_tibble() %>%
  janitor::clean_names() %>%
  mutate(stage = factor(stage, levels = 1:4, labels = paste("Stage", 1:4)))

pbc_s <- pbc %>% filter(!is.na(stage), !is.na(bili))

pbc_s %>%
  ggplot(aes(x = stage, y = log(bili), fill = stage)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_brewer(palette = "Blues") +
  labs(x = NULL, y = "log(Bilirubin)") +
  theme_bw() +
  theme(legend.position = "none")

fit_bili <- aov(log(bili) ~ stage, data = pbc_s)
summary(fit_bili)

TukeyHSD(fit_bili) %>% tidy() %>% arrange(adj.p.value)

augment(fit_bili) %>%
  ggplot(aes(sample = .resid)) +
  stat_qq() +
  stat_qq_line(colour = "red") +
  labs(title = "QQ-plot of residuals") +
  theme_bw()
```

Run the chunk above. The ANOVA is highly significant (`F(3, 408) ≈
15.1, p ≈ 2.6e-09`). Tukey's HSD shows the same pattern seen with the
Kruskal-Wallis post-hoc test in
[Nonparametric Tests](nonparametric-tests.qmd): Stage 4 differs
significantly from every other stage (all adj.p < 0.0001), while
Stages 1, 2, and 3 do not differ significantly from each other
(adj.p between 0.32 and 0.84). The residuals are approximately normal
on the log scale, confirming the transformation was appropriate.
:::

### Exercise 2 (Semi-guided): ANOVA for Albumin by Stage

*Estimated time: ~15 minutes*

Using `survival::pbc`, test whether serum albumin differs across the
four disease stages.

1. Plot albumin by stage.
2. Run `aov(albumin ~ stage, ...)` and report the key output numbers.
3. Run Tukey's HSD. Which pairs are significantly different?
4. Check the residual QQ-plot.
5. Does the direction of the albumin effect make biological sense for a
   disease where liver function progressively declines?

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r ex2}
pbc_s <- pbc %>% filter(!is.na(stage), !is.na(albumin))

pbc_s %>%
  ggplot(aes(x = stage, y = albumin, fill = stage)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1.5) +
  scale_fill_brewer(palette = "Blues") +
  labs(title = "Serum albumin by disease stage",
       x = NULL, y = "Albumin (g/dL)") +
  theme_bw() +
  theme(legend.position = "none")

fit_alb <- aov(albumin ~ stage, data = pbc_s)
summary(fit_alb)

TukeyHSD(fit_alb) %>% tidy() %>% arrange(adj.p.value)

augment(fit_alb) %>%
  ggplot(aes(sample = .resid)) +
  stat_qq() +
  stat_qq_line(colour = "red") +
  labs(title = "QQ-plot of residuals") +
  theme_bw()
```

Albumin falls with disease stage, consistent with declining liver
synthetic function. The ANOVA is significant and post-hoc tests
show early vs. late stage differences drive the result.
:::

### Exercise 3 (Open-ended)

*Estimated time: 15--30 minutes*

From your own data, identify a continuous outcome and a categorical
grouping variable with three or more groups. Run a one-way ANOVA,
check assumptions with residual plots, and run Tukey's HSD. Write a
complete Methods and Results paragraph as you would for a journal paper,
including the F-statistic, degrees of freedom, p-value, and the
specific group pairs that differ.


## Comprehension Check

*Estimated time: ~10 minutes (self-test)*

1. What is the null hypothesis in a one-way ANOVA? What does a
   significant result tell you, and what does it not tell you?
2. Why is running multiple pairwise t-tests not a valid substitute for
   ANOVA?
3. In a two-way ANOVA, the interaction term `supp:dose` is significant.
   What does this mean, and how should you proceed?
4. Your residual QQ-plot shows clear non-normality and you have n = 8
   per group. What should you do?
5. "ANOVA tests variances": true or false, and why?

::: {.callout-note collapse="true"}
### Answers

1. The null hypothesis is $H_0: \mu_1 = \mu_2 = \ldots = \mu_k$:
   all group population means are equal. A significant result means
   *at least one* group mean differs from at least one other. It does
   not identify which pairs differ: that requires a post-hoc test.

2. Running $\binom{k}{2}$ independent t-tests inflates the family-wise
   error rate. With k = 4 groups there are 6 comparisons; at α = 0.05,
   the probability of at least one false positive is 1 − 0.95⁶ = 26%.
   ANOVA tests all groups simultaneously, keeping the overall Type I
   error at 5%.

3. A significant interaction means the effect of one factor depends on
   the level of the other. Reporting only main effects is misleading:
   you must show the interaction plot and describe how the effect of
   dose changes across supplement types (or vice versa).

4. With small, non-normal groups, switch to Kruskal-Wallis: a
   nonparametric ANOVA alternative that works on ranks and does not
   require normality.

5. False. ANOVA tests whether group *means* are equal, using a technique
   that partitions total variance into between-group and within-group
   components. It assumes equal within-group variances, but the
   hypothesis being tested is about means, not variances.
:::


## How to Report

::: {.callout-note}
### Reporting ANOVA Results

**In a Methods section:**

> "Differences in [outcome] across [k] groups were tested using a
> one-way ANOVA. Post-hoc pairwise comparisons were performed using
> Tukey's honestly significant difference (HSD) test to control the
> family-wise error rate at 5%."

**In a Results section:**

> "[Outcome] differed significantly across groups (F(2, 27) = 4.85,
> p = 0.016, η² = 0.26). Post-hoc comparisons showed that [group A]
> had significantly lower [outcome] than [group B] (mean difference:
> X.XX, 95% CI: L.L to U.U, adjusted p = 0.012)."

**For two-way ANOVA with a significant interaction:**

> "There was a significant interaction between [factor 1] and
> [factor 2] (F(df, df) = X, p = Y). The effect of [factor 1] on
> [outcome] differed by level of [factor 2] (see Figure X)."

**Always include:**

- F-statistic with numerator and denominator df, e.g., F(2, 27)
- Overall p-value
- Effect size η² (SS_between / SS_total)
- Post-hoc test used, adjusted p-values for significant pairs
- Group means and SDs in a table or figure
:::


## Further Reading

- @Bland-2015: *An Introduction to Medical Statistics*: clear
  treatment of one-way and two-way ANOVA for clinical researchers
- @Dalgaard-2008: *Introductory Statistics with R*: practical ANOVA
  chapter with worked examples
- `?aov`, `?TukeyHSD`, `?oneway.test` in R
