# Basic Statistics for Clinical Research {#sec-basic-statistics}

```{r setup}
#| include: false
#| eval: true
source(here::here("R/functions.R"))
source(here::here("R/load-packages.R"))
```

## Introduction

This session introduces the essential statistical concepts used in
clinical research. It focuses on interpretation, reproducible code, and
how basic statistics support real research decisions.

## Learning outcomes

At the end of this session you will be able to:

- Summarize clinical data using descriptive statistics.
- Visualize distributions and compare groups with clear plots.
- Compute and interpret confidence intervals, t-tests, and ANOVA.
- Fit basic regression and correlation models.
- Use contingency tables for categorical clinical outcomes.
- Apply power calculations for planning clinical studies.

## 1. Why basic statistics matter in research

Clinical research begins with data that describe people, treatments,
and outcomes. Before using complex models, researchers must answer
simple questions:

- What is the typical value of this lab measurement?
- How variable is it across patients?
- Do two groups have different average outcomes?
- Are two clinical measurements correlated?
- Can we quantify uncertainty around our estimates?

These questions are the foundation of credible research.

## 2. Descriptive statistics with clinical data

We begin with the `pbc` dataset, a standard clinical dataset from the
`survival` package.

```{r pbc-summary}
pbc <- survival::pbc %>%
  as_tibble() %>%
  janitor::clean_names() %>%
  mutate(
    sex = factor(sex, levels = c("f", "m"), labels = c("Female", "Male")),
    stage = factor(stage)
  )

pbc %>%
  summarize(
    n = n(),
    mean_bili = mean(bili, na.rm = TRUE),
    sd_bili = sd(bili, na.rm = TRUE),
    median_bili = median(bili, na.rm = TRUE),
    iqr_bili = IQR(bili, na.rm = TRUE)
  )
```

Notice how we use `na.rm = TRUE` to ignore missing values. This is
essential for research datasets where missingness is common.

### 2.1 Visualizing distribution

```{r pbc-plot}
pbc %>%
  ggplot(aes(x = bili)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 30) +
  labs(
    title = "Distribution of serum bilirubin",
    x = "Bilirubin (mg/dL)",
    y = "Count"
  ) +
  theme_minimal()
```

### 2.2 Summary by group

Next, compare bilirubin values across disease stage.

```{r pbc-group-summary}
pbc %>%
  group_by(stage) %>%
  summarize(
    n = n(),
    mean_bili = mean(bili, na.rm = TRUE),
    sd_bili = sd(bili, na.rm = TRUE),
    median_bili = median(bili, na.rm = TRUE)
  ) %>%
  ungroup()
```

Use group summaries to identify whether clinical values differ across
important patient strata.

## 3. Transformations and normality

Clinical variables are often skewed. Log-transforming can make a
variable easier to interpret and analyze.

```{r pbc-log-transform}
pbc <- pbc %>%
  mutate(log_bili = log(bili))

pbc %>%
  ggplot(aes(sample = log_bili)) +
  stat_qq() +
  stat_qq_line(color = "firebrick") +
  labs(
    title = "QQ plot of log-transformed bilirubin",
    x = "Theoretical quantiles",
    y = "Sample quantiles"
  ) +
  theme_minimal()
```

### 3.1 Interpretation

The Q-Q plot shows whether the log-transformed variable approximates a
normal distribution. Research methods often rely on this assumption, so
examining plots like these is good practice.

## 4. Confidence intervals

A confidence interval quantifies uncertainty around an estimate. In
research, we use it to express how precise our estimate is.

```{r pbc-ci}
ci_bili <- pbc %>%
  summarize(
    estimate = mean(log_bili, na.rm = TRUE),
    se = sd(log_bili, na.rm = TRUE) / sqrt(sum(!is.na(log_bili))),
    df = sum(!is.na(log_bili)) - 1
  ) %>%
  mutate(
    t_crit = qt(0.975, df),
    lower = estimate - t_crit * se,
    upper = estimate + t_crit * se,
    lower_orig = exp(lower),
    upper_orig = exp(upper),
    estimate_orig = exp(estimate)
  )

ci_bili
```

This example shows a 95% confidence interval for the mean bilirubin on
the original scale.

## 5. Comparing two groups with a t-test

The `ToothGrowth` dataset illustrates a basic comparison between two
groups.

```{r toothgrowth-setup}
tg <- datasets::ToothGrowth %>%
  as_tibble() %>%
  mutate(dose = factor(dose))

tg %>%
  filter(dose == 0.5) %>%
  group_by(supp) %>%
  summarize(
    n = n(),
    mean_len = mean(len),
    sd_len = sd(len)
  )
```

```{r toothgrowth-ttest}
tg_low <- tg %>% filter(dose == 0.5)

test_result <- t.test(len ~ supp, data = tg_low)

broom::tidy(test_result)
```

### 5.1 Interpretation

The t-test compares average tooth length between the two supplement
groups. A smaller p-value suggests evidence of a true difference in
means.

## 6. One-way ANOVA

One-way ANOVA extends the comparison to three or more groups.

```{r plantgrowth-anova}
pg <- datasets::PlantGrowth %>% as_tibble()

pg_aov <- aov(weight ~ group, data = pg)
summary(pg_aov)
```

Use `TukeyHSD()` to see which groups differ.

```{r plantgrowth-tukey}
TukeyHSD(pg_aov)
```

### 6.1 Interpretation

ANOVA tests whether at least one group mean differs from the others.
Tukey's test shows which specific group pairs are different.

## 7. Correlation and linear regression

Correlation measures the relationship between two continuous variables.

```{r mtcars-correlation}
mtcars <- datasets::mtcars %>%
  as_tibble(rownames = "model") %>%
  mutate(am = factor(am, labels = c("Automatic", "Manual")))

correlation <- cor.test(mtcars$wt, mtcars$mpg)
correlation
```

A simple linear model quantifies the association.

```{r mtcars-regression}
lm_mpg <- lm(mpg ~ wt, data = mtcars)
broom::tidy(lm_mpg)
```

### 7.1 Interpretation

The regression slope tells us the expected change in mpg for each
unit increase in weight.

## 8. Categorical analysis and chi-square tests

Contingency tables help analyze categorical outcomes.

```{r hair-eye}
hair_eye_table <- margin.table(HairEyeColor, c(1, 2))

hair_eye_table

chi_result <- stats::chisq.test(hair_eye_table)

chi_result
```

### 8.1 Interpretation

A chi-square test evaluates whether hair color and eye color are
associated beyond random chance.

## 9. Power and sample size

Study planning uses power calculations to determine how many subjects
are needed.

```{r power-calculation}
power.t.test(power = 0.8, delta = 0.5, sd = 1)
power.t.test(n = 20, sd = 1, power = 0.8)
```

These calculations show the relationship between sample size, effect
size, and statistical power.

## 10. Exercises

1. Create a table of mean bilirubin by sex in the `pbc` dataset.
2. Fit a linear model predicting `mpg` from `wt` and interpret the slope.
3. Use `PlantGrowth` to compute and interpret a Tukey pairwise comparison.
4. Create a contingency table from `HairEyeColor` and explain the chi-square
   test result in one sentence.

## Key takeaways

Basic statistics are not just formulas. They are tools for asking and
answering clinical research questions with data. Every analysis should
connect back to interpretation and research decisions.
