# Basic Data Visualization {#sec-basic-visualization}

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(patchwork)
library(survival)
library(palmerpenguins)
```

## When Do You Use This? {.callout-tip}

> Before reporting any statistical result, you must plot the data. A well-chosen graph shows you things a table cannot: whether a distribution is skewed, whether two groups overlap, whether there is a trend, and whether any values look unusual. This session teaches the core plot types you will use in every analysis.

## Learning Objectives

After completing this session you will be able to:

- Explain the grammar of graphics: data, aesthetics, and geometries
- Create histograms, density plots, boxplots, scatter plots, and bar charts with `ggplot2`
- Map variables to color, fill, shape, and facets
- Add axis labels, titles, and themes appropriate for publications
- Choose the right plot type for a given research question

---

## Background: The Grammar of Graphics

`ggplot2` is built on the **grammar of graphics**: every plot is constructed from layers.

| Component | What it does | Example |
|---|---|---|
| `ggplot(data, aes(...))` | Set the data and aesthetic mappings | `ggplot(pbc, aes(x = age, y = bili))` |
| `geom_*()` | Choose the geometric shape | `geom_point()`, `geom_histogram()` |
| `scale_*()` | Control axis/color scales | `scale_color_manual()` |
| `facet_*()` | Split into subplots | `facet_wrap(~ sex)` |
| `labs()` | Add titles and axis labels | `labs(title = "...", x = "...")` |
| `theme_*()` | Control visual appearance | `theme_bw()`, `theme_classic()` |

You build a plot by adding layers with `+`. Each layer can see the same data, or its own.

---

## Worked Example 1: Clinical Data (PBC)

We use the `pbc` dataset throughout --- 418 patients with primary biliary cholangitis, a real clinical trial.

```{r load-pbc}
pbc <- survival::pbc |>
  as_tibble() |>
  janitor::clean_names() |>
  mutate(
    sex   = factor(sex, levels = c("f", "m"), labels = c("Female", "Male")),
    stage = factor(stage, levels = 1:4, labels = paste("Stage", 1:4))
  )
```

### Histogram: Visualizing a Single Continuous Variable

Use a histogram to understand the distribution of one continuous variable.

```{r histogram}
ggplot(pbc, aes(x = bili)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of serum bilirubin",
    x     = "Bilirubin (mg/dL)",
    y     = "Count"
  ) +
  theme_bw()
```

The distribution is right-skewed: most patients have low bilirubin, but a few have very high values.

**Choosing the number of bins:** too few bins hide the shape; too many make it noisy. `bins = 25` to `bins = 40` works for most datasets with 100--500 observations.

### Density Plot: Smoother Version of a Histogram

```{r density}
ggplot(pbc, aes(x = bili, fill = sex)) +
  geom_density(alpha = 0.4) +
  scale_fill_manual(values = c("Female" = "steelblue", "Male" = "tomato")) +
  labs(
    title = "Bilirubin distribution by sex",
    x     = "Bilirubin (mg/dL)",
    y     = "Density",
    fill  = "Sex"
  ) +
  theme_bw()
```

### Boxplot: Comparing Groups

Use a boxplot when you want to compare a continuous variable across two or more groups.

```{r boxplot}
pbc |>
  filter(!is.na(stage)) |>
  ggplot(aes(x = stage, y = bili, fill = stage)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1.2) +
  scale_fill_brewer(palette = "Blues") +
  labs(
    title = "Bilirubin by disease stage",
    x     = NULL,
    y     = "Bilirubin (mg/dL)"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

::: callout-note
**Reading a boxplot:**
- The horizontal line = median (50th percentile)
- The box = interquartile range (IQR): 25th to 75th percentile
- The whiskers extend to 1.5 x IQR beyond the box
- Points beyond the whiskers = potential outliers

Adding `geom_jitter()` on top shows individual observations, which is important with small sample sizes.
:::

### Scatter Plot: Relationship Between Two Continuous Variables

Use a scatter plot to examine whether two continuous variables are associated.

```{r scatter}
ggplot(pbc, aes(x = age, y = bili, color = sex)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = c("Female" = "steelblue", "Male" = "tomato")) +
  labs(
    title    = "Age and serum bilirubin by sex",
    subtitle = "Lines show linear trend with 95% confidence band",
    x        = "Age (years)",
    y        = "Bilirubin (mg/dL)",
    color    = "Sex"
  ) +
  theme_bw()
```

### Bar Chart: Counts of a Categorical Variable

Use a bar chart when you want to show counts or proportions across categories.

```{r bar-chart}
pbc |>
  filter(!is.na(stage)) |>
  ggplot(aes(x = stage, fill = sex)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c("Female" = "steelblue", "Male" = "tomato")) +
  labs(
    title = "Sex distribution by disease stage",
    x     = NULL,
    y     = "Proportion",
    fill  = "Sex"
  ) +
  theme_bw()
```

`position = "fill"` rescales bars to 100%, making proportional comparisons easier than raw counts.

---

## Worked Example 2: Animal Data (Penguins)

The `palmerpenguins` dataset is a clean multi-species penguin dataset, useful as a stand-in for animal experiment data with groups and continuous measurements.

```{r penguin-setup}
penguins_clean <- penguins |> drop_na()
```

### Combining Plot Types: Violin + Boxplot

```{r violin}
ggplot(penguins_clean, aes(x = species, y = body_mass_g, fill = species)) +
  geom_violin(alpha = 0.4, trim = FALSE) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Body mass by penguin species",
    x     = NULL,
    y     = "Body mass (g)"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

A violin plot shows the full distribution shape (like a smooth histogram rotated sideways). Combining it with an inner boxplot gives both the distribution shape and the summary statistics.

### Faceting: Small Multiples

Use `facet_wrap()` to repeat the same plot for each level of a grouping variable.

```{r facet}
ggplot(penguins_clean, aes(x = bill_length_mm, y = flipper_length_mm,
                            color = species)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ sex) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Bill length vs flipper length by species and sex",
    x     = "Bill length (mm)",
    y     = "Flipper length (mm)",
    color = "Species"
  ) +
  theme_bw()
```

### Combining Plots with `patchwork`

```{r patchwork}
p1 <- ggplot(penguins_clean, aes(x = bill_length_mm)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "white") +
  labs(title = "Bill length", x = "mm", y = "Count") +
  theme_bw()

p2 <- ggplot(penguins_clean, aes(x = flipper_length_mm)) +
  geom_histogram(bins = 25, fill = "tomato", color = "white") +
  labs(title = "Flipper length", x = "mm", y = "Count") +
  theme_bw()

p1 + p2   # side by side
```

`patchwork` uses `+` to place plots side by side, `/` to stack them, and `(p1 + p2) / p3` for more complex arrangements.

---

## Publication-Ready Themes

For journal submissions, choose a clean theme and increase the base font size.

```{r themes}
# theme_bw()      -- white background, grey gridlines  (good default)
# theme_classic() -- no gridlines, clean axes (common in biology journals)
# theme_minimal() -- minimal decoration

ggplot(pbc |> filter(!is.na(stage)),
       aes(x = stage, y = albumin, fill = stage)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Serum albumin by disease stage",
       x = NULL, y = "Albumin (g/dL)") +
  theme_classic(base_size = 14) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"))
```

::: callout-tip
**For publications:**
- Use `theme_classic()` or `theme_bw()` with `base_size = 12` or `14`
- Set `legend.position` to `"bottom"` or `"none"` to save horizontal space
- Use `scale_fill_brewer()` or `scale_color_brewer()` for accessible colour palettes
- Export with `ggsave("figure1.pdf", width = 6, height = 4)` for vector-format output
:::

---

## Where to Find Data Like This {.callout-note}

**PBC and other clinical trial datasets:**
Built into the `survival` package --- no download needed. `?survival::pbc` for the variable descriptions.

**Palmer penguins:**
```{r}
#| eval: false
install.packages("palmerpenguins")
library(palmerpenguins)
data("penguins")
```
Gorman et al. (2014), collected at Palmer Station Antarctica.

**Plot inspiration and code examples:**
- [R Graph Gallery](https://r-graph-gallery.com) --- searchable by plot type
- [ggplot2 extension gallery](https://exts.ggplot2.tidyverse.org) --- specialist plot types

---

## What Can Go Wrong {.callout-warning}

**Overplotting in scatter plots**
When many points fall on the same location, the plot looks like a blob. Fix with `alpha = 0.3` to make points semi-transparent, or `geom_jitter()` for discrete data.

**Bar charts of means without showing spread**
A bar chart that shows only group means hides the distribution. Use boxplots, violin plots, or scatter overlays instead.

**Default grey background in presentations**
`ggplot2`'s default grey theme looks poor on projector slides. Switch to `theme_bw()` or `theme_classic()`.

**Choosing colours that are not colourblind-safe**
About 8% of men have red-green colour blindness. Use `scale_colour_brewer(palette = "Set1")` or the `viridis` scale: `scale_colour_viridis_d()`.

---

## Exercises

### Exercise 1 (Guided): Distribution of a Lab Value

Using `survival::pbc`:

1. Create a histogram of `protime` (prothrombin time) with 25 bins.
2. Create a density plot of `protime` coloured by `sex`.
3. Create a boxplot of `protime` by `stage`.
4. Does the distribution look normal or skewed? Report the mean and median to check.

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r}
pbc |>
  ggplot(aes(x = protime)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "white") +
  labs(title = "Prothrombin time", x = "PT (seconds)", y = "Count") +
  theme_bw()

ggplot(pbc, aes(x = protime, fill = sex)) +
  geom_density(alpha = 0.4) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(title = "Prothrombin time by sex", x = "PT (seconds)") +
  theme_bw()

pbc |> filter(!is.na(stage)) |>
  ggplot(aes(x = stage, y = protime, fill = stage)) +
  geom_boxplot(alpha = 0.7) +
  labs(x = NULL, y = "PT (seconds)") +
  theme_bw() + theme(legend.position = "none")

# Check skewness
pbc |> summarise(mean_pt = mean(protime, na.rm = TRUE),
                  median_pt = median(protime, na.rm = TRUE))
```

The mean is slightly higher than the median --- mild right skew, consistent with a lab value.
:::

### Exercise 2 (Semi-guided): Penguin Comparison

Using `palmerpenguins::penguins`:

1. Create a scatter plot of `bill_length_mm` vs `bill_depth_mm`, coloured by species.
2. Add `geom_smooth(method = "lm")` for each species.
3. Facet by `island`.
4. In one sentence: does bill length predict bill depth in the same direction for all three species?

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r}
penguins |>
  drop_na() |>
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ island) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Bill dimensions by species and island",
       x = "Bill length (mm)", y = "Bill depth (mm)") +
  theme_bw()
```

Within each species, bill length and bill depth are positively associated (longer bills are also deeper). However, the overall cross-species correlation appears negative (Chinstraps have longer but shallower bills than Adelies) --- this is an example of Simpson's Paradox.
:::

### Exercise 3 (Open-ended)

Take a dataset from your own research or download one from the Mouse Phenome Database. Create three plots: one showing the distribution of a continuous variable, one comparing that variable across groups, and one scatter plot of two continuous variables. Choose `theme_classic()` and write a one-paragraph figure legend for each plot as you would for a journal submission.

---

## Comprehension Check

1. In `ggplot(data, aes(x = age, y = bili, color = sex))`, what does `color = sex` do?
2. What is the difference between `color` and `fill` aesthetics?
3. You want to show the same scatter plot separately for men and women without using the `color` aesthetic. What `ggplot2` function do you use?
4. Why should you avoid bar charts of means without showing the underlying data spread?
5. Which two arguments in `geom_histogram()` control the appearance of the bars?

::: {.callout-note collapse="true"}
### Answers
1. It maps the `sex` variable to the colour of each point, so different sexes appear in different colours. `sex` must be a factor or character column for the colours to be discrete.
2. `color` controls the outline/border colour (for points, lines). `fill` controls the interior colour (for bars, boxes, violin plots). Histograms use `fill`; scatter points use `color`.
3. `facet_wrap(~ sex)` or `facet_grid(. ~ sex)` --- these create a separate panel for each level of `sex`.
4. A bar chart showing only the group mean hides whether the data is normally distributed, how much spread there is, and whether there are outliers. A plot that only shows means can make a large-variance group look the same as a low-variance group.
5. `bins` (number of bins) and `binwidth` (width of each bin in data units). Also `fill` for interior colour and `color` for border colour.
:::

## Further Reading

- [ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org) --- Wickham's full book, free online
- [R Graph Gallery](https://r-graph-gallery.com) --- searchable examples by plot type
- [patchwork documentation](https://patchwork.data-imaginist.com) --- composing multi-panel figures
- [Fundamentals of Data Visualization](https://clauswilke.com/dataviz/) --- Claus Wilke's guide to visualization principles
