---
title: "Bayesian Inference"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(broom)
```

## When Do You Use This? {#sec-bayesian .callout-tip}

> You are running a small mouse experiment with only 8 animals per group. A frequentist t-test gives p = 0.08 — technically "not significant", but you have strong prior evidence from three previous experiments that the effect is real. Bayesian inference lets you incorporate that prior knowledge formally, combine it with your new data, and produce a probability statement about the effect size: "given everything we know, there is a 92% probability that the treatment reduces tumour volume by more than 10%."

## Learning Objectives

After completing this session you will be able to:

- Explain the difference between Bayesian and frequentist inference
- Define prior, likelihood, and posterior in plain language
- Interpret a Bayesian credible interval and contrast it with a frequentist confidence interval
- Fit a simple Bayesian regression model in R using `rstanarm` or `brms`
- Choose between weakly informative and informative priors and justify your choice
- Report a Bayesian analysis in a methods section

## Background: The Bayesian Framework

### Prior, Likelihood, Posterior

Bayesian inference updates a prior belief about a parameter $\theta$ using observed data $y$ to produce a posterior distribution:

$$P(\theta | y) \propto P(y | \theta) \cdot P(\theta)$$

| Term | What it means | Example |
|---|---|---|
| **Prior** $P(\theta)$ | Your belief about $\theta$ before seeing the data | "Based on prior experiments, I expect a 15–25% reduction in tumour volume" |
| **Likelihood** $P(y | \theta)$ | How probable the observed data are given $\theta$ | The standard statistical model (same as frequentist) |
| **Posterior** $P(\theta | y)$ | Updated belief after combining prior and data | The distribution of plausible effect sizes given all evidence |

The posterior is the answer Bayesian inference provides: not a single number, but a full distribution over plausible parameter values.

### Frequentist vs. Bayesian: the key distinction

| | Frequentist | Bayesian |
|---|---|---|
| **Parameter** | Fixed, unknown constant | Random variable with a distribution |
| **Data** | Random (repeating the experiment would give different data) | Fixed (you have the data you have) |
| **95% CI / CrI** | "95% of intervals from this procedure contain the true value" | "Given the data, there is a 95% probability $\theta$ lies in this interval" |
| **p-value** | P(data this extreme \| $H_0$ is true) | Not used — probability of hypotheses is direct |
| **Prior information** | Not incorporated | Formally incorporated |

::: {.callout-note}
### Bayesian credible interval vs. frequentist confidence interval

The statement most researchers *want* to make — "there is a 95% probability the effect lies between A and B" — is only valid for a **Bayesian credible interval**. A frequentist confidence interval does *not* license this interpretation. This is one of the most common statistical misconceptions in biomedical research.
:::

### When Bayesian methods are particularly useful

- **Small samples:** The prior stabilises estimates when data are sparse
- **Sequential analysis:** Bayesian updating is natural for adaptive trial designs
- **Incorporating prior knowledge:** Meta-analyses, repeat experiments, informative priors from existing literature
- **Complex hierarchical models:** Bayesian MCMC samplers handle these more naturally than frequentist alternatives
- **Decision making under uncertainty:** Posterior predictive distributions translate directly into clinical decision probabilities

---

## Example 1: Estimating a Proportion (Beta-Binomial)

The simplest Bayesian calculation: you observe $k$ successes in $n$ trials and want to estimate the true success probability $p$.

**Prior:** $p \sim \text{Beta}(\alpha_0, \beta_0)$

A Beta(1, 1) prior is uniform (no preference). Beta(2, 8) encodes a prior belief that $p$ is around 20%.

**Posterior:** $p | k, n \sim \text{Beta}(\alpha_0 + k, \; \beta_0 + (n - k))$

### Clinical application: estimating a response rate

You run a pilot trial. 6 of 20 patients respond to a new therapy. What is the Bayesian estimate of the true response rate?

```{r beta-binomial}
#| eval: true

# Observed data
k <- 6   # successes (responders)
n <- 20  # total patients

# Prior: weakly informative — centred at 20% response (from prior literature)
alpha_prior <- 2
beta_prior  <- 8

# Posterior parameters
alpha_post <- alpha_prior + k
beta_post  <- beta_prior + (n - k)

# Posterior mean and 95% credible interval
post_mean <- alpha_post / (alpha_post + beta_post)
cri_95    <- qbeta(c(0.025, 0.975), alpha_post, beta_post)

cat("Posterior mean:", round(post_mean, 3), "\n")
cat("95% Credible Interval:", round(cri_95, 3), "\n")
```

```{r beta-binomial-plot}
#| eval: true
#| fig.cap: "Prior (dashed) and posterior (solid) distributions for the response rate. The posterior is narrower, shifted towards the data."

p_grid <- seq(0, 1, length.out = 500)

prior_dens <- dbeta(p_grid, alpha_prior, beta_prior)
post_dens  <- dbeta(p_grid, alpha_post, beta_post)

tibble(p = p_grid, Prior = prior_dens, Posterior = post_dens) %>%
  pivot_longer(-p, names_to = "Distribution", values_to = "Density") %>%
  ggplot(aes(x = p, y = Density, colour = Distribution, linetype = Distribution)) +
  geom_line(linewidth = 1) +
  scale_linetype_manual(values = c("Prior" = "dashed", "Posterior" = "solid")) +
  labs(
    title = "Bayesian estimation of response rate",
    x = "True response rate (p)",
    y = "Density"
  ) +
  theme_minimal()
```

---

## Example 2: Bayesian Linear Regression (PBC Data)

We fit a Bayesian linear regression using `rstanarm`, estimating the association between bilirubin and albumin in the PBC dataset.

::: {.callout-note}
### Install rstanarm first
`rstanarm` requires a C++ compiler. If you completed setup (including Rtools on Windows), run:
```r
install.packages("rstanarm")
```
On first use it compiles Stan internally — this takes 3–5 minutes once.
:::

```{r bayes-regression}
#| eval: false
#| message: false
#| warning: false

library(rstanarm)

pbc <- survival::pbc %>%
  as_tibble() %>%
  janitor::clean_names() %>%
  filter(!is.na(bili), !is.na(albumin))

# Bayesian linear regression with weakly informative priors (default in rstanarm)
# Default priors: normal(0, 2.5) on scaled coefficients
bayes_fit <- stan_glm(
  albumin ~ log(bili),
  data   = pbc,
  family = gaussian(),
  seed   = 42,
  refresh = 0  # suppress iteration output
)

# Summarise the posterior
print(bayes_fit, digits = 3)
```

```{r bayes-posterior}
#| eval: false

# 90% credible interval for the log(bilirubin) coefficient
posterior_interval(bayes_fit, prob = 0.90)

# Posterior predictive check: do simulated data resemble observed data?
pp_check(bayes_fit, plotfun = "hist", nreps = 5)
```

### Interpreting the output

The key quantities from a Bayesian regression summary:

| Output | Interpretation |
|---|---|
| `Median` | Posterior median — point estimate analogous to the frequentist coefficient |
| `MAD_SD` | Median absolute deviation — uncertainty in the posterior (analogous to SE) |
| `10%` / `90%` | 80% credible interval — the range of values with 80% posterior probability |
| `Rhat` | Chain convergence diagnostic — values close to 1.0 indicate convergence |

**Compare to frequentist:** Run `lm(albumin ~ log(bili), data = pbc)` alongside and compare coefficients. With weakly informative priors and ample data, the Bayesian and frequentist estimates are nearly identical. The difference appears with small samples or informative priors.

---

## What Can Go Wrong

::: {.callout-warning}
### Common Mistakes in Bayesian Analysis

**Prior sensitivity.** With small samples, the posterior is sensitive to the prior choice. Always report your priors explicitly and consider running a sensitivity analysis: how much do results change if you use a more or less informative prior?

**Misinterpreting the credible interval.** A 95% credible interval *does* support the statement "there is a 95% probability $\theta$ is in this range" — but only conditional on the model and prior being correct. It is not a licence to ignore model assumptions.

**Convergence failure.** MCMC samplers can fail to converge. Check `Rhat` < 1.05 for all parameters. If `Rhat` is large, increase iterations or reparametrise the model.

**Treating a vague prior as "objective".** There is no truly uninformative prior for all quantities. A uniform prior on a probability is informative on the log-odds scale. Be explicit about your prior choices and why you made them.

**Using Bayesian methods to rescue a non-significant result.** If you run a frequentist analysis that gives p = 0.08 and then switch to a Bayesian analysis hoping to get a significant result, you are p-hacking. Pre-specify your analytical approach.
:::

## How to Report

::: {.callout-note}
### Reporting a Bayesian Analysis

**In a Methods section:**

> "We used Bayesian linear regression implemented in `rstanarm` (v2.x; Goodrich et al., 20xx) to estimate the association between [predictor] and [outcome]. Weakly informative priors were used for all coefficients (normal(0, 2.5) on scaled predictors). Model convergence was assessed by inspecting trace plots and confirming $\hat{R}$ < 1.05 for all parameters. Results are reported as posterior median (95% credible interval)."

**In a Results section:**

> "The posterior median association between log(bilirubin) and albumin was -0.XX (95% CrI: -0.XX to -0.XX), indicating that a one-unit increase in log(bilirubin) is associated with a decrease of 0.XX g/dL in albumin, with 95% posterior probability that the true association lies between -0.XX and -0.XX."

**Always report:**
- The software and version used
- The prior distributions chosen and why
- $\hat{R}$ values to confirm convergence
- Posterior median and a credible interval (state the width: 90% or 95%)
- Whether a sensitivity analysis to the prior was performed
:::

## Exercises

### Exercise 1: Prior sensitivity

A drug has been tested in 4 previous studies. The pooled estimate of responders is 30%. You run a new study with 15 patients and observe 7 responders.

1. Set up a Beta prior that encodes the 30% prior response rate with moderate confidence (Beta(6, 14)).
2. Calculate the posterior mean and 95% credible interval.
3. Compare with a uniform prior (Beta(1, 1)). How much do the posteriors differ?
4. Interpret the difference in plain language.

```{r ex1}
#| eval: false

# Your code here
alpha_informative <- ___
beta_informative  <- ___
k <- 7; n <- 15

# Posterior with informative prior
alpha_post_inf <- alpha_informative + k
# ...
```

### Exercise 2: Bayesian vs. frequentist for a small experiment

Simulate a small mouse experiment (n = 5 per group) where the true mean difference is 2 units (SD = 3):

```{r ex2}
#| eval: false

set.seed(123)
control   <- rnorm(5, mean = 10, sd = 3)
treatment <- rnorm(5, mean = 12, sd = 3)
```

1. Run a frequentist t-test. What is the p-value?
2. Fit a Bayesian t-test using `rstanarm::stan_glm()` with a weakly informative prior.
3. What does the posterior say about the probability that the treatment effect > 0?
4. Which analysis gives more useful information for a decision about whether to continue with the treatment?

### Exercise 3: Credible interval interpretation

A Bayesian analysis of a clinical trial gives a posterior median treatment effect of 5.2 mmHg reduction in systolic blood pressure, with a 95% credible interval of (1.1, 9.3).

1. Write a one-sentence interpretation of this result.
2. A reviewer writes: "This is not significant because the CI does not exclude zero." Is this criticism valid? Explain.
3. What is the posterior probability that the treatment reduces blood pressure by at least 3 mmHg? (Hint: you need the posterior distribution, not just the CrI.)

## Comprehension Check

1. What is the key difference between a frequentist p-value and a Bayesian posterior probability?
2. You set a prior Beta(2, 2) for a success probability. What belief does this encode? What happens to the posterior if you collect 100 successes and 10 failures?
3. What does a 95% credible interval mean, and how does this differ from a 95% confidence interval?
4. In the Beta-Binomial model, what is the role of the prior when the sample size is very large?
5. A colleague says "the posterior is just a smoothed version of the data." When is this roughly true, and when is it not?

::: {.callout-note collapse="true"}
### Answers
1. A p-value is the probability of observing results as extreme as those seen *if the null hypothesis is true* --- it does not tell you the probability that the null is true or false. A Bayesian posterior probability directly states the probability that a parameter takes a particular value (or range), given the data and prior.
2. Beta(2, 2) encodes mild belief that the success probability is near 0.5 (symmetric, slightly peaked). With 100 successes and 10 failures, the posterior is Beta(102, 12), strongly concentrated near 0.89. The prior has almost no influence when n is large.
3. A 95% credible interval means there is a 95% posterior probability that the true parameter lies within the interval --- a direct probability statement. A frequentist 95% confidence interval means that if you repeated the experiment many times, 95% of such intervals would contain the true value; any single CI either contains it or does not.
4. When n is very large, the likelihood dominates and the posterior is almost entirely determined by the data regardless of the prior.
5. This is roughly true when the prior is weakly informative and n is large. It is not true when n is small (the prior strongly shapes the posterior) or when the prior and likelihood are in conflict.
:::

## Further Reading

- @Gelman-2013 --- *Bayesian Data Analysis* (3rd ed.) — the definitive textbook
- @McElreath-2020 --- *Statistical Rethinking* — the most accessible introduction; lecture videos free on YouTube
- `rstanarm` vignette: `vignette("rstanarm", package = "rstanarm")`
- `brms` package documentation: [paul-buerkner.github.io/brms](https://paul-buerkner.github.io/brms/)
- [Bayes Factors and model comparison](https://easystats.github.io/bayestestR/) — `bayestestR` package
