---
title: "Categorical Data and Chi-Square Tests"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(janitor)
library(survival)
library(boot)
if (!requireNamespace("gtsummary", quietly = TRUE)) {
  message("Package 'gtsummary' not available; related tables will be skipped during render.")
} else {
  library(gtsummary)
}
library(broom)
```

::: {.callout-note}
### Session at a Glance

**Total core time: about 115 minutes (~2 hours).**

| Section | Time | Type |
|---|---|---|
| When Do You Use This? + Learning Objectives | 5 min | Reading |
| The Key Idea: Comparing Observed to Expected | 15 min | Reading |
| Example 1: Clinical Data (PBC) | 30 min | Worked example |
| Example 2: Melanoma Data | 15 min | Worked example |
| What Can Go Wrong | 10 min | Reading |
| Exercises 1 & 2 | 30 min | Practice |
| Comprehension Check | 10 min | Self-test |

In a taught course, this session fits one teaching block. For
self-paced study, split it into two sittings, e.g. the key idea and
Example 1 first, Example 2 onward later. Exercise 3 is open-ended and
works best as a follow-up using a dataset from your own field.
:::

::: {.callout-tip}
### How to Use the Code in This Session

Every code chunk below is written so you can copy it into your R
console (or an R script) and run it yourself, in the order shown.
Where you see a "Run It Yourself" box, stop and run the code above it
before reading the explanation that follows.
:::

## When Do You Use This? {#sec-categorical}

::: {.callout-tip}
You want to know whether adverse event rates differ between two treatment arms: both outcomes are binary (event: yes/no). Or you want to know whether disease stage at diagnosis differs between men and women. These are questions about **counts and proportions**, not means. Chi-square and Fisher's exact tests are the right tools.
:::

::: {.callout-note collapse="true"}
### Teacher Note

Ask the room for a binary or categorical outcome from their own work:
response/no response, complication/no complication, genotype categories.
Keep one of those examples in mind through the session as an alternative
to the PBC and melanoma examples below.
:::

## Learning Objectives

After completing this session you will be able to:

- Identify when chi-square or Fisher's exact test is the right choice
- Explain what "expected counts" are and why they matter
- Construct a contingency table and run `chisq.test()` and `fisher.test()` in R
- Use `prop.test()` to compare two proportions with a confidence interval
- Build publication-quality frequency tables with `gtsummary`
- Report an association correctly: test result *plus* the actual proportions

## The Key Idea: Comparing Observed to Expected

*Estimated time: ~15 minutes (reading)*

::: {.callout-note collapse="true"}
### Teacher Note

Board exercise: with the 2x2 table below on the board, ask the room to
compute the four expected counts by hand using
$E = \frac{\text{row total} \times \text{column total}}{n}$ before you
reveal that they are all 25. Then ask: "given the observed counts of
30, 70, 20, 80, does this table look close to that, or far from it?"
This primes the chi-square statistic as a literal measure of distance
between observed and expected.
:::

Before we touch any R code, let's build the intuition.

Suppose you run a clinical trial. At the end, you count how many people in the treatment group had a side effect versus the placebo group:

| | Side effect | No side effect | Total |
|---|---|---|---|
| Treatment | 30 | 70 | 100 |
| Placebo | 20 | 80 | 100 |

Are these groups really different, or could this just be chance? The chi-square test answers that by asking: **if side effects had nothing to do with which group you were in, what counts would we expect?**

If treatment and side effects were completely unrelated, side effects would be spread across groups in proportion to group size. With 50 side effects total out of 200 patients, each group of 100 would be expected to have exactly 25. The **expected count** for each cell is:

$$E = \frac{\text{row total} \times \text{column total}}{n}$$

Chi-square then measures how far the actual table is from this "no relationship" version:

$$\chi^2 = \sum \frac{(O - E)^2}{E}$$

A large chi-square means the observed table looks nothing like what we'd expect under independence, evidence of a real association. A small chi-square means the table is consistent with chance alone.

This is the same logic from the [hypothesis testing session](hypothesis-testing.qmd): we are asking "how surprising is this data if the null hypothesis (no association) were true?" The chi-square distribution with $(r-1)(c-1)$ degrees of freedom converts that statistic into a p-value.

## Example 1: Clinical Data (PBC)

*Estimated time: ~30 minutes (worked example)*

::: {.callout-note collapse="true"}
### Teacher Note

The PBC dataset has appeared in several earlier sessions (always as
continuous outcomes: albumin, bilirubin). Here every variable of
interest is categorical (sex, stage) or binary (death). Ask the room:
"if you wanted to test whether *albumin level* differed by sex, would
you use chi-square?" The answer is no — chi-square is for the
*sex/stage* question here, not for a continuous outcome like albumin.
:::

### Is Sex Associated with Disease Stage?

The PBC dataset (`survival::pbc`) records 418 patients with primary biliary cirrhosis. Before running anything, take a moment to think:

> **Think before you run:** Do you expect sex and disease stage to be associated? In a randomised trial, baseline characteristics should not differ between arms by design. But this is an observational cohort. Is there a clinical reason men and women might present at different stages?

```{r pbc-setup}
pbc <- survival::pbc %>%
  as_tibble() %>%
  clean_names() %>%
  mutate(
    sex   = factor(sex, levels = c("f", "m"), labels = c("Female", "Male")),
    trt   = factor(trt, levels = c(1, 2), labels = c("D-penicillamine", "Placebo")),
    stage = factor(stage, levels = 1:4, labels = paste("Stage", 1:4)),
    died  = as.integer(status == 2)
  )
```

```{r sex-stage-table}
table(pbc$sex, pbc$stage)
```

```{r chisq-test}
chisq.test(table(pbc$sex, pbc$stage))
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above (you'll see a warning about the chi-square
approximation — some expected counts are small, which we'll check
properly with Fisher's exact test next). You should get
`X-squared ≈ 0.88`, `df = 3`, `p ≈ 0.83`. Far above 0.05: no evidence
that the distribution of disease stage differs between men and women.
Does this match what you predicted in "Think before you run"?
:::

**Reading the output:**

| Output element | What it means |
|---|---|
| `X-squared = ...` | The chi-square statistic: how far observed counts are from expected |
| `df = 3` | Degrees of freedom: (2 rows − 1) × (4 cols − 1) = 1 × 3 = 3 |
| `p-value = ...` | Probability of seeing this much difference by chance alone |

A significant p-value tells you the distribution of stages differs between men and women. It does **not** tell you which stages drive the difference: you need the table and the plot for that.

**Visualise:**

```{r sex-stage-plot}
pbc %>%
  filter(!is.na(sex), !is.na(stage)) %>%
  ggplot(aes(x = stage, fill = sex)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Disease stage by sex",
       x = NULL, y = "Proportion", fill = "Sex") +
  theme_bw()
```

### Fisher's Exact Test for a Sparse Table

Chi-square relies on a large-sample approximation. When any expected cell count is below 5, that approximation breaks down and the p-value becomes unreliable. Fisher's exact test computes the exact probability without any approximation: it is the right choice for sparse tables, regardless of sample size.

```{r fisher-example}
pbc_14 <- pbc %>%
  filter(!is.na(sex), stage %in% c("Stage 1", "Stage 4"))

tab_14 <- table(pbc_14$sex, pbc_14$stage)
tab_14

# Always check expected counts before choosing the test
chisq.test(tab_14)$expected

# If any expected count < 5, use Fisher's exact
fisher.test(tab_14)
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. The expected count for Male/Stage 1 is only
≈ 2.5, below the rule-of-thumb minimum of 5 — chi-square would not be
reliable here. Fisher's exact test gives `p ≈ 0.72`: no evidence of
association between sex and stage in this restricted Stage 1 vs
Stage 4 comparison.
:::

### Comparing Two Proportions with a CI

`prop.test()` compares two proportions and gives a confidence interval for the **difference**, the quantity you actually want to report. If the 95% CI for the difference includes zero, there is no statistically significant difference at the 5% level.

This is exactly the [equivalence between confidence intervals and hypothesis tests](hypothesis-testing.qmd) from the previous session: the CI and the p-value always agree.

```{r prop-test}
pbc_rand <- pbc %>% filter(!is.na(trt), !is.na(died))

deaths_by_trt <- pbc_rand %>%
  group_by(trt) %>%
  summarise(n = n(), deaths = sum(died))

prop.test(x = deaths_by_trt$deaths, n = deaths_by_trt$n)
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. You should get `X-squared ≈ 0.077`, `p ≈ 0.78`,
and a 95% CI for the difference in death rates of about (-0.093,
0.137) — straddling zero. The randomised treatment groups do not differ
in mortality, consistent with the comparable baseline albumin and
bilirubin values seen in earlier sessions.
:::

### Publication-Quality Table with gtsummary

```{r tbl1}
#| eval: !expr requireNamespace("gtsummary", quietly = TRUE)
pbc %>%
  filter(!is.na(trt)) %>%
  select(trt, sex, stage, died) %>%
  tbl_summary(
    by = trt,
    statistic = list(all_categorical() ~ "{n} ({p}%)")
  ) %>%
  add_p() %>%
  bold_labels()
```

`add_p()` automatically selects chi-square or Fisher's exact depending on expected cell counts: no manual checking required for this table format.

## Example 2: Melanoma Data

*Estimated time: ~15 minutes (worked example)*

::: {.callout-note collapse="true"}
### Teacher Note

Unlike the PBC sex/stage comparison above, this example produces a
significant result. Use the contrast deliberately: a non-significant
chi-square (PBC) and a significant one (melanoma) side by side help
students see that "the test didn't find anything" and "the test found
something" are both normal outcomes, not a sign of doing it wrong.
:::

```{r melanoma-setup}
mel <- boot::melanoma %>%
  as_tibble() %>%
  mutate(
    ulceration = factor(ulcer, levels = c(0, 1), labels = c("No", "Yes")),
    sex        = factor(sex, levels = c(0, 1), labels = c("Female", "Male")),
    status_cat = factor(status, levels = c(1, 2, 3),
                        labels = c("Died (melanoma)", "Alive", "Died (other)"))
  )
```

### Is Ulceration Associated with Sex?

```{r mel-chi}
tab_mel <- table(mel$sex, mel$ulceration)
tab_mel

chisq.test(tab_mel)
tidy(chisq.test(tab_mel))
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. The table is 79/47 (Female: no/yes ulceration) vs
36/43 (Male: no/yes). With Yates' correction, `X-squared ≈ 5.11`,
`df = 1`, `p ≈ 0.024` — below 0.05. Unlike the PBC sex/stage example,
this one is statistically significant: ulceration is more common among
male patients in this sample.
:::

### Relative Risk and Odds Ratio

For a 2×2 table, the p-value only tells you whether an association exists. You need an effect size measure to say how strong it is. Two measures are common:

- **Relative risk (RR):** The proportion in group A divided by the proportion in group B. Directly interpretable: "Men are 1.4× more likely to have ulceration."
- **Odds ratio (OR):** More common in case-control studies and logistic regression. Numerically equals the RR only when events are rare.

```{r rr-or}
a <- tab_mel["Male", "Yes"]
b <- tab_mel["Male", "No"]
c <- tab_mel["Female", "Yes"]
d <- tab_mel["Female", "No"]

rr <- (a / (a + b)) / (c / (c + d))
or <- (a * d) / (b * c)

cat("Relative Risk:", round(rr, 2), "\n")
cat("Odds Ratio:   ", round(or, 2), "\n")
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. You should get `RR ≈ 1.46` and `OR ≈ 2.01`. Men in
this sample are 1.46× as likely to have ulceration as women. The OR
(2.01) is noticeably larger than the RR (1.46) — a reminder that the OR
exaggerates the association relative to the RR when the outcome is not
rare (ulceration occurs in roughly a third to a half of patients here).
:::

For logistic regression-based ORs with 95% CIs, see the [Logistic Regression](logistic-regression.qmd) session.

::: {.callout-note}
## Where to Find Data Like This

**PBC dataset:** `survival::pbc`: stage, sex, treatment, status: all categorical.

**boot::melanoma:** `boot::melanoma`: sex, ulceration, survival status: classic 2×2 scenarios.

**MASS::birthwt:** Binary outcome (low birth weight) with categorical predictors.

**NHANES (nhanes package):** Large population survey with many categorical health variables.
:::

::: {.callout-warning}
## What Can Go Wrong

*Estimated time: ~10 minutes (reading)*

**Using chi-square with small expected counts.**
When any expected cell count is < 5, the chi-square approximation is unreliable. Use `fisher.test()`. R warns you when this occurs: do not ignore that warning.

**Confusing counts with proportions.**
`chisq.test()` requires a table of *counts*, not proportions. Dividing cells by totals before running the test gives wrong results.

**Yates' continuity correction.**
R applies Yates' correction by default for 2×2 tables (`correct = TRUE`), making the test more conservative. Use `correct = FALSE` to match Pearson's chi-square, but be consistent in your reporting.

**Interpreting significance as strong association.**
With large samples, chi-square detects tiny, clinically irrelevant differences. Supplement the p-value with an effect size: **Cramér's V** for larger tables, or relative risk/odds ratio for 2×2 tables.

**Forgetting that chi-square is non-directional.**
A significant chi-square tells you the variables are associated, not which group has a higher proportion. Always report the contingency table or a bar chart alongside the test.
:::

::: {.callout-note collapse="true"}
### Teacher Note

Quick true/false check, drawing on the points above:

- "R prints 'Chi-squared approximation may be incorrect' but still gives a p-value, so it's fine to report it." (**False** -- switch to `fisher.test()`.)
- "`chisq.test()` works the same whether you give it counts or proportions." (**False** -- it needs counts.)
- "A chi-square p-value of 0.001 tells you which group has the higher proportion." (**False** -- chi-square is non-directional; check the table or proportions to see direction.)
:::

## Exercises

::: {.callout-note collapse="true"}
### Teacher Note

Pair students up for these exercises. Before running Exercise 1, ask each pair to predict: based on what they've seen of the PBC dataset in earlier sessions, do they expect sex to be associated with survival? Exercise 2 revisits the same `mel$sex` x `mel$ulceration` table used in Example 2, but asks students to compute the odds ratio by hand -- a useful check that they understand how the four cells of a 2x2 table map onto the OR formula.
:::

### Exercise 1 (Guided): Sex and Mortality in PBC

*Estimated time: ~15 minutes*

1. Create a 2×2 table of `sex` by `died` in `pbc`.
2. Run a chi-square test. Check that all expected counts are ≥ 5.
3. Run `prop.test()` and report the proportion who died in each sex group with the 95% CI for the difference.
4. Plot the proportions.

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r}
tab_sx <- table(pbc$sex, pbc$died)
tab_sx

chisq.test(tab_sx)$expected   # check all ≥ 5
tidy(chisq.test(tab_sx))

prop.test(tab_sx)

pbc %>%
  filter(!is.na(sex)) %>%
  ggplot(aes(x = sex, fill = factor(died, labels = c("Alive", "Died")))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "Proportion", fill = NULL) +
  theme_bw()
```
:::

### Exercise 2 (Semi-guided): Ulceration and Sex in Melanoma

*Estimated time: ~15 minutes*

Using `boot::melanoma`, test whether ulceration status differs by sex.

1. Build a 2×2 contingency table.
2. Check expected counts and choose chi-square or Fisher's exact.
3. Compute the odds ratio manually and interpret it in one plain-language sentence.

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r}
tab_m <- table(mel$sex, mel$ulceration)
chisq.test(tab_m)$expected   # check for < 5

tidy(chisq.test(tab_m))

a <- tab_m["Male", "Yes"];  b <- tab_m["Male", "No"]
c <- tab_m["Female", "Yes"]; d <- tab_m["Female", "No"]
cat("OR:", round((a * d) / (b * c), 2))
```
:::

### Exercise 3 (Open-ended)

*Estimated time: 15--30 minutes*

Find a binary or nominal outcome in your own data. Construct a contingency table with a grouping variable of interest. Run the appropriate test, check assumptions, and write a two-sentence result suitable for a paper, including the actual proportions, not just the p-value.

## Comprehension Check

*Estimated time: ~10 minutes (self-test)*

1. When should you use Fisher's exact test instead of chi-square?
2. R prints: "Chi-squared approximation may be incorrect." What does this mean and what should you do?
3. A chi-square test gives p = 0.001. Does this tell you which group has a higher proportion of events?
4. What is the difference between a relative risk and an odds ratio? When are they approximately equal?
5. You have a 3×4 contingency table (3 groups, 4 categories). What are the degrees of freedom?

::: {.callout-note collapse="true"}
### Answers

1. When any expected cell count is < 5. Fisher's exact test does not rely on the large-sample approximation and is valid for any sample size.
2. At least one expected cell count is < 5, so the chi-square approximation is unreliable. Switch to `fisher.test()`.
3. No. Chi-square is non-directional: it tells you whether an association exists, not its direction or magnitude. Examine the contingency table or proportions to see which group has higher rates.
4. Relative risk = (proportion in group A) / (proportion in group B). Odds ratio = (odds in group A) / (odds in group B). They are approximately equal when events are rare (below 10--15% in both groups). For common outcomes, the OR exaggerates the association relative to the RR.
5. Degrees of freedom = (rows − 1) × (columns − 1) = (3 − 1) × (4 − 1) = 2 × 3 = 6.
:::

## How to Report

::: {.callout-note}
**In a methods section:** "The association between [categorical variable 1] and [categorical variable 2] was assessed using the chi-square test of independence. Fisher's exact test was used where any expected cell count was below 5."

**In results:** "There was a significant association between [var 1] and [var 2] (χ²(*df*) = X, *p* = Y). [Group A] had a higher proportion of [outcome] than [group B] (X% vs Y%)."

**Always include:**

- The chi-square statistic with degrees of freedom: χ²(2) = 8.4
- The p-value
- The actual proportions: the test result alone is not informative
- Whether Fisher's exact was used (and why)
- An effect size for 2×2 tables: odds ratio or relative risk with 95% CI

For models that estimate odds ratios with confidence intervals directly, see the [Logistic Regression](logistic-regression.qmd) session.
:::

## Further Reading

- @Agresti-2019: *An Introduction to Categorical Data Analysis*, the standard reference
- @Bland-2015: Chi-square and Fisher's test in *An Introduction to Medical Statistics*
- `?chisq.test`, `?fisher.test`, `?prop.test` in R: built-in documentation with examples
