---
title: "Categorical Data and Chi-Square Tests"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(survival)
if (!requireNamespace("gtsummary", quietly = TRUE)) {
  message("Package 'gtsummary' not available; related tables will be skipped during render.")
} else {
  library(gtsummary)
}
library(broom)
```

## When Do You Use This? {#sec-categorical .callout-tip}

> You want to know whether adverse event rates differ between two treatment arms in a clinical trial - both outcomes are binary (event: yes/no). Or you want to know whether disease stage at diagnosis differs between men and women. These are questions about **counts and proportions**, not means, and they call for chi-square or Fisher's exact tests.

## Learning Objectives

After completing this session you will be able to:

- Identify when chi-square or Fisher's exact test is the right choice
- Construct a contingency table and calculate expected cell counts
- Run `chisq.test()` and `fisher.test()` in R and interpret the output
- Use `prop.test()` to compare two proportions with a confidence interval
- Build publication-quality frequency tables with `gtsummary`
- Know when to apply Yates' continuity correction


## Background: Testing Categorical Associations

When both variables are categorical, we test whether the distribution of one variable differs across levels of the other - or equivalently, whether the two variables are **associated**.

**The 2?2 contingency table:**

|  | Outcome: Yes | Outcome: No | Row total |
|---|---|---|---|
| Group A | $a$ | $b$ | $a+b$ |
| Group B | $c$ | $d$ | $c+d$ |
| Column total | $a+c$ | $b+d$ | $n$ |

**Chi-square test:** Under independence, each cell's expected count is

$$E_{ij} = \frac{\text{row total}_i \times \text{col total}_j}{n}$$

The test statistic is $\chi^2 = \sum \frac{(O-E)^2}{E}$, with degrees of freedom $(r-1)(c-1)$.

**Fisher's exact test:** Used when expected cell counts are small (< 5). It computes the exact probability of seeing a table as extreme as observed under independence.

**Rule of thumb:** Use chi-square when all expected counts = 5; use Fisher's exact when any expected count < 5.


## Example 1: Clinical Data (PBC)

### Is Sex Associated with Disease Stage?

```{r pbc-setup}
pbc <- survival::pbc %>%
  as_tibble() %>%
  janitor::clean_names() %>%
  mutate(
    sex   = factor(sex, levels = c("f", "m"), labels = c("Female", "Male")),
    trt   = factor(trt, levels = c(1, 2), labels = c("D-penicillamine", "Placebo")),
    stage = factor(stage, levels = 1:4, labels = paste("Stage", 1:4)),
    died  = as.integer(status == 2)
  )
```

```{r sex-stage-table}
# Contingency table
table(pbc$sex, pbc$stage)
```

```{r chisq-test}
chisq.test(table(pbc$sex, pbc$stage))
```

A significant chi-square test tells you that the distribution of stage differs between men and women (or equivalently, sex and stage are associated). It does not tell you the direction or size.

**Visualise:**

```{r sex-stage-plot}
pbc %>%
  filter(!is.na(sex), !is.na(stage)) %>%
  ggplot(aes(x = stage, fill = sex)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Disease stage by sex",
       x = NULL, y = "Proportion", fill = "Sex") +
  theme_bw()
```

### Fisher's Exact Test for a Sparse Table

```{r fisher-example}
# Stage 1 vs Stage 4 only (small cell counts may arise)
pbc_14 <- pbc %>%
  filter(!is.na(sex), stage %in% c("Stage 1", "Stage 4"))

tab_14 <- table(pbc_14$sex, pbc_14$stage)
tab_14

# Check expected counts
chisq.test(tab_14)$expected

# If any expected count < 5, use Fisher's exact
fisher.test(tab_14)
```

### Comparing Two Proportions with a CI

`prop.test()` compares two proportions and gives a confidence interval for the difference.

```{r prop-test}
# Proportion who died in each treatment group
pbc_rand <- pbc %>% filter(!is.na(trt), !is.na(died))

deaths_by_trt <- pbc_rand %>%
  group_by(trt) %>%
  summarise(n = n(), deaths = sum(died))

prop.test(x = deaths_by_trt$deaths, n = deaths_by_trt$n)
```

**Interpretation:** The 95% CI for the difference in proportions includes/excludes zero, indicating whether the groups differ in mortality.

### Publication-Quality Table with gtsummary

```{r tbl1}
#| eval: requireNamespace("gtsummary", quietly = TRUE)
pbc %>%
  filter(!is.na(trt)) %>%
  select(trt, sex, stage, died) %>%
  tbl_summary(
    by = trt,
    statistic = list(all_categorical() ~ "{n} ({p}%)")
  ) %>%
  add_p() %>%
  bold_labels()
```

`add_p()` automatically selects chi-square or Fisher's exact depending on expected cell counts.


## Example 2: Animal Data (boot::melanoma)

```{r melanoma-setup}
library(boot)
mel <- boot::melanoma %>%
  as_tibble() %>%
  mutate(
    ulceration = factor(ulcer, levels = c(0, 1), labels = c("No", "Yes")),
    sex        = factor(sex, levels = c(0, 1), labels = c("Female", "Male")),
    status_cat = factor(status, levels = c(1, 2, 3),
                        labels = c("Died (melanoma)", "Alive", "Died (other)"))
  )
```

### Is Ulceration Associated with Sex?

```{r mel-chi}
tab_mel <- table(mel$sex, mel$ulceration)
tab_mel

chisq.test(tab_mel)
tidy(chisq.test(tab_mel))
```

### Relative Risk and Odds Ratio

For a 2?2 table, you can calculate the relative risk (RR) and odds ratio (OR) by hand:

```{r rr-or}
a <- tab_mel["Male", "Yes"]
b <- tab_mel["Male", "No"]
c <- tab_mel["Female", "Yes"]
d <- tab_mel["Female", "No"]

rr <- (a / (a + b)) / (c / (c + d))
or <- (a * d) / (b * c)

cat("Relative Risk:", round(rr, 2), "\n")
cat("Odds Ratio:   ", round(or, 2), "\n")
```

For logistic regression-based OR with CIs, see the [Logistic Regression](logistic-regression.qmd) session.


## Where to Find Data Like This {.callout-note}

**PBC dataset:** `survival::pbc` - stage, sex, treatment, status: all categorical.

**boot::melanoma:** `boot::melanoma` - sex, ulceration, survival status: classic 2?2 scenarios.

**MASS::birthwt:** Contains binary outcomes (low birth weight) and categorical predictors.

**NHANES (nhanes package):** Large population survey with many categorical health variables and demographic strata.


## What Can Go Wrong {.callout-warning}

**Using chi-square with small expected counts**
When any expected cell count is < 5, the chi-square approximation is unreliable. Use Fisher's exact test. The output of `chisq.test()` warns you when this occurs.

**Confusing counts with proportions**
`chisq.test()` requires a table of **counts**, not proportions. Dividing cells by totals before running the test gives incorrect results.

**Yates' continuity correction**
R applies Yates' correction by default for 2?2 tables (`correct = TRUE`). This makes the test more conservative. It is often dropped in practice; use `correct = FALSE` to match Pearson's chi-square, but be consistent in your reporting.

**Interpreting significance as strong association**
With large samples, chi-square detects tiny, clinically irrelevant differences. Supplement the p-value with a measure of effect size: Cram?r's V (for tables), or relative risk / odds ratio (for 2?2 tables).

**Forgetting that chi-square is non-directional**
A significant chi-square tells you the variables are associated; not which group has a higher proportion. Always report the contingency table or proportions alongside the test.


## Exercises

### Exercise 1 (Guided): Sex and Mortality in PBC

1. Create a 2?2 table of `sex` by `died` in `pbc`.
2. Run a chi-square test. Check that expected counts are = 5.
3. Run `prop.test()` and report the proportion who died in each sex group, with the 95% CI for the difference.
4. Plot the proportions.

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r}
tab_sx <- table(pbc$sex, pbc$died)
tab_sx

# Check expected counts
chisq.test(tab_sx)$expected

# Chi-square
tidy(chisq.test(tab_sx))

# prop.test
prop.test(tab_sx)

# Plot
pbc %>%
  filter(!is.na(sex)) %>%
  ggplot(aes(x = sex, fill = factor(died, labels = c("Alive", "Died")))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "Proportion", fill = NULL) + theme_bw()
```
:::

### Exercise 2 (Semi-guided): Treatment and Adverse Events (Melanoma)

Using `boot::melanoma`, test whether ulceration status differs by sex.

1. Build a 2?2 contingency table.
2. Check expected counts and choose chi-square or Fisher's exact.
3. Compute the odds ratio manually and interpret it.

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r}
tab_m <- table(mel$sex, mel$ulceration)
chisq.test(tab_m)$expected   # check for < 5

# All = 5: use chi-square
tidy(chisq.test(tab_m))

# Odds ratio
a <- tab_m["Male", "Yes"];  b <- tab_m["Male", "No"]
c <- tab_m["Female", "Yes"]; d <- tab_m["Female", "No"]
cat("OR:", round((a * d) / (b * c), 2))
```
:::

### Exercise 3 (Open-ended)

Find a binary or nominal outcome in your own data. Construct a contingency table with a grouping variable of interest. Run the appropriate test, check assumptions, and write a two-sentence result suitable for a paper.


## Comprehension Check

1. When should you use Fisher's exact test instead of a chi-square test?
2. You run `chisq.test()` and R prints a warning: "Chi-squared approximation may be incorrect." What does this mean, and what should you do?
3. A chi-square test gives p = 0.001. Does this tell you which group has a higher proportion of events?
4. What is the difference between a relative risk (RR) and an odds ratio (OR)? When are they approximately equal?
5. You have a 3?4 contingency table (3 groups, 4 ordered categories). What are the degrees of freedom for the chi-square test?

::: {.callout-note collapse="true"}
### Answers
1. When any expected cell count is < 5. Fisher's exact test does not rely on the large-sample chi-square approximation and is valid for any sample size.
2. The warning means at least one expected cell count is < 5, so the chi-square approximation is unreliable. Switch to `fisher.test()`.
3. No. Chi-square is a non-directional test; it tells you whether an association exists, not its direction or size. You must examine the contingency table or plot the proportions to see which group has higher rates.
4. Relative risk = (proportion in group A) / (proportion in group B). Odds ratio = (odds in group A) / (odds in group B). They are approximately equal when the event is rare (< 10?15% in both groups). For common outcomes, the OR exaggerates the association relative to the RR.
5. Degrees of freedom = (rows - 1) - (columns - 1) = (3 - 1) - (4 - 1) = 6.
:::


## How to Report

### Reporting Chi-square and Fisher Tests

::: {.callout-note}
**In a methods section:** "The association between [categorical variable 1] and [categorical variable 2] was assessed using the chi-square test of independence. Fisher's exact test was used where any expected cell count was below 5."

**In results:** "There was a significant association between [var 1] and [var 2] (\u03c7²(df) = X, p = Y). [Group A] had a higher proportion of [outcome] than [group B] (X% vs Y%)."

**What to always include:**

- Chi-square statistic with degrees of freedom (e.g., \u03c7²(2) = 8.4)
- p-value
- The actual proportions/percentages --- the test result alone is not informative
- Whether Fisher's exact test was used (and why)

**Effect size:** For 2\u00d72 tables, also report odds ratio or relative risk with 95% CI to quantify the magnitude of the association.
:::

## Further Reading

- @Agresti-2019 --- *An Introduction to Categorical Data Analysis*, the standard reference
- @Bland-2015 --- Chi-square and Fisher's test in *An Introduction to Medical Statistics*
- `?chisq.test`, `?fisher.test`, `?prop.test` in R: built-in documentation with examples
