# Clinical Research Methods in R {#sec-clinical-methods}

```{r setup}
#| include: false
#| eval: true
source(here::here("R/functions.R"))
source(here::here("R/load-packages.R"))
```

## Introduction

This session introduces clinical research modeling concepts such as
comparative analysis, cohort studies, and hypothesis testing for research.

## Learning outcomes

- Understand study design and research questions.
- Translate clinical data questions into analysis plans.
 - Translate clinical data questions into analysis plans.
 - Implement analytic workflows with modern R tools for data wrangling and visualization.

## Methods covered in this session

- Correlation analysis for variable association.
- One-way and two-way ANOVA for group comparisons.
- Multiple comparison adjustments (Tukey, Dunnett, Bonferroni, Holm).
- Chi-square tests for categorical data relationships.
- Logistic regression for binary outcomes and risk interpretation.
- Model diagnostics and interpretation of fitted models.

## One-way and two-way ANOVA with multiple comparisons

ANOVA tests whether group means differ; use `glht` from `multcomp` to
extract adjusted comparisons (Tukey, Dunnett).

```{r anova-glht}
pg <- datasets::PlantGrowth %>% as_tibble()

pg_lm <- lm(weight ~ group, data = pg)
summary(pg_lm)

library(multcomp)
pg_glht <- glht(pg_lm, linfct = mcp(group = "Tukey"))
summary(pg_glht)
confint(pg_glht)
```

How to explain to a clinician

- **Report:** "ANOVA tests whether any group differs; Tukey shows pairwise differences with CIs adjusted for multiplicity."  
- **Context:** emphasize effect sizes and which pairwise contrasts are clinically meaningful rather than only p-values.

## Logistic regression and odds ratios

Logistic regression models the log-odds of a binary outcome; exponentiated coefficients are odds ratios.

```{r logistic-example}
mtcars <- datasets::mtcars %>% as_tibble(rownames = "model") %>%
	mutate(manual = if_else(am == 1, 1, 0))

fit_log <- glm(manual ~ wt + hp, family = binomial, data = mtcars)
summary(fit_log)

# Odds ratios and 95% CI
exp(coef(fit_log))
exp(confint.default(fit_log))

# Predicted probabilities for example profiles
new_profiles <- tibble(wt = c(2.5, 3.5), hp = c(100, 150))
new_profiles$pred_prob <- predict(fit_log, newdata = new_profiles, type = "response")
new_profiles
```

How to explain to a clinician

- **Report:** present the odds ratio with 95% CI and an example probability change: "A patient with wt=X and hp=Y has predicted probability Z% (95% CI [L%, U%])."  
- **Context:** clarify that odds ratios are multiplicative changes in odds and that risk differences may be easier to communicate when baseline risk is known.

## Exercises (short)

- Fit the logistic model above and report the odds ratio for a 1-unit change in `wt` with its 95% CI; write one sentence interpreting the clinical meaning.

**Hint:** use `glm(..., family=binomial)`, `exp(coef())` and `exp(confint.default())`.

**Additional Exercises**

- See `dev_code_data/excercise8.Rmd` for a longitudinal / repeated-measures example using `lmm()` and mixed models; reproduce the analysis and examine mean differences and model summaries.

::: {.callout-tip collapse="true" title="Exercise 8 — Note"}
- Exercise 8 (longitudinal LMM): `dev_code_data/excercise8.Rmd` and `Exercise8.pdf`.  
- No pre-prepared solution file is included for Exercise 8 in `dev_code_data` — students should run the Rmd and inspect outputs.
:::

## Instructor notes

- Target audience: clinicians with basic R familiarity; emphasize interpretation over p-values.
- Estimated time: 75–90 minutes (lecture + hands-on exercise).
- Pre-run: ensure `dev_code_data/package-loading.R` or `R/load-packages.R` is sourced so required packages are available.
- Suggested assessment: short write-up interpreting logistic regression output for a clinical example.

## Worked example — logistic regression (concise)

```{r clinical-worked-logistic, echo=TRUE}
# Minimal clinical-style logistic regression example using `mtcars`
library(dplyr)
mtcars2 <- datasets::mtcars %>%
	as_tibble(rownames = "model") %>%
	mutate(manual = factor(if_else(am == 1, "Manual", "Automatic")))

fit_log <- glm(manual ~ wt + hp, family = binomial(), data = mtcars2)

# Tidy coefficients and compute odds ratios
broom::tidy(fit_log) %>%
	mutate(odds_ratio = exp(estimate))

# Predicted probability for two example profiles
new_profiles <- tibble(wt = c(2.5, 3.5), hp = c(100, 150))
predict(fit_log, newdata = new_profiles, type = "response")
```

## Short exercise (clinical case)

- Run `dev_code_data/excercise6.R` to fit a logistic model for a binary outcome, compute odds ratios with 95% CIs, and write a one-sentence clinical interpretation.

::: {.callout-tip collapse="true" title="Exercise — Solution"}
- Solution script: `dev_code_data/Solution6.R`  
- Solution (PDF): `dev_code_data/Solution6.pdf`  
:::

## Further reading

- Clinical methods and applied workflows: @Wickham-Grolemund-2017; @Matloff-2011.
- Model selection and interactive model building: @Henderson-Velleman-1981.

## Resources

- Exercises and lecture notes: `dev_code_data/lesson_basic_statistic_course.R`, `dev_code_data/Lecture1.pdf`.
