# Correlation and Association {#sec-correlation}

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(survival)
library(broom)
library(GGally)
```

## When Do You Use This? {.callout-tip}

> Before fitting a regression model, you want to understand the pairwise relationships in your data: does albumin decline as bilirubin rises? Is there a linear relationship between platelet count and prothrombin time? Correlation quantifies the direction and strength of these relationships before you commit to a model.

## Learning Objectives

After completing this session you will be able to:

- Distinguish Pearson, Spearman, and Kendall correlation coefficients and choose appropriately
- Calculate correlations and test their significance in R
- Visualise correlation matrices with `GGally::ggpairs()`
- Explain why correlation does not imply causation
- Recognise spurious correlation from confounding

---

## Background: Measuring Linear Association

**Pearson's r** measures the strength and direction of a *linear* relationship between two continuous variables:

$$r = \frac{\sum (x_i - \bar{x})(y_i - \bar{y})}{\sqrt{\sum (x_i - \bar{x})^2 \sum (y_i - \bar{y})^2}}$$

$r$ ranges from −1 (perfect negative) to +1 (perfect positive). $r = 0$ means no *linear* association.

**When not to use Pearson's r:**
- One or both variables are heavily skewed or contain outliers
- The relationship is clearly non-linear (e.g., exponential)
- Data are ordinal (ranked categories)

**Spearman's $\rho$** and **Kendall's $\tau$** work on the *ranks* of the data. They detect any monotonic relationship (not just linear) and are robust to outliers.

| Coefficient | Assumes | Best for |
|---|---|---|
| Pearson $r$ | Linearity, approximately bivariate normal | Continuous, symmetric variables |
| Spearman $\rho$ | Monotonic relationship | Skewed data, small n, ordinal |
| Kendall $\tau$ | Monotonic relationship | Small n, many ties |

---

## Worked Example 1: Clinical Data (PBC)

### Pairwise Correlations

```{r pbc-setup}
pbc <- survival::pbc |>
  as_tibble() |>
  janitor::clean_names() |>
  filter(!is.na(albumin), !is.na(bili), !is.na(protime), !is.na(platelet))
```

```{r cor-pair}
# Pearson correlation: albumin vs bilirubin
cor.test(pbc$albumin, log(pbc$bili), method = "pearson")
```

```{r cor-spearman}
# Spearman: more appropriate given bilirubin's skew
cor.test(pbc$albumin, pbc$bili, method = "spearman")
tidy(cor.test(pbc$albumin, pbc$bili, method = "spearman"))
```

**Why log-transform before Pearson?** Bilirubin is heavily right-skewed. Pearson's r is sensitive to extreme values; log-transforming makes the relationship more linear and the correlation estimate more meaningful.

### Scatter Plot with Smooth

```{r scatter}
pbc |>
  ggplot(aes(x = log(bili), y = albumin)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "lm", color = "tomato", se = TRUE) +
  labs(title = "Albumin vs log(bilirubin) in PBC",
       x = "log(Bilirubin)", y = "Albumin (g/dL)") +
  theme_bw()
```

### Correlation Matrix

```{r cor-matrix}
pbc_num <- pbc |>
  mutate(log_bili = log(bili)) |>
  select(albumin, log_bili, protime, platelet)

cor(pbc_num, use = "complete.obs", method = "spearman") |>
  round(2)
```

```{r ggpairs}
#| fig-height: 6
ggpairs(
  pbc_num,
  lower = list(continuous = wrap("points", alpha = 0.3, size = 0.8)),
  upper = list(continuous = wrap("cor", method = "spearman")),
  diag  = list(continuous = wrap("densityDiag"))
) +
  theme_bw(base_size = 10)
```

`GGally::ggpairs()` gives you scatterplots (lower triangle), Spearman correlations (upper triangle), and density plots (diagonal) in one call.

### Testing Multiple Correlations

```{r multi-cor}
# tidy approach: all pairs
pbc_num |>
  pivot_longer(-albumin, names_to = "variable", values_to = "value") |>
  group_by(variable) |>
  summarise(
    r    = cor(albumin, value, use = "complete.obs", method = "spearman"),
    tidy(cor.test(albumin, value, method = "spearman"))
  ) |>
  select(variable, r, p.value) |>
  mutate(p_adj = p.adjust(p.value, method = "holm"))
```

---

## Worked Example 2: Animal Data (palmerpenguins)

The penguin dataset provides a clean example of correlation *within groups* versus *across groups* — a useful lesson about confounding.

```{r penguins}
library(palmerpenguins)

penguins_clean <- penguins |>
  drop_na(bill_length_mm, bill_depth_mm, body_mass_g, flipper_length_mm, species)
```

```{r penguin-overall}
# Overall correlation (ignoring species)
cor.test(penguins_clean$bill_length_mm, penguins_clean$bill_depth_mm,
         method = "pearson")
```

```{r penguin-species}
# Within-species correlations
penguins_clean |>
  group_by(species) |>
  summarise(
    r = cor(bill_length_mm, bill_depth_mm, method = "pearson"),
    n = n()
  )
```

```{r penguin-plot}
penguins_clean |>
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c("steelblue", "tomato", "seagreen")) +
  labs(title = "Simpson's Paradox: bill dimensions in penguins",
       subtitle = "Overall correlation is negative; within-species it is positive",
       x = "Bill length (mm)", y = "Bill depth (mm)") +
  theme_bw()
```

This is **Simpson's Paradox**: the overall (marginal) correlation is negative, but within each species it is positive. The species variable confounds the relationship. Always check whether a third variable explains an unexpected correlation.

---

## Where to Find Data Like This {.callout-note}

**PBC dataset:** `survival::pbc` — multiple continuous clinical variables (albumin, bilirubin, protime, platelet, etc.) with known clinical correlations.

**palmerpenguins:** `palmerpenguins::penguins` — body measurements with a clear grouping variable for demonstrating confounding.

**NHANES:** `nhanesA` or `nhanes` package — population-level continuous variables (blood pressure, cholesterol, BMI, age).

---

## What Can Go Wrong {.callout-warning}

**Correlation ≠ causation**
Two variables can correlate strongly because they share a common cause, not because one causes the other. Always think mechanistically.

**Using Pearson's r with skewed data or outliers**
A single extreme outlier can drive a large r. Always plot your data before computing a correlation. Use Spearman's ρ for robustness.

**Multiple comparisons in correlation matrices**
Testing all $p(p-1)/2$ pairs from a correlation matrix inflates Type I error. Adjust p-values with `p.adjust()` when reporting multiple correlations.

**Range restriction**
If you select subjects based on one of the variables (e.g., only patients with bilirubin > 5), the correlation in that subset will differ from the population correlation — usually attenuated.

**Ecological correlation**
Correlations computed on group averages (e.g., countries, hospitals) are typically stronger than individual-level correlations. Do not interpret ecological correlations as individual-level associations.

---

## Exercises

### Exercise 1 (Guided): Albumin and Prothrombin Time

1. Plot albumin vs prothrombin time in `pbc`. Does the relationship look linear?
2. Compute Pearson and Spearman correlations. Which is higher? Why?
3. Compute Pearson correlation after log-transforming prothrombin time.
4. Report: r, 95% CI, and p-value for your chosen method.

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r}
# 1. Plot
pbc |>
  ggplot(aes(x = protime, y = albumin)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "loess", color = "tomato") +
  labs(x = "Prothrombin time (s)", y = "Albumin (g/dL)") + theme_bw()

# 2. Both correlations
tidy(cor.test(pbc$albumin, pbc$protime, method = "pearson"))
tidy(cor.test(pbc$albumin, pbc$protime, method = "spearman"))

# 3. After log transform
tidy(cor.test(pbc$albumin, log(pbc$protime), method = "pearson"))
```
:::

### Exercise 2 (Semi-guided): Penguin Correlation Matrix

Using `palmerpenguins::penguins`, produce a correlation matrix for the four continuous measurements.

1. Compute a Spearman correlation matrix using `cor()`.
2. Plot with `ggpairs()`.
3. Run the analysis separately for each species. Do any correlations change sign?

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r}
peng <- penguins |> drop_na()

# 1. Spearman matrix
peng_num <- peng |> select(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g)
cor(peng_num, method = "spearman") |> round(2)

# 2. ggpairs
ggpairs(peng_num,
        upper = list(continuous = wrap("cor", method = "spearman")),
        lower = list(continuous = wrap("points", alpha = 0.3, size = 0.8))) +
  theme_bw()

# 3. By species
peng |>
  group_by(species) |>
  summarise(r_bl_bd = cor(bill_length_mm, bill_depth_mm, method = "spearman"))
```
:::

### Exercise 3 (Open-ended)

Using your own data, identify two continuous variables you expect to be related. Plot them, compute the appropriate correlation, and test the significance. Write a two-sentence result. Now think of a third variable that might confound the relationship — how would you check for it?

---

## Comprehension Check

1. When should you use Spearman's ρ instead of Pearson's r?
2. You compute r = 0.85, p < 0.001 between two blood biomarkers. Can you conclude one biomarker causes changes in the other?
3. A correlation matrix of 10 variables has 45 pairwise tests. Three are significant at p < 0.05. Is this noteworthy?
4. You observe a strong negative correlation between overall bill length and bill depth in penguins, but a positive correlation within each species. What is this called, and what causes it?
5. You want to correlate a Likert-scale pain score (1–10) with serum CRP. Which coefficient should you use?

::: {.callout-note collapse="true"}
### Answers
1. Use Spearman when (1) one or both variables are skewed or contain outliers, (2) the relationship is monotonic but not necessarily linear, or (3) one variable is ordinal.
2. No. Correlation describes the strength of association, not direction of causation. The two biomarkers may both reflect a third underlying process (confounding), or the direction of causality could run either way.
3. Not necessarily. With 45 tests at α = 0.05, you expect 45 × 0.05 = 2.25 false positives by chance. Three significant results is only slightly above the expected false-positive count. Adjust for multiple testing and scrutinise the findings carefully.
4. This is Simpson's Paradox. The species variable is a confounder: different species have distinct bill morphologies, and within each species the relationship is positive. The overall correlation is dominated by between-species variation, which moves in the opposite direction.
5. Spearman's ρ (or Kendall's τ). Pain scores are ordinal — the difference between 4 and 5 is not necessarily the same as between 7 and 8 — so rank-based methods are more appropriate than Pearson's r.
:::

## Further Reading

- @Bland-2015 --- Correlation in *An Introduction to Medical Statistics*
- @Dalgaard-2008 --- Chapter on correlation and regression
- `?cor.test` in R — documentation covering all three methods
- `GGally::ggpairs()` vignette for visualising multivariate relationships
