# Exploring clinical and medical data with R {#explore-clinical-data}

```{r setup}
#| include: false
#| eval: true
source(here::here("R/functions.R"))
source(here::here("R/load-packages.R"))

```

## Introduction

In this session, we'll dive into publicly available clinical and medical
datasets, utilizing essential statistical tools to analyze and interpret
complex data using R. With the help of `dplyr` for data manipulation and
`ggplot2` for visualization, participants will gain practical experience
working with real-world datasets. This hands-on approach will not only
enhance analytical skills but also contribute significantly to
advancements in medical research. For some dtasets we have genrated
simualted data when the data was not available.

::: callout-important
Please install and load the `dplyr` `ggplt2` packages before proceeding.
If you have not done so already please do so.
:::

::: callout-tip
`iris` data set is a famous real world data set that is available in R.
It is a data set that contains information about iris. It contains 150
observations and 5 variables. The variables are sepal length, sepal
width, petal length, petal width, and species. The species variable has
three levels: setosa, versicolor, and virginica. The data set is
available in R and can be loaded using the data() function. The data set
is also available in CSV format on the internet.
:::

::: {.callout-tip collapse="true" title="Want to know what the summary output means?"}
:::

::: {.callout-tip collapse="true" title="Excercise 8: Click to see solution"}
:::

::: callout-tip
:::

# Exploring clinical and medical data with R

The field of clinical research is rich with complex and varied data,
encompassing everything from patient demographics to treatment outcomes
and beyond. Analyzing this data to uncover trends, patterns, and
statistical insights is crucial for advancing medical knowledge and
improving patient care. However, the sheer volume and complexity of
clinical datasets can pose significant challenges, especially when it
comes to data manipulation and visualization.

This educational guide is designed to address these challenges by
introducing readers to the power of `dplyr` and `ggplot2`, two of the
most popular packages in the `tidyverse` suite of tools for data science
in R. Through a series of examples using clinical datasets, this guide
will demonstrate how to transition from base R techniques to a more
`tidyverse`-centric approach for data analysis and visualization.

Our journey will begin with an exploration of several clinical datasets,
each offering unique insights into different aspects of patient data. We
will start with basic data manipulations, such as filtering,
summarizing, and transforming data, before moving on to in depth
statistics analysis and interpretation to uncover trends and statistics
that lie within the data.

Whether you are a beginner in R or looking to expand your data analysis
toolkit, this guide aims to equip you with the skills and knowledge to
effectively analyze and visualize clinical data using `dplyr` and
`ggplot2`. Let's embark on this journey together, unlocking the
potential of clinical datasets to inform and inspire the future of
healthcare.

## Dataset Overview

### Migraine Dataset

The `migraine` dataset is a clinical data collection used to analyze the
distribution of migraine medication dosages among patients. It includes
two main variables: the dosage levels of the medication (measured in
milligrams) and the number of patients treated at each dosage level.
This dataset is particularly useful for understanding how different
dosages are administered in practice and for investigating the
dose-response relationship in migraine treatment. It serves as a
valuable resource for practicing data manipulation, visualization, and
statistical analysis in R, making it an ideal tool for learners and
researchers interested in clinical data analysis.

```{r}
library(dplyr)
library(ggplot2)
```

```{r load-migraine}
data(migraine)
head(migraine)
```

### Interpretation

Interpretation of the Columns **dose**: This column helps in analyzing
the dose-response relationship, which is crucial for determining the
optimal dosage of medication that is effective in treating migraines
with minimal side effects.

**painfree**: By indicating how many patients became pain-free at each
dosage level, this column allows for the assessment of the medication's
efficacy. Higher numbers in this column, relative to the total number of
patients treated (ntrt), suggest a more effective dose.

**ntrt**: Reflecting the total number of patients treated at each dosage
level, this column is essential for understanding the distribution of
treatment across the study population. It helps in assessing the
popularity or research focus on specific dosage levels.

### Analytical Insights

With these columns, the dataset can be analyzed to uncover insights such
as:

-   The effectiveness of different dosages in providing relief from
    migraine symptoms.
-   The dosage level at which the highest proportion of patients
    reported being pain-free.
-   The relationship between dosage levels and the occurrence of
    effective relief, which can inform dosage recommendations and
    further research.

### Explore the Data

### Count number of patients in each doese group

```{r migraine-barplot-counts}
# Start by specifying the dataset and then piping it into ggplot for visualization
migraine %>%
  ggplot(aes(x = factor(dose), # Convert the dose variable to a factor for discrete x-axis values
             y = ntrt,        # Use the number of patients for the y-axis values
             fill = factor(dose))) + # Color code bars based on the dose for visual distinction
  geom_bar(stat = "identity") + # Create a bar plot with heights equal to y-axis values (ntrt)
  geom_text(aes(label = ntrt),  # Add text labels on bars showing the number of patients
            vjust = -0.5,      # Adjust text vertically to appear just above the bars
            position = position_dodge(width = 0.9)) + # Ensure text labels are centered over bars
  labs(x = "Dose (mg)",         # Label for the x-axis
       y = "Number of patients",# Label for the y-axis
       fill = "Dose") +         # Label for the legend, explaining the color fill represents dose levels
  theme_minimal()               # Use a minimalistic theme for a clean and focused visualization

```

### Convert patients count into frequencies

While the bar plot above provides a clear view of the distribution of
patients across different treatment doses, it can be further enhanced by
transforming the raw patient counts into percentages. This
transformation offers several advantages:

1.  **Facilitates Fair Comparison**: Transforming raw patient counts
    into percentages allows for equitable comparison across different
    treatment doses, highlighting trends and preferences in treatment
    practices without the bias of unequal group sizes.

2.  **Enhances Interpretability**: Using percentages makes the data more
    intuitive and accessible, enabling stakeholders to quickly grasp the
    distribution of patients across treatment doses without needing to
    parse raw numbers.

3.  **Informs Clinical and Research Decisions**: The visualized
    distribution offers insights into prescribing behaviors and patient
    demographics, guiding research focus on prevalent doses and
    optimizing patient care strategies based on observed treatment
    patterns.

```{r migraine-barplot-frequencies}
# Start with the migraine dataset
migraine %>% 
  # Add a new column 'frequency' that calculates the percentage of patients for each dose
  mutate(frequency = ntrt / sum(ntrt) * 100) %>%
  # Initialize a ggplot with 'dose' on the x-axis and the calculated 'frequency' on the y-axis
  ggplot(aes(x = factor(dose), y = frequency, fill = factor(dose))) +
  # Create a bar plot where the height of each bar corresponds to 'frequency' value
  geom_bar(stat = "identity") +
  # Add text labels on each bar to show the frequency percentage, formatted to one decimal place
  geom_text(aes(label = sprintf("%.1f%%", frequency)), # Format the frequency values as percentages
            vjust = -0.5, # Vertically adjust the labels to sit above the bars
            position = position_dodge(width = 0.9)) + # Position the text correctly even when bars are dodged
  # Set labels for the axes and the legend
  labs(x = "Dose (mg)", y = "Patients (%)", fill = "Dose") +
  # Apply a minimalistic theme for a clean and uncluttered plot appearance
  theme_minimal()

```

To gain analytical insights we can analyze the `migraine` dataset
focusing on the effectiveness of different dosages, identifying the
dosage level with the highest proportion of patients reporting relief
(pain-free), and exploring the relationship between dosage levels and
effective relief, we can use below steps and code snippets for each of
these analyses:

1.  **Calculate the effectiveness of different dosages**:

Effectiveness can be quantified as the proportion of patients who are
pain-free at each dosage level.

```{r}
# Calculate the effectiveness of each dose in the migraine dataset
migraine_effectiveness <- migraine %>%
  # Create a new column 'efficiency' by dividing the number of pain-free patients by total number of patients
  mutate(efficiency = painfree / ntrt) %>%
  # Select only the 'dose' and the newly created 'efficiency' columns for further analysis
  dplyr::select(dose, efficiency)

# Identify the dose with the highest effectiveness
highest_effective_dose <- migraine_effectiveness %>%
  # Arrange the dataset in descending order of efficiency to bring the highest value to the top
  arrange(desc(efficiency)) %>%
  # Select the first row, which now represents the dose with the highest efficiency
  slice(1)

# Output the dose with the highest effectiveness
print(highest_effective_dose)



```

2.  **Visual relationship between dosage levels and the occurrence of
    effective relief**:

```{r}
# Create a plot to visualize the effectiveness of different dosages in the migraine dataset
ggplot(migraine, aes(x = dose, y = painfree / ntrt)) + # Initialize plot with dose on x-axis and proportion pain-free on y-axis
  geom_line(group = 1, color = "blue") + # Draw a line connecting the points for visualizing trends, with a specified group and color
  geom_point(color = "red") + # Add points to represent each data value, colored red for visibility
  geom_text(aes(label = dose), vjust = -0.5, color = "black") + # Add text labels for each dose above the corresponding point
  labs(x = "Dose (mg)", y = "Proportion Pain-Free", # Customize axis labels
       title = "Effectiveness of Different Dosages in Providing Relief from Migraine Symptoms") + # Set the title of the plot
  theme_minimal() # Use a minimal theme for a clean and modern look

```

-   **Visual Relationship**: The graph illustrates how the proportion of
    pain-free patients varies with different dosages. This can help in
    identifying the optimal dosage that balances effectiveness with the
    potential for side effects, guiding dosage recommendations and
    future research.

These analyses provide insights into the dataset's representation of
medication effectiveness for migraine relief across different dosages,
offering a foundation for more detailed statistical analysis and
clinical decision-making.

### Galton Dataset

The Galton dataset is a historical collection of data gathered by
Francis Galton, a pioneer in the field of eugenics and statistical
correlation. This dataset specifically contains measurements of the
heights of parents and their adult children. It is often used to study
heredity and the relationship between the physical characteristics of
parents and their offspring, particularly focusing on the concept of
regression toward the mean. The dataset is a classic example used in
statistical analyses to demonstrate correlation, linear regression, and
other statistical concepts related to inheritance.

```{r load-galton}
data(Galton)
head(Galton)
```

```{r}
# Summary statistics for the heights of parents and children in the Galton dataset
summary(Galton)
```

### Interpretation

The summary of the Galton dataset provides key statistical metrics, such
as the mean (average height), median (middle value of the height data),
range (difference between the smallest and largest values), and other
measures of dispersion like the quartiles. These metrics offer insights
into the distribution of heights within the dataset, indicating trends,
variability, and central tendencies of the data. Initially, the heights
are recorded in inches, a common practice during Galton's time. However,
to enhance the interpretability and visualization of the data, it may be
beneficial to convert the units to meters, a more universally recognized
unit of measurement for height. This conversion can make the data more
accessible and relatable to a broader audience, facilitating a clearer
understanding of the distribution of heights among parents and children.

### Converting units from inches to meters

```{r}
# Convert heights from inches to meters for both children and parents
galton <- Galton %>%
  mutate(
    child = child * 0.0254,  # Convert child heights to meters
    parent = parent * 0.0254  # Convert parent heights to meters
  )
```

```{r}
# Applying jitter to children's height to add a small random variation
# This helps in visualizing overlapping data points in the plot
child2 <- jitter(galton$child)
```

```{r}
# Plotting the density histogram of children's heights in meters
ggplot() + 
  geom_histogram(
    aes(x = child2, y = ..density..),  # Map child2 to x, automatically calculate density for y
    binwidth = 0.001,  # Set binwidth to ensure meaningful distribution representation
    fill = "grey"  # Set the color of histogram bars to grey for visual neutrality
  ) +
  labs(
    x = "Height (m)",  # Label the x-axis as "Height (m)"
    y = "Density"  # Label the y-axis as "Density" to indicate density plot
  ) +
  theme_minimal()  # Apply a minimalistic theme for a clean and modern look
```

```{r}
# Plotting the frequency histogram of children's heights in meters
# This plot shows the frequency of data points within each bin, providing a count-based view
ggplot() + 
  geom_histogram(
    aes(x = child2),  # Map child2 to x
    binwidth = 0.001,  # Maintain the same binwidth as the density plot for consistency
    fill = "grey"  # Match the fill color with the density histogram
  ) +
  labs(
    x = "Height (m)",  # Consistent axis labeling with the density histogram
    y = "Frequency"  # Indicate that this plot shows the frequency of observations
  ) +
  theme_minimal()  # Same minimal theme for a cohesive look across plots


```

### ToothGrowth Dataset

The ToothGrowth dataset is a fascinating collection of data that
explores the impact of Vitamin C on tooth growth in guinea pigs. It
details the length of odontoblasts (cells responsible for tooth growth)
in response to varying doses of Vitamin C, providing a unique insight
into nutritional effects on biological growth processes. The dataset
includes three key variables: len (tooth length), supp (supplement type:
VC or OJ), and dose (Vitamin C dose in milligrams per day). Analyzing
this dataset can offer valuable insights into the relationship between
Vitamin C intake and tooth growth, with potential implications for
dental health and nutrition.

```{r}
# Load the ToothGrowth dataset into the R session
data(ToothGrowth)

# Convert the 'dose' column from numeric to a factor to reflect its categorical nature
ToothGrowth <- ToothGrowth %>%
  mutate(dose = as.factor(dose))

# Display the first few rows of the dataset to understand its structure and contents
head(ToothGrowth)

```

```{r}
# Generate a summary of the ToothGrowth dataset to get an overview of the data
summary(ToothGrowth)
# The summary provides a quick overview of the dataset, including the range, median, and quartiles of tooth length for each combination of supplement type and Vitamin C dose.
```

### The effect different doses of Vitamin C on tooth growth:

```{r toothgrowth-analysis}
# Use ggplot2 to visualize the effect of different doses of Vitamin C on tooth growth
# and distinguish the results by supplement type
ggplot(ToothGrowth, aes(x = dose, y = len, color = dose, shape = supp)) +
  geom_jitter(width = 0.1) + # Add jitter to visually separate overlapping points
  labs(x = "Dose (mg/day)", y = "Tooth Length (microns)", color = "Dose", shape = "Supplement Type") +
  theme_minimal() # Apply a minimalistic theme for a clear and focused visualization

```

### Melanoma Dataset

The melanoma dataset is a crucial resource in medical research,
particularly in the study of melanoma, a serious form of skin cancer.
This dataset compiles measurements of tumor thickness from a variety of
patients diagnosed with melanoma. Tumor thickness, measured in
millimeters (mm) or as specified, is a significant prognostic factor in
melanoma treatment and research, as it's directly correlated with the
likelihood of melanoma spreading (metastasis). Understanding the
distribution and central tendencies of tumor thickness can aid in better
treatment planning, risk assessment, and patient prognosis.

```{r load-melanoma}
data(melanoma)
head(melanoma)
```

### Interpretation and Visualizations

We'll explore the distribution of tumor thickness, calculate basic
statistics, and visualize the data:

### Histogram and Statistics

```{r melanoma-analysis}
# Using ggplot2 to create a histogram that visualizes the distribution of tumor thickness in the dataset
ggplot(melanoma, aes(x = thick)) +
  geom_histogram(binwidth = 1, fill = "grey", color = "black") + # Setting binwidth to 1 for detailed granularity
  labs(x = "Tumor thickness (1/100 mm)", y = "Frequency") + # Labeling x-axis as tumor thickness and y-axis as frequency of occurrence
  theme_minimal() # Applying a minimal theme for a clean and focused visualization
```

The histogram provides a visual representation of how tumor thickness
values are distributed among the patients in the melanoma dataset. Each
bar in the histogram represents a range of tumor thickness, and the
height of the bar indicates the frequency of observations within that
range. The visualization allows one to quickly grasp the most common
ranges of tumor thickness, how the thickness values are spread out, and
whether there are any peaks or unusual patterns in the data distribution
(such as skewness). In this plot we can see that the distribution is
right-skewed, with a long tail to the right, indicating that there are a
few cases with very high tumor thickness. On the left side, the
distribution is more concentrated, indicating that most cases have lower
tumor thickness.

```{r melanoma-analysis}
# Calculate and summarize the mean and standard deviation of tumor thickness in the dataset
melanoma %>% 
  summarise(mean_thickness = mean(thick), # Calculate the mean thickness
            sd_thickness = sd(thick)) # Calculate the standard deviation of thickness
```

#### Explanation of mean_thickness:

The mean_thickness represents the average tumor thickness across all
measured melanoma cases in the dataset. It's a central measure that
gives a general idea of tumor thickness, providing insight into the
typical severity of cases within the dataset. The mean is particularly
useful for understanding the overall distribution and can be a starting
point for further statistical analysis, such as comparing the average
tumor thickness across different subgroups (e.g., based on treatment
type, patient demographics).

#### Explanation of sd_thickness:

sd_thickness represents the standard deviation of tumor thickness, which
quantifies the amount of variation or dispersion in the thickness
measurements. A higher standard deviation indicates greater variability
in tumor thickness, suggesting a wider range of severity among melanoma
cases. Understanding the standard deviation is crucial for assessing the
consistency or spread of tumor thickness values, as it provides context
for the mean and helps identify potential outliers or extreme cases.

### Visualizing Outliers

```{r melanoma-analysis}
# Creating a boxplot to visualize the distribution of tumor thickness and identify potential outliers
p <- ggplot(melanoma, aes(y = thick)) +
  geom_boxplot(fill = "lightblue", color = "black") + # Customizing boxplot appearance
  labs(y = "Tumor thickness (1/100 mm)") + # Labeling y-axis as tumor thickness
  theme_minimal() # Applying a minimal theme for a clean and focused visualization


# Calculate the quartiles and median
Q1 <- quantile(melanoma$thick, 0.25) # Calculate the quartiles and median
Median <- median(melanoma$thick) # Calculate the quartiles and median
Q3 <- quantile(melanoma$thick, 0.75) # Calculate the quartiles and median

# Assuming you've  calculated the potential outlier values
OutlierValue <- max(subset(melanoma$thick, melanoma$thick > Q3 + 1.5 * IQR(melanoma$thick))) # Calculate the potential outlier value using the IQR method


# Then, manually add annotations
p + annotate("text", x = 1, y = Q1, label = paste("Q1 (25th percentile):", round(Q1, 2)), vjust = 0, hjust = 1.73) + # Add annotation for Q1
  annotate("text", x = 1, y = Median, label = paste("Median:", round(Median, 2)), vjust = -0.3, hjust = 6) + # Add annotation for median
  annotate("text", x = 1, y = Q3, label = paste("Q3 (75th percentile):", round(Q3, 2)), vjust = 1.55, hjust = 1.65) + # Add annotation for Q3
  annotate("text", x = 1, y = OutlierValue, label = "Outliers", vjust = 2, hjust = 8.5) # Add annotation for potential outliers
```

#### Log Transformation of raw data

Log transformation is a widely used technique in data analysis to handle
skewed data or data that spans several orders of magnitude. By applying
a logarithmic scale to the data, several benefits are achieved, which
can improve the analysis and visualization of the data. Here's an
explanation of what log transformation does and why it might be applied,
especially in the context of visualizing tumor thickness in the
`melanoma` dataset:

### Purpose of Log Transformation

1.  **Normalizing Data**: Many statistical methods assume that the data
    follows a normal (Gaussian) distribution. Skewed data violate this
    assumption, which can affect the validity of statistical tests and
    models. Log transformation can help in reducing skewness, making the
    data more symmetrical or "normal."

2.  **Reducing the Impact of Outliers**: In datasets with significant
    outliers or vast ranges of values, outliers can disproportionately
    affect the mean and standard deviation. Log transformation reduces
    the relative impact of very large values, which can make patterns
    more apparent and statistical analysis more robust.

3.  **Equalizing Variances**: Heteroscedasticity, where the variability
    of a variable is unequal across the range of values, can be
    problematic for regression analysis and other statistical
    techniques. Log transformation can help stabilize variances,
    especially in cases where the variance increases with the mean.

4.  **Linearizing Relationships**: Log transformation can convert
    multiplicative relationships into additive relationships, which
    makes linear modeling techniques more applicable and interpretation
    simpler. This is particularly useful in regression models where the
    relationship between variables is multiplicative rather than
    additive.

### Effect of Log Transformation

-   **On Distribution**: Applying a log transformation to skewed data
    can make the distribution more symmetric, making it resemble a
    normal distribution. This transformation is particularly effective
    for data that follows a log-normal distribution.

-   **On Scale**: Log transformation changes the scale of the data from
    absolute to relative changes. Differences in the log-transformed
    data represent percentage changes rather than absolute differences,
    which can make certain patterns more visible and easier to
    interpret.

-   **On Visualization**: Visualizing log-transformed data, as in the
    provided `ggplot` code for tumor thickness, can reveal patterns that
    are not evident in the raw data. It can make the data more
    manageable to display and interpret, especially when the data spans
    several orders of magnitude. Histograms of log-transformed data can
    show a more balanced view of the distribution, highlighting the
    central tendency and variability without being dominated by
    outliers.

### Comparison with Raw Data

-   **Skewness**: Raw data might be heavily skewed with a long tail in
    one direction. Log transformation can reduce this skewness, making
    the data more symmetric.

-   **Interpretability**: Patterns and relationships that are obscured
    in raw data due to the scale or outliers may become more evident
    after log transformation.

-   **Statistical Analysis**: Statistical analyses performed on
    log-transformed data can be more valid and interpretable, especially
    if the transformation helps meet the assumptions of the analysis
    methods being used.

In summary, log transformation is a powerful tool for making data more
amenable to analysis and visualization, especially for skewed data or
data with outliers. By applying it to the tumor thickness in the
melanoma dataset, the analysis can become more robust and insightful,
revealing patterns that might not be visible in the raw data.

```{r}
# Visualize the log-transformed tumor thickness
ggplot(melanoma, aes(x = log(thick))) +
  geom_histogram(binwidth = 0.1, fill = "grey", color = "black") +
  labs(x = "Log of tumor thickness (1/100 mm)", y = "Frequency") +
  theme_minimal()
```

```{r}
# Calculate mean and standard deviation of log-transformed thickness
melanoma %>%
  summarise(mean_log_thickness = mean(log(thick)),
            sd_log_thickness = sd(log(thick)))

```

#### Normal Range Calculation

Understanding Normal Range Calculation Log Transformation: The thickness
measurements are first log-transformed. This transformation is applied
to handle skewness in the data, reducing the impact of outliers and
making the distribution more symmetric and closer to normal (Gaussian).
This step is crucial for datasets where the raw measurements span
several orders of magnitude or are highly skewed.

**Calculating Mean and SD**: The mean (average) and standard deviation
(a measure of dispersion) of these log-transformed measurements are
computed. The mean provides a central value, while the SD indicates how
spread out the measurements are around this mean.

**Defining the Normal Range**: The normal range is defined as the
interval from mean - 2*SD to mean + 2*SD. This range is significant
because, in a normal distribution, it encompasses approximately 95% of
the data. Thus, this range can be interpreted as covering the typical,
expected variation in log-transformed thickness measurements.

**Rounding for Clarity**: The calculated lower and upper bounds of the
normal range are rounded to two decimal places. This step simplifies the
presentation and interpretation of these bounds.

**Why We Need It** Identifying Outliers: By establishing what
constitutes a "normal" range, it becomes easier to identify measurements
that fall outside this range as potential outliers. These outliers may
warrant further investigation or may be subject to different
considerations in analysis.

**Data Normalization:** Establishing a normal range on a log scale is
part of data normalization, making it easier to compare measurements
across different scales or distributions.

**Statistical Analysis:** Many statistical tests assume normally
distributed data. By transforming data and establishing a normal range,
analyses may become more robust and less sensitive to the effects of
non-normality.

**Clinical or Biological Relevance:** In medical or biological contexts,
defining a normal range can help in distinguishing between healthy and
pathological states, or in assessing the severity of a condition based
on how far a measurement deviates from the norm.

#### Explanation

This process of calculating the normal range of log-transformed
thickness measurements is a preparatory step for further statistical
analysis, providing a basis for understanding the distribution of the
data, identifying outliers, and making informed decisions in subsequent
analyses. It tailors the analysis to the characteristics of the data,
ensuring that the conclusions drawn are both accurate and relevant to
the context of the dataset, such as studying the characteristics of
melanoma or any other condition where thickness measurements are
significant.

#### Calculate normal range of log-transformed thickness

```{r}
# Calculate the normal range (mean ± 2*SD) of the log-transformed thickness measurements
# from the melanoma dataset and round the results to 2 decimal places.
normal_range_log <- melanoma %>%
  # Use summarise to calculate the lower and upper bounds of the normal range
  summarise(
    lower = mean(log(thick)) - 2*sd(log(thick)), # Calculate the lower bound
    upper = mean(log(thick)) + 2*sd(log(thick))  # Calculate the upper bound
  ) %>%
  # Round the calculated lower and upper bounds to 2 decimal places
  mutate(across(everything(), round, 2))

print(normal_range_log)
```

```{r}
# Display quantiles of tumor thickness
quantile(melanoma$thick)
```

```{r}
# Visualize tumor thickness by ulceration status
ggplot(melanoma, aes(x = factor(ulc), y = thick)) +
  geom_boxplot() +
  labs(x = "Ulceration", y = "Tumor thickness (1/100 mm)") +
  theme_minimal()
```

### Statistical Testing

```{r}
# Binomial test example
# confidence interval for a prevalence when observing 12 cases out of 1071
binom.test(x = 12, n = 1071)
```

The `binom.test` function in R performs an exact binomial test of the
null hypothesis that the probability of success in a Bernoulli
experiment is `p = 0.5` against the alternative hypothesis that the
probability of success is not equal to `0.5`. This test is useful for
determining if the observed proportion of successes differs
significantly from what would be expected under a specific null
hypothesis probability.

### Test Outcome Interpretation:

-   **Data**: The test was conducted with `12` successes observed out of
    `1071` trials.

-   **P-value**: The p-value is less than `2.2e-16`, which is extremely
    small. This indicates that the observed data (12 successes out of
    1071 trials) would be highly unlikely if the true probability of
    success were actually `0.5`. Given conventional alpha levels (e.g.,
    `0.05`), we would reject the null hypothesis that the true
    probability of success is `0.5`, suggesting that the actual
    probability of success is significantly different from `0.5`.

-   **Alternative Hypothesis**: The test's alternative hypothesis is
    that the true probability of success is not equal to `0.5`. The
    extremely low p-value supports this alternative hypothesis,
    indicating that the observed proportion of successes is
    significantly different from what would be expected under the null
    hypothesis.

-   **Confidence Interval**: The `95%` confidence interval for the true
    probability of success is between approximately `0.58%` and `1.95%`.
    This range provides an estimate of where the true probability of
    success lies, with `95%` confidence. It's clear from this interval
    that the observed proportion of successes (`1.12%`) is significantly
    lower than `50%`, further supporting the conclusion that the true
    probability of success is far from `0.5`.

-   **Sample Estimates**: The estimated probability of success, based on
    the sample, is approximately `1.12%`. This estimate is the observed
    proportion of successes in the sample and is consistent with the
    calculated confidence interval.

### Summary:

The binomial test results strongly suggest that the observed proportion
of successes (approximately `1.12%`) significantly deviates from the
`50%` expected under the null hypothesis. Given the evidence, we reject
the null hypothesis in favor of the alternative, concluding that the
true probability of success in this context is significantly different
(and much lower) than `0.5`. This outcome could reflect on the specific
context of the Bernoulli trials being tested, indicating that the event
in question is quite rare or uncommon in the population from which the
sample was drawn.

```{r}
# T-test on log-transformed thickness
t.test(log(melanoma$thick))
```

This test evaluates whether the average (mean) log-transformed tumor
thickness in the sample significantly differs from a theoretical mean,
which, by default, is `0` unless otherwise specified. Here's an
interpretation of the test results:

### Test Outcome Interpretation:

-   **Data**: The test was conducted on the log-transformed thickness
    values of the melanoma tumors.

-   **Test Statistic (t)**: The calculated t-statistic is `73.899`. This
    value represents the ratio of the deviation of the sample mean from
    the null hypothesis mean (in this case, `0`) to the standard error
    of the mean. A high t-value indicates a significant difference
    between the sample mean and the null hypothesis mean.

-   **Degrees of Freedom (df)**: The degrees of freedom for the test are
    `204`, which typically corresponds to the sample size minus one
    (`n - 1`). This suggests the original dataset had `205`
    observations.

-   **P-value**: The p-value is less than `2.2e-16`, indicating that the
    probability of observing a t-statistic as extreme as `73.899` under
    the null hypothesis is extremely low. Given conventional
    significance levels (e.g., `0.05` or `0.01`), this p-value leads to
    the rejection of the null hypothesis.

-   **Alternative Hypothesis**: The alternative hypothesis posited is
    that the true mean of the log-transformed tumor thickness is not
    equal to `0`. The extremely low p-value supports this hypothesis,
    indicating a statistically significant difference between the sample
    mean and the null hypothesis mean of `0`.

-   **Confidence Interval**: The `95%` confidence interval for the true
    mean of the log-transformed tumor thickness is between approximately
    `5.084` and `5.363`. This interval does not include `0`, further
    supporting the conclusion that the mean log-transformed thickness is
    significantly different from `0`.

-   **Sample Estimates**: The estimated mean of the log-transformed
    tumor thickness, based on the sample, is approximately `5.223`. This
    is the average value of the log-transformed thickness measurements
    in the dataset.

### Summary:

The results from the one-sample t-test on the log-transformed tumor
thickness measurements strongly indicate that the average
log-transformed thickness significantly deviates from `0`. This finding
is not only statistically significant (p-value \< 2.2e-16) but also
practically significant, as the confidence interval and the sample mean
(5.223) suggest that the true mean log thickness of melanoma tumors is
considerably greater than `0`. This analysis highlights the importance
of log transformation for normalizing data and making it suitable for
parametric tests like the t-test, providing meaningful insights into the
data's central tendency that might not be as apparent or interpretable
on the original scale.

```{r}
# extracting the 95% confidence interval for log-transformed thickness
ci_log_thickness <- t.test(log(melanoma$thick))$conf.int
print(ci_log_thickness)
```

```{r}
# Convert the confidence interval to the original scale
# Back-transformed confidence interval
exp(ci_log_thickness)
```

Applying the exponential function `exp()` to the confidence interval
(CI) of log-transformed tumor thickness data converts these values back
to their original scale, millimeters (mm) in this case. This
back-transformation makes the results more interpretable in a clinical
context.

For a log-transformed CI of `[5.083980, 5.362701]`, the back-transformed
interval is calculated as follows:

-   **Lower Limit**: `exp(5.083980)` transforms the lower bound back to
    millimeters, indicating the minimum average tumor thickness.
-   **Upper Limit**: `exp(5.362701)` transforms the upper bound back to
    millimeters, indicating the maximum average tumor thickness.

The resulting interval provides an estimated range for the true mean
tumor thickness in its original units (mm), facilitating a more
intuitive understanding of tumor size and aiding in clinical
decision-making. This approach highlights the significance of the
original scale values for practical applications, offering a direct
insight into tumor characteristics with 95% confidence.

---
title: "Basic Statistics for Health Researchers - Exercise Day 1"
date: "25 October 2021"
---

## Course Overview

This set of exercises is designed to warm up your skills in statistical
analysis using R, following the R-demo from Lecture 1. Before beginning,
ensure to:

1.  Read and run the code provided in the R-demo available on the course
    webpage.
2.  Verify that the output matches the results presented on the slides.
3.  Feel free to insert your own comments into the script to enhance
    understanding.

## Exercises with the Sickle Cell Disease (SCD) Data

### Warming Up

For today's exercise, we will be working with the Sickle Cell Disease
(SCD) dataset, which you can find in the data folder or
[here](%22data/scd_simulated_data.rda%22).

### Question 1: Data Loading and Exploration

-   Load the SCD dataset into R.
-   Use `head()` to display the first few lines of the data.
-   Provide a summary of the data using `summary()`.
-   Identify which variables contain missing values.

::: {.callout-tip collapse="true" title="Question 1: Click to see solution"}
```{r}
# Load the SCD dataset
load("../data/scd_simulated_data.rda")
```

```{r}
# assign it to an object d
d <- scd_simulated_data # Using a shorter name for convenience
```

```{r}
# Display the first few lines of the dataset
head(d)
```

```{r}
# Get summary statistics for all variables
summary(d)
```
:::

### Question 2: Counting Subjects

-   Use the `table()` function to count:
    1.  The number of male and female subjects.
    2.  How many subjects have Sickle Cell Disease (SCD)?

::: {.callout-tip collapse="true" title="Question 2: Click to see solution"}
```{r}
# Count of male and female subjects
table(d$sex)
# 1 = males, 2 = females
```

```{r}
# Count of subjects with and without SCD
table(d$scd)
# 0 = no, 1 = yes
```
:::

### Question 3: Visualizing Systolic Pressure

1.  Create two boxplots to compare the distribution of Systolic
    pressure. Ensure to fine-tune the plot with appropriate titles and
    axis labels.
2.  Describe the interpretation of each element of the boxplot (boxes,
    whiskers, dots, etc.).
3.  Evaluate and interpret the plot. does it aligns with your
    expectations?

::: {.callout-tip collapse="true" title="Question 3: Click to see solution"}
1: Creating Boxplots

```{r}

# Assuming 'd' is your dataframe and it contains 'Psys' as Systolic Pressure and 'SCD' indicating the disease presence
ggplot(d, aes(x = factor(scd, levels = c(0, 1), labels = c("No SCD", "Yes SCD")), y = psys)) +
  geom_boxplot() +
  labs(x = "Sickle Cell Disease Presence", y = "Systolic Pressure (mmHg)", title = "Systolic Pressure Distribution by SCD Status") +
  theme_minimal()
```

2: Boxplot Interpretation **Box:** The box represents the interquartile
range (IQR), extending from the first quartile (Q1, 25th percentile) to
the third quartile (Q3, 75th percentile). This range contains the middle
50% of the data. The line within the box marks the median (Q2, 50th
percentile), indicating the central tendency of the data. **Whiskers:**
The lines extending from the top and bottom of the box (the "whiskers")
typically represent the range of the data, extending up to 1.5 times the
IQR from the box's edges. Points beyond the whiskers are often
considered outliers. **Dots:** Individual points beyond the whiskers are
usually plotted as dots and represent outliers, indicating data points
that fall outside the expected range based on the IQR.

3: Interpratation The plot shows the distribution of systolic pressure
across two groups: individuals without Sickle Cell Disease (SCD) and
those with SCD. The central line in each box represents the median value
of systolic pressure for each group. The boxes themselves represent the
interquartile range (IQR), containing the middle 50% of the data. The
whiskers show the range of the data, excluding outliers. It seems the
overall range of systolic pressure is slightly wider in the "Yes SCD"
group, and there are a few outliers in both groups.

Is this expected?

Generally, if SCD affects the blood in a way that could influence blood
pressure, we might expect to see differences in the distribution between
the two groups. The fact that the medians are similar suggests that SCD
might not have a strong impact on the median systolic pressure, but the
wider variability in the "Yes SCD" group could indicate that SCD has
other effects on blood pressure that might not be captured by the median
alone.

For example, SCD can lead to complications like anemia, which could
affect blood pressure. Also, treatment for SCD might involve medications
that affect blood pressure. Without more context, it's hard to say
whether this plot matches expectations. It would be important to compare
these results with medical literature or clinical expectations.
:::

### Question 4: Body Mass Index (BMI) Analysis

1.  Create a new variable `BMI` for each subject and add it to the data.
2.  Generate a histogram to describe the distribution of BMI. Assess if
    it appears normally distributed.
3.  Calculate descriptive statistics for BMI: mean, standard deviation
    (sd), minimum, maximum, first and third quartile, and median.
4.  Discuss which of these statistics you would prefer to report in a
    publication and why.
5.  Categorize subjects into BMI groups as per WHO standards and count
    how many subjects fall into each category.
6.  Create a barplot to visualize the frequencies of each BMI group.
7.  Compare the BMI distribution between subjects with and without SCD
    using barplots. Note any differences observed.

::: {.callout-tip collapse="true" title="Question 4: Click to see solution"}
1.  Creating a New BMI Variable

```{r}
# Adding a new BMI variable to the data frame
d <- d %>%
  mutate(BMI = weight/(height/100)^2) # Height is converted to meters by dividing by 100
```

2.  Generating a Histogram

```{r}
# Generating a histogram for BMI distribution
ggplot(d, aes(x=BMI)) + # Initialize a ggplot with 'd' as the data and 'BMI' as the x-axis aesthetic
  geom_histogram(binwidth=1, color="black", fill="white") + # Add a histogram layer with bins of width 1, black borders, and white fill color
  xlab("Body Mass Index") + # Set the label for the x-axis as "Body Mass Index"
  ggtitle("Distribution of Body Mass Index") # Add a title to the plot
```

3.  Calculating Descriptive Statistics

```{r}
# Calculating descriptive statistics for BMI
d %>%
  summarise(
    mean_BMI = mean(BMI), # Calculate the mean (average) of the BMI values
    sd_BMI = sd(BMI), # Calculate the standard deviation of the BMI values, which measures the amount of variation
    min_BMI = min(BMI), # Find the minimum BMI value in the dataset
    max_BMI = max(BMI), # Find the maximum BMI value in the dataset
    first_quartile = quantile(BMI, 0.25), # Calculate the first quartile (25th percentile) of the BMI values
    third_quartile = quantile(BMI, 0.75), # Calculate the third quartile (75th percentile) of the BMI values
    median_BMI = median(BMI) # Calculate the median (50th percentile) of the BMI values
  )
```

4.  Preferred Statistics for Publication

```{r}
# Categorizing subjects into BMI groups as per WHO standards
d <- d %>%
  mutate(
    # The cut function is used to divide the BMI into categories based on WHO standards.
    BMIgroup = cut(
      BMI, # The numeric vector to be divided into intervals
      breaks = c(0, 18.5, 25, 30, Inf), # Define the boundaries of the categories
      labels = c("Underweight", "Normal", "Overweight", "Obese"), # Labels for the categories
      include.lowest = TRUE, # Include the lowest value in the first interval
      right = FALSE # Intervals are left-closed (i.e., the right endpoint is not included)
    )
  )
```

5.  Counting Subjects in Each BMI Category

```{r}
# Counting the number of subjects in each BMI category
d %>%
  count(BMIgroup)
```

6.  Creating a Bar Plot for BMI Categories

```{r}
# Creating a bar plot for the frequencies of each BMI group
# Use the ggplot function to initialize the plot with the data frame 'd' and
# set the aesthetic mappings.
ggplot(d, aes(x=BMIgroup)) + # Initialize the plot specifying 'd' as the data and 'BMIgroup' as the x-axis
  geom_bar( # Add bars to represent counts of each BMI category
    color="black", # Outline color of the bars
    fill="grey", # Fill color of the bars
    show.legend = FALSE # Do not show legend for bars
  ) +
  geom_text( # Add text labels on top of the bars
    stat = 'count', # Use the count statistic to determine the number to display
    aes(label = ..count..), # Use the computed count for the label
    vjust = -0.5, # Vertically adjust text to be just above the bars
    color = "black" # Text color
  ) +
  xlab("Body Mass Index Categories") + # Customize the x-axis label
  ylab("Number of Subjects") + # Customize the y-axis label
  ggtitle("Frequencies of BMI Categories") # Add a title to the plot
```

7.  Comparing the BMI Distribution by SCD Status

```{r}
# Comparing the BMI distribution between subjects with and without SCD
d %>%
  group_by(scd, BMIgroup) %>%
  summarise(count = n()) %>%
  mutate(proportion = count / sum(count)) %>%
  ggplot(aes(x=BMIgroup, y=proportion, fill=as.factor(scd))) +
  geom_bar(stat="identity", position="dodge") +
  scale_fill_manual(values=c("blue", "red"), labels=c("No SCD", "Yes SCD")) +
  xlab("Body Mass Index Categories") +
  ylab("Proportion of Subjects") +
  ggtitle("BMI Distribution by SCD Status")
```
:::

### Question 5: Diastolic Pressure Analysis

1.  Calculate the mean and standard deviation of diastolic pressure
    among subjects with and without SCD.
2.  Determine the 95% confidence interval of the mean diastolic pressure
    for both groups. Discuss your conclusions.
3.  Calculate the 95% prediction intervals for diastolic pressure in
    both groups. Provide an interpretation in plain English.

### Question 6: Systolic Pressure and Gender Analysis

1.  Compute the 95% confidence interval of the mean systolic pressure
    for subjects with and without SCD.
2.  Perform the same analysis for female subjects only.
3.  Use a dotplot (stripchart) to visually compare systolic pressure
    among females with and without SCD.
4.  Generate QQplots to assess the normal distribution of systolic
    pressure among females in both groups.
5.  Count how many observations come from women with and without SCD.
6.  Based on the above analyses, evaluate the reliability of the 95%
    confidence intervals.

### Question 7: Mean Arterial Pressure (MAP)

1.  Add a new variable `MAP` to the data, representing Mean Arterial
    Pressure.
2.  Visualize the distribution of MAP among subjects without SCD using a
    histogram and QQplot.
3.  Produce a "Wally plot" for a comparative analysis of MAP
    distribution.
4.  Estimate the normal range of MAP and compare it with established
    standards.

