---
execute: 
  warning: false
---

# Data Frames and Lists {#sec-df-lists}

```{r setup}
#| include: false
#| eval: true
source(here::here("R/functions.R"))
source(here::here("R/load-packages.R"))
```

# Introduction

In this section, we will analyze two datasets related to dietary effects
on milk protein levels in cows and gene expression data for the
alpha-synuclein gene. The analysis includes data cleaning,
visualization, hypothesis testing, and statistical adjustments for
multiple comparisons. The tools used are `dplyr`, `ggplot2`, and
`tidyverse`.

## Analysis of Milk Protein Levels

### Data Preparation

We analyze milk protein levels at specific time points after calving for
cows on different diets. The objective is to assess dietary effects on
protein levels and compute statistical summaries and comparisons.

```{r}
# Create a sample dairy dataset for analysis
milk_data <- tibble::tribble(
  ~Cow, ~Diet, ~Time, ~protein,
  "C1", "lupins", 6, 3.1,
  "C2", "barley+lupins", 6, 3.7,
  "C3", "lupins", 6, 2.9,
  "C4", "barley+lupins", 6, 3.4,
  "C5", "lupins", 6, 3.0,
  "C6", "barley+lupins", 6, 3.8,
  "C7", "lupins", 6, 3.2,
  "C8", "barley+lupins", 6, 3.5,
  "C9", "lupins", 6, 3.3,
  "C10", "barley+lupins", 6, 3.6
) %>%
  mutate(Diet = factor(Diet))

# Data summary
milk_data %>% summary()

```

##Exploratory Data Analysis

###Visualization of Protein Levels by Diet

```{r}
ggplot(milk_data, aes(x = Diet, y = protein, color = Diet)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.6) +
  labs(title = "Protein Levels by Diet", x = "Diet", y = "Protein Level (%)") +
  theme_minimal()
```

```{r}
milk_summary <- milk_data %>%
  group_by(Diet) %>%
  summarise(
    mean_protein = mean(protein, na.rm = TRUE),
    sd_protein = sd(protein, na.rm = TRUE),
    n = n(),
    ci_lower = mean_protein - qt(0.975, df = n - 1) * sd_protein / sqrt(n),
    ci_upper = mean_protein + qt(0.975, df = n - 1) * sd_protein / sqrt(n)
  )
milk_summary

```

## Hypothesis Testing

### We test if there is a significant difference in protein levels between diets.

```{r}
t_test_results <- t.test(protein ~ Diet, data = milk_data)
t_test_results
```

```{r}
ggplot(milk_data, aes(x = protein, fill = Diet)) +
  geom_histogram(alpha = 0.7, position = "identity", bins = 10) +
  labs(title = "Distribution of Protein Levels by Diet", x = "Protein Level (%)", y = "Frequency") +
  scale_fill_manual(values = c("lupins" = "blue", "barley+lupins" = "green")) +
  theme_minimal()

```

```{r}
# Adjust for multiple comparisons
p_adjusted <- p.adjust(t_test_results$p.value, method = "bonferroni")
p_adjusted

```
