# Exploring Clinical and Medical Data with R {#sec-explore-clinical}

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(patchwork)
library(survival)
library(gtsummary)
library(skimr)
```

## When Do You Use This? {.callout-tip}

> You download a clinical dataset for the first time. Before running a single test you need to understand what you have: how many patients, which variables, how much missing data, whether distributions look plausible, and whether there are obvious data quality issues. This session teaches that systematic exploration process.

## Learning Objectives

After completing this session you will be able to:

- Load and inspect a clinical dataset in R
- Identify missing values, implausible values, and data quality issues
- Produce a descriptive summary table appropriate for Table 1 in a paper
- Visualize distributions, group differences, and dose-response relationships
- Apply log transformation to right-skewed lab values and explain why

---

## Background: Why Explore Before You Analyse?

Exploratory data analysis (EDA) is not optional — it is the first analytic step for every dataset. Rushing to a regression or a t-test without first understanding the data leads to errors that are hard to detect later.

The systematic EDA workflow is:

1. **Structure** — how many rows, columns, variable types?
2. **Missing data** — which variables have gaps, and do they appear at random?
3. **Distributions** — continuous variables: symmetric or skewed? Categorical: any rare levels?
4. **Plausibility** — are any values biologically implausible?
5. **Group comparisons** — do key variables differ across groups as expected?

---

## Worked Example 1: Primary Biliary Cholangitis (PBC)

The `pbc` dataset records 418 patients from a randomised trial of a treatment for primary biliary cholangitis (PBC), a chronic liver disease. It is built into the `survival` package.

### Load and Inspect

```{r pbc-load}
pbc <- survival::pbc |>
  as_tibble() |>
  janitor::clean_names() |>
  mutate(
    sex   = factor(sex, levels = c("f", "m"), labels = c("Female", "Male")),
    trt   = factor(trt,   levels = c(1, 2),    labels = c("D-penicillamine", "Placebo")),
    stage = factor(stage, levels = 1:4,         labels = paste("Stage", 1:4))
  )
```

```{r pbc-glimpse}
glimpse(pbc)
```

`glimpse()` shows the number of rows and columns, and the first few values of every variable. Use it before doing anything else.

### Missing Data

```{r pbc-missing}
pbc |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  arrange(desc(n_missing))
```

Several variables have missing data. Note that `trt` is `NA` for patients who were not randomised — they are valid observational cases but should be excluded from the treatment comparison.

### Summary Table (Table 1)

```{r pbc-table1}
pbc |>
  filter(!is.na(trt)) |>
  select(trt, age, sex, stage, bili, albumin, protime, alk_phos) |>
  tbl_summary(
    by       = trt,
    missing  = "no",
    statistic = list(all_continuous() ~ "{median} ({p25}, {p75})")
  ) |>
  add_overall() |>
  add_p() |>
  bold_labels()
```

::: callout-note
**Reading Table 1:**
Continuous variables are summarised as median (IQR) because most are right-skewed. Categorical variables are shown as count (%). The p-values compare the two treatment groups — we would expect them to be non-significant for baseline variables in a properly randomised trial.
:::

### Distributions of Key Lab Values

```{r pbc-distributions}
pbc |>
  select(bili, albumin, protime, alk_phos) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~ variable, scales = "free") +
  labs(title = "Distributions of PBC lab values", x = NULL, y = "Count") +
  theme_bw()
```

All four lab values are right-skewed. This is typical for biochemical markers — a few patients have very high values. When you run models with these variables, consider log-transforming them.

### Log Transformation

Right-skewed distributions violate the normality assumption required by many statistical tests. Log transformation compresses the long right tail.

```{r pbc-log-transform}
p_raw <- ggplot(pbc, aes(x = bili)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(title = "Raw bilirubin", x = "mg/dL", y = "Count") +
  theme_bw()

p_log <- ggplot(pbc, aes(x = log(bili))) +
  geom_histogram(bins = 30, fill = "tomato", color = "white") +
  labs(title = "Log-transformed bilirubin", x = "log(mg/dL)", y = "Count") +
  theme_bw()

wrap_plots(p_raw, p_log)
```

After log transformation, the distribution is approximately symmetric. When you report results from a model fitted on log-bilirubin, back-transform to the original scale using `exp()` for clinical interpretability.

```{r pbc-log-summary}
pbc |>
  summarise(
    mean_raw    = mean(bili, na.rm = TRUE),
    median_raw  = median(bili, na.rm = TRUE),
    mean_log    = mean(log(bili), na.rm = TRUE),
    back_transf = exp(mean(log(bili), na.rm = TRUE))
  )
```

The `back_transf` value is the **geometric mean** — the appropriate average for log-normal data.

### Group Comparison by Stage

```{r pbc-stage}
pbc |>
  filter(!is.na(stage)) |>
  ggplot(aes(x = stage, y = log(bili), fill = stage)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
  scale_fill_brewer(palette = "Blues") +
  labs(
    title = "Log-bilirubin by disease stage",
    x     = NULL,
    y     = "Log bilirubin (log mg/dL)"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

Bilirubin clearly increases with disease stage — a sensible finding that validates the data.

---

## Worked Example 2: Vitamin C and Tooth Growth (Animal Experiment)

The `ToothGrowth` dataset records the effect of Vitamin C dose and delivery method on the length of odontoblasts (tooth growth cells) in 60 guinea pigs. It is built into base R.

```{r toothgrowth-load}
tg <- ToothGrowth |>
  as_tibble() |>
  mutate(
    dose = factor(dose, levels = c(0.5, 1.0, 2.0)),
    supp = factor(supp, levels = c("VC", "OJ"),
                  labels = c("Ascorbic acid", "Orange juice"))
  )
```

### Structure and Summary

```{r toothgrowth-skim}
skim(tg)
```

10 observations per dose-supplement combination — a balanced 3 x 2 factorial design.

### Dose-Response Relationship

```{r toothgrowth-dose-response}
tg |>
  ggplot(aes(x = dose, y = len, fill = supp)) +
  geom_boxplot(position = position_dodge(0.8), alpha = 0.7) +
  geom_jitter(aes(group = supp), position = position_jitterdodge(jitter.width = 0.1),
              alpha = 0.5, size = 2) +
  scale_fill_manual(values = c("Ascorbic acid" = "steelblue", "Orange juice" = "tomato")) +
  labs(
    title = "Effect of Vitamin C on tooth growth in guinea pigs",
    x     = "Dose (mg/day)",
    y     = "Odontoblast length (μm)",
    fill  = "Supplement"
  ) +
  theme_bw()
```

### Interpreting This Figure

- Tooth growth increases with dose for both supplement types.
- At lower doses (0.5 and 1.0 mg/day), orange juice produces more growth than ascorbic acid.
- At the highest dose (2.0 mg/day), the two supplements are approximately equivalent.
- This is an interaction — the effect of supplement type depends on dose level.

---

## Worked Example 3: Melanoma Survival Data

The `boot::melanoma` dataset contains data on 205 patients with malignant melanoma who had a tumour removed at Odense University Hospital between 1962 and 1977.

```{r melanoma-load}
# boot::melanoma; if boot is not installed, survival::melanoma is an alternative
melanoma <- boot::melanoma |>
  as_tibble() |>
  mutate(
    status    = factor(status, levels = c(2, 1, 3),
                       labels = c("Alive", "Died from melanoma", "Other death")),
    ulceration = factor(ulcer, levels = c(0, 1), labels = c("Absent", "Present"))
  )
```

### Tumour Thickness Distribution

```{r melanoma-thickness}
p1 <- ggplot(melanoma, aes(x = thickness)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "white") +
  labs(title = "Raw tumour thickness", x = "Thickness (1/100 mm)", y = "Count") +
  theme_bw()

p2 <- ggplot(melanoma, aes(x = log(thickness))) +
  geom_histogram(bins = 25, fill = "tomato", color = "white") +
  labs(title = "Log-transformed thickness", x = "log(thickness)", y = "Count") +
  theme_bw()

wrap_plots(p1, p2)
```

### Thickness by Ulceration Status

```{r melanoma-ulceration}
melanoma |>
  ggplot(aes(x = ulceration, y = thickness, fill = ulceration)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.4) +
  scale_fill_manual(values = c("Absent" = "steelblue", "Present" = "tomato")) +
  labs(
    title = "Tumour thickness by ulceration status",
    x     = NULL,
    y     = "Tumour thickness (1/100 mm)"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

Ulcerated tumours tend to be thicker — consistent with the known relationship between tumour thickness (Breslow depth) and aggressiveness.

---

## Where to Find Data Like This {.callout-note}

**Built-in R datasets used in this session:**
- `survival::pbc` — primary biliary cholangitis clinical trial, 418 patients
- `boot::melanoma` — melanoma surgery outcomes, 205 patients
- `datasets::ToothGrowth` — guinea pig tooth growth factorial experiment

**Other clinical datasets in the `survival` package:**
`lung`, `veteran`, `colon`, `diabetic`, `udca` — all real datasets from clinical trials and registries.

**NHANES (National Health and Nutrition Examination Survey):**
```{r}
#| eval: false
install.packages("nhanesA")
library(nhanesA)
nhanes("DEMO_J")   # demographic data from 2017-2018 cycle
```

**Mouse Phenome Database:** phenome.jax.org — download CSV files of mouse phenotype data directly into R.

---

## What Can Go Wrong {.callout-warning}

**Removing rows with `na.omit()` without thinking**
`na.omit()` drops any row with *any* missing value. In a dataset with many columns, this can discard most of the data. Instead, use `filter(!is.na(variable_of_interest))` to drop only rows where the key variable is missing.

**Treating ordered categories as numbers**
Disease stage (1, 2, 3, 4) looks like a number, but treating it as continuous implies equal spacing between stages. Use `factor(stage)` unless you have a specific reason to model it as numeric.

**Plotting raw lab values without checking for outliers first**
One extreme outlier can collapse the entire axis scale. Always check the range with `range(x, na.rm = TRUE)` before plotting, and consider whether very extreme values are plausible or data entry errors.

**Forgetting to back-transform**
If you fit a model on log(bilirubin), your coefficients are on the log scale. Report `exp(estimate)` with `exp(CI)` when writing up. The units are now geometric means and ratios, not raw differences.

---

## Exercises

### Exercise 1 (Guided): Full EDA of PBC Dataset

Using `survival::pbc`:

1. How many patients are in the dataset? How many were randomised?
2. Which variables have more than 10% missing data?
3. Create a histogram and boxplot for `albumin`. Does it look normally distributed?
4. Create a Table 1 comparing `albumin`, `bili`, and `protime` by treatment group.

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r}
# 1. Patient counts
nrow(pbc)
sum(!is.na(pbc$trt))   # randomised patients

# 2. Missing > 10%
pbc |>
  summarise(across(everything(), ~ mean(is.na(.)))) |>
  pivot_longer(everything(), names_to = "var", values_to = "pct_missing") |>
  filter(pct_missing > 0.10) |>
  mutate(pct_missing = scales::percent(pct_missing, 0.1))

# 3. Albumin distribution
ggplot(pbc, aes(x = albumin)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "white") +
  labs(x = "Albumin (g/dL)", y = "Count") + theme_bw()

# 4. Table 1
pbc |>
  filter(!is.na(trt)) |>
  select(trt, albumin, bili, protime) |>
  tbl_summary(by = trt,
              statistic = list(all_continuous() ~ "{median} ({p25}, {p75})")) |>
  add_p() |> bold_labels()
```
:::

### Exercise 2 (Semi-guided): ToothGrowth Exploration

Using `ToothGrowth`:

1. How many observations are there per dose-supplement combination?
2. Calculate the mean and SD of `len` for each group.
3. Create a line plot showing mean length ± SD by dose, with separate lines for each supplement.
4. Write one sentence describing the interaction you see.

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r}
tg <- ToothGrowth |> as_tibble() |>
  mutate(dose = factor(dose), supp = factor(supp, labels = c("Ascorbic acid", "OJ")))

# 1. Counts
count(tg, dose, supp)

# 2. Summary statistics
tg |>
  group_by(dose, supp) |>
  summarise(mean_len = mean(len), sd_len = sd(len), .groups = "drop")

# 3. Line plot
tg |>
  group_by(dose, supp) |>
  summarise(mean = mean(len), sd = sd(len), .groups = "drop") |>
  ggplot(aes(x = dose, y = mean, color = supp, group = supp)) +
  geom_line(size = 1) +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.1) +
  geom_point(size = 3) +
  scale_color_manual(values = c("steelblue", "tomato")) +
  labs(title = "Mean tooth length ± SD by dose and supplement",
       x = "Dose (mg/day)", y = "Length (μm)", color = "Supplement") +
  theme_bw()
```
:::

### Exercise 3 (Open-ended)

Download a dataset from your own research or from NHANES. Perform a complete EDA: (1) inspect structure and missing data, (2) produce a Table 1 with `gtsummary`, (3) create at least two visualisations, and (4) write a half-page data report as you would for a Methods section.

---

## Comprehension Check

1. You run `skim(pbc)` and see `n_missing = 106` for the `trt` variable. What is the most likely explanation, and what should you do before running a treatment comparison?
2. A histogram of albumin values shows a roughly symmetric, bell-shaped distribution. A histogram of bilirubin values shows a long right tail. How should you handle these differently when fitting a linear model?
3. Why is the geometric mean (back-transformed from log scale) more appropriate than the arithmetic mean for a right-skewed variable like bilirubin?
4. You have 10 animals per group in a 2 × 3 factorial experiment (2 supplements × 3 dose levels). How many observations total? What does "balanced design" mean?
5. A colleague creates a Table 1 that shows means and standard deviations for all continuous variables, including bilirubin (which is heavily right-skewed). What is wrong with this, and what should they report instead?

::: {.callout-note collapse="true"}
### Answers
1. The 106 missing values in `trt` are patients who enrolled in the observational portion of the study and were not randomised. They are valid cases for some analyses but should be excluded with `filter(!is.na(trt))` when comparing treatment arms.
2. Albumin is approximately normally distributed, so it can be entered into a model on its original scale. Bilirubin is right-skewed, so log-transform it before modelling: `log(bili)`. After modelling, back-transform coefficients and confidence intervals with `exp()`.
3. The arithmetic mean is pulled upward by extreme high values in a right-skewed distribution, so it does not represent the "typical" patient. The geometric mean (exp(mean(log(x)))) is the value that, when raised to the power *n*, equals the product of all observations — it is a more robust central measure for log-normal data.
4. 10 × 2 × 3 = 60 total observations. A balanced design has the same number of observations in every cell of the design, which simplifies analysis and maximises statistical power.
5. For a right-skewed variable, the mean and SD are misleading because extreme outliers inflate both. The correct summary is median and interquartile range (IQR), as the median is resistant to extreme values.
:::

## Further Reading

- [R for Data Science, Chapter 11: Exploratory Data Analysis](https://r4ds.hadley.nz/eda) --- Wickham & Grolemund
- [gtsummary documentation](https://www.danieldsjoberg.com/gtsummary/) --- Table 1 and beyond
- [skimr documentation](https://docs.ropensci.org/skimr/) --- quick dataset summaries
