---
title: "Help and Documentation"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
```

## When Do You Use This? {#sec-help-docs .callout-tip}

> You are halfway through an analysis and get an error message you don't understand. Or you can't remember the exact arguments for a function. This session is your reference guide: how to get help inside R, how to read error messages, how to find the right function for a task, and where to turn when you're stuck.

## Learning Objectives

After completing this session you will be able to:

- Use R's built-in help system to look up any function
- Interpret common R error and warning messages
- Search for packages and functions you don't know the name of
- Find the right online resources for R and statistics questions
- Locate the relevant session in this course for any statistical method


## Getting Help Inside R

### The `?` Operator and `help()`

```{r help-basics}
#| eval: false
# Access help for a function you know the name of
?mean
help(mean)

# Access help for an operator
?`+`

# Access help for a whole package
help(package = "survival")
```

### Help When You Don't Know the Name

```{r help-search}
#| eval: false
# Fuzzy search: returns all help pages matching the keyword
??regression
help.search("regression")

# Find a function by name fragment
apropos("glm")
```

### Reading a Help Page

Every R help page has the same structure:

| Section | What it tells you |
|---------|------------------|
| **Description** | What the function does in one sentence |
| **Usage** | Function signature — all arguments with defaults |
| **Arguments** | What each argument expects |
| **Value** | What the function returns |
| **Examples** | Copy-paste these first to understand the function |
| **See Also** | Related functions — often more useful than the main function |

**Tip:** Jump straight to **Examples** when learning a new function. Run them interactively to see what the function does before applying it to your own data.


## Common Error Messages

### Object Not Found

```r
Error in plot(my_data) : object 'my_data' not found
```

**Cause:** The object does not exist in your environment. You may have forgotten to run the line that creates it, or you misspelled the name. R is case-sensitive: `mydata` ? `MyData`.

**Fix:** Check `ls()` to list all objects in your environment. Re-run the code that creates the object.


### Unexpected Symbol / Unexpected Input

```r
Error: unexpected ')' in "mean(x, na.rm = TRUE))"
```

**Cause:** A syntax error — mismatched brackets, a missing comma, or an unclosed string.

**Fix:** Count your opening and closing brackets. In RStudio, click inside the bracket to see its matching partner highlighted.


### Package Not Found

```r
Error in library(ggplot2) : there is no package called 'ggplot2'
```

**Cause:** The package is not installed.

**Fix:**

```{r install-fix}
#| eval: false
install.packages("ggplot2")   # install once
library(ggplot2)               # load every session
```


### NA / NaN Warnings

```r
Warning message: NAs produced by coercion
```

**Cause:** R cannot convert a non-numeric value (e.g., `"N/A"`, `"."`, or `""`) to a number. These become `NA`.

**Fix:** Check what non-numeric strings are in your column before coercing:

```{r na-debug}
x <- c("1", "2", "N/A", "4")
which(is.na(as.numeric(x)))   # which elements became NA?
x[is.na(as.numeric(x))]       # what are they?
```


### Subscript Out of Bounds

```r
Error in x[[5]] : subscript out of bounds
```

**Cause:** You are trying to access an element that does not exist (e.g., the 5th column of a data frame that only has 4 columns).

**Fix:** Check dimensions: `dim(df)`, `length(x)`, `nrow(df)`, `ncol(df)`.


### Factor Level Mismatch

```r
Error in model.frame.default : factor has new levels: XYZ
```

**Cause:** You are predicting with new data that contains a factor level not seen during model fitting.

**Fix:** Ensure new data factor levels match training data. Use `droplevels()` or re-encode factor levels consistently.


## Warnings vs Errors

| Message type | Meaning | Action needed? |
|-------------|---------|---------------|
| `Error` | R stopped — no result produced | Yes — must fix |
| `Warning` | R produced a result, but something may be wrong | Check — often important |
| `Message` | Informational only (e.g., from packages loading) | No — usually safe to ignore |

**Key warning to never ignore:**

```r
Warning message: In glm.fit() : fitted probabilities numerically 0 or 1 occurred
```

This means your logistic regression has complete separation — a predictor perfectly separates outcomes. The model is unreliable; see @sec-logistic-regression.


## Useful Functions for Debugging

```{r debug-functions}
#| eval: false
# What is in my environment?
ls()

# What class is this object?
class(my_object)
typeof(my_object)

# What is the structure?
str(my_data)
glimpse(my_data)

# Check for missing values
sum(is.na(my_data))
colSums(is.na(my_data))

# View first/last rows
head(my_data)
tail(my_data)

# Check column names and types
names(my_data)
sapply(my_data, class)
```


## Online Resources

### Finding Answers

| Resource | Best for | URL |
|----------|---------|-----|
| **Stack Overflow** | Specific error messages and code questions | [stackoverflow.com/questions/tagged/r](https://stackoverflow.com/questions/tagged/r) |
| **RDocumentation** | Searchable R function and package documentation | [rdocumentation.org](https://www.rdocumentation.org) |
| **CRAN Task Views** | Finding the right package for a statistical task | [cran.r-project.org/web/views/](https://cran.r-project.org/web/views/) |
| **Posit Community** | Tidyverse and RStudio questions | [community.rstudio.com](https://community.rstudio.com) |
| **R-bloggers** | Tutorials and use cases | [r-bloggers.com](https://www.r-bloggers.com) |

### Asking Good Questions

When posting to Stack Overflow or asking a colleague, always provide a **minimal reproducible example (reprex)**:

```{r reprex}
#| eval: false
library(reprex)
reprex({
  x <- c(1, 2, NA, 4)
  mean(x)               # what did you get?
  mean(x, na.rm = TRUE) # what do you expect?
})
```

A reprex is a small, self-contained snippet of code that demonstrates the problem. Remove anything not essential to reproducing the error.


## Course Method Map

The table below maps every statistical method covered in this course to its session. Use this to quickly navigate when you need to recall a method.

| Method | Session |
|--------|---------|
| Setting up R and packages | @sec-packages-libraries |
| Vectors and basic data types | @sec-vectors |
| Data wrangling (dplyr, tidyr) | @sec-tidy-data-wrangling |
| Descriptive statistics | @sec-descriptive-statistics |
| Basic visualisation (ggplot2) | @sec-basic-visualization |
| Advanced visualisation | @sec-advanced-viz |
| Exploring clinical data | @sec-explore-clinical |
| Sampling and estimation | @sec-sampling |
| Hypothesis testing framework | @sec-hypothesis-testing |
| t-tests | @sec-t-tests |
| Non-parametric tests | @sec-nonparametric |
| ANOVA | @sec-anova |
| Categorical data and chi-square | @sec-categorical |
| Power and sample size | @sec-power |
| Correlation and association | @sec-correlation |
| Linear regression | @sec-linear-regression |
| Multiple regression | @sec-multiple-regression |
| Logistic regression | @sec-logistic-regression |
| Model building and diagnostics | @sec-model-building |
| Missing data imputation | @sec-imputation |
| Mixed models | @sec-mixed-models |
| Survival analysis | @sec-survival |
| Causal inference and DAGs | @sec-causal-inference |
| Mendelian randomisation | @sec-mr |
| Statistics for omics | @sec-omics |
| Statistical decision guide | @sec-statistical-decision-guide |
| Clinical research methods | @sec-clinical-methods |


## Comprehension Check

1. You run `lm(y ~ x, data = mydata)` and get `Error: object 'y' not found`. What are two likely causes?
2. What is the difference between an R error and a warning? Give an example of a warning you should not ignore.
3. You want to do a survival analysis but don't know which function to use. How would you search for relevant functions in R?
4. A colleague's code throws `Error in model.frame : factor has new levels`. What does this mean, and how would you fix it?
5. You get a message `Warning: 152 rows removed due to missing values`. Should you be concerned? What would you check?

::: {.callout-note collapse="true"}
### Answers
1. (a) The column `y` does not exist in `mydata` — check column names with `names(mydata)`. A common cause is a typo or capitalisation mismatch (R is case-sensitive: `Y` ? `y`). (b) The object `mydata` itself does not exist in the environment — run `ls()` to check, and re-run the code that creates `mydata`.
2. An **error** stops execution and produces no output — you must fix it before R will continue. A **warning** means R completed the computation but flagged something potentially problematic. Warnings you should not ignore include: fitted probabilities of 0 or 1 in logistic regression (complete separation); NAs introduced by coercion (unexpected missing data); rank-deficient fits (collinearity in regression). A warning you can usually ignore: `package 'X' was built under R version Y.Z` — just means the package was built with a slightly different R version.
3. Two approaches: (1) `help.search("survival")` or `??survival` searches all help pages for the keyword; (2) Look at the **CRAN Task View for Survival Analysis** at [https://cran.r-project.org/web/views/Survival.html](https://cran.r-project.org/web/views/Survival.html) — this lists all packages and key functions for survival analysis. The `survival` package and its `Surv()`, `survfit()`, `coxph()` functions are the standard starting point.
4. The model was trained on data where a factor had certain levels (e.g., `race`: "White", "Black"), but the new data contains a level not seen during training (e.g., "Other"). R cannot encode this new level. Fix: ensure all factor levels in new data are present in the training data — use `levels(new_data$race) <- levels(training_data$race)` or use consistent factor encoding before splitting into train/test sets.
5. Yes, be concerned. 152 missing values could mean: (a) a column has systematic missingness — check `colSums(is.na(df))` to identify which variable; (b) the missing data is not missing at random (MCAR) — if missingness is related to the outcome or exposure, complete-case analysis will be biased. Steps: identify which variables have missing data, check whether missingness is associated with the outcome, and consider whether imputation is needed (see @sec-imputation).
:::

## Common Pitfalls When Getting Help

::: {.callout-warning}
**Posting an incomplete question.** The most common reason a Stack Overflow question goes unanswered is that it lacks a reproducible example. Always include: the exact error message, the R version and key package versions, and a minimal dataset that demonstrates the problem (use `dput()` or a built-in dataset).

**Searching for the wrong thing.** Error messages in R are often generic. Search for the *function name* and *package name* alongside the error text, not the error alone. For example: "lme4 singular fit random effects" is a better search than "model failed to converge."

**Copying code without understanding it.** If you paste a solution and it works, spend 60 seconds asking why. Change one thing to confirm your mental model. Copy-paste dependency is the main reason R skills don't compound.

**Not checking the package version.** Many Stack Overflow answers are outdated. Always confirm that the answer applies to your version (`packageVersion("pkg")`). Behaviour changes across versions --- especially for `dplyr`, `ggplot2`, and `tidymodels`.

**Skipping `?function_name` before asking.** The help page often contains a working example that answers the question. Always check the Examples section at the bottom of the help page first.
:::

## Further Reading

- @Grolemund-2015 --- *Hands-On Programming with R* — getting started, debugging, and help
- @Wickham-2019 --- *Advanced R* — chapter on debugging and performance: [adv-r.hadley.nz](https://adv-r.hadley.nz/)
- [Happy Git and GitHub for the useR](https://happygitwithr.com/) — version control with R
- [R for Data Science (2nd ed.)](https://r4ds.hadley.nz/) — canonical reference for the tidyverse workflow
