---
title: "Hypothesis Testing: Concepts and Framework"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(survival)
```

## When Do You Use This? {#sec-hypothesis-testing}

::: {.callout-tip}
You have data from two groups of patients and you want to know whether
they genuinely differ, or whether the difference you observe could
simply be due to the luck of which patients ended up in each group. Or
you have a new treatment and you want to know whether its effect is real
or just noise. Hypothesis testing gives you a principled, reproducible
framework for making these decisions from data, while keeping the rate
of false positives under control.

This session builds directly on the previous one. In [Sampling,
Estimation, and Confidence Intervals](sampling-and-estimation.qmd) you
learned how to estimate a population quantity and quantify your
uncertainty. Hypothesis testing is the next step: instead of just
estimating, you are now making a yes/no decision --- is the effect real?
:::

## Learning Objectives

After completing this session you will be able to:

- State null and alternative hypotheses for a comparison
- Explain what a p-value means --- and what it does not mean
- Describe Type I and Type II errors and how they relate to $\alpha$ and
  $\beta$
- Explain statistical power and the factors that determine it
- Apply multiple testing corrections and explain when each is
  appropriate
- Relate hypothesis tests to confidence intervals


## Background: The Logic of Hypothesis Testing

### The criminal trial analogy

Before any formulas, here is the underlying logic in a context you
already understand.

In a criminal trial, the defendant is presumed innocent until proven
guilty. The court does not ask "is this person guilty?" --- it asks "is
the evidence against an innocent person this damning, or more so?" If
the answer is yes (the evidence would be very unlikely if the person
were truly innocent), you reject the presumption of innocence and
convict.

Hypothesis testing follows the exact same logic:

- The **null hypothesis** ($H_0$) is the default position --- usually
  "there is no effect" or "the groups are the same." This is the
  presumption of innocence.
- You do not ask "is the effect real?" You ask: "How likely is it that
  I would observe data *this extreme or more extreme* if the null
  hypothesis were actually true?"
- If that probability is very small (below your threshold $\alpha$),
  you reject the null hypothesis.

That probability is the **p-value**. And just as a conviction does not
prove guilt beyond all doubt --- only beyond a reasonable doubt --- a
small p-value does not prove your hypothesis is true. It only says the
data are sufficiently inconsistent with the null.

### The five steps

Every hypothesis test follows the same sequence:

1. **State hypotheses**: define $H_0$ (null) and $H_1$ (alternative)
   before looking at the data.
2. **Explore the data**: plot first. Always. You may find obvious
   violations of assumptions, unexpected outliers, or a result that
   clearly does not need a test.
3. **Check assumptions and choose the test**: select the test
   appropriate for your data type, design, and sample size.
4. **Compute the test statistic and p-value**: run the test in R, read
   the output line by line.
5. **Interpret**: report the test statistic, p-value, effect size, and
   confidence interval. State your conclusion in plain language.

### What a p-value is --- and is not

::: {.callout-warning}
### The most important thing to understand about p-values

**What a p-value IS:**
The probability of observing a test statistic as extreme as (or more
extreme than) the one you got, *assuming the null hypothesis is true*.

**What a p-value is NOT:**

- The probability the null hypothesis is true
- The probability your result is due to chance
- A measure of the size or importance of the effect
- A pass/fail grade for your study

The most common misstatement in science is: "p = 0.03 means there is a
3% chance the result is a fluke." This is wrong. The p-value is
computed *under the assumption* that the null is true. It cannot tell
you the probability that the null is true --- that requires Bayesian
reasoning.

**The correct statement:** "If the null hypothesis were true, the
probability of observing a test statistic this extreme or more extreme
is 3%."
:::

A large p-value (say, p = 0.6) does not mean there is no effect. It
means your data are consistent with the null hypothesis. There may be
a real effect that your study was too small to detect. This is why
"p > 0.05" and "no effect" are not the same thing.

### Type I and Type II errors

Every decision rule makes errors. Here are the two kinds:

|  | $H_0$ is true | $H_0$ is false |
|---|---|---|
| **Reject $H_0$** | **Type I error** (false positive) --- probability $= \alpha$ | Correct (true positive) --- probability $= \text{power}$ |
| **Fail to reject $H_0$** | Correct (true negative) | **Type II error** (false negative) --- probability $= \beta$ |

**Power** $= 1 - \beta$ is the probability of detecting a real effect
of a given size. A study with 80% power has a 20% chance of missing a
real effect.

Think of a smoke alarm. A Type I error is the alarm going off when
there is no fire (someone burnt toast). A Type II error is the alarm
staying silent when there actually is a fire. Lowering the
sensitivity of the alarm ($\alpha$) reduces false positives but
increases missed fires.

Power increases when:

- Sample size increases --- the most reliable lever
- Effect size is larger --- big effects are easier to detect
- Variability is smaller --- cleaner measurements help
- $\alpha$ is larger --- but this increases the Type I error rate

### The connection to confidence intervals

A 95% confidence interval and a two-sided hypothesis test at $\alpha =
0.05$ are two ways of saying the same thing:

- If the 95% CI for a difference **excludes** zero, the test is
  significant at p < 0.05.
- If the 95% CI **includes** zero, the test is not significant.

The CI is usually more informative than the p-value because it tells
you both whether the effect is significant and *how large it plausibly
is*. Report both.


## Example 1: Clinical Data (PBC)

The question: does serum albumin differ between patients who received
D-penicillamine versus those who received placebo?

This is a clinical trial, so the groups were randomised. Before we test
the primary outcome (which comes after treatment), it is standard
practice to check **baseline comparability** --- were the groups similar
on key characteristics before the intervention began? Albumin at
baseline should be similar between groups if the randomisation worked.

Think about this for a moment before running the test: what result do
you *expect* to find when testing baseline albumin in a well-randomised
trial?

```{r pbc-setup}
pbc <- survival::pbc %>%
  as_tibble() %>%
  janitor::clean_names() %>%
  mutate(
    trt = factor(trt, levels = c(1, 2), labels = c("D-penicillamine", "Placebo"))
  )
```

### Step 1: State the hypotheses

$$H_0: \mu_{\text{D-pen}} = \mu_{\text{placebo}}$$
$$H_1: \mu_{\text{D-pen}} \neq \mu_{\text{placebo}}$$

Two-sided, because we have no reason to expect a specific direction. The
groups were randomly assigned.

### Step 2: Explore the data

Always plot before testing. Look at the distribution and ask: are there
obvious outliers? Does the spread look similar between groups?

```{r explore}
pbc %>%
  filter(!is.na(trt), !is.na(albumin)) %>%
  group_by(trt) %>%
  summarise(
    n      = n(),
    mean   = mean(albumin),
    sd     = sd(albumin),
    median = median(albumin)
  )

pbc %>%
  filter(!is.na(trt), !is.na(albumin)) %>%
  ggplot(aes(x = trt, y = albumin, fill = trt)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1.5) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(
    title = "Serum albumin by treatment group (baseline)",
    x     = NULL,
    y     = "Albumin (g/dL)"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

The two groups look nearly identical. The means differ by less than
0.05 g/dL. Before even running the test, the plot strongly suggests a
null result --- and that is exactly what we want to see for a baseline
variable.

### Step 3: Check assumptions and choose the test

The two-sample t-test assumes:

1. Observations are independent (patients are not paired or related)
2. The sampling distribution of the mean is approximately normal (CLT
   holds with ~150 patients per group)
3. Welch's t-test (the R default) relaxes the equal-variance
   assumption, so we do not need to test for that

### Step 4: Perform the test

```{r ttest}
t.test(albumin ~ trt,
       data      = pbc %>% filter(!is.na(trt)),
       var.equal = FALSE)   # Welch's t-test (default; always use this)
```

::: {.callout-note}
### Reading the Output Line by Line

| Output | What it means |
|---|---|
| **t = 0.45** | Test statistic: the difference between means divided by the standard error. A value near zero means the difference is about the size we would expect from random sampling alone. |
| **df = 310** | Welch's degrees of freedom. Not simply n₁ + n₂ − 2 because it accounts for the possibility of unequal variances. |
| **p-value = 0.65** | If there were truly no difference in albumin, we would see a gap this large or larger 65% of the time just from sampling variation. Strong evidence that the groups are similar. |
| **95% CI: −0.09 to 0.14** | The difference in means (D-pen minus placebo) is plausibly between −0.09 and +0.14 g/dL. Because the interval contains zero, the difference is not statistically significant. |
| **Means: 3.52 vs 3.50** | The actual values. The 0.02 g/dL gap is both statistically and clinically negligible. |
:::

### Step 5: Interpret

The result is exactly what we hoped for: no statistically significant
difference in baseline albumin between groups (p = 0.65, 95% CI for
difference: −0.09 to 0.14 g/dL). This confirms that randomisation
produced two comparable groups, which is reassuring for the validity of
the trial.

::: {.callout-tip}
### Writing the result

> "Serum albumin at baseline was similar between the D-penicillamine
> group (mean 3.52 g/dL) and placebo group (mean 3.50 g/dL; Welch's
> t-test: *t*(310) = 0.45, p = 0.65, mean difference 0.02 g/dL, 95%
> CI: −0.09 to 0.14)."

Note: always report the means for both groups, the test statistic with
degrees of freedom, the exact p-value, and the 95% CI for the
difference. "p = 0.65" alone tells the reader almost nothing.
:::


## Example 2: Multiple Testing

When you run many tests at once, the p < 0.05 threshold becomes
dangerous. If you test 20 genes and none are truly differentially
expressed, you still expect $20 \times 0.05 = 1$ false positive by
chance alone. This is not a statistical quirk --- it is a direct
consequence of the definition of $\alpha$.

Imagine you are testing 20 different drugs against placebo. All 20 are
inert sugar pills. If you declare victory whenever p < 0.05, you would
on average find one "effective" drug per trial just by luck. This is
the multiple testing problem.

```{r multiple-testing}
set.seed(2024)

# Simulate 20 p-values where the null is true for every test
p_values <- runif(20, min = 0, max = 1)

# How many look "significant" by chance?
sum(p_values < 0.05)

# Bonferroni: divide alpha by the number of tests
p_bonf <- p.adjust(p_values, method = "bonferroni")
sum(p_bonf < 0.05)

# Benjamini-Hochberg FDR: controls proportion of false discoveries
p_bh <- p.adjust(p_values, method = "BH")
sum(p_bh < 0.05)
```

::: {.callout-note}
### Which correction to use?

| Method | Controls | Use when |
|---|---|---|
| **Bonferroni** | Family-wise error rate (probability of *any* false positive) | False positives are very costly: clinical decisions, regulatory submissions, few pre-specified tests |
| **Benjamini–Hochberg (FDR)** | False discovery rate (proportion of discoveries that are false) | Screening many hypotheses: genomics, proteomics, GWAS |
| **None** | Nothing | Single pre-specified primary hypothesis only |

Bonferroni is conservative --- it will miss real effects when you run
many tests. For genomic studies with thousands of tests, use FDR.
For a clinical trial with five pre-specified endpoints, use Bonferroni
(or better: specify one primary endpoint in your protocol).
:::


## What Can Go Wrong

**P-hacking: running tests until something is significant**

Every additional test you run on the same data increases the false
positive rate. If you test all 50 variables in your dataset and report
only the five that came out significant, your Type I error rate is no
longer 5% --- it is much higher. This is p-hacking, even when done
unintentionally. Specify your primary analysis in advance and treat
everything else as exploratory.

**Confusing statistical and clinical significance**

A study of 50,000 patients could detect a 0.1 mmHg reduction in blood
pressure as statistically significant (p < 0.001). That would not make
it clinically meaningful. Typical antihypertensive treatments aim for
reductions of 5--10 mmHg. Statistical significance tells you the
effect is distinguishable from zero. Clinical significance tells you
whether the effect is large enough to matter. Always report the effect
size and CI, not just the p-value.

**Using one-sided tests opportunistically**

A one-sided test is more powerful but only valid when you specified the
direction of the effect *before collecting data*. Using a one-sided
test because the data happened to go the expected way is p-hacking --- it
halves the p-value without any legitimate justification.

::: {.callout-warning}
### Common misinterpretations

**"p = 0.049 and p = 0.051 are meaningfully different results"**
They are not. Both represent essentially the same strength of evidence.
The 0.05 threshold is a convention, not a natural boundary. Report
exact p-values and let readers draw their own conclusions.

**"p > 0.05 means no effect"**
It means your sample did not provide enough evidence to reject the null
at the chosen threshold. An underpowered study will produce p > 0.05
even when a real effect exists. "No significant effect" and "no effect"
are not the same statement.

**"Wide confidence intervals make a result non-significant"**
A CI that excludes zero is significant regardless of width. A narrow CI
that includes zero is not significant. Width reflects precision;
significance reflects where the interval sits relative to zero.

**"A non-significant result is not worth reporting"**
A well-powered null result is scientifically valuable --- it provides
an upper bound on how large a real effect could plausibly be. Non-
significant results from adequately powered studies should be reported.
:::


## Exercises

### Exercise 1 (Guided): Hypothesis Test for Bilirubin

Using `survival::pbc`, test whether bilirubin (`bili`) differs between
the D-penicillamine and placebo groups at baseline.

1. State $H_0$ and $H_1$.
2. Plot bilirubin by group. Is the distribution symmetric? What
   transformation should you consider, and why?
3. Run a Welch's t-test on `log(bili)`. Report the result in one
   complete sentence including the test statistic, p-value, and 95% CI.
4. Does the 95% CI include zero? Is this result expected for a baseline
   variable in a randomised trial?

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r ex1}
pbc %>%
  filter(!is.na(trt)) %>%
  ggplot(aes(x = trt, y = log(bili), fill = trt)) +
  geom_boxplot(alpha = 0.7) +
  labs(x = NULL, y = "log(Bilirubin)") +
  theme_bw() +
  theme(legend.position = "none")

t.test(log(bili) ~ trt,
       data = pbc %>% filter(!is.na(trt)))
```

The two groups have similar log-bilirubin at baseline. This is expected
for a randomised trial: randomisation should produce comparable groups
on all baseline characteristics, not just the ones we planned to test.
Write-up: "There was no statistically significant difference in
bilirubin between treatment groups at baseline (report t, df, p, and
CI from your output)."
:::

### Exercise 2 (Semi-guided): The Multiple Testing Problem in Practice

Simulate 100 hypothesis tests: 90 where the null is true and 10 where
there is a real effect.

```{r ex2-setup}
#| eval: false
set.seed(42)
p_null <- replicate(90, t.test(rnorm(30, 0, 1), rnorm(30, 0, 1))$p.value)
p_real <- replicate(10, t.test(rnorm(30, 0.8, 1), rnorm(30, 0, 1))$p.value)
all_p   <- c(p_null, p_real)
is_real <- c(rep(FALSE, 90), rep(TRUE, 10))
```

1. How many raw p-values are < 0.05? How many of those are true
   positives (from real effects) and how many are false positives?
2. Apply Bonferroni correction. How many survive?
3. Apply BH FDR. How many survive?
4. Which correction would you use for a clinical trial with 5
   pre-specified endpoints? For a GWAS?

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r ex2-solution}
set.seed(42)
p_null <- replicate(90, t.test(rnorm(30, 0, 1), rnorm(30, 0, 1))$p.value)
p_real <- replicate(10, t.test(rnorm(30, 0.8, 1), rnorm(30, 0, 1))$p.value)
all_p   <- c(p_null, p_real)
is_real <- c(rep(FALSE, 90), rep(TRUE, 10))

cat("Raw < 0.05:", sum(all_p < 0.05), "total (",
    sum(all_p < 0.05 & !is_real), "false positives,",
    sum(all_p < 0.05 &  is_real), "true positives)\n")
cat("Bonferroni:", sum(p.adjust(all_p, "bonferroni") < 0.05), "\n")
cat("BH FDR:    ", sum(p.adjust(all_p, "BH")         < 0.05), "\n")
```

Clinical trial (5 endpoints): Bonferroni, or better, pre-specify one
primary endpoint and treat the rest as secondary. GWAS: BH FDR, or
the conventional genome-wide significance threshold of 5 × 10⁻⁸
(which is effectively a very conservative Bonferroni for ~1 million
independent SNPs).
:::

### Exercise 3 (Open-ended)

Find a published paper in your field that reports multiple hypothesis
tests. Does it apply any multiple testing correction? Is that
correction appropriate for the study design? Write a short paragraph
critique, noting what the authors did and what you would have done
differently.


## Comprehension Check

1. A study reports p = 0.04. State the correct interpretation of this
   value in one sentence.
2. What is a Type I error? What probability does $\alpha = 0.05$ set for
   this error?
3. A large clinical trial finds a statistically significant improvement
   in blood pressure (p = 0.001, mean difference = 0.5 mmHg). Is this
   finding clinically important? What additional information do you need
   to decide?
4. You test 500 genes for differential expression. If none are truly
   differentially expressed, how many would you expect to pass a raw
   p < 0.05 threshold by chance?
5. When would you prefer Bonferroni over FDR correction, and vice versa?

::: {.callout-note collapse="true"}
### Answers

1. If the null hypothesis were true (no real effect), the probability
   of observing a test statistic as extreme as this one or more extreme
   is 0.04 (4%). It does not mean there is a 4% chance the null is true.

2. A Type I error is incorrectly rejecting a null hypothesis that is
   actually true (a false positive). Setting $\alpha = 0.05$ means we
   accept a 5% chance of making this error on any single test.

3. Statistical significance does not equal clinical importance. A 0.5
   mmHg reduction in blood pressure is almost certainly not clinically
   meaningful --- typical antihypertensive treatments aim for 5--10 mmHg
   reductions. You need the 95% CI for the difference (to see the range
   of plausible effect sizes) and a clinical judgement about the minimum
   relevant effect in this context.

4. $500 \times 0.05 = 25$ false positives are expected by chance alone.
   This is why multiple testing corrections are essential in omics
   studies and anywhere you are testing hundreds or thousands of
   hypotheses simultaneously.

5. **Bonferroni**: when any false positive is very costly --- clinical
   trial with pre-specified endpoints, regulatory decision-making,
   confirmatory studies. **FDR (BH)**: when you are screening many
   hypotheses and can tolerate some false discoveries in exchange for
   more power --- genomics, proteomics, metabolomics, any exploratory
   high-dimensional study.
:::


## How to Report

::: {.callout-note}
### Reporting Hypothesis Test Results

**In a Methods section:**

> "A two-sample Welch's *t*-test was used to compare [outcome] between
> [group A] and [group B]. The significance threshold was set at
> α = 0.05. Where multiple comparisons were made, p-values were
> adjusted using the Benjamini–Hochberg method to control the false
> discovery rate."

**In a Results section:**

> "Mean [outcome] was X.XX (SD Y.YY) in [group A] and X.XX (SD Y.YY)
> in [group B]; the difference was [significant / not significant]
> (*t*(df) = X.XX, p = .YYY, 95% CI for difference: L.LL to U.UU)."

**Always include:**

- Means and SDs (or medians and IQRs) for both groups
- Test statistic with degrees of freedom (*t*(df) = ...)
- Exact p-value (not "p < 0.05"; use "p < 0.001" when p is very small)
- 95% CI for the difference

**Common errors:**

- Writing "p = 0.000" --- use "p < 0.001"
- Reporting only "p < 0.05" without the actual value
- Omitting the CI --- this is often more informative than the p-value
- Writing "the result was significant" without giving the numbers
:::


## Further Reading

- @Bland-Altman-1994 --- BMJ "Statistics notes" series: short,
  accessible articles on hypothesis testing and p-values
- [Nature Methods Points of Significance](https://www.nature.com/collections/qghhqm)
  --- authoritative short articles on p-values, power, and replication
  in biology
- @Benjamini-Hochberg-1995 --- the original FDR paper; readable and
  worth understanding
