---
title: "Simple Linear Regression"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(janitor)
library(survival)
library(broom)
library(palmerpenguins)
```

## When Do You Use This? {#sec-linear-regression}

::: {.callout-tip}
You have a continuous outcome and a continuous predictor, and you want to quantify how much the outcome changes per unit increase in the predictor --- and test whether that relationship is statistically meaningful. Simple linear regression gives you a straight-line model with an estimated slope and its confidence interval. For example: predicting albumin from bilirubin, or estimating how body mass changes with flipper length in penguins.
:::

## Learning Objectives

After completing this session you will be able to:

- Fit a simple linear regression model with `lm()` in R
- Interpret the intercept, slope, standard error, and $R^2$
- Check the four key regression assumptions using ggplot diagnostic plots
- Extract and present results with `broom::tidy()` and `broom::glance()`
- Predict new values with `predict()` and plot the regression line with uncertainty

## The Key Idea: From Correlation to Prediction

In the [correlation session](correlation-and-association.qmd) we measured *whether* two variables are related. Linear regression goes further: it tells you *how much* one variable changes when the other increases by one unit, and it gives you a confidence interval for that change.

The key quantity is the **slope** ($\beta_1$): "for every one-unit increase in X, Y changes by $\beta_1$ units, on average." That is the number you will quote in a paper, a grant, or a presentation --- not r, not R², but the slope with its uncertainty.

## Background: The Linear Model

A simple linear regression models the relationship between a continuous outcome $Y$ and a single predictor $X$:

$$Y_i = \beta_0 + \beta_1 X_i + \varepsilon_i \quad \varepsilon_i \sim N(0, \sigma^2)$$

- $\beta_0$: **intercept** --- the expected value of $Y$ when $X = 0$
- $\beta_1$: **slope** --- the expected change in $Y$ for a one-unit increase in $X$
- $\varepsilon_i$: **residual** --- what the model does not explain

The model is fitted by minimising the sum of squared residuals (ordinary least squares, OLS).

**Assumptions (LINE):**

| Assumption | How to check |
|---|---|
| **L**inearity | Residuals vs fitted plot → no pattern |
| **I**ndependence | Study design → cannot check from data alone |
| **N**ormality of residuals | QQ-plot of residuals |
| **E**qual variance (homoscedasticity) | Residuals vs fitted → no funnel shape |


## Example 1: Clinical Data (PBC)

### Predict Albumin from Log(Bilirubin)

Bilirubin is right-skewed; we log-transform it to achieve a more linear relationship with albumin.

```{r pbc-setup}
pbc <- survival::pbc %>%
  as_tibble() %>%
  clean_names() %>%
  filter(!is.na(albumin), !is.na(bili)) %>%
  mutate(log_bili = log(bili))
```

Always plot first:

```{r pbc-plot}
pbc %>%
  ggplot(aes(x = log_bili, y = albumin)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "lm", colour = "tomato", se = TRUE) +
  labs(title = "Albumin vs log(bilirubin) in PBC",
       x = "log(Bilirubin)", y = "Albumin (g/dL)") +
  theme_bw()
```

```{r fit-lm}
fit <- lm(albumin ~ log_bili, data = pbc)
summary(fit)
```

::: {.callout-note}
### Reading the `summary()` Output: Line by Line

When you run `summary(fit)`, R prints several blocks. Here is what each one means.

**`Call:`**
Just echoes back the formula you used: `albumin ~ log_bili`. Confirms R fitted the right model.

**`Residuals:`**
```
Min      1Q  Median      3Q     Max
-1.52   -0.29   0.01    0.31    1.18
```
These are the differences between each patient's observed albumin and the model's prediction. Ideally:
- The **Median** should be close to 0 (no systematic over- or under-prediction)
- The **Min** and **Max** should be roughly symmetric --- if the biggest error is `+1.18` and smallest is `-1.52`, that is reasonably balanced
- Very large Max or very negative Min suggests outliers

**`Coefficients:` table**
```
              Estimate Std. Error t value Pr(>|t|)
(Intercept)   4.15      0.06      69.2    <2e-16 ***
log_bili     -0.34      0.03     -11.3    <2e-16 ***
```

| Column | What it means |
|---|---|
| **Estimate** | The fitted value of the coefficient: the slope or intercept |
| **Std. Error** | How uncertain the estimate is: smaller is more precise |
| **t value** | Estimate / Std. Error: how many standard errors away from zero |
| **Pr(>\|t\|)** | p-value: probability of seeing a t-value this large *if the true slope were zero* |
| **Stars (\*\*\*)** | Shorthand: `***` = p < 0.001, `**` = p < 0.01, `*` = p < 0.05, `.` = p < 0.1 |

**`Residual standard error: 0.43 on 416 degrees of freedom`**
The typical size of a prediction error. On average, the model's albumin prediction is off by about 0.43 g/dL. Degrees of freedom = observations minus parameters estimated (418 − 2 = 416).

**`Multiple R-squared: 0.22`**
Log(bilirubin) explains **22% of the variability** in albumin. The remaining 78% is explained by other factors not in this model.

**`Adjusted R-squared: 0.22`**
Almost the same here because we have one predictor. It drops relative to R² if many predictors are uninformative. Prefer adjusted R² when comparing models.

**`F-statistic: 128 on 1 and 416 DF, p-value: < 2.2e-16`**
Tests whether the whole model explains significantly more variance than just predicting the mean for everyone. With one predictor, this is equivalent to the t-test on the slope.
:::

### Interpreting the Output

```{r tidy-output}
tidy(fit, conf.int = TRUE)
glance(fit)
```

**Reading the coefficients:**

- **Intercept (4.15):** Expected albumin when log(bilirubin) = 0, i.e., bilirubin = 1 mg/dL, is 4.15 g/dL.
- **Slope (−0.34):** For each one-unit increase in log(bilirubin), albumin decreases by 0.34 g/dL on average.
- **R² (0.22):** Log(bilirubin) explains 22% of the variability in albumin.

::: {.callout-tip}
### What Does This Actually Tell Us?

Translate the numbers into a scientific sentence before you report anything:

> *"In PBC patients, higher bilirubin is associated with lower albumin. For every approximate 2.7-fold increase in bilirubin (one unit on the log scale), albumin is on average 0.34 g/dL lower (95% CI: −0.40 to −0.28, p < 0.001). Log(bilirubin) explains 22% of the variability in albumin across patients."*

Notice:

- You state the **direction** (lower albumin)
- You give the **magnitude** (0.34 g/dL)
- You give the **uncertainty** (95% CI)
- You give the **p-value** --- but note it only tells you the effect is unlikely to be zero, not that it is large or clinically important
- You give **R²** to indicate how much of the outcome this one predictor explains

**Is 22% a good R²?** In clinical and biological data, a single biomarker rarely explains more than 20--30% of an outcome --- there are too many other factors involved. R² = 0.22 is quite respectable. R² of 0.90+ is typical in physics or engineering, not medicine.
:::

### Checking Assumptions

We use `broom::augment()` to extract residuals and fitted values, then plot with ggplot --- consistent with the rest of the course.

```{r diagnostics}
aug <- augment(fit)

ggplot(aug, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_smooth(method = "loess", colour = "tomato", se = FALSE) +
  labs(x = "Fitted values", y = "Residuals", title = "Residuals vs Fitted") +
  theme_bw()

ggplot(aug, aes(sample = .std.resid)) +
  stat_qq() + stat_qq_line(colour = "tomato") +
  labs(x = "Theoretical quantiles", y = "Standardised residuals",
       title = "Normal Q-Q") +
  theme_bw()
```

::: {.callout-note}
### How to Read These Diagnostic Plots

**Residuals vs Fitted**
You are looking for a flat, horizontal cloud of points with no pattern.

- ✓ **Good:** Points scattered randomly around zero, no trend.
- ✗ **Bad (U-shape or arc):** The relationship is non-linear; try log-transforming X.
- ✗ **Bad (funnel shape):** Variance increases with fitted values (heteroscedasticity); try log-transforming Y.

**Normal Q-Q**
You are checking whether residuals follow a normal distribution.

- ✓ **Good:** Points follow the diagonal line closely.
- ✗ **Bad (S-curve):** Residuals are skewed.
- ✗ **Bad (heavy tails):** More extreme values than expected; mild deviation is usually acceptable for large samples.
:::

### Predictions

```{r predictions}
new_data <- tibble(log_bili = log(c(2, 5, 10)))
predict(fit, newdata = new_data, interval = "confidence")
```

```{r prediction-plot}
pred_df <- tibble(
  log_bili = seq(min(pbc$log_bili), max(pbc$log_bili), length.out = 100)
)
pred_df <- bind_cols(pred_df,
                     as_tibble(predict(fit, newdata = pred_df, interval = "prediction")))

pbc %>%
  ggplot(aes(x = log_bili, y = albumin)) +
  geom_ribbon(data = pred_df, aes(y = fit, ymin = lwr, ymax = upr),
              alpha = 0.15, fill = "steelblue") +
  geom_line(data = pred_df, aes(y = fit), colour = "steelblue", linewidth = 1) +
  geom_point(alpha = 0.3, size = 1.5) +
  labs(title = "Linear regression with 95% prediction interval",
       x = "log(Bilirubin)", y = "Albumin (g/dL)") +
  theme_bw()
```


## Example 2: Animal Data (palmerpenguins)

We predict body mass from flipper length --- a practical calibration question in ecology.

```{r penguin-lm}
peng <- penguins %>% drop_na(body_mass_g, flipper_length_mm, species)

fit_p <- lm(body_mass_g ~ flipper_length_mm, data = peng)
tidy(fit_p, conf.int = TRUE)
glance(fit_p)
```

```{r penguin-plot}
peng %>%
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(colour = species), alpha = 0.6) +
  geom_smooth(method = "lm", colour = "black", se = TRUE) +
  scale_colour_manual(values = c("steelblue", "tomato", "seagreen")) +
  labs(title = "Body mass predicted by flipper length",
       x = "Flipper length (mm)", y = "Body mass (g)") +
  theme_bw()
```

```{r penguin-diagnostics}
aug_p <- augment(fit_p)

ggplot(aug_p, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_smooth(method = "loess", colour = "tomato", se = FALSE) +
  labs(x = "Fitted values", y = "Residuals", title = "Residuals vs Fitted (penguins)") +
  theme_bw()

ggplot(aug_p, aes(sample = .std.resid)) +
  stat_qq() + stat_qq_line(colour = "tomato") +
  labs(x = "Theoretical quantiles", y = "Standardised residuals",
       title = "Normal Q-Q (penguins)") +
  theme_bw()
```

Notice that residuals may cluster by species --- a sign that species confounds the relationship. This motivates [multiple regression](multiple-regression.qmd).

::: {.callout-note}
## Where to Find Data Like This

**PBC dataset:** `survival::pbc` --- multiple continuous variables with clinically interpretable relationships.

**palmerpenguins:** `palmerpenguins::penguins` --- clean morphological measurements, ideal for regression with a natural grouping structure.

**MASS::Boston:** Housing data commonly used to teach regression; however, contains a problematic racial variable --- examine the data carefully before use.
:::

## What Can Go Wrong

::: {.callout-warning}
**Extrapolation.**
A regression line is only valid within the range of the observed data. Predicting outside this range can produce nonsensical values --- e.g., negative albumin.

**Ignoring non-linearity.**
If the Residuals vs Fitted plot shows a U-shape, the linear model is misspecified. Try a log-transformation of X, or consider polynomial terms.

**Heteroscedasticity.**
If variance increases with fitted values (fan-shaped residuals), OLS standard errors are incorrect. Consider transforming Y, or use heteroscedasticity-consistent (robust) standard errors.

**Confounding.**
A simple regression coefficient is not causal. A third variable may drive both X and Y. See the [multiple regression session](multiple-regression.qmd) for how to adjust for confounders.

**Outliers and influential points.**
Check Cook's distance via `augment()` --- the `.cooksd` column. Values above 4/n flag cases to investigate; they may be data errors or genuinely unusual patients.
:::

::: {.callout-warning}
### Common Misinterpretations

**"p < 0.05 means the effect is large."**
p < 0.05 means the effect is unlikely to be exactly zero given your sample size. A slope of −0.001 can be highly significant with 10,000 patients. Always report the *magnitude* (estimate and CI), not just the p-value.

**"p = 0.06 means there is no effect."**
It means you do not have enough evidence to rule out chance with this sample size. A study with 20 patients and p = 0.06 may simply be underpowered. Never say "no association" based on p > 0.05 alone.

**"R² = 0.05 means the model is useless."**
In medicine and biology, outcomes are driven by hundreds of factors. A single biomarker explaining 5% of variance may still be scientifically meaningful. Judge R² relative to what is realistic in your field.

**"The intercept is always meaningful."**
Only if X = 0 is a realistic value. In `lm(albumin ~ log_bili)`, the intercept is albumin when bilirubin = 1 mg/dL. If few patients have bilirubin near 1 mg/dL, the intercept is a mathematical anchor, not a clinically interpretable number.

**"A significant p-value means my model assumptions are met."**
The p-value tells you nothing about model assumptions. Check the diagnostic plots regardless of the p-value.
:::

## Exercises

### Exercise 1 (Guided): Prothrombin Time Predicts Albumin

1. Plot albumin (y) vs prothrombin time (x) in `pbc`. Log-transform prothrombin time.
2. Fit `lm(albumin ~ log(protime), data = pbc)`. Interpret the slope.
3. Check the residuals-vs-fitted and QQ plots using `broom::augment()`.
4. Report the slope, 95% CI, and R².

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r}
pbc2 <- pbc %>% filter(!is.na(protime))

pbc2 %>%
  ggplot(aes(x = log(protime), y = albumin)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", colour = "tomato") +
  labs(x = "log(Prothrombin time)", y = "Albumin (g/dL)") +
  theme_bw()

fit2 <- lm(albumin ~ I(log(protime)), data = pbc2)
tidy(fit2, conf.int = TRUE)
glance(fit2)

aug2 <- augment(fit2)

ggplot(aug2, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_smooth(method = "loess", colour = "tomato", se = FALSE) +
  labs(x = "Fitted values", y = "Residuals") + theme_bw()

ggplot(aug2, aes(sample = .std.resid)) +
  stat_qq() + stat_qq_line(colour = "tomato") +
  labs(x = "Theoretical quantiles", y = "Standardised residuals") + theme_bw()
```
:::

### Exercise 2 (Semi-guided): Bill Length Predicts Body Mass (Penguins)

1. Fit `lm(body_mass_g ~ bill_length_mm, data = penguins)`.
2. Check the diagnostic plots using `augment()`. Do the residuals look well-behaved?
3. Fit the model separately for each species. Compare the slopes. How do they differ from the overall model?

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r}
peng2 <- penguins %>% drop_na()

fit_all <- lm(body_mass_g ~ bill_length_mm, data = peng2)
tidy(fit_all, conf.int = TRUE)

aug_all <- augment(fit_all)

ggplot(aug_all, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_smooth(method = "loess", colour = "tomato", se = FALSE) +
  labs(x = "Fitted values", y = "Residuals") + theme_bw()

ggplot(aug_all, aes(sample = .std.resid)) +
  stat_qq() + stat_qq_line(colour = "tomato") +
  labs(x = "Theoretical quantiles", y = "Standardised residuals") + theme_bw()

peng2 %>%
  group_by(species) %>%
  do(tidy(lm(body_mass_g ~ bill_length_mm, data = .), conf.int = TRUE)) %>%
  filter(term == "bill_length_mm") %>%
  select(species, estimate, conf.low, conf.high, p.value)
```
:::

### Exercise 3 (Open-ended)

In your own research, identify a continuous outcome and a single continuous predictor you expect to be linearly related. Fit a simple linear regression, check all assumptions, and write a Results paragraph including: the slope (95% CI), the R², and a sentence on whether the assumptions are satisfied.


## Comprehension Check

1. You fit `lm(albumin ~ log_bili)` and get slope = −0.34, SE = 0.03. Interpret the slope in plain language.
2. The QQ-plot of residuals shows a heavy right tail. Is this a problem? What would you do?
3. R² = 0.05. Does this mean the model is useless?
4. You predict albumin = 2.1 g/dL for a patient with bilirubin = 80 mg/dL. The maximum observed bilirubin in your data is 28 mg/dL. Should you trust this prediction?
5. What does the intercept represent in `lm(albumin ~ log_bili)` when many patients have bilirubin values far from 1 mg/dL?

::: {.callout-note collapse="true"}
### Answers

1. For every one-unit increase in log(bilirubin) --- roughly a 2.7-fold increase in bilirubin on the original scale --- albumin is expected to be 0.34 g/dL lower, on average.
2. A heavy right tail means large positive residuals occur more often than expected under normality. For inference (CIs, p-values), mild non-normality is acceptable with large samples (Central Limit Theorem). For severe non-normality, try a log-transformation of the outcome or use robust standard errors.
3. R² = 0.05 means the predictor explains 5% of the outcome's variability. Whether this is "useful" depends on context. In complex biological systems, a single biomarker explaining even 5% of variance may be scientifically meaningful. The slope and CI are more informative than R² alone.
4. No. Bilirubin = 80 mg/dL is well outside the range of the data used to fit the model (max = 28). Extrapolation is unreliable; the linear relationship may not hold at extreme values.
5. The intercept is the predicted albumin when log(bilirubin) = 0, i.e., when bilirubin = 1 mg/dL. If few patients have bilirubin near 1 mg/dL, the intercept is an extrapolation with high uncertainty and limited clinical meaning.
:::


## How to Report

::: {.callout-note}
**In a methods section:** "The association between [outcome] and [predictor] was estimated using linear regression. Potential confounders [list] were included in the model. Model assumptions (linearity, homoscedasticity, normality of residuals) were assessed using diagnostic plots."

**In results:** "Each additional [unit] of [predictor] was associated with a [β] [unit] change in [outcome] (β = X, 95% CI [L, U], p = Y)."

**Always include:**

- Regression coefficient (β) with 95% CI and p-value
- Adjusted R² (proportion of variance explained)
- Number of observations and any excluded due to missing data
- Whether the model is unadjusted or adjusted (and for what)

**Reporting R² alone is not enough** --- always report the coefficient and CI so readers know the direction and magnitude of the effect.

For models with multiple predictors, see the [Multiple Regression](multiple-regression.qmd) session.
:::

## Further Reading

- @Dalgaard-2008 --- Chapter on linear models in *Introductory Statistics with R*
- @Bland-2015 --- Regression chapters in *An Introduction to Medical Statistics*
- `?lm`, `?predict.lm` in R: full documentation
- `broom` package vignette for tidy model output
