---
title: "Logistic Regression"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(janitor)
library(survival)
library(broom)
library(boot)
library(pROC)
```

::: {.callout-important}
## Before You Start

Logistic regression shares the same structure as linear regression but models a binary outcome. If you have not done so yet, complete the [Linear Regression](linear-regression.qmd) and [Multiple Regression](multiple-regression.qmd) sessions first.
:::

::: {.callout-note}
### Session at a Glance

**Total core time: ~150 minutes** (plus extra time for Exercise 3, which is open-ended).

| Section | Time | Type |
|---|---|---|
| The Key Idea: Why Not Just Use Linear Regression? | ~10 min | Concept |
| Background: The Logistic Model | ~15 min | Concept |
| Example 1: Clinical Data (PBC) | ~45 min | Walkthrough |
| Example 2: Melanoma Data | ~15 min | Walkthrough |
| What Can Go Wrong | ~10 min | Reading |
| Exercise 1 (Guided) | ~15 min | Practice |
| Exercise 2 (Semi-guided) | ~15 min | Practice |
| Exercise 3 (Open-ended) | 15--30 min | Practice |
| Comprehension Check | ~10 min | Self-test |

If you are working through this self-paced and have limited time, the
Example 1 walkthrough (PBC data) is the core of this session -- it covers
everything from fitting a model to checking discrimination with a ROC
curve. Example 2 (melanoma) repeats the same steps on a smaller dataset
and can be skimmed or done as homework. Exercise 3 is open-ended: budget
extra time if you want to apply this to your own data.
:::

::: {.callout-tip}
### How to Use the Code in This Session

Every code chunk below is written so you can copy it into your R
console (or an R script) and run it yourself, in the order shown.
Where you see a "Run It Yourself" box, stop and run the code above it
before reading the explanation that follows.
:::

## When Do You Use This? {#sec-logistic-regression}

::: {.callout-tip}
Your outcome is binary (yes/no, alive/dead, event/no event) and you want to model how one or more predictors change the probability of that outcome. Logistic regression gives you odds ratios with confidence intervals and lets you adjust for confounders. For example: predicting whether a patient died from liver disease, or whether a tumour responded to treatment.
:::

::: {.callout-note collapse="true"}
### Teacher Note

If learners have just finished [Multiple Regression](multiple-regression.qmd), remind them that `glm()` uses the same model-fitting syntax as `lm()` -- the difference is `family = binomial` and the fact that the coefficients live on the log-odds scale until you exponentiate them. Ask learners to predict: in earlier sessions, higher bilirubin was associated with *lower* albumin. Would you expect higher bilirubin to be associated with *higher* or *lower* odds of death in PBC patients, and why?
:::

## Learning Objectives

After completing this session you will be able to:

- Explain the logistic model and why ordinary linear regression is not appropriate for binary outcomes
- Fit a logistic regression with `glm(..., family = binomial)` in R
- Interpret odds ratios and convert them to probabilities
- Assess model discrimination with ROC curves and the C-statistic
- Build a multivariable logistic model and adjust for confounders

## The Key Idea: Why Not Just Use Linear Regression?

*Estimated time: ~10 minutes (concept)*

Consider predicting whether a patient dies (0 or 1) from their bilirubin level. If you fit a straight line, nothing stops it from predicting a probability of −0.2 or 1.7, which makes no sense. Probabilities must stay between 0 and 1.

Logistic regression solves this by modelling the **log-odds** (logit) of the outcome instead of the probability directly. The log-odds can range freely from −∞ to +∞, and when you transform it back, the probability is always between 0 and 1.

The trade-off: coefficients are now on the log-odds scale, which is hard to read directly. We exponentiate them to get **odds ratios (OR)**, and those are what you report in a paper.

## Background: The Logistic Model

*Estimated time: ~15 minutes (concept)*

A binary outcome $Y \in \{0, 1\}$ is modelled as:

$$\log\left(\frac{P(Y=1)}{1-P(Y=1)}\right) = \beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p$$

Exponentiating the coefficients gives **odds ratios**:

$$OR = e^{\beta_j}$$

An OR > 1 means the odds of the event increase with $X_j$; OR < 1 means they decrease.

The **predicted probability** for a patient is:

$$\hat{P} = \frac{1}{1 + e^{-(\hat{\beta}_0 + \hat{\beta}_1 X_1 + \cdots)}}$$

**Key distinction:** The OR is *not* the relative risk (RR). For rare outcomes (< 10%), they are approximately equal. For common outcomes, the OR exaggerates the association. Always specify which measure you are reporting.


## Example 1: Clinical Data (PBC)

*Estimated time: ~45 minutes (walkthrough)*

::: {.callout-note collapse="true"}
### Teacher Note

This example has six natural stopping points: (1) fit `fit_log` and read the `summary()` output, (2) extract odds ratios with `tidy()`, (3) compute predicted probabilities, (4) fit the multivariable model `fit_multi`, (5) assess discrimination with the ROC curve and C-statistic, and (6) check `glance()` output. For a self-paced learner with limited time, stops 1, 2, and 5 give the core story (single-predictor model, odds ratio, discrimination); stops 3, 4, and 6 can be done as homework or skimmed.
:::

### Predict Mortality from Bilirubin

```{r pbc-setup}
pbc <- survival::pbc %>%
  as_tibble() %>%
  clean_names() %>%
  mutate(
    died      = as.integer(status == 2),
    log_bili  = log(bili),
    log_proto = log(protime),
    sex       = factor(sex, levels = c("f", "m"), labels = c("Female", "Male")),
    stage     = factor(stage, levels = 1:4, labels = paste("Stage", 1:4)),
    trt       = factor(trt, levels = c(1, 2), labels = c("D-penicillamine", "Placebo"))
  ) %>%
  filter(!is.na(bili), !is.na(protime), !is.na(albumin), !is.na(stage))
```

> **Think before you run:** Among the 410 PBC patients in this dataset, about 38% (156 patients) died from liver disease during follow-up. Do you expect higher bilirubin to be associated with higher mortality? What direction would you predict for the OR?

Always visualise first:

```{r pbc-plot}
pbc %>%
  ggplot(aes(x = log_bili, y = died)) +
  geom_jitter(height = 0.05, alpha = 0.3, size = 1.5) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),
              colour = "tomato", se = TRUE) +
  labs(title = "Probability of death by log(bilirubin)",
       x = "log(Bilirubin)", y = "Died (0/1)") +
  theme_bw()
```

```{r glm-fit}
fit_log <- glm(died ~ log_bili, data = pbc, family = binomial)
summary(fit_log)
```

::: {.callout-note}
### Reading the `summary()` Output for Logistic Regression

Logistic regression output looks similar to linear regression but the numbers mean different things.

**`Coefficients:` table**
```
              Estimate Std. Error z value Pr(>|z|)
(Intercept)  -1.22      0.15     -8.41    <2e-16 ***
log_bili      1.14      0.13      8.74    <2e-16 ***
```

| Column | What it means |
|---|---|
| **Estimate** | The log-odds coefficient. Not directly interpretable: exponentiate to get an odds ratio. |
| **Std. Error** | Uncertainty in the estimate. |
| **z value** | Estimate / Std. Error (note: z, not t; logistic regression uses a normal approximation). |
| **Pr(>\|z\|)** | p-value: same interpretation as in linear regression. |

**`Null deviance: 544.73 on 409 degrees of freedom`**
The deviance of a model with *no predictors* (intercept only). This is the baseline.

**`Residual deviance: 442.71 on 408 degrees of freedom`**
The deviance of *your* model. A large drop from null to residual deviance (here: about 102 units, for the cost of a single predictor) means `log_bili` explains a lot.

**`AIC: 446.71`**
Lower AIC = better fit (penalised for complexity). Use AIC to compare models, not to interpret a single model in isolation.

::: {.callout-tip}
### What Does This Actually Tell Us?

The coefficient for `log_bili` is 1.14 on the log-odds scale. To interpret it:

1. **Exponentiate** to get an odds ratio: $e^{1.14} \approx 3.13$
2. **Translate:** Each one-unit increase in log(bilirubin) (roughly a 2.7-fold increase in bilirubin) multiplies the odds of death by about **3.1**.
3. **Write it up:** *"Higher bilirubin is strongly associated with mortality in PBC patients. Each 2.7-fold increase in bilirubin is associated with about 3.1 times higher odds of death (OR = 3.13, 95% CI: 2.45--4.09, p < 0.001)."*

Remember: odds ratios are **not** the same as relative risk. An OR of 3.13 does not mean the risk triples. For rare outcomes (< 10%), OR ≈ RR; for common outcomes (like death here, at 38%), OR exaggerates the association.
:::
:::

::: {.callout-tip}
### Run It Yourself

Run the `glm-fit` chunk above and check your `summary(fit_log)` output
against the table above: `(Intercept)` ≈ −1.22 and `log_bili` ≈ 1.14
(both with `p < 2e-16`), residual deviance ≈ 442.71 on 408 df, and
AIC ≈ 446.71. **Predict before you check:** since `log_bili` has a
positive coefficient, do you expect the odds ratio (once you
exponentiate it) to be greater than 1 or less than 1?
:::

### Extracting Odds Ratios

```{r or-table}
tidy(fit_log, conf.int = TRUE, exponentiate = TRUE) %>%
  select(term, estimate, conf.low, conf.high, p.value)
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. You should get `log_bili` with `estimate` ≈ 3.13
(95% CI: 2.45--4.09, `p.value` ≈ 2.41e-18) and `(Intercept)` ≈ 0.295
(95% CI: 0.220--0.389). The intercept's "odds ratio" here is just the
baseline odds of death when `log_bili = 0` (i.e. bilirubin = 1
mg/dL) -- it is not usually reported on its own. **Check:** does the
95% CI for `log_bili` include 1? What does that tell you about
statistical significance?
:::

### Predicted Probabilities

```{r predicted-probs}
new_d <- tibble(log_bili = log(c(1, 2, 5, 10, 20)))
new_d %>%
  mutate(pred_prob = predict(fit_log, newdata = new_d, type = "response"))
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. For bilirubin = 1, 2, 5, 10, and 20 mg/dL, the
predicted probability of death is approximately 0.228, 0.395, 0.650,
0.804, and 0.900. Notice how the probability rises steeply but then
levels off as it approaches 1 -- this is the S-shaped logistic curve
from the "Key Idea" section, not a straight line. **Check:** does the
*difference* in predicted probability between bilirubin = 1 and
bilirubin = 2 (about 0.17) equal the difference between bilirubin = 10
and bilirubin = 20 (about 0.10)? Why not, if the odds ratio per
log-unit is constant?
:::

### Multivariable Logistic Model

```{r multi-logistic}
fit_multi <- glm(died ~ log_bili + log_proto + albumin + stage,
                 data = pbc, family = binomial)

tidy(fit_multi, conf.int = TRUE, exponentiate = TRUE) %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. `log_bili` now has OR ≈ 2.53 (95% CI: 1.93--3.36,
p < 0.001) -- still significant, though attenuated from 3.13 once we
adjust for prothrombin time, albumin, and stage. `albumin` has
OR ≈ 0.755 (95% CI: 0.399--1.42, p = 0.383, not significant), and only
`stageStage 4` reaches significance versus Stage 1 (OR ≈ 6.52, 95% CI:
1.42--50.2, p = 0.033).

Look closely at `log_proto`: OR ≈ 1523 with a 95% CI of roughly
67.6--43432. **This is a red flag, not a strong effect.** A 95% CI that
spans several orders of magnitude usually signals near-separation or
an unstable estimate -- here, prothrombin time has a narrow range in
this dataset, so a moderate coefficient on the *log* scale translates
into a huge OR per one-unit change in log(protime) (which corresponds
to a 2.7-fold change in protime -- far outside the range actually
observed). Always check the *range* of a predictor before reporting an
OR for "a one-unit increase."
:::

### Model Discrimination: ROC Curve and C-statistic

The **C-statistic** (area under the ROC curve) measures how well the model separates cases from non-cases. C = 0.5 is random; C = 1.0 is perfect discrimination.

```{r roc}
pred_prob <- predict(fit_multi, type = "response")
roc_obj   <- roc(pbc$died, pred_prob, quiet = TRUE)

cat("C-statistic:", round(auc(roc_obj), 3), "\n")

ggroc(roc_obj, colour = "steelblue", linewidth = 1) +
  geom_abline(intercept = 1, slope = 1, linetype = "dashed", colour = "grey60") +
  labs(title = paste("ROC curve - AUC =", round(auc(roc_obj), 3)),
       x = "1 - Specificity", y = "Sensitivity") +
  theme_bw()
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. You should get a C-statistic of about 0.825,
meaning the model assigns a higher predicted probability of death to
the patient who died, versus the patient who survived, in about 83% of
all possible (died, survived) pairs. **Check:** is 0.825 closer to the
0.75 considered "moderate" or the 0.90+ considered "excellent"
discrimination in the Comprehension Check question below? Remember
that this C-statistic still says nothing about *calibration*.
:::

### Goodness of Fit

```{r hosmer}
glance(fit_multi)
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. You should see `null.deviance` ≈ 545 (409 df),
`deviance` ≈ 402 (403 df), `AIC` ≈ 416, and `nobs` = 410. Compare this
AIC to the single-predictor `fit_log` model's AIC of 446.71: the
multivariable model is about 31 AIC points lower, indicating a
meaningfully better fit despite adding three more predictors --
consistent with the higher C-statistic (0.825 vs the discrimination
you would get from `log_bili` alone).
:::

A large drop from null to residual deviance indicates good model fit. Compare AIC between models to choose the best specification.


## Example 2: Melanoma Data

*Estimated time: ~15 minutes (walkthrough)*

::: {.callout-note collapse="true"}
### Teacher Note

This example repeats the same steps as Example 1 on a smaller dataset
with a different multivariable model -- use it to check whether
learners can apply the workflow independently. Before running the
chunk below, ask: which of `log_thick`, `ulceration`, and `sex` do you
expect to be the strongest predictor of melanoma death, and in which
direction?
:::

We predict melanoma mortality from tumour thickness and ulceration.

```{r mel-setup}
mel <- boot::melanoma %>%
  as_tibble() %>%
  mutate(
    died       = as.integer(status == 1),
    ulceration = factor(ulcer, levels = c(0, 1), labels = c("No", "Yes")),
    sex        = factor(sex, levels = c(0, 1), labels = c("Female", "Male")),
    log_thick  = log(thickness)
  )

fit_mel <- glm(died ~ log_thick + ulceration + sex,
               data = mel, family = binomial)

tidy(fit_mel, conf.int = TRUE, exponentiate = TRUE) %>%
  select(term, estimate, conf.low, conf.high, p.value)
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. You should get `log_thick` with OR ≈ 1.84 (95% CI:
1.21--2.89, p ≈ 0.006) and `ulcerationYes` with OR ≈ 2.85 (95% CI:
1.34--6.21, p ≈ 0.007) -- both significant, with ulceration roughly
doubling the odds of death on top of tumour thickness. `sexMale` has
OR ≈ 1.43 (95% CI: 0.713--2.84, p ≈ 0.31), not statistically
significant once thickness and ulceration are accounted for.
**Check:** does the direction of the `log_thick` and `ulceration`
effects match what you predicted in the Teacher Note above?
:::

```{r mel-plot}
mel %>%
  ggplot(aes(x = log_thick, y = died, colour = ulceration)) +
  geom_jitter(height = 0.05, alpha = 0.4, size = 1.5) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = TRUE) +
  scale_colour_manual(values = c("steelblue", "tomato")) +
  labs(title = "Melanoma mortality by tumour thickness and ulceration",
       x = "log(Thickness)", y = "Died from melanoma (0/1)") +
  theme_bw()
```

::: {.callout-note}
## Where to Find Data Like This

**PBC dataset:** `survival::pbc`: binary mortality outcome with multiple continuous and categorical predictors.

**boot::melanoma:** Binary mortality with tumour characteristics; good for a smaller example.

**MASS::birthwt:** Low birth weight (binary) with maternal predictors, commonly used in logistic regression teaching.

**medicaldata package:** Multiple clean clinical trial datasets with binary outcomes.
:::

## What Can Go Wrong

*Estimated time: ~10 minutes (reading)*

::: {.callout-warning}
**Separation (complete or quasi-complete).**
If a predictor perfectly separates the outcome, the algorithm fails to converge and produces extreme coefficients with huge SEs. Check with a contingency table or histogram. Use Firth's penalised logistic regression (`logistf` package) as a remedy.

**Interpreting ORs as relative risks for common outcomes.**
When the outcome prevalence is > 15--20%, OR > RR substantially. If you report ORs, state them clearly and note that they do not equal relative risks.

**Events per variable (EPV) rule of thumb.**
Logistic models require approximately 10--20 events per predictor variable for stable estimates. Our `fit_multi` model above has 156 deaths and 6 predictors (`log_bili`, `log_proto`, `albumin`, and three stage dummies), giving EPV ≈ 26 -- comfortably above the rule of thumb. If you had a dataset with, say, 50 events and 6 predictors, EPV ≈ 8 would be borderline, and you would want to reduce predictors or use penalised methods. (Note that the unstable `log_proto` estimate we saw earlier was not an EPV problem -- it was a scale problem, caused by the narrow real-world range of prothrombin time.)

**Not checking calibration.**
A high C-statistic does not guarantee good calibration (that predicted probabilities match observed proportions). Use a calibration plot (`val.prob()` from the `rms` package) for clinical prediction models.
:::

::: {.callout-warning}
### Common Misinterpretations

**"OR = 3.13 means the risk is 3.13 times higher."**
An odds ratio of 3.13 means the *odds* increased about 3-fold, not the risk. When the outcome is common (> 15% -- and death occurs in 38% of this PBC sample), the OR substantially overestimates the relative risk. If you want relative risk, use a Poisson model with robust standard errors or a log-binomial model.

**"C-statistic = 0.80 means my model is good."**
The C-statistic measures discrimination: can the model rank a case above a non-case? It says nothing about calibration: are predicted probabilities accurate? A model can discriminate well but be badly miscalibrated. Always check both.

**"My predictor has OR = 1.02, so it has no effect."**
On the odds ratio scale, even small ORs can be meaningful for very common exposures. Conversely, a large OR for a rare predictor may have little population impact. Report the OR alongside the absolute risk change at clinically relevant predictor values.

**"I adjusted for many variables so my result is unconfounded."**
Adjusting for more variables does not automatically remove confounding. You can only adjust for confounders you *measured*. Unmeasured confounders remain a threat in all observational studies.

**"Quasi-complete separation means my predictor is too good."**
Separation causes coefficient inflation and numerical instability. It usually means you have too few events. Use Firth's penalised logistic regression (`logistf::logistf()`).
:::

::: {.callout-note collapse="true"}
### Teacher Note: True or False?

Pose these statements to learners (true/false, with justification):

1. "An OR of 3.13 means patients with higher bilirubin are 3.13 times more likely to die." (**False** -- this confuses odds with probability/risk.)
2. "A C-statistic of 0.825 means the model's predicted probabilities are well calibrated." (**False** -- C-statistic measures discrimination, not calibration.)
3. "Because `log_proto` had OR ≈ 1523 in `fit_multi`, prothrombin time is the most important predictor of death." (**False** -- the huge OR and CI spanning orders of magnitude signal an unstable, scale-dependent estimate, not a strong real effect.)
4. "Adding `stage` to the model means we have fully adjusted for disease severity." (**False** -- we have adjusted for *measured* severity as captured by histologic stage; unmeasured aspects of severity could still confound.)
:::

## Exercises

::: {.callout-note collapse="true"}
### Teacher Note

Before learners run Exercise 1, ask them to predict: of `log_bili`,
`albumin`, `log_proto`, and `stage`, which single predictor do they
expect to have the strongest association with death (largest OR away
from 1, smallest p-value)? After they run it, discuss why `log_proto`'s
huge OR is misleading (same issue as in `fit_multi` above) even though
it has the smallest p-value.
:::

### Exercise 1 (Guided): Univariable Predictors of Mortality

*Estimated time: ~15 minutes (practice)*

Using `pbc`:

1. Fit separate logistic regressions for each of: `log_bili`, `albumin`, `log_proto`, `stage`.
2. Report OR (95% CI) and p-value for each.
3. Which single predictor has the largest OR?

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r}
predictors <- c("log_bili", "albumin", "log_proto")

results <- map_dfr(predictors, function(pred) {
  f <- as.formula(paste("died ~", pred))
  m <- glm(f, data = pbc, family = binomial)
  tidy(m, conf.int = TRUE, exponentiate = TRUE) %>%
    filter(term != "(Intercept)") %>%
    mutate(predictor = pred)
})

results %>%
  select(predictor, estimate, conf.low, conf.high, p.value) %>%
  arrange(p.value)

fit_s <- glm(died ~ stage, data = pbc, family = binomial)
tidy(fit_s, conf.int = TRUE, exponentiate = TRUE)
```

Sorted by p-value, the univariable results are: `log_proto`
(OR ≈ 43331, 95% CI: 2392--986265, p ≈ 3.4e-12), `log_bili`
(OR ≈ 3.13, 95% CI: 2.45--4.09, p ≈ 2.4e-18), and `albumin`
(OR ≈ 0.258, 95% CI: 0.151--0.430, p ≈ 3.6e-7 -- albumin is
*protective*, so higher albumin is associated with lower odds of
death). By p-value alone, `log_bili` looks "strongest," but
`log_proto`'s astronomically large OR and CI are the same
scale-instability problem flagged earlier, not evidence of a uniquely
powerful predictor.

For `fit_s` (`died ~ stage`), only `stageStage 4` is significant versus
Stage 1 (OR ≈ 12.4, 95% CI: 3.42--80.2, p ≈ 0.001); `stageStage 2`
(OR ≈ 3.00, p ≈ 0.16) and `stageStage 3` (OR ≈ 4.04, p ≈ 0.07) are not.
This mirrors what we saw for `stage` inside `fit_multi`.
:::

### Exercise 2 (Semi-guided): Multivariable Melanoma Model

*Estimated time: ~15 minutes (practice)*

Using `boot::melanoma`:

1. Fit a univariable model: `died ~ log_thick`.
2. Add `ulceration`. How does the OR for thickness change?
3. Add `sex`. Is it a significant predictor?
4. Compare all three models by AIC.

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r}
m1 <- glm(died ~ log_thick,               data = mel, family = binomial)
m2 <- glm(died ~ log_thick + ulceration,  data = mel, family = binomial)
m3 <- glm(died ~ log_thick + ulceration + sex, data = mel, family = binomial)

bind_rows(
  tidy(m1, exponentiate = TRUE) %>% mutate(model = "1: thickness"),
  tidy(m2, exponentiate = TRUE) %>% mutate(model = "2: + ulceration"),
  tidy(m3, exponentiate = TRUE) %>% mutate(model = "3: + sex")
) %>%
  filter(term == "log_thick") %>%
  select(model, estimate, p.value)

AIC(m1, m2, m3)
```

The OR for `log_thick` shrinks each time a new predictor is added:
about 2.53 (model 1, p ≈ 3.4e-6) → about 1.91 (model 2, after adding
`ulceration`, p ≈ 0.003) → about 1.84 (model 3, after adding `sex`,
p ≈ 0.006). Thickness remains a significant predictor throughout, but
part of its univariable association is shared with ulceration --
thicker tumours are more likely to be ulcerated. Comparing AIC: model 1
≈ 219.3, model 2 ≈ 213.4, model 3 ≈ 214.3. Adding `ulceration` improves
the model (AIC drops by about 6), but adding `sex` on top makes AIC
slightly *worse* (it rises by about 1), consistent with `sexMale` not
being a significant predictor in `fit_mel` above (p ≈ 0.31). Model 2 is
the best fit by AIC.
:::

### Exercise 3 (Open-ended)

*Estimated time: 15--30 minutes (practice)*

Using your own data, identify a binary outcome and at least two predictors. Fit a multivariable logistic regression. Report: ORs with 95% CIs, C-statistic, and a calibration check. Write a one-paragraph Methods section and a one-paragraph Results section as you would for a clinical paper.


## Comprehension Check

*Estimated time: ~10 minutes (self-test)*

1. You fit logistic regression and get a coefficient of 0.8 for log_bili. What is the odds ratio, and what does it mean?
2. Your outcome occurs in 40% of the study sample. A colleague reports OR = 2.5 as "roughly a doubling of risk." Is this correct?
3. You have 30 events and want to include 8 predictors. What is the concern?
4. C-statistic = 0.75. Is this a good model? What additional checks should you do?
5. Your model fails to converge. The output shows a coefficient of 25 with SE = 1000 for one predictor. What is likely happening?

::: {.callout-note collapse="true"}
### Answers

1. OR = exp(0.8) = 2.23. Each one-unit increase in log(bilirubin) is associated with 2.23 times higher odds of the event (e.g., death), holding other predictors constant.
2. No. When the outcome is common (40%), OR ≠ RR. OR = 2.5 for a 40% baseline corresponds to a RR considerably less than 2.5. The colleague is conflating the odds ratio with the risk ratio. Use the formula $RR = OR / (1 - P_0 + P_0 \times OR)$ to convert.
3. With EPV = 30/8 = 3.75, the model is severely underpowered. Coefficients will be unstable and confidence intervals very wide. Reduce the number of predictors to at most 2--3, or use Firth's penalised logistic regression.
4. C = 0.75 indicates moderate discrimination. However, a high C-statistic does not guarantee good calibration. Also produce a calibration plot (observed vs predicted proportions across deciles of risk) to confirm the model gives accurate probabilities.
5. This is (quasi-)complete separation: the predictor almost perfectly predicts the outcome, so the algorithm tries to push the coefficient to ±∞ and fails. Check by tabulating the predictor against the outcome. Remedy: use Firth's penalised logistic regression (`logistf::logistf()`).
:::


## How to Report

::: {.callout-note}
**In a methods section:** "Logistic regression was used to examine the association between [predictor] and [binary outcome]. Results are expressed as odds ratios (OR) with 95% confidence intervals. [Confounders] were included as covariates."

**In results:** "Patients with [predictor level] had X-fold higher odds of [outcome] compared with [reference] (OR = X, 95% CI [L, U], p = Y)."

**Always include:**

- Odds ratio (OR): not the log-odds coefficient
- 95% CI for the OR
- Reference category (e.g., "compared with Stage 1")
- Number of events and total sample size (e.g., "87 deaths among 312 patients")

**AUC/ROC:** If you report a ROC curve, include AUC with 95% CI. Do not report only p-values without ORs and CIs.
:::

## Further Reading

- @Hosmer-2013: *Applied Logistic Regression*, the standard reference
- @Bland-2015: Logistic regression in *An Introduction to Medical Statistics*
- `?glm` in R: full documentation for generalised linear models
- `pROC` package vignette for ROC analysis
- `logistf` package for Firth's penalised logistic regression
