# Model Building and Diagnostics {#sec-model-building}

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(survival)
library(broom)
library(car)
```

::: {.callout-important}
## Before You Start
This session covers model selection and diagnostics for regression models. Complete **Linear Regression**, **Multiple Regression**, and **Logistic Regression** before starting here.
:::

## When Do You Use This? {.callout-tip}

> You have fitted a regression model, but before trusting its conclusions you need to verify that the model is appropriate: are the assumptions met? Are there influential outliers? Is the model overfitted? Model building and diagnostics are the quality-control steps between fitting a model and reporting it.

## Learning Objectives

After completing this session you will be able to:

- Apply a systematic model-building strategy for linear and logistic regression
- Diagnose assumption violations using residual plots, QQ-plots, and leverage diagnostics
- Detect influential observations with Cook's distance and leverage statistics
- Evaluate model performance: $R^2$, AIC, cross-validation RMSE
- Apply penalised regression (ridge, lasso) when variable selection is needed

---

## Background: A Model-Building Framework

Good regression modelling is a structured process, not a p-value fishing expedition.

**Steps:**

1. **Define the research question** — outcome, predictors, and target population
2. **Explore the data** — distributions, missingness, correlations
3. **Pre-specify the model** — driven by domain knowledge, not by initial results
4. **Fit and assess assumptions** — residuals, influential points
5. **Compare models** — use AIC/BIC or likelihood ratio tests, not step-wise
6. **Validate** — internal (bootstrap) or external (hold-out data)
7. **Report** — all results, including assumption checks

**The selection problem:** Choosing predictors based on their p-values from the same data you use to estimate coefficients inflates the apparent significance and overfits the model. Pre-specification and penalised methods address this.

---

## Worked Example 1: Linear Regression Diagnostics (PBC)

### Fit a Model

```{r pbc-fit}
pbc <- survival::pbc |>
  as_tibble() |>
  janitor::clean_names() |>
  filter(!is.na(albumin), !is.na(bili), !is.na(protime), !is.na(stage)) |>
  mutate(
    log_bili  = log(bili),
    log_proto = log(protime),
    stage     = factor(stage, levels = 1:4, labels = paste("Stage", 1:4))
  )

fit <- lm(albumin ~ log_bili + log_proto + stage, data = pbc)
```

### Residual Diagnostics

```{r residuals}
par(mfrow = c(2, 2))
plot(fit)
par(mfrow = c(1, 1))
```

**Reading the four plots:**

- **Residuals vs Fitted:** Should be a horizontal cloud around zero. Any pattern suggests non-linearity or omitted predictors.
- **Normal QQ:** Residuals should follow the diagonal. Systematic deviation from the line indicates non-normality.
- **Scale-Location (√|residuals| vs fitted):** Should be flat. An upward slope indicates heteroscedasticity.
- **Residuals vs Leverage:** Points in the upper/lower right are high-leverage *and* high-residual (influential). Cook's distance contours (dashed lines) mark the boundary.

### Augmenting Data with Diagnostics

```{r augment}
diag_df <- augment(fit) |>
  mutate(.row = row_number())

# High Cook's distance
diag_df |>
  arrange(desc(.cooksd)) |>
  select(.row, albumin, log_bili, .fitted, .resid, .cooksd, .hat) |>
  head(10)
```

```{r cook-plot}
diag_df |>
  ggplot(aes(x = .row, y = .cooksd)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_hline(yintercept = 4 / nrow(pbc), linetype = "dashed", color = "tomato") +
  labs(title = "Cook's distance — influential observations",
       x = "Observation", y = "Cook's distance") +
  theme_bw()
```

Points above the 4/n threshold merit investigation. They may be data errors or genuinely extreme but valid cases.

### Checking Multicollinearity

```{r vif}
vif(fit)
```

### Testing for Heteroscedasticity

```{r ncvtest}
ncvTest(fit)   # Breusch-Pagan test from car package
```

A significant p-value indicates heteroscedasticity. Consider transforming Y or using heteroscedasticity-consistent (HC) standard errors.

---

## Worked Example 2: Logistic Regression Diagnostics (PBC Mortality)

```{r logistic-fit}
pbc2 <- pbc |>
  mutate(
    died = as.integer(status == 2),
    sex  = factor(sex, levels = c("f", "m"), labels = c("Female", "Male"))
  ) |>
  filter(!is.na(died))

fit_log <- glm(died ~ log_bili + log_proto + albumin + stage,
               data = pbc2, family = binomial)
```

### Residual and Influence Diagnostics

```{r logistic-diag}
diag_log <- augment(fit_log, type.predict = "response")

# Deviance residuals vs fitted probabilities
diag_log |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(se = FALSE, color = "tomato") +
  labs(title = "Deviance residuals vs fitted probabilities",
       x = "Fitted probability", y = "Deviance residual") +
  theme_bw()
```

```{r logistic-cook}
diag_log |>
  mutate(row = row_number()) |>
  ggplot(aes(x = row, y = .cooksd)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_hline(yintercept = 4 / nrow(pbc2), linetype = "dashed", color = "tomato") +
  labs(title = "Cook's distance — logistic model",
       x = "Observation", y = "Cook's distance") +
  theme_bw()
```

### AIC Comparison: Model Selection Without P-value Fishing

```{r aic-compare}
m1 <- glm(died ~ log_bili,                         data = pbc2, family = binomial)
m2 <- glm(died ~ log_bili + log_proto,             data = pbc2, family = binomial)
m3 <- glm(died ~ log_bili + log_proto + albumin,   data = pbc2, family = binomial)
m4 <- glm(died ~ log_bili + log_proto + albumin + stage, data = pbc2, family = binomial)

AIC(m1, m2, m3, m4) |>
  tibble::rownames_to_column("model") |>
  arrange(AIC)
```

Choose the model with the lowest AIC that aligns with your pre-specified hypotheses.

---

## Worked Example 3: Penalised Regression (Lasso)

When you have many candidate predictors and want data-driven selection, lasso (L1 penalisation) shrinks some coefficients to exactly zero.

```{r lasso}
library(glmnet)

pbc_mat <- pbc |>
  select(albumin, log_bili, log_proto, platelet, bili, protime) |>
  drop_na()

X <- model.matrix(albumin ~ log_bili + log_proto + platelet, data = pbc_mat)[, -1]
y <- pbc_mat$albumin

# Cross-validated lasso
set.seed(2024)
cv_lasso <- cv.glmnet(X, y, alpha = 1)
plot(cv_lasso)

# Coefficients at the optimal lambda
coef(cv_lasso, s = "lambda.min")
```

Lasso selects predictors by shrinking small coefficients to zero. It is useful for exploratory modelling with many candidates, but coefficients are biased (shrunk toward zero) and should not be interpreted as effect estimates without post-selection inference.

---

## Where to Find Data Like This {.callout-note}

**PBC dataset:** `survival::pbc` — multiple correlated continuous and categorical predictors; good for demonstrating overfitting, VIF, and model selection.

**MASS::Boston (or updated `BostonHousing2` from `mlbench`):** Housing data with correlated predictors, commonly used for penalised regression examples.

**Any of your prior datasets** — model building and diagnostics are always applied to real data from a research question, not synthetic datasets.

---

## What Can Go Wrong {.callout-warning}

**Stepwise selection**
Automated forward/backward stepwise selection based on p-values produces overfit models with inflated effect estimates and incorrect standard errors. Avoid unless used purely for exploration, not inference.

**Ignoring assumption violations**
A model that violates assumptions (non-linearity, heteroscedasticity, non-normality of residuals) gives incorrect CIs and p-values. Check diagnostics before reporting.

**Removing outliers without justification**
Influential observations may be genuine extreme cases, not errors. Removing them without documenting and justifying the decision is a form of selective reporting.

**Confusing in-sample fit with predictive performance**
$R^2$ and AIC measure in-sample fit. A model that fits the data well may not predict new observations accurately (overfitting). Use cross-validation or a hold-out test set to estimate predictive performance.

---

## Exercises

### Exercise 1 (Guided): Full Diagnostic Workflow

Fit `lm(albumin ~ log_bili + log_proto + stage, data = pbc)` and:

1. Produce the four standard diagnostic plots.
2. Identify the top 5 observations by Cook's distance. Are they plausible outliers?
3. Check VIF. Is multicollinearity a concern?
4. Run `ncvTest()`. Is heteroscedasticity detected?

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r}
fit_ex <- lm(albumin ~ log_bili + log_proto + stage, data = pbc)

# 1. Diagnostic plots
par(mfrow = c(2, 2)); plot(fit_ex); par(mfrow = c(1, 1))

# 2. Top Cook's distances
augment(fit_ex) |>
  mutate(.row = row_number()) |>
  arrange(desc(.cooksd)) |>
  select(.row, albumin, log_bili, .resid, .cooksd) |>
  head(5)

# 3. VIF
vif(fit_ex)

# 4. Heteroscedasticity test
ncvTest(fit_ex)
```
:::

### Exercise 2 (Semi-guided): AIC Model Selection for Logistic Regression

Using `pbc2`, fit four logistic models predicting mortality, adding one predictor at a time. Choose the best model by AIC. Then check Cook's distance for the chosen model.

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r}
a1 <- glm(died ~ log_bili, data = pbc2, family = binomial)
a2 <- glm(died ~ log_bili + albumin, data = pbc2, family = binomial)
a3 <- glm(died ~ log_bili + albumin + log_proto, data = pbc2, family = binomial)
a4 <- glm(died ~ log_bili + albumin + log_proto + stage, data = pbc2, family = binomial)

AIC(a1, a2, a3, a4) |> tibble::rownames_to_column() |> arrange(AIC)

# Cook's for best model
augment(a4, type.predict = "response") |>
  arrange(desc(.cooksd)) |>
  head(5)
```
:::

### Exercise 3 (Open-ended)

Take a model from your own research and apply the full diagnostic workflow: residual plots, QQ-plot, influence analysis, VIF. Write a one-paragraph Supplementary Methods section documenting your model-checking process and how you handled any issues found.

---

## Comprehension Check

1. The Residuals vs Fitted plot shows a clear U-shape. What does this indicate, and how should you fix it?
2. One observation has Cook's distance = 0.8 (the threshold 4/n = 0.01). Should you automatically remove it?
3. All VIFs are 1.1 except one predictor that has VIF = 9.2. What does this mean, and what are your options?
4. You compare three models with AIC: 312, 308, 307. Which should you prefer?
5. A lasso model with 20 candidate predictors shrinks 14 coefficients to zero, leaving 6 non-zero predictors. Can you interpret those 6 coefficients as the "true" effects?

::: {.callout-note collapse="true"}
### Answers
1. A U-shape indicates non-linearity — the linear model misses a curved relationship. Check whether X needs to be log-transformed, squared, or modelled with a spline. You can also try a LOESS smoother on the data to visualise the true shape.
2. No — influential observations should be investigated, not automatically removed. Check whether the observation is a data entry error (correct it) or a genuinely extreme but valid case (keep it and report sensitivity analyses with and without that observation).
3. VIF = 9.2 for one predictor indicates it is highly collinear with other predictors. Its coefficient estimate is unreliable (large SE, may flip sign). Options: (1) remove the collinear predictor if it is not the predictor of primary interest; (2) combine correlated predictors into a composite score; (3) use ridge regression.
4. The third model (AIC = 307) is preferred — it has the lowest AIC, indicating the best balance of fit and complexity. However, the difference from model 2 (AIC = 308) is only 1, which is not a meaningful improvement. Models within 2 AIC units of the minimum are considered roughly equivalent.
5. No. Lasso coefficients are biased toward zero by the penalty. After lasso selects variables, you should refit the model using ordinary regression on those selected variables to obtain unbiased coefficient estimates and valid CIs. Even then, some caution is needed because variable selection inflates Type I error — ideally, use pre-specified variables and treat lasso as exploratory.
:::

## Further Reading

- @Harrell-2015 --- *Regression Modeling Strategies* — the definitive reference for clinical regression
- @Fox-2019 --- *An R Companion to Applied Regression* — covers all diagnostics in R
- `?plot.lm`, `?augment`, `?vif` in R
- `glmnet` package vignette for penalised regression
