---
title: "Nonparametric Tests"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(survival)
library(broom)
```

::: {.callout-note}
### Session at a Glance

**Total core time: about 125 minutes (~2 hours).**

| Section | Time | Type |
|---|---|---|
| When Do You Use This? + Learning Objectives | 5 min | Reading |
| Background: The Idea of Ranks | 20 min | Reading |
| Example 1: Small Samples (Lab Data) | 25 min | Worked example |
| Example 2: Clinical Data (PBC) | 25 min | Worked example |
| What Can Go Wrong | 10 min | Reading |
| Exercises 1 & 2 | 30 min | Practice |
| Comprehension Check | 10 min | Self-test |

In a taught course, this session fits one teaching block. For
self-paced study, split it into two sittings, e.g. Background and
Example 1 first, Example 2 onward later. Exercise 3 is open-ended and
works best as a follow-up using a dataset from your own field.
:::

::: {.callout-tip}
### How to Use the Code in This Session

Every code chunk below is written so you can copy it into your R
console (or an R script) and run it yourself, in the order shown.
Where you see a "Run It Yourself" box, stop and run the code above it
before reading the explanation that follows.
:::

## When Do You Use This? {#sec-nonparametric}

::: {.callout-tip}
You have 8 mice per group and the outcome (tumour volume, cytokine
concentration, cell count) is heavily right-skewed with a few extreme
values. Or you are measuring pain on a 0--10 scale and the values clump
at the low end. The t-tests from the previous session assume the
sampling distribution of the mean is normal, and with small non-normal
samples the Central Limit Theorem cannot guarantee that.

Nonparametric tests are the alternative. They work on the *ranks* of the
data rather than the raw values. This makes them robust to skew and
outliers. The trade-off: they are slightly less powerful than t-tests
when the normality assumption genuinely holds.
:::

::: {.callout-note collapse="true"}
### Teacher Note

Ask the room for an outcome from their own research that is skewed or
has extreme values: length of stay, cytokine levels, viral load, time
to event. Use one of those examples throughout the session in place of
(or alongside) the tumour-volume and bilirubin examples below — it
makes the "ranks are robust to outliers" argument land harder when it's
their own data.
:::

## Learning Objectives

After completing this session you will be able to:

- Decide when a nonparametric test is more appropriate than a t-test
- Explain what "testing on ranks" means and why it helps with
  non-normal data
- Run the Wilcoxon rank-sum and signed-rank tests in R and read the
  output
- Run the Kruskal-Wallis test and appropriate post-hoc comparisons
- Report nonparametric results correctly, including the right summary
  statistics


## Background: The Idea of Ranks

*Estimated time: ~20 minutes (reading)*

::: {.callout-note collapse="true"}
### Teacher Note

Board exercise: write the five tumour volumes (5, 1, 42, 3, 7) on the
board and ask the room to rank them by hand before revealing the table
below. Then ask: "if 42 were actually 420, what would change?" Let
someone answer "nothing" before you confirm it — the moment of
realising that ranks are completely insensitive to *how* extreme an
outlier is, only to its *order*, is the conceptual core of this whole
session.
:::

### What ranks are and why they help

Suppose five mice have tumour volumes of 5, 1, 42, 3, and 7 mm³. Their
ranks are:

| Volume | Rank |
|---|---|
| 1 | 1 |
| 3 | 2 |
| 5 | 3 |
| 7 | 4 |
| 42 | 5 |

Now suppose the outlier was a measurement error and the true value was
420 instead of 42. The ranks do not change at all: 420 is still rank
5. A t-test would give a very different answer; a Wilcoxon test would
not. This is the core property: nonparametric tests are insensitive to
the exact magnitude of extreme values.

The Wilcoxon rank-sum test takes all observations from both groups,
ranks them together from smallest to largest, and then checks whether
one group's ranks are systematically higher than the other's. The null
hypothesis is that the two groups are drawn from the same distribution:
no tendency for one group to produce larger values.

### When to use nonparametric tests

Use a nonparametric test when:

1. **Sample size is small** ($n < 30$ per group) **and the data are
   clearly non-normal**: check a QQ-plot; if points deviate
   substantially from the line, switch to nonparametric
2. **Extreme outliers** that reflect real biology (not data errors):
   a mouse tumour genuinely 10× larger than the rest is real information
   that should not drive your result
3. **Ordinal outcomes**: pain scales, Likert items, disease stages;
   the numeric labels do not have consistent spacing

When *not* to use them by default: if $n \geq 30$ and data are
reasonably symmetric, the t-test is valid and more powerful. Using
nonparametric tests as a universal "safe" choice costs statistical power
unnecessarily.

| Parametric | Nonparametric equivalent | Use when |
|---|---|---|
| One-sample t-test | Wilcoxon signed-rank | Small n, symmetric but non-normal |
| Two-sample t-test | Wilcoxon rank-sum (Mann-Whitney) | Two independent groups, small n or skewed |
| Paired t-test | Wilcoxon signed-rank (paired) | Paired data, small n or non-normal differences |
| One-way ANOVA | Kruskal-Wallis | Three or more groups, normality violated |


## Example 1: Small Samples (Lab Data)

*Estimated time: ~25 minutes (worked example)*

::: {.callout-note collapse="true"}
### Teacher Note

Before running the `tg-small` chunk, ask the room to predict (show of
hands) whether the QQ-plots will look "normal enough" for a t-test with
only n = 10 per group. Then run the chunk, look at the plots together,
and connect this back to the QQ-plot discussion in
[T-tests](t-tests.qmd): the decision to use a nonparametric test is made
*before* seeing the test result, based on the plot and sample size, not
by picking whichever test gives a smaller p-value.
:::

### Step 1: Check normality, then decide

The decision between a t-test and a Wilcoxon test should be made *before*
running the test, based on a QQ-plot and the sample size. Here we
demonstrate this decision with `ToothGrowth` at the 0.5 mg/day dose
where each group has only 10 guinea pigs.

```{r tg-small}
#| eval: true

tg_05 <- ToothGrowth %>%
  as_tibble() %>%
  mutate(
    supp = factor(supp, levels = c("VC", "OJ"),
                  labels = c("Ascorbic acid", "Orange juice"))
  ) %>%
  filter(dose == 0.5)

tg_05 %>%
  ggplot(aes(sample = len)) +
  stat_qq() +
  stat_qq_line(colour = "red") +
  facet_wrap(~supp) +
  labs(
    title = "QQ-plot: tooth length at 0.5 mg/day (n = 10 per group)",
    subtitle = "Points should follow the red line if data are approximately normal"
  ) +
  theme_bw()
```

With n = 10 and some departure from the line, the Wilcoxon test is the
safer choice here. The t-test would also be defensible (the deviations
are not severe), but when in doubt with small samples, use the
nonparametric option.

### Two-sample Wilcoxon rank-sum test

```{r tg-wilcox}
#| eval: true

res_wilcox <- wilcox.test(len ~ supp, data = tg_05, exact = FALSE)
tidy(res_wilcox)

tg_05 %>%
  group_by(supp) %>%
  summarise(
    n      = n(),
    median = median(len),
    q1     = quantile(len, 0.25),
    q3     = quantile(len, 0.75)
  )
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. You should get `W ≈ 19.5` and `p ≈ 0.023`. The
group medians are 7.15 (IQR: 5.95--10.9) for ascorbic acid and 12.2
(IQR: 9.7--16.2) for orange juice — orange juice produces longer teeth
at this dose, and the gap is large relative to the spread. Before
moving on, predict: will the t-test on the same data (next chunk) reach
the same conclusion?
:::

::: {.callout-note}
### Reading the Wilcoxon Output

| Output | What it means |
|---|---|
| **statistic (W)** | The sum of ranks in the first group. A large W means the first group tends to have higher values. |
| **p.value** | The probability of observing a rank distribution this extreme if both groups came from the same population. |

Always report **medians and IQR**, not means and SD, alongside a
Wilcoxon test. The test is based on ranks, so the rank-based summary
(median) is the appropriate measure of location.
:::

How does this compare to using the t-test on the same data?

```{r tg-compare}
#| eval: true

bind_rows(
  tidy(t.test(len ~ supp, data = tg_05)) %>%
    mutate(test = "Welch t-test"),
  tidy(wilcox.test(len ~ supp, data = tg_05, exact = FALSE)) %>%
    mutate(test = "Wilcoxon rank-sum")
) %>%
  select(test, statistic, p.value)
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. The Welch t-test gives `t ≈ -3.17, p ≈ 0.006`, and
the Wilcoxon gives `W ≈ 19.5, p ≈ 0.023`. Both are well below 0.05, so
both tests agree: orange juice produces significantly longer teeth than
ascorbic acid at this dose.
:::

Both reach the same conclusion here. They often agree. When they
disagree, trust the Wilcoxon for small non-normal samples, and
investigate why they differ (usually an outlier is driving the t-test
result).

### Paired Wilcoxon signed-rank test

The `sleep` dataset has 10 patients measured under two drugs: a paired
design. For the paired Wilcoxon, what matters is the normality of the
*within-patient differences*, not the raw values.

```{r sleep-wilcox}
#| eval: true

sleepdat <- sleep %>%
  as_tibble() %>%
  rename(drug = group, extra_hours = extra) %>%
  mutate(drug = factor(drug, labels = c("Drug 1", "Drug 2")))

sleep_wide <- sleepdat %>%
  tidyr::pivot_wider(
    names_from  = drug,
    values_from = extra_hours,
    id_cols     = ID
  )

diffs <- sleep_wide[["Drug 2"]] - sleep_wide[["Drug 1"]]

tibble(diff = diffs) %>%
  ggplot(aes(sample = diff)) +
  stat_qq() +
  stat_qq_line(colour = "red") +
  labs(
    title    = "QQ-plot: within-patient differences in sleep hours",
    subtitle = "Normality of differences, not raw values, is what matters for paired tests"
  ) +
  theme_bw()
```

```{r sleep-wilcox-test}
#| eval: true

res_sleep <- wilcox.test(
  sleep_wide[["Drug 2"]], sleep_wide[["Drug 1"]],
  paired = TRUE
)
tidy(res_sleep)
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above (you'll see two warnings about ties and zeroes —
these are expected with small integer-valued data and do not invalidate
the test). You should get `V ≈ 45` and `p ≈ 0.009`. This is well below
0.05: Drug 2 gives significantly more extra sleep than Drug 1.
:::

With n = 10 differences, the QQ-plot shows reasonable normality:
a paired t-test would also be valid. The Wilcoxon and t-test conclusions
agree here, which is reassuring.


## Example 2: Clinical Data (PBC)

*Estimated time: ~25 minutes (worked example)*

::: {.callout-note collapse="true"}
### Teacher Note

The PBC dataset has been used for treatment-group comparisons in
[T-tests](t-tests.qmd) and [Hypothesis Testing](hypothesis-testing.qmd)
with `albumin` (a roughly symmetric variable, no significant
difference between groups). Here we switch to `bilirubin`, which is
heavily skewed in the opposite direction: ask the room why the same
randomised comparison (D-penicillamine vs placebo) might call for a
different test depending on which variable you look at.
:::

Bilirubin in the PBC dataset is a genuine case for nonparametric tests:
heavily right-skewed, extreme outliers, and meaningful biological
variability. No log transformation changes the underlying problem:
the distribution is better described by its median and spread than its
mean.

```{r pbc-setup}
#| eval: true

pbc <- survival::pbc %>%
  as_tibble() %>%
  janitor::clean_names() %>%
  mutate(
    sex   = factor(sex, levels = c("f", "m"), labels = c("Female", "Male")),
    trt   = factor(trt, levels = c(1, 2), labels = c("D-penicillamine", "Placebo")),
    stage = factor(stage, levels = 1:4, labels = paste("Stage", 1:4))
  )
```

### Wilcoxon rank-sum: bilirubin by treatment group

```{r pbc-wilcox}
#| eval: true

pbc_rand <- pbc %>% filter(!is.na(trt), !is.na(bili))

pbc_rand %>%
  ggplot(aes(x = trt, y = bili, fill = trt)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1.5) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  scale_y_log10() +
  labs(
    title = "Bilirubin by treatment group (log scale)",
    x     = NULL,
    y     = "Bilirubin (mg/dL, log scale)"
  ) +
  theme_bw() +
  theme(legend.position = "none")

tidy(wilcox.test(bili ~ trt, data = pbc_rand))

pbc_rand %>%
  group_by(trt) %>%
  summarise(
    n      = n(),
    median = median(bili),
    q1     = quantile(bili, 0.25),
    q3     = quantile(bili, 0.75)
  )
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. You should get `W ≈ 12007` and `p ≈ 0.84`. The
medians are 1.4 mg/dL (IQR: 0.8--3.2, n = 158) for D-penicillamine and
1.3 mg/dL (IQR: 0.725--3.6, n = 154) for placebo — very similar
distributions, as expected for a randomised baseline comparison.
:::

No significant difference at baseline (median bilirubin 1.4 mg/dL,
IQR: 0.8--3.2, in the D-penicillamine group vs 1.3 mg/dL, IQR:
0.725--3.6, in the placebo group; Wilcoxon rank-sum: W = 12007, p =
0.84), consistent with successful randomisation.

### Kruskal-Wallis: three or more groups

The Kruskal-Wallis test is the nonparametric analogue of one-way ANOVA.
It tests whether at least one group differs from the others, but does
not tell you which pairs differ. A significant result requires post-hoc
pairwise comparisons with multiple testing correction.

Does bilirubin differ across the four disease stages?

```{r kruskal}
#| eval: true

pbc_stage <- pbc %>% filter(!is.na(stage), !is.na(bili))

kruskal.test(bili ~ stage, data = pbc_stage)
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. You should get `chi-squared ≈ 46.6`, `df = 3`,
and `p ≈ 4e-10` — a tiny p-value, far below 0.05. At least one disease
stage differs from the others in bilirubin.
:::

The overall test is highly significant: bilirubin clearly rises with
disease stage, as expected biologically. Now identify which specific
stage pairs account for this:

```{r kruskal-posthoc}
#| eval: true

pairwise.wilcox.test(
  pbc_stage$bili,
  pbc_stage$stage,
  p.adjust.method = "bonferroni"
)
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. You should get a table of pairwise p-values.
Stage 1 vs Stage 2 (p = 1.000) and Stage 2 vs Stage 3 (p = 0.168) are
not significant after Bonferroni correction, and Stage 1 vs Stage 3 is
borderline (p = 0.029). Every comparison involving Stage 4 is highly
significant (p < 0.0001). In words: bilirubin in Stage 4 is clearly
different from every earlier stage, while the early stages are harder
to tell apart from each other.
:::

```{r kruskal-plot}
#| eval: true

pbc_stage %>%
  ggplot(aes(x = stage, y = bili, fill = stage)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
  scale_fill_brewer(palette = "Blues") +
  scale_y_log10() +
  labs(
    title = "Bilirubin by disease stage (log scale)",
    x     = NULL,
    y     = "Bilirubin (mg/dL)"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

The post-hoc table shows which stage pairs are significantly different
after Bonferroni correction. Early stages (1 and 2) tend not to differ
from each other, but later stages (3 and 4) separate clearly.

::: {.callout-note}
### After Kruskal-Wallis: what to report

A significant Kruskal-Wallis result tells you "at least one group
differs." To say more, you need post-hoc pairwise tests. Always apply a
multiple testing correction (`"bonferroni"` is conservative but clear;
`"holm"` is slightly more powerful). Report the overall test result and
then summarise which pairs differed.
:::


## What Can Go Wrong

*Estimated time: ~10 minutes (reading)*

::: {.callout-note collapse="true"}
### Teacher Note

Run a quick true/false round on the "Common misinterpretations" callout
below: read each bold statement aloud and ask for a thumbs up/down
before revealing the explanation. The "Wilcoxon makes no assumptions"
one is usually the most surprising to students who learned it as "the
test you use when you can't make assumptions."
:::

**Using nonparametric tests as the default "safe" choice**

Nonparametric tests are not always better. When data are approximately
normal, the t-test is more powerful: it extracts more information
from the same data. Switching to nonparametric by default when a t-test
would be valid is a quiet way to reduce your power to detect real
effects. Make the decision from the data: check the QQ-plot and the
sample size.

**Reporting means alongside a Wilcoxon result**

The Wilcoxon test is based on ranks, which correspond to the median as
a measure of location. If you report means and SDs in your results table
and then cite a Wilcoxon p-value in the text, you are mixing
incompatible summaries. Report medians and IQR when using nonparametric
tests.

**Skipping post-hoc correction after Kruskal-Wallis**

Running six pairwise Wilcoxon tests without correction after a
Kruskal-Wallis gives a false positive rate far above 5%. Always specify
a correction method in `pairwise.wilcox.test()`.

**Interpreting a non-significant result as "no difference"**

With small samples, nonparametric tests have limited power. A p > 0.05
from a Wilcoxon test with n = 8 per group means you could not detect a
difference, not that no difference exists. Report the medians, IQR, and
sample size so readers can judge whether the study was adequately powered.

::: {.callout-warning}
### Common misinterpretations

**"Nonparametric tests make no assumptions"**
They do. The Wilcoxon rank-sum test assumes the two distributions have
the same shape and only differ in location (shift). It is more robust
than the t-test, but not assumption-free.

**"I used a Wilcoxon test because my data failed Shapiro-Wilk"**
Shapiro-Wilk is a poor decision tool (see [T-tests](t-tests.qmd)). Use
a QQ-plot and your knowledge of the outcome instead. If bilirubin is
right-skewed in every study ever published, you do not need a formal
normality test to justify a Wilcoxon.

**"The Wilcoxon test compares medians"**
Not exactly. It tests whether one group's values tend to be larger than
the other's: technically, whether P(X > Y) = 0.5. This is close to
comparing medians when the distributions have the same shape, but not
in general.
:::


## Exercises

::: {.callout-note collapse="true"}
### Teacher Note

Pair students for Exercises 1 and 2: one person runs the code while the
other predicts the result before each output appears (medians, then
test statistic, then p-value). Exercise 1 has a genuine surprise in
store (see the solution), which works best if students commit to a
prediction first.
:::

### Exercise 1 (Guided): Wilcoxon Test for Bilirubin by Sex

*Estimated time: ~15 minutes*

Using `survival::pbc`, test whether bilirubin differs between male and
female patients.

1. Plot bilirubin by sex using a log scale on the y-axis.
2. Compute medians and IQR for each sex.
3. Run a Wilcoxon rank-sum test.
4. For comparison, run a t-test on `log(bili)`. Do both tests reach
   the same conclusion?

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r ex1}
pbc_sex <- pbc %>% filter(!is.na(sex), !is.na(bili))

pbc_sex %>%
  ggplot(aes(x = sex, y = bili, fill = sex)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
  scale_y_log10() +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(x = NULL, y = "Bilirubin (mg/dL, log scale)") +
  theme_bw() +
  theme(legend.position = "none")

pbc_sex %>%
  group_by(sex) %>%
  summarise(
    n      = n(),
    median = median(bili),
    q1     = quantile(bili, 0.25),
    q3     = quantile(bili, 0.75)
  )

tidy(wilcox.test(bili ~ sex, data = pbc_sex))

tidy(t.test(log(bili) ~ sex, data = pbc_sex))
```

Run the chunk above. Median bilirubin is 1.3 mg/dL (IQR: 0.7--3.4) for
females (n = 374) and 2.05 mg/dL (IQR: 1.3--3.5) for males (n = 44).
The Wilcoxon test gives `W ≈ 6574, p ≈ 0.029` — significant at the 0.05
level. But the t-test on `log(bili)` gives `t ≈ -1.73, df ≈ 64,
p ≈ 0.089` — *not* significant.

This time the two tests **disagree**: the Wilcoxon says males and
females differ significantly, the t-test on log-bilirubin does not
quite reach 0.05. This is a useful real example, not an error. A few
things are happening at once: the male group is small (n = 44) and the
two group sizes are very unequal (44 vs 374), which affects the t-test's
power and its degrees of freedom (Welch's correction gives df ≈ 64, far
below the 416 you might naively expect). The Wilcoxon, working on ranks
across the whole sample, is less sensitive to that imbalance here. When
two reasonable methods disagree like this, do not pick whichever gives
the answer you want — report both results and let the reader judge,
or note that the result is borderline and would benefit from a larger
male sample.
:::

### Exercise 2 (Semi-guided): Kruskal-Wallis Across Disease Stages

*Estimated time: ~15 minutes*

Test whether bilirubin differs across disease stages 1--4 in `pbc`.

1. Run `kruskal.test()`. Is the overall test significant?
2. Run `pairwise.wilcox.test()` with Bonferroni correction. Which stage
   pairs are significantly different?
3. Create a boxplot (log scale) to visualise the pattern.
4. Does the pattern make biological sense? What would you expect for a
   progressive disease?

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r ex2}
pbc_s <- pbc %>% filter(!is.na(stage), !is.na(bili))

kruskal.test(bili ~ stage, data = pbc_s)

pairwise.wilcox.test(pbc_s$bili, pbc_s$stage, p.adjust.method = "bonferroni")

pbc_s %>%
  ggplot(aes(x = stage, y = bili, fill = stage)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
  scale_fill_brewer(palette = "Blues") +
  scale_y_log10() +
  labs(
    title = "Bilirubin by disease stage (log scale)",
    x = NULL, y = "Bilirubin (mg/dL)"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

For a progressive disease, we expect bilirubin to rise with stage,
and it does. The pairwise tests show that adjacent early stages may not
differ significantly, but the gap between early and late stages is
substantial and highly significant.
:::

### Exercise 3 (Open-ended)

*Estimated time: 15--30 minutes*

From your own data, find a continuous variable that is clearly
non-normal (cytokine levels, viral load, cell counts, enzyme
concentrations). Run both a t-test and a Wilcoxon test on the same
comparison. Do they agree? If they disagree, investigate why: look
at the data and identify what is driving the difference.


## Comprehension Check

*Estimated time: ~10 minutes (self-test)*

1. In what two situations is a nonparametric test preferred over a
   t-test?
2. What does it mean to "test on ranks"? Give a concrete example of
   why this makes a test robust to outliers.
3. You run a Kruskal-Wallis test and get p = 0.02 across four groups.
   What should you do next?
4. You used a Wilcoxon test. Should you report means or medians in your
   results table?
5. A colleague says "I always use Wilcoxon because it doesn't assume
   normality, so it's safer." Is this a good strategy?

::: {.callout-note collapse="true"}
### Answers

1. When (1) sample size is small ($n < 30$ per group) and data are
   visibly non-normal on a QQ-plot, or (2) the outcome has extreme
   outliers that represent real biology rather than measurement error.
   Also appropriate for ordinal outcomes.

2. All observations are ranked from smallest to largest, and the test
   compares the distribution of ranks between groups rather than the
   raw values. Example: values of 1, 3, 5, 7, 42 become ranks 1, 2,
   3, 4, 5. Changing 42 to 420 does not change the ranks, so the test
   result is unaffected by that extreme value.

3. Follow up with pairwise Wilcoxon tests using `pairwise.wilcox.test()`
   with a multiple testing correction (Bonferroni or Holm). The
   Kruskal-Wallis result tells you "at least one group differs";
   post-hoc tests identify which specific pairs.

4. Medians and IQR. The Wilcoxon test is based on ranks, and the
   median is the rank-based measure of location. Reporting means
   alongside a Wilcoxon result is internally inconsistent.

5. Not a good general strategy. When data are approximately normal,
   the t-test is more powerful. Using Wilcoxon by default costs
   statistical power unnecessarily. Make the decision based on the
   QQ-plot and sample size, not out of caution.
:::


## How to Report

::: {.callout-note}
### Reporting Nonparametric Results

**In a Methods section:**

> "Because [outcome] was markedly right-skewed with extreme outliers
> (confirmed by QQ-plot), the Wilcoxon rank-sum test was used to
> compare groups and results are presented as median (IQR). For
> comparisons across three or more groups, the Kruskal-Wallis test was
> used, followed by pairwise Wilcoxon tests with Bonferroni correction."

**In a Results section:**

> "Median [outcome] was X (IQR: A--B) in [group 1] and Y (IQR: C--D)
> in [group 2]; groups did not differ significantly (Wilcoxon
> rank-sum: W = Z, p = P)."
>
> "Bilirubin differed significantly across disease stages (Kruskal-
> Wallis: H(3) = XX.X, p < 0.001). Post-hoc pairwise comparisons
> with Bonferroni correction showed significant differences between
> Stages 1 and 4 (p = 0.001) and Stages 2 and 4 (p = 0.003)."

**Always include:**

- Medians and IQR for each group (not means and SD)
- Test statistic (W for Wilcoxon, H for Kruskal-Wallis with df)
- Exact p-value
- Post-hoc method and correction when multiple comparisons are made
:::


## Further Reading

- @Bland-2015: *An Introduction to Medical Statistics*: clear
  chapters on nonparametric methods written for clinicians
- @Hollander-1999: *Nonparametric Statistical Methods*: comprehensive
  reference
- `?wilcox.test` and `?kruskal.test` in R: built-in help with worked
  examples
