---
title: "Statistics for Omics Data"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(broom)
```

## When Do You Use This? {#sec-omics .callout-tip}

> You run a transcriptomics experiment: 20,000 genes measured in 10 treated and 10 control mouse liver samples. You want to find which genes are differentially expressed. Applying 20,000 t-tests would generate ~1,000 false positives by chance alone. Omics statistics solves the multiple testing problem, borrows information across genes to stabilise variance estimates, and ranks results by a combination of effect size and significance.

## Learning Objectives

After completing this session you will be able to:

- Explain why multiple testing requires correction and apply FDR methods
- Use `limma` or `DESeq2` to analyse differential expression
- Interpret volcano plots and MA plots
- Perform pathway enrichment analysis with `clusterProfiler`
- Understand the key differences between bulk RNA-seq, microarray, and proteomics analysis workflows


## Background: The Multiple Testing Problem

### Why Ordinary p-values Fail at Scale

With $m = 20,000$ tests and $\alpha = 0.05$, the expected number of false positives is $0.05 \times 20,000 = 1,000$ even when no gene is truly differentially expressed. The **family-wise error rate (FWER)** controls the probability of *any* false positive: Bonferroni correction sets $\alpha^* = 0.05 / m$. This is very conservative for omics.

The **false discovery rate (FDR)** controls the *proportion* of discoveries that are false positives ? a more powerful approach for exploratory omics. The Benjamini-Hochberg (BH) procedure adjusts p-values to **q-values**: a q-value of 0.05 means 5% of genes called significant at that threshold are expected to be false positives.

$$\text{q-value}_i = \frac{p_{(i)} \cdot m}{i}$$

(where $p_{(i)}$ are p-values sorted ascending and $m$ is total tests).

### Moderated Tests in Genomics

With small sample sizes (n = 3?6 per group), gene-wise variance estimates are noisy. `limma` (microarray/RNA-seq) and `DESeq2` (RNA-seq counts) stabilise variance estimates by:
- **Empirical Bayes shrinkage:** Borrowing information across genes to shrink each gene's variance toward a common prior
- **Negative binomial model (DESeq2):** Accounting for overdispersion in count data

This produces more powerful and stable differential expression (DE) calls than simple t-tests.


## Example 1: Simulated RNA-seq Differential Expression

We simulate a count matrix and apply a limma-voom workflow.

```{r sim-counts}
set.seed(2024)
n_genes   <- 5000
n_samples <- 20          # 10 treated, 10 control

# Simulate log2 expression (microarray-style for simplicity)
group <- factor(rep(c("Control", "Treated"), each = 10))

# ~200 genes truly DE with effect size 1 log2 unit
true_de     <- sample(n_genes, 200)
effect_size <- rep(0, n_genes)
effect_size[true_de] <- rnorm(200, mean = 1, sd = 0.5)

# Expression matrix: genes x samples
expr <- matrix(
  rnorm(n_genes * n_samples, mean = 8, sd = 1.5),
  nrow = n_genes
)
# Add treatment effect for DE genes
expr[true_de, group == "Treated"] <- expr[true_de, group == "Treated"] + effect_size[true_de]
rownames(expr) <- paste0("Gene_", seq_len(n_genes))

cat("Expression matrix:", nrow(expr), "genes x", ncol(expr), "samples\n")
```

### Simple t-test (naive approach)

```{r naive-ttest}
# 5000 independent t-tests
pvals_naive <- apply(expr, 1, function(x) {
  t.test(x[group == "Treated"], x[group == "Control"])$p.value
})

# How many significant at p < 0.05?
cat("Significant at p < 0.05 (uncorrected):", sum(pvals_naive < 0.05), "\n")
cat("Expected false positives:", round(0.05 * n_genes), "\n")
```

### Applying FDR Correction

```{r fdr}
qvals <- p.adjust(pvals_naive, method = "BH")

# How many at FDR < 5%?
cat("Significant at FDR < 5%:", sum(qvals < 0.05), "\n")
cat("True positives in simulation:", length(true_de), "\n")
```

### Volcano Plot

```{r volcano}
# Compute per-gene fold change and -log10(p)
fc <- apply(expr, 1, function(x) mean(x[group == "Treated"]) - mean(x[group == "Control"]))

de_results <- tibble(
  gene  = rownames(expr),
  fc    = fc,
  pval  = pvals_naive,
  qval  = qvals,
  sig   = qval < 0.05 & abs(fc) > 0.5
)

de_results %>%
  ggplot(aes(x = fc, y = -log10(pval), color = sig)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(values = c("grey70", "tomato"),
                     labels = c("Not significant", "FDR < 5% & |FC| > 0.5")) +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed") +
  geom_hline(yintercept = -log10(max(pvals_naive[qvals < 0.05])),
             linetype = "dashed", color = "steelblue") +
  labs(title = "Volcano plot: differential expression",
       x = "log2 fold change (Treated vs Control)",
       y = "-log10(p-value)", color = "") +
  theme_bw() +
  theme(legend.position = "bottom")
```


## Example 2: Limma Workflow (Empirical Bayes)

```{r limma}
library(limma)

# Design matrix
design <- model.matrix(~ group)

# Fit linear model per gene
fit  <- lmFit(expr, design)
fit2 <- eBayes(fit)     # empirical Bayes moderation

# Extract top DE genes
top_genes <- topTable(fit2, coef = "groupTreated", n = Inf, sort.by = "P") %>%
  tibble::rownames_to_column("gene") %>%
  as_tibble()

head(top_genes, 10)
```

```{r limma-volcano}
top_genes %>%
  mutate(sig = adj.P.Val < 0.05 & abs(logFC) > 0.5) %>%
  ggplot(aes(x = logFC, y = -log10(P.Value), color = sig)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(values = c("grey70", "tomato")) +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed") +
  labs(title = "Volcano plot (limma eBayes)",
       x = "log2 fold change", y = "-log10(p-value)", color = "FDR < 5%") +
  theme_bw()
```

**Comparing naive t-test vs limma:** The limma moderated t-statistics are more stable for genes with low expression variance, and the eBayes step improves ranking of true positives.


## Example 3: Pathway Enrichment

Once you have a gene list, the next question is: which biological pathways are enriched?

```{r enrichment}
#| eval: false
library(clusterProfiler)
library(org.Mm.eg.db)   # mouse gene annotation

# Significant DE genes
sig_genes <- top_genes %>%
  filter(adj.P.Val < 0.05, logFC > 0.5) %>%
  pull(gene)

# Over-representation analysis (ORA) using Gene Ontology
ego <- enrichGO(
  gene         = sig_genes,
  universe     = top_genes$gene,
  OrgDb        = org.Mm.eg.db,
  keyType      = "SYMBOL",
  ont          = "BP",         # Biological Process
  pAdjustMethod = "BH",
  qvalueCutoff  = 0.05
)
dotplot(ego, showCategory = 15)
```

**Gene Set Enrichment Analysis (GSEA):** Instead of a binary DE/non-DE split, GSEA uses the full ranked gene list to test whether genes in a pathway tend to be up- or down-regulated.


## Where to Find Data Like This {.callout-note}

**GEO (Gene Expression Omnibus):** [https://www.ncbi.nlm.nih.gov/geo/](https://www.ncbi.nlm.nih.gov/geo/) ? thousands of publicly available microarray and RNA-seq datasets.

**Bioconductor:** `BiocManager::install("GEOquery")` to download GEO datasets directly into R.

**`pasilla` and `airway` packages:** Standard RNA-seq benchmark datasets included with Bioconductor ? used in DESeq2 and edgeR vignettes.

**`curatedMetagenomicData`:** Processed microbiome datasets ready for R analysis.


## What Can Go Wrong {.callout-warning}

**Batch effects**
Technical variation from different sequencing runs, plates, or RNA extraction batches can dominate biological signal. Visualise PCA before and after correction. Use `ComBat` (from `sva`) or include batch as a covariate in the design matrix.

**Ignoring library size normalisation**
Raw counts differ between samples due to sequencing depth. Always normalise (TMM in edgeR, DESeq2 size factors, or CPM) before comparing samples.

**Using fold change alone**
A 10-fold change is meaningless if expression is near-zero in both groups (high noise, low absolute levels). Always report both fold change and significance. Filter low-expression genes before analysis.

**Misinterpreting enrichment results**
ORA depends on the gene universe (background). If you use all measured genes as universe, results are more stringent than if you use only protein-coding genes. Report your universe and software version.


## Exercises

### Exercise 1 (Guided): FDR Simulation

1. Simulate 10,000 independent t-tests where the null is true for all genes (`rnorm` both groups with same mean).
2. Count how many are significant at p < 0.05.
3. Apply BH correction. How many survive FDR < 5%?
4. Repeat with 500 truly DE genes (mean shift = 1 in treated). What is the sensitivity and FDR of the BH-corrected calls?

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r}
set.seed(2024)
m  <- 10000
n  <- 10     # samples per group
m_de <- 500  # true positives

# Simulate
ctrl <- matrix(rnorm(m * n, 0, 1), m, n)
trt  <- matrix(rnorm(m * n, 0, 1), m, n)
trt[1:m_de, ] <- trt[1:m_de, ] + 1    # true DE

pv <- sapply(seq_len(m), function(i) t.test(trt[i,], ctrl[i,])$p.value)
qv <- p.adjust(pv, "BH")

cat("Sig at p < 0.05 (uncorrected):", sum(pv < 0.05), "\n")
cat("Sig at FDR < 5%:", sum(qv < 0.05), "\n")
cat("True positives recovered:", sum(qv[1:m_de] < 0.05), "/", m_de, "\n")
cat("False positives:", sum(qv[(m_de+1):m] < 0.05), "\n")
```
:::

### Exercise 2 (Semi-guided): Apply Limma to the Simulation

Using the `expr` matrix from Example 1:

1. Build the design matrix for `group`.
2. Fit `lmFit()` and apply `eBayes()`.
3. Compare the number of DE genes (FDR < 5%) from limma vs the naive t-test.
4. Plot the moderated t-statistics vs naive t-statistics for the top 50 genes by naive p-value.

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r}
design2 <- model.matrix(~ group)
fit_ex  <- lmFit(expr, design2)
fit_ex2 <- eBayes(fit_ex)
tbl_ex  <- topTable(fit_ex2, coef = 2, n = Inf, sort.by = "P") %>%
  tibble::rownames_to_column("gene") %>%
  as_tibble()

cat("Limma FDR < 5%:", sum(tbl_ex$adj.P.Val < 0.05), "\n")
cat("Naive FDR < 5%:", sum(qvals < 0.05), "\n")
```
:::

### Exercise 3 (Open-ended)

Download a public RNA-seq dataset from GEO using `GEOquery` (e.g., `GSE93819` ? mouse liver after high-fat diet). Apply a full differential expression workflow:

1. Download and load the data.
2. Filter low-expression genes (at least 10 counts in at least 3 samples).
3. Normalise with `DESeq2` or `edgeR`.
4. Run DE analysis.
5. Produce a volcano plot and report the top 20 DE genes.
6. Perform Gene Ontology enrichment and interpret the biological pathways.


## Comprehension Check

1. You run 15,000 t-tests and apply Bonferroni correction. Three genes remain significant. A colleague says this is "too conservative" for omics. What is the alternative, and what does it control?
2. A gene has fold change = 8 and p = 0.8. Should you report it as significant?
3. PCA of your RNA-seq samples shows two clear clusters that match the sequencing batch, not the biological groups. What is the problem, and how do you fix it?
4. Your enrichment analysis returns "ribosome biogenesis" as the top pathway. Is this a meaningful result?
5. A paper reports 2,000 DE genes at FDR < 5%. What does this mean about the likely number of false positives in their list?

::: {.callout-note collapse="true"}
### Answers
1. The Bonferroni correction controls the family-wise error rate (FWER) ? the probability that *any* test produces a false positive. For omics, this is very conservative and has low power. The alternative is the **false discovery rate (FDR)**, controlled by the Benjamini-Hochberg procedure. FDR allows a small *proportion* of false positives among significant results (e.g., 5% FDR means ~5% of called DE genes are expected to be false positives). BH is more powerful and more appropriate for exploratory omics analyses.
2. No. p = 0.8 means the fold change is not statistically distinguishable from noise with the available sample size. A large fold change with a non-significant p-value typically reflects low statistical power ? either the gene is highly variable or the sample size is too small. Report both fold change and significance; do not highlight genes that are not statistically significant.
3. This is a **batch effect** ? technical variation from sequencing batches dominates the biological signal. Without correction, any DE analysis will be confounded: genes that vary between batches will appear differentially expressed between biological groups if batches are correlated with groups. Fixes: (1) include batch as a covariate in the design matrix (`~ batch + group`); (2) apply `ComBat` batch correction if groups are not perfectly confounded with batches; (3) if batches are perfectly confounded with groups, the study is uninterpretable and must be repeated.
4. Ribosome biogenesis appearing as the top pathway may be meaningful (rapidly proliferating treated cells produce more ribosomes) or may be an artefact (ribosomal genes are often highly expressed and variable, dominating enrichment analyses). Check: are ribosomal genes actually in your DE gene list, or are they just abundant in your background? Consider removing ribosomal genes from the universe or interpreting with caution.
5. At FDR 5%, approximately 5% ? 2000 = 100 genes in the list are expected to be false positives. The FDR is a statement about the long-run proportion of false positives among discoveries, not a guarantee about this specific list. Importantly, the FDR does not indicate the direction or magnitude of effects ? the 1,900 true positives may range widely in effect size, and many may be biologically trivial despite statistical significance.
:::

## How to Report

::: {.callout-note}
### Reporting Omics Statistics Results

**In a Methods section:**

> "Differential expression analysis was performed using `limma`/`DESeq2` (cite). Genes with an FDR-adjusted p-value < 0.05 and |log₂ fold-change| ≥ 1 were considered differentially expressed. Multiple testing correction used the Benjamini–Hochberg method."

**In a Results section:**

> "Of XXXX genes tested, XXX were significantly differentially expressed (FDR < 0.05, |log₂FC| ≥ 1): XXX up-regulated and XXX down-regulated in [condition A] relative to [condition B]."

**Always report:**
- The statistical method and R package with version
- The FDR threshold used (and why, if not 0.05)
- Both the statistical threshold (FDR) and the biological threshold (fold-change cutoff)
- Number of features tested (important for interpreting multiple testing correction)

**Common reporting errors:**
- Reporting raw (unadjusted) p-values without FDR correction in a high-throughput analysis
- Using an FDR threshold of 0.1 or 0.2 without acknowledging the higher false positive rate
- Treating an FDR-adjusted p-value as if it were a Bonferroni-corrected p-value
- Omitting the fold-change threshold — statistical significance alone does not imply biological relevance
:::

## Further Reading

- @Ritchie-2015 --- *limma powers differential expression analyses for RNA-sequencing and microarray studies* (NAR)
- @Love-2014 --- *DESeq2: moderated estimation of fold change and dispersion for RNA-seq data* (Genome Biology)
- @Benjamini-1995 --- *Controlling the False Discovery Rate* (JRSS-B ? the original BH paper)
- Bioconductor workflows: [https://bioconductor.org/packages/release/BiocViews.html#___Workflow](https://bioconductor.org/packages/release/BiocViews.html#___Workflow)
