---
title: "Power Analysis and Sample Size"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(pwr)
```

::: {.callout-note}
### Session at a Glance

**Total core time: about 120 minutes (~2 hours).**

| Section | Time | Type |
|---|---|---|
| When Do You Use This? + Learning Objectives | 5 min | Reading |
| The Key Idea: A Trade-Off You Cannot Escape | 10 min | Reading |
| Background: Power and Its Components | 15 min | Reading |
| Example 1: Mouse Experiment (Two-Sample t-test) | 25 min | Worked example |
| Example 2: Clinical Trial (Two Proportions) | 15 min | Worked example |
| What Can Go Wrong | 10 min | Reading |
| Exercises 1 & 2 | 30 min | Practice |
| Comprehension Check | 10 min | Self-test |

In a taught course, this session fits one teaching block. For
self-paced study, split it into two sittings, e.g. the Key Idea,
Background, and Example 1 first, Example 2 onward later. Exercise 3 is
open-ended and works best as a follow-up using a study you are
actually planning.
:::

::: {.callout-tip}
### How to Use the Code in This Session

Every code chunk below is written so you can copy it into your R
console (or an R script) and run it yourself, in the order shown.
Where you see a "Run It Yourself" box, stop and run the code above it
before reading the explanation that follows.
:::

## When Do You Use This? {#sec-power}

::: {.callout-tip}
You are designing a mouse experiment to test whether a new drug reduces tumour volume. You expect a 20% reduction and want to know: *how many mice do I need?* Power analysis answers this **before you collect a single data point**, and prevents you from running an underpowered study where a real effect would go undetected.
:::

::: {.callout-note collapse="true"}
### Teacher Note

Ask the room: has anyone designed an experiment or study and picked a
sample size without a formal power calculation (e.g. "what we could
afford" or "what the lab always uses")? Keep that example in mind --
at the end of Example 1 you can return to it and ask whether a power
calculation would have suggested a larger or smaller n.
:::

## Learning Objectives

After completing this session you will be able to:

- Define statistical power and explain the relationship between power, sample size, effect size, and significance level
- Calculate required sample sizes for t-tests and proportion comparisons in R
- Compute Cohen's d and Cohen's h as standardised effect sizes
- Produce a power curve and interpret it
- Identify the key planning decisions and their consequences for study design


## The Key Idea: A Trade-Off You Cannot Escape

*Estimated time: ~10 minutes (reading)*

Before the formulas, here is the honest picture. You are designing a study where you will later run a hypothesis test. That test can make two kinds of mistakes (both discussed in the [hypothesis testing session](hypothesis-testing.qmd)):

- **False positive (Type I error):** You declare an effect when there isn't one. Controlled by α.
- **False negative (Type II error):** You miss a real effect. Its probability is β. Power = 1 − β.

The brutal reality: you cannot simultaneously make both errors rare without collecting more data. If you want to be very sure you won't miss a real effect (high power), you need a large sample. If your sample is small, you will miss real effects, no matter how sophisticated your analysis.

Power analysis makes this trade-off explicit *before* you run the study, so you can make an informed decision about how many participants or animals to recruit.

::: {.callout-note collapse="true"}
### Teacher Note

Put it to the room directly: "If I want α = 0.01 *and* power = 99%,
what's the catch?" Lead them to the answer -- a much larger sample
size than for α = 0.05 and power = 80%. There is no free lunch: the
only lever that relaxes the trade-off without raising α or accepting
lower power is collecting more data (or finding a way to reduce
variability, e.g. a paired design).
:::

## Background: Power and Its Components

*Estimated time: ~15 minutes (reading)*

**Statistical power** ($1 - \beta$) is the probability of detecting a true effect when one exists. A study with 80% power will correctly reject $H_0$ in 80% of repeated experiments when the true effect is real.

Four quantities are connected; fix three and you can determine the fourth:

| Quantity | Symbol | Typical choice |
|---|---|---|
| Significance level | $\alpha$ | 0.05 |
| Power | $1-\beta$ | 0.80 |
| Effect size | $d$, $h$, etc. | From pilot data or literature |
| Sample size | $n$ | **What you want to find** |

**Effect size:** The standardised difference between groups, removing units.

For a two-sample t-test:
$$d = \frac{\mu_1 - \mu_2}{\sigma_{\text{pooled}}}$$

Cohen's rough benchmarks: small $d = 0.2$, medium $d = 0.5$, large $d = 0.8$.

**What happens when you change each quantity:**

| Change | Effect on required n |
|---|---|
| More stringent $\alpha$ (0.05 → 0.01) | ↑ n needed |
| Higher power (80% → 90%) | ↑ n needed |
| Smaller effect size | ↑ n needed |
| One-sided vs two-sided test | ↓ n if one-sided |

::: {.callout-note collapse="true"}
### Teacher Note

Before moving to Example 1, ask students to predict: if the effect
size is halved (e.g. d = 0.75 becomes d = 0.375), does the required n
roughly double, or roughly quadruple? Because n is roughly
proportional to $1/d^2$ for a two-sample t-test, halving d roughly
quadruples n. This sets up the sensitivity analysis later in this
example.
:::

## Example 1: Mouse Experiment (Two-Sample t-test)

*Estimated time: ~25 minutes (worked example)*

You are testing whether a new drug reduces tumour volume. From a pilot experiment
(n = 5 per group), you estimate:

- Expected reduction in mean tumour volume: **150 mm3**
- Pooled SD from pilot data: **200 mm3** (so Cohen's d = 150/200 = 0.75)
- alpha = 0.05, power = 0.80, two-sided

```{r power-ttest}
#| eval: true

# Base R
power.t.test(
  delta     = 150,    # expected difference in means (mm3)
  sd        = 200,    # pooled SD from pilot
  sig.level = 0.05,
  power     = 0.80,
  type      = "two.sample",
  alternative = "two.sided"
)
```

```{r power-pwr}
#| eval: true

library(pwr)

d <- 150 / 200   # Cohen's d
pwr.t.test(
  d         = d,
  sig.level = 0.05,
  power     = 0.80,
  type      = "two.sample",
  alternative = "two.sided"
)
```

Both functions agree: approximately **29 mice per group** (round up). Always round up.

::: {.callout-tip}
### Run It Yourself

Run the two chunks above. Both `power.t.test()` and `pwr.t.test()`
should give `n ≈ 28.9` for d = 0.75, α = 0.05, power = 0.80,
two-sided -- agreement is expected, since `pwr.t.test()` is just
another implementation of the same formula. Round up to **29 mice per
group** (58 mice total).
:::

### Power Curve: n vs Power

Plot power as a function of sample size to see the trade-off and justify your choice:

```{r power-curve}
#| eval: true

n_seq <- seq(5, 80, by = 2)

power_vals <- map_dbl(n_seq, ~ pwr.t.test(
  n         = .x,
  d         = d,
  sig.level = 0.05,
  type      = "two.sample",
  alternative = "two.sided"
)$power)

tibble(n = n_seq, power = power_vals) %>%
  ggplot(aes(x = n, y = power)) +
  geom_line(linewidth = 1, colour = "steelblue") +
  geom_hline(yintercept = 0.80, linetype = "dashed", colour = "tomato") +
  annotate("text", x = 78, y = 0.81, label = "80% power", hjust = 1, colour = "tomato") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "Power curve: two-sample t-test (d = 0.75, mouse tumour experiment)",
    x = "n per group", y = "Power"
  ) +
  theme_bw()
```

::: {.callout-tip}
### Run It Yourself

Before running the chunk above, predict: at roughly what value of n
will the curve cross the dashed 80% line? Then run it and check --
it should cross at **n ≈ 29**, matching the result from `power.t.test()`
and `pwr.t.test()` above.
:::

### Sensitivity Analysis: What If the Pilot SD Was Underestimated?

Pilot studies often underestimate variability. Show how required n grows if SD is larger:

```{r sensitivity}
#| eval: true

# d decreases as SD grows (fixed delta = 150)
sd_seq <- seq(150, 350, by = 25)

n_required <- map_dbl(sd_seq, ~ ceiling(pwr.t.test(
  d         = 150 / .x,
  sig.level = 0.05,
  power     = 0.80,
  type      = "two.sample"
)$n))

tibble(sd = sd_seq, n_per_group = n_required) %>%
  ggplot(aes(x = sd, y = n_per_group)) +
  geom_line(linewidth = 1, colour = "steelblue") +
  geom_point(size = 3) +
  geom_vline(xintercept = 200, linetype = "dashed", colour = "tomato") +
  annotate("text", x = 202, y = 200, label = "Pilot estimate", hjust = 0, colour = "tomato") +
  labs(
    title = "n required as SD increases (delta = 150, alpha = 0.05, power = 80%)",
    x = "Pooled SD (mm3)", y = "n per group"
  ) +
  theme_bw()
```

If the true SD is 300 (not 200), you need more than twice as many mice. **Be conservative with your SD estimate.**


## Example 2: Clinical Trial (Two Proportions)

*Estimated time: ~15 minutes (worked example)*

::: {.callout-note collapse="true"}
### Teacher Note

This example switches from a continuous outcome (tumour volume) to a
binary one (treatment response). Ask the room: do they expect the
required sample size for a 40% vs 65% response rate to be smaller or
larger than the 29 mice per group needed in Example 1? (It will turn
out to be larger -- proportions generally need bigger samples than a
moderate-to-large continuous effect size like d = 0.75.)
:::

You are planning a clinical trial of a new treatment. The control response rate is 40%;
you want to detect an increase to 65% with 80% power at alpha = 0.05.

```{r power-prop}
#| eval: true

# Cohen's h for proportions
h <- ES.h(0.65, 0.40)
cat("Cohen's h:", round(h, 3), "\n")

# Sample size
pwr.2p.test(
  h         = h,
  sig.level = 0.05,
  power     = 0.80,
  alternative = "two.sided"
)
```

```{r power-prop-base}
#| eval: true

# Cross-check with base R
power.prop.test(
  p1        = 0.40,
  p2        = 0.65,
  sig.level = 0.05,
  power     = 0.80
)
```

::: {.callout-tip}
### Run It Yourself

Run the chunks above. You should get `Cohen's h ≈ 0.506`, and both
`pwr.2p.test()` and `power.prop.test()` give `n ≈ 61.3--61.4` per
group -- round up to **62 per group (124 total)**. Notice this is
more than double the 29 mice per group needed in Example 1, even
though both effects (d = 0.75, h ≈ 0.51) would be called "medium to
large" -- proportions and means are not directly comparable on the
same scale.
:::

### Power Curve for Proportions

```{r power-prop-curve}
#| eval: true

n_seq2 <- seq(5, 100, by = 5)

pow2 <- map_dbl(n_seq2, ~ pwr.2p.test(
  n         = .x,
  h         = h,
  sig.level = 0.05,
  alternative = "two.sided"
)$power)

tibble(n = n_seq2, power = pow2) %>%
  ggplot(aes(x = n, y = power)) +
  geom_line(linewidth = 1, colour = "steelblue") +
  geom_hline(yintercept = 0.80, linetype = "dashed", colour = "tomato") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "Power curve: two-proportion test (h = 0.51, 40% vs 65% response)",
    x = "n per group", y = "Power"
  ) +
  theme_bw()
```


::: {.callout-note}
## Where to Find Data Like This

**Your own pilot data:** The best source for SD estimates. Even n = 5 per group gives a useful starting point.

**`survival::pbc`:** A published clinical dataset: useful for deriving realistic SD and mean estimates for planning a new study in a similar patient population (see Exercise 1).

**`pwr` package:** `pwr.t.test()`, `pwr.2p.test()`, `pwr.chisq.test()`, `pwr.anova.test()`: covers all common designs.

**G\*Power software:** Free standalone tool (gpower.hhu.de): popular in clinical and behavioural research, good for cross-checking R results.
:::


::: {.callout-warning}
## What Can Go Wrong

*Estimated time: ~10 minutes (reading)*

**Using a too-optimistic effect size.**
Published effect sizes are often inflated by publication bias. If you power your study using an unrealistically large d, you will be underpowered when the true effect is smaller. Be conservative, or use the lower bound of a pilot confidence interval.

**Powering for the wrong test.**
If you plan to use a Wilcoxon test (because your data will be non-normal), power calculations based on `power.t.test()` are approximate. Simulation-based approaches are more accurate for nonparametric tests.

**Forgetting to account for dropout or missing data.**
Calculate the required $n$, then inflate by an expected dropout fraction. If you expect 15% dropout, divide your required $n$ by 0.85.

**Multiple primary outcomes.**
If you have several primary outcomes, adjust $\alpha$ (e.g., Bonferroni) before running the power calculation. Powering for each outcome at $\alpha = 0.05$ without correction is overly optimistic.

**Confusing power for a one-sided vs two-sided test.**
A one-sided test requires fewer participants for the same power. Only use a one-sided test if you have a strong, pre-registered directional hypothesis and would not care about an effect in the other direction.
:::

::: {.callout-note collapse="true"}
### Teacher Note

Quick true/false check, drawing on the points above:

- "A Cohen's d of 1.2 reported in a single published paper with n = 10 per group is a safe basis for your power calculation." (**False** -- small studies tend to overestimate effect sizes; be conservative.)
- "If your power calculation gives n = 25 per group and you expect 15% dropout, dividing 25 by 0.85 tells you how many to recruit per group." (**True** -- about 30 per group.)
- "A one-sided test needs a larger sample size than a two-sided test for the same power." (**False** -- a one-sided test needs fewer, but is only valid with a strong, pre-registered directional hypothesis.)
:::

## Exercises

::: {.callout-note collapse="true"}
### Teacher Note

Pair students up for these exercises. Exercise 1 uses the PBC dataset
seen in earlier sessions to illustrate a common real-world situation:
no pilot data of your own, but a published dataset from a similar
population. Before running the solution, ask pairs to predict whether
the required n will be closer to the 29 from Example 1 or the 62 from
Example 2 -- it turns out to be larger than both, which is a useful
surprise to discuss (smaller effect size on the log-bilirubin scale
than either worked example).
:::

### Exercise 1 (Guided): Using Existing Data to Plan a New Study

*Estimated time: ~15 minutes*

A common situation: you want to plan a new clinical study but have no pilot data. You can use a published dataset with a similar patient population to estimate the pooled SD.

Using `survival::pbc`, estimate the pooled SD of log(bilirubin) in the two treatment groups. Then calculate the sample size needed to detect a 0.4-unit difference on the log scale.

1. Load `pbc`, filter to randomised patients, compute pooled SD of `log(bili)` by treatment group.
2. Compute Cohen's d assuming delta = 0.4.
3. Run `pwr.t.test()` at alpha = 0.05, power = 0.80.
4. Plot the power curve. At what n does power first exceed 80%?

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r}
pbc <- survival::pbc %>%
  as_tibble() %>%
  janitor::clean_names() %>%
  mutate(trt = factor(trt))

pbc_rand <- pbc %>% filter(!is.na(trt), !is.na(bili))

sd_vals <- pbc_rand %>%
  group_by(trt) %>%
  summarise(sd = sd(log(bili)), n = n())
print(sd_vals)

# Pooled SD
n1 <- sd_vals$n[1]; s1 <- sd_vals$sd[1]
n2 <- sd_vals$n[2]; s2 <- sd_vals$sd[2]
sd_pool <- sqrt(((n1 - 1) * s1^2 + (n2 - 1) * s2^2) / (n1 + n2 - 2))
d_bili  <- 0.4 / sd_pool
cat("Pooled SD:", round(sd_pool, 3), " Cohen's d:", round(d_bili, 3), "\n")

# Sample size
pwr.t.test(d = d_bili, sig.level = 0.05, power = 0.80, type = "two.sample")

# Power curve
n_seq <- seq(5, 150, by = 5)
pw <- map_dbl(n_seq, ~ pwr.t.test(n = .x, d = d_bili, sig.level = 0.05,
                                   type = "two.sample")$power)
tibble(n = n_seq, power = pw) %>%
  ggplot(aes(n, power)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_hline(yintercept = 0.8, linetype = "dashed", colour = "tomato") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "n per group", y = "Power") + theme_bw()
```

Run the chunk above. You should get pooled SD ≈ 1.033 and Cohen's d
≈ 0.387 -- a smaller effect size than either worked example, which is
why `pwr.t.test()` returns `n ≈ 105.7`: about **106 per group**. The
power curve (plotted up to n = 150 here, since the required n is
above 100) crosses the 80% line at that same value, n ≈ 106.
:::

### Exercise 2 (Semi-guided): Mouse Tumour Response Experiment

*Estimated time: ~15 minutes*

You are planning an in vivo experiment. The control group is expected to have a 30% tumour response rate; the treated group 55%.

1. Compute Cohen's h using `ES.h(0.55, 0.30)`.
2. Calculate n per group at alpha = 0.05, power = 0.80 using `pwr.2p.test()`.
3. You expect 20% cage dropout. How many mice should you enrol per group?
4. How does the required n change if you increase power to 90%?

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r}
h_mouse <- ES.h(0.55, 0.30)
cat("Cohen's h:", round(h_mouse, 3), "\n")

# n at 80%
res80 <- pwr.2p.test(h = h_mouse, sig.level = 0.05, power = 0.80)
n80 <- ceiling(res80$n)
cat("n per group (80% power):", n80, "\n")

# With 20% dropout
cat("Enrol per group:", ceiling(n80 / 0.80), "\n")

# At 90% power
res90 <- pwr.2p.test(h = h_mouse, sig.level = 0.05, power = 0.90)
cat("n per group (90% power):", ceiling(res90$n), "\n")
```
:::

### Exercise 3 (Open-ended)

*Estimated time: 15--30 minutes*

For your next planned experiment or study, define:

1. The primary outcome and the comparison you will make
2. Expected means/proportions and SD in each group (from pilot data or literature)
3. $\alpha$ and desired power

Run the power calculation and produce a power curve. Write one paragraph suitable for an ethics committee submission justifying your sample size.


## Comprehension Check

*Estimated time: ~10 minutes (self-test)*

1. What is statistical power? What does "80% power" mean in plain language?
2. You calculate that you need n = 25 per group. You expect 10% dropout. How many participants should you recruit?
3. A researcher uses a Cohen's d of 1.2 from a single published paper with n = 10 per group to power their study. What is the concern?
4. All else equal, does a one-sided test require more or fewer participants than a two-sided test for the same power?
5. Your power calculation gives n = 12 per group, but you can only recruit n = 8. What are your options?

::: {.callout-note collapse="true"}
### Answers
1. Statistical power is the probability of correctly rejecting the null hypothesis when the alternative is true. "80% power" means that if the true effect is as assumed, your study will detect it (i.e., return a significant result) in 80% of repeated experiments of the same design.
2. Divide by the expected retention fraction: $25 / 0.90 = 28$, so recruit 28 per group (round up).
3. Single small studies overestimate effect sizes due to sampling variability and publication bias. Using d = 1.2 from a study with n = 10 is likely to be inflated. The true d may be considerably smaller, and the actual study will be underpowered.
4. A one-sided test requires fewer participants. It allocates the full a to one tail, making the critical threshold easier to reach - but only valid when you have a strong directional hypothesis and would not act on an effect in the opposite direction.
5. Options: (1) Accept lower power (< 80%) and acknowledge the limitation; (2) increase the minimum effect size you are trying to detect (e.g., only detect a large effect); (3) decrease α (less stringent threshold, which increases power); (4) use a paired design or block on a covariate to reduce within-group variability; (5) pool data across experiments.
:::

## How to Report

::: {.callout-note}
### Reporting a Power Calculation

**In a Methods section (grant application or ethics submission):**

> "Sample size was calculated a priori using the `pwr` package in R (v4.x). Assuming an effect size of Cohen's *d* = 0.5 (medium effect), α = 0.05 (two-sided), and 80% power, a minimum of XX participants per group (YY total) is required. Effect size was estimated from [pilot data / prior literature (cite)]. A 10% dropout allowance was added, giving a final target of ZZ participants per group."

**Key elements to include:**
- Software or formula used for the calculation
- Assumed effect size and its justification (pilot data or literature)
- α level and whether one- or two-sided
- Target power (usually 80% or 90%)
- Any attrition/dropout allowance added

**Common reporting errors:**
- Using a "medium" Cohen's *d* = 0.5 without clinical justification: reviewers will question this
- Reporting only the total N, not per-group N
- Not reporting the assumed effect size  -  makes the calculation unverifiable
- Post-hoc power calculations on observed effect sizes  -  these are misleading and now widely discouraged
:::

## Further Reading

- @Cohen-1988: *Statistical Power Analysis for the Behavioral Sciences* - the original reference for effect sizes and power tables
- @Bland-2015: Sample size chapters in *An Introduction to Medical Statistics*
- `vignette("pwr-vignette", package = "pwr")` - comprehensive `pwr` package tutorial
- G*Power manual at gpower.hhu.de
