---
title: "Power Analysis and Sample Size"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(pwr)
```

## When Do You Use This? {#sec-power .callout-tip}

> You are designing a mouse experiment to test whether a new drug reduces tumour volume. You expect a 20% reduction and want to know: *how many mice do I need?* Power analysis answers this before you collect a single data point ? and prevents you from running an underpowered study where a real effect would go undetected.

## Learning Objectives

After completing this session you will be able to:

- Define statistical power and explain the relationship between power, sample size, effect size, and significance level
- Calculate required sample sizes for t-tests and proportion comparisons in R
- Compute Cohen's d and Cohen's h as standardised effect sizes
- Produce a power curve and interpret it
- Identify the key planning decisions and their consequences for study design


## Background: Power and Its Components

**Statistical power** ($1 - \beta$) is the probability of detecting a true effect when one exists. A study with 80% power will correctly reject $H_0$ in 80% of repeated experiments when the true effect is real.

Four quantities are connected; fix three and you can determine the fourth:

| Quantity | Symbol | Typical choice |
|---|---|---|
| Significance level | $\alpha$ | 0.05 |
| Power | $1-\beta$ | 0.80 |
| Effect size | $d$, $h$, etc. | From pilot data or literature |
| Sample size | $n$ | **What you want to find** |

**Effect size:** The standardised difference between groups, removing units.

For a two-sample t-test:
$$d = \frac{\mu_1 - \mu_2}{\sigma_{\text{pooled}}}$$

Cohen's rough benchmarks: small $d = 0.2$, medium $d = 0.5$, large $d = 0.8$.

**What happens when you change each quantity:**

| Change | Effect on required n |
|---|---|
| More stringent $\alpha$ (0.05 ? 0.01) | ? n needed |
| Higher power (80% ? 90%) | ? n needed |
| Smaller effect size | ? n needed |
| One-sided vs two-sided test | ? n if one-sided |


## Example 1: Mouse Experiment (Two-Sample t-test)

You are testing whether a new drug reduces tumour volume. From a pilot experiment
(n = 5 per group), you estimate:

- Expected reduction in mean tumour volume: **150 mm3**
- Pooled SD from pilot data: **200 mm3** (so Cohen's d = 150/200 = 0.75)
- alpha = 0.05, power = 0.80, two-sided

```{r power-ttest}
#| eval: true

# Base R
power.t.test(
  delta     = 150,    # expected difference in means (mm3)
  sd        = 200,    # pooled SD from pilot
  sig.level = 0.05,
  power     = 0.80,
  type      = "two.sample",
  alternative = "two.sided"
)
```

```{r power-pwr}
#| eval: true

library(pwr)

d <- 150 / 200   # Cohen's d
pwr.t.test(
  d         = d,
  sig.level = 0.05,
  power     = 0.80,
  type      = "two.sample",
  alternative = "two.sided"
)
```

Both functions agree: approximately **29 mice per group** (round up). Always round up.

### Power Curve: n vs Power

Plot power as a function of sample size to see the trade-off and justify your choice:

```{r power-curve}
#| eval: true

n_seq <- seq(5, 80, by = 2)

power_vals <- map_dbl(n_seq, ~ pwr.t.test(
  n         = .x,
  d         = d,
  sig.level = 0.05,
  type      = "two.sample",
  alternative = "two.sided"
)$power)

tibble(n = n_seq, power = power_vals) |>
  ggplot(aes(x = n, y = power)) +
  geom_line(linewidth = 1, colour = "steelblue") +
  geom_hline(yintercept = 0.80, linetype = "dashed", colour = "tomato") +
  annotate("text", x = 78, y = 0.81, label = "80% power", hjust = 1, colour = "tomato") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "Power curve: two-sample t-test (d = 0.75, mouse tumour experiment)",
    x = "n per group", y = "Power"
  ) +
  theme_bw()
```

### Sensitivity Analysis: What If the Pilot SD Was Underestimated?

Pilot studies often underestimate variability. Show how required n grows if SD is larger:

```{r sensitivity}
#| eval: true

# d decreases as SD grows (fixed delta = 150)
sd_seq <- seq(150, 350, by = 25)

n_required <- map_dbl(sd_seq, ~ ceiling(pwr.t.test(
  d         = 150 / .x,
  sig.level = 0.05,
  power     = 0.80,
  type      = "two.sample"
)$n))

tibble(sd = sd_seq, n_per_group = n_required) |>
  ggplot(aes(x = sd, y = n_per_group)) +
  geom_line(linewidth = 1, colour = "steelblue") +
  geom_point(size = 3) +
  geom_vline(xintercept = 200, linetype = "dashed", colour = "tomato") +
  annotate("text", x = 202, y = 200, label = "Pilot estimate", hjust = 0, colour = "tomato") +
  labs(
    title = "n required as SD increases (delta = 150, alpha = 0.05, power = 80%)",
    x = "Pooled SD (mm3)", y = "n per group"
  ) +
  theme_bw()
```

If the true SD is 300 (not 200), you need more than twice as many mice. **Be conservative with your SD estimate.**


## Example 2: Clinical Trial (Two Proportions)

You are planning a clinical trial of a new treatment. The control response rate is 40%;
you want to detect an increase to 65% with 80% power at alpha = 0.05.

```{r power-prop}
#| eval: true

# Cohen's h for proportions
h <- ES.h(0.65, 0.40)
cat("Cohen's h:", round(h, 3), "\n")

# Sample size
pwr.2p.test(
  h         = h,
  sig.level = 0.05,
  power     = 0.80,
  alternative = "two.sided"
)
```

```{r power-prop-base}
#| eval: true

# Cross-check with base R
power.prop.test(
  p1        = 0.40,
  p2        = 0.65,
  sig.level = 0.05,
  power     = 0.80
)
```

### Power Curve for Proportions

```{r power-prop-curve}
#| eval: true

n_seq2 <- seq(5, 100, by = 5)

pow2 <- map_dbl(n_seq2, ~ pwr.2p.test(
  n         = .x,
  h         = h,
  sig.level = 0.05,
  alternative = "two.sided"
)$power)

tibble(n = n_seq2, power = pow2) |>
  ggplot(aes(x = n, y = power)) +
  geom_line(linewidth = 1, colour = "steelblue") +
  geom_hline(yintercept = 0.80, linetype = "dashed", colour = "tomato") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "Power curve: two-proportion test (h = 0.51, 40% vs 65% response)",
    x = "n per group", y = "Power"
  ) +
  theme_bw()
```


## Where to Find Data Like This {.callout-note}

**Your own pilot data:** The best source for SD estimates. Even n = 5 per group gives a useful starting point.

**`survival::pbc`:** A published clinical dataset: useful for deriving realistic SD and mean estimates for planning a new study in a similar patient population (see Exercise 1).

**`pwr` package:** `pwr.t.test()`, `pwr.2p.test()`, `pwr.chisq.test()`, `pwr.anova.test()`: covers all common designs.

**G\*Power software:** Free standalone tool (gpower.hhu.de): popular in clinical and behavioural research, good for cross-checking R results.


## What Can Go Wrong {.callout-warning}

**Using a too-optimistic effect size**
Published effect sizes are often inflated by publication bias. If you power your study using an unrealistically large d, you will be underpowered when the true effect is smaller. Be conservative or use the lower bound of a pilot confidence interval.

**Powering for the wrong test**
If you plan to use a Wilcoxon test (because your data will be non-normal), power calculations based on `power.t.test()` are approximate. The `pwr` package and simulation-based approaches are more accurate for nonparametric tests.

**Forgetting to account for dropout or missing data**
Calculate the required $n$, then inflate by an expected dropout fraction. If you expect 15% dropout, divide your required $n$ by 0.85.

**Multiple primary outcomes**
If you have several primary outcomes, adjust $\alpha$ (e.g., Bonferroni) before running the power calculation. Powering for each outcome at $\alpha = 0.05$ without correction is overly optimistic.

**Confusing power for a one-sided vs two-sided test**
A one-sided test requires fewer participants for the same power. Only use a one-sided test if you have a strong, pre-registered directional hypothesis and would not care about an effect in the other direction.


## Exercises

### Exercise 1 (Guided): Using Existing Data to Plan a New Study

A common situation: you want to plan a new clinical study but have no pilot data. You can use a published dataset with a similar patient population to estimate the pooled SD.

Using `survival::pbc`, estimate the pooled SD of log(bilirubin) in the two treatment groups. Then calculate the sample size needed to detect a 0.4-unit difference on the log scale.

1. Load `pbc`, filter to randomised patients, compute pooled SD of `log(bili)` by treatment group.
2. Compute Cohen's d assuming delta = 0.4.
3. Run `pwr.t.test()` at alpha = 0.05, power = 0.80.
4. Plot the power curve. At what n does power first exceed 80%?

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r}
pbc <- survival::pbc |>
  as_tibble() |>
  janitor::clean_names() |>
  mutate(trt = factor(trt))

pbc_rand <- pbc |> filter(!is.na(trt), !is.na(bili))

sd_vals <- pbc_rand |>
  group_by(trt) |>
  summarise(sd = sd(log(bili)), n = n())
print(sd_vals)

# Pooled SD
n1 <- sd_vals$n[1]; s1 <- sd_vals$sd[1]
n2 <- sd_vals$n[2]; s2 <- sd_vals$sd[2]
sd_pool <- sqrt(((n1 - 1) * s1^2 + (n2 - 1) * s2^2) / (n1 + n2 - 2))
d_bili  <- 0.4 / sd_pool
cat("Pooled SD:", round(sd_pool, 3), " Cohen's d:", round(d_bili, 3), "\n")

# Sample size
pwr.t.test(d = d_bili, sig.level = 0.05, power = 0.80, type = "two.sample")

# Power curve
n_seq <- seq(5, 100, by = 5)
pw <- map_dbl(n_seq, ~ pwr.t.test(n = .x, d = d_bili, sig.level = 0.05,
                                   type = "two.sample")$power)
tibble(n = n_seq, power = pw) |>
  ggplot(aes(n, power)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_hline(yintercept = 0.8, linetype = "dashed", colour = "tomato") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "n per group", y = "Power") + theme_bw()
```
:::

### Exercise 2 (Semi-guided): Mouse Tumour Response Experiment

You are planning an in vivo experiment. The control group is expected to have a 30% tumour response rate; the treated group 55%.

1. Compute Cohen's h using `ES.h(0.55, 0.30)`.
2. Calculate n per group at alpha = 0.05, power = 0.80 using `pwr.2p.test()`.
3. You expect 20% cage dropout. How many mice should you enrol per group?
4. How does the required n change if you increase power to 90%?

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r}
h_mouse <- ES.h(0.55, 0.30)
cat("Cohen's h:", round(h_mouse, 3), "\n")

# n at 80%
res80 <- pwr.2p.test(h = h_mouse, sig.level = 0.05, power = 0.80)
n80 <- ceiling(res80$n)
cat("n per group (80% power):", n80, "\n")

# With 20% dropout
cat("Enrol per group:", ceiling(n80 / 0.80), "\n")

# At 90% power
res90 <- pwr.2p.test(h = h_mouse, sig.level = 0.05, power = 0.90)
cat("n per group (90% power):", ceiling(res90$n), "\n")
```
:::

### Exercise 3 (Open-ended)

For your next planned experiment or study, define:

1. The primary outcome and the comparison you will make
2. Expected means/proportions and SD in each group (from pilot data or literature)
3. $\alpha$ and desired power

Run the power calculation and produce a power curve. Write one paragraph suitable for an ethics committee submission justifying your sample size.


## Comprehension Check

1. What is statistical power? What does "80% power" mean in plain language?
2. You calculate that you need n = 25 per group. You expect 10% dropout. How many participants should you recruit?
3. A researcher uses a Cohen's d of 1.2 from a single published paper with n = 10 per group to power their study. What is the concern?
4. All else equal, does a one-sided test require more or fewer participants than a two-sided test for the same power?
5. Your power calculation gives n = 12 per group, but you can only recruit n = 8. What are your options?

::: {.callout-note collapse="true"}
### Answers
1. Statistical power is the probability of correctly rejecting the null hypothesis when the alternative is true. "80% power" means that if the true effect is as assumed, your study will detect it (i.e., return a significant result) in 80% of repeated experiments of the same design.
2. Divide by the expected retention fraction: $25 / 0.90 = 28$, so recruit 28 per group (round up).
3. Single small studies overestimate effect sizes due to sampling variability and publication bias. Using d = 1.2 from a study with n = 10 is likely to be inflated. The true d may be considerably smaller, and the actual study will be underpowered.
4. A one-sided test requires fewer participants. It allocates the full a to one tail, making the critical threshold easier to reach ? but only valid when you have a strong directional hypothesis and would not act on an effect in the opposite direction.
5. Options: (1) Accept lower power (< 80%) and acknowledge the limitation; (2) increase the effect size you are trying to detect (e.g., only detect a large effect); (3) decrease a (less stringent threshold, increasing power); (4) use a paired design or block on a covariate to reduce within-group variability; (5) pool data across experiments.
:::

## How to Report

::: {.callout-note}
### Reporting a Power Calculation

**In a Methods section (grant application or ethics submission):**

> "Sample size was calculated a priori using the `pwr` package in R (v4.x). Assuming an effect size of Cohen's *d* = 0.5 (medium effect), α = 0.05 (two-sided), and 80% power, a minimum of XX participants per group (YY total) is required. Effect size was estimated from [pilot data / prior literature (cite)]. A 10% dropout allowance was added, giving a final target of ZZ participants per group."

**Key elements to include:**
- Software or formula used for the calculation
- Assumed effect size and its justification (pilot data or literature)
- α level and whether one- or two-sided
- Target power (usually 80% or 90%)
- Any attrition/dropout allowance added

**Common reporting errors:**
- Using a "medium" Cohen's *d* = 0.5 without clinical justification  -  reviewers will question this
- Reporting only the total N, not per-group N
- Not reporting the assumed effect size  -  makes the calculation unverifiable
- Post-hoc power calculations on observed effect sizes  -  these are misleading and now widely discouraged
:::

## Further Reading

- @Cohen-1988 --- *Statistical Power Analysis for the Behavioral Sciences* ? the original reference for effect sizes and power tables
- @Bland-2015 --- Sample size chapters in *An Introduction to Medical Statistics*
- `vignette("pwr-vignette", package = "pwr")` ? comprehensive `pwr` package tutorial
- G*Power manual at gpower.hhu.de
