# Reproducible Research {#sec-reproducible-research}

```{r setup}
#| include: false
#| eval: true
library(here)
```

## When Do You Use This? {.callout-tip}

> You finish an analysis and send your R script to a collaborator. They run it and get different results --- or it does not run at all. This happens more often than you would expect. This session teaches the three habits that prevent it: portable file paths, locked package versions, and self-contained reports.

## Learning Objectives

After completing this session you will be able to:

- Use `here::here()` to write file paths that work on any computer
- Initialize, snapshot, and restore a project environment with `renv`
- Create and render a Quarto document that imports, analyses, and reports data
- Explain the difference between a script and a reproducible report
- Describe the basic Git workflow for version-controlling an analysis

---

## Why Reproducibility Matters

A result is reproducible if someone else --- or your future self --- can re-run your code and get the same output. This is not just a technical nicety. Journals increasingly require code and data submission. Grant reviewers may ask for analysis scripts. And you will almost certainly need to re-run your own analysis when you change the data or a reviewer asks you to use a different cutoff.

The most common reproducibility failures in practice are:

1. **Hard-coded file paths** --- `read_csv("C:/Users/myname/Desktop/data.csv")` does not work on anyone else's machine.
2. **Package version mismatch** --- a function's behavior changed in a newer package version.
3. **Out-of-order code execution** --- running code chunks in a Jupyter notebook or R script in the wrong order produces results that cannot be replicated.
4. **Missing or undocumented data transformations** --- the analysis script loads data that was pre-processed in a separate undocumented step.

Each section below addresses one of these failures directly.

---

## 1. Portable File Paths with `here`

The `here` package solves the problem of hard-coded paths. It finds the root of your project (the folder containing your `.Rproj` file) and builds all paths relative to that root.

```{r}
#| eval: false
install.packages("here")
library(here)

# Instead of this:
data <- read_csv("C:/Users/myname/Documents/project/data/measurements.csv")

# Write this:
data <- read_csv(here::here("data", "measurements.csv"))
```

The `here::here("data", "measurements.csv")` call builds the correct full path regardless of which computer runs it, which operating system is used, or which subdirectory the script lives in.

::: callout-tip
**Rule:** Always use `here::here()` for file paths. Never hard-code an absolute path. This single habit eliminates most path-related errors when sharing code.
:::

### Check That `here` Finds Your Project Root

```{r}
#| eval: false
library(here)
here()         # prints the project root path
here("data")   # prints the path to the data folder
```

If `here()` returns an unexpected path, check that you have an `.Rproj` file in your project root. Opening the project in RStudio via the `.Rproj` file sets the working directory correctly.

---

## 2. Managing Package Versions with `renv`

The `renv` package creates a **project-specific library** of packages. Instead of relying on the packages installed globally on your machine, your project locks the exact version of every package it uses. Anyone who restores your `renv.lock` file gets the exact same package versions.

### Initial Setup

```{r}
#| eval: false
install.packages("renv")

# In your project, run once to initialise:
renv::init()
```

`renv::init()` does three things: it creates a project-level library in `renv/library/`, it takes a snapshot of all packages your code currently uses, and it writes an `renv.lock` file recording every package and its version.

### Saving the State: Snapshot

After you install a new package or update an existing one, record the change:

```{r}
#| eval: false
# After installing or updating packages:
renv::snapshot()
```

This updates `renv.lock`. Commit `renv.lock` to Git (see the Git section below) so collaborators and your future self can restore the environment exactly.

### Restoring an Environment

When someone else clones your project (or you return to it on a new machine), they run one command:

```{r}
#| eval: false
renv::restore()
```

This reads `renv.lock` and installs every package at the exact version recorded.

### The `renv` Workflow in Practice

```
1. renv::init()          -- once, when starting the project
2. install.packages()    -- as you add packages
3. renv::snapshot()      -- after each change to packages
4. Commit renv.lock      -- so the lockfile is version-controlled
5. renv::restore()       -- when setting up the project on a new machine
```

::: callout-note
You do not need to commit the entire `renv/library/` folder to Git --- that would be huge. Commit only `renv.lock` and the `renv/` configuration files. The `.gitignore` that `renv::init()` creates handles this automatically.
:::

---

## 3. Quarto for Reproducible Reports

A Quarto document (`.qmd`) combines prose, code, and output in a single file. When you **render** the document, R runs the code chunks from top to bottom in a clean session and embeds the output into the report. This means the document is always self-consistent --- you cannot manually edit a table to show different numbers from what the code produces.

### Anatomy of a Quarto Document

```markdown
---
title: "My Analysis"
author: "Your Name"
date: today
format: html
---

## Introduction

This report analyses weight and glucose data from the mouse experiment.

\`\`\`{r setup}
#| include: false
library(tidyverse)
library(here)
\`\`\`

## Load Data

\`\`\`{r load-data}
mice <- read_csv(here::here("data", "mouse_data.csv"))
glimpse(mice)
\`\`\`

## Results

\`\`\`{r plot}
ggplot(mice, aes(x = group, y = body_weight)) +
  geom_boxplot()
\`\`\`
```

### Key Chunk Options

| Option | What it does |
|---|---|
| `#| eval: false` | Show the code but do not run it |
| `#| echo: false` | Run the code but do not show it |
| `#| include: false` | Run the code, hide both code and output |
| `#| warning: false` | Suppress warnings from appearing in the output |
| `#| fig-cap: "..."` | Add a caption to a figure |

### The `freeze` Setting

For large course books like this one, `execute: freeze: auto` in `_quarto.yml` means R code only re-runs when you change the source. This makes building fast. You can force a re-run of one file with:

```bash
quarto render sessions/my-session.qmd --no-freeze
```

---

## 4. Complete Reproducible Analysis Example

Here is a minimal complete example that follows the reproducible workflow: portable paths, declared packages, and a single document that imports, analyses, and reports.

```{r}
#| eval: false

# File: sessions/my-analysis.qmd
# All packages declared at the top
library(tidyverse)
library(here)
library(broom)

# Load data using here::here()
mice <- read_csv(here::here("data", "mouse_weights.csv"))

# Inspect
glimpse(mice)

# Tidy: reshape if needed
mice_long <- mice |>
  pivot_longer(starts_with("Week"),
               names_to = "week",
               values_to = "body_weight") |>
  mutate(week_num = as.numeric(str_remove(week, "Week")),
         group    = factor(group, levels = c("control", "high_fat")))

# Summarise
summary_table <- mice_long |>
  group_by(group, week_num) |>
  summarise(mean_weight = mean(body_weight), .groups = "drop")

# Plot
ggplot(summary_table, aes(x = week_num, y = mean_weight,
                           color = group, group = group)) +
  geom_line() +
  geom_point() +
  theme_bw()

# Model: is body weight different between groups at Week 12?
week12 <- mice_long |> filter(week_num == 12)

model <- lm(body_weight ~ group, data = week12)
tidy(model, conf.int = TRUE)
```

When you render this document, every step from data import to final table runs in order in a clean R session. You get a report that is provably consistent with the code.

---

## 5. Git Basics for Version Control

Git records every change to your files over time. You can always go back to an earlier version. For a research project, git means you never lose work and you have a complete history of how the analysis evolved.

### The Three Commands You Need First

```bash
# Initialise a git repository in your project folder
git init

# Stage all changed files (prepare them for saving)
git add .

# Commit: save a snapshot with a message describing what you did
git commit -m "Add initial mouse weight analysis"
```

### What to Commit

- R scripts and `.qmd` files
- `_quarto.yml`, `index.qmd`
- `renv.lock`
- Small data files (< 50 MB)

### What NOT to Commit

- Large data files (use `.gitignore`)
- The `renv/library/` folder
- Rendered HTML output (optional — it is large and regeneratable)
- API keys or passwords

::: callout-note
**GitHub** is a website that stores git repositories online. Pushing your repository to GitHub means collaborators can access the code, and you have an off-site backup. Creating a GitHub account and running `git push origin main` after each set of commits is strongly recommended.
:::

---

## Example: Recommended Project Structure

A clean project folder structure makes it easy for anyone (including you six months from now) to understand what is what:

```
my-project/
├── _quarto.yml           # Quarto book or document config
├── index.qmd             # Landing page
├── sessions/             # Session .qmd files
├── data/                 # Raw data (never modify these files)
├── R/
│   ├── functions.R       # Custom functions
│   └── load-packages.R   # Package loading (source at top of sessions)
├── renv/                 # renv configuration
├── renv.lock             # Locked package versions (commit this)
├── .gitignore            # What git should ignore
└── my-project.Rproj      # RStudio project file
```

::: callout-tip
**Golden rule:** Raw data files go in `data/` and are never modified. If you need to clean the data, write a cleaning script that reads from `data/` and writes to `data/derived/` or similar. This way you can always go back to the original.
:::

---

## Exercises

### Exercise 1 (Guided): Set Up a Reproducible Project

1. Create a new folder on your computer called `practice-project`.
2. Open RStudio and create a new R Project in that folder (File > New Project > Existing Directory).
3. Open the R console and run `renv::init()`.
4. Create a new Quarto document (File > New File > Quarto Document). Save it as `analysis.qmd` in the project root.
5. Add a code chunk that loads `tidyverse` and `here`, creates a small tibble with 3 rows and 2 columns, and prints it.
6. Render the document. Check that the output appears.
7. Run `renv::snapshot()`. Open `renv.lock` and find the entry for `ggplot2`. What version is recorded?

### Exercise 2 (Semi-guided): Fix a Broken Path

This code will fail because the path is hard-coded:

```{r}
#| eval: false
data <- read_csv("C:/Users/myname/Documents/my-project/data/measurements.csv")
```

1. Rewrite this line using `here::here()` so it works on any computer.
2. Where does the `here` package look for the project root?
3. What file in the project folder signals the root to `here`?

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r}
#| eval: false
library(here)
data <- read_csv(here::here("data", "measurements.csv"))
```

`here` looks for a `.Rproj` file (RStudio project file) to identify the project root. It also checks for other markers like a `DESCRIPTION` file or a `.git` folder.
:::

### Exercise 3 (Open-ended)

Take any analysis script you have previously written (or the mouse weight example from the Tidy Data session). Convert it into a self-contained Quarto document that:

- Loads all packages at the top in a setup chunk
- Imports data using `here::here()`
- Produces at least one table and one plot
- Can be rendered from scratch in a clean R session without error

---

## Comprehension Check

1. What does `renv::restore()` do?
2. What is the difference between `renv::snapshot()` and `renv::restore()`?
3. You share your `.qmd` file with a colleague. They cannot run it because `read_csv(here::here("data", "mice.csv"))` returns an error saying the file does not exist. What is the most likely cause?
4. What does the `#| echo: false` chunk option do?
5. Why should you never manually edit a results table in a Quarto document after rendering?

::: {.callout-note collapse="true"}
### Answers
1. `renv::restore()` reads the `renv.lock` file and installs every package at the exact version recorded in the lockfile. It is used to recreate an environment on a new machine or after a fresh clone.
2. `renv::snapshot()` reads your current project code and records the packages and versions you are currently using into `renv.lock`. `renv::restore()` goes the other direction: it reads `renv.lock` and installs those exact versions.
3. The data file is probably not in the `data/` folder on your colleague's machine. You need to share the data file along with the code --- or the colleague needs to download it from the same source using a script.
4. `#| echo: false` runs the code chunk normally but hides the code from the rendered output. Only the output (plots, tables, printed values) is shown.
5. If you manually edit a results table, the table will no longer match the code that supposedly produced it. The whole point of a reproducible report is that every number in the output is provably generated by the code. Manually edited results cannot be traced back to the analysis.
:::

## Further Reading

- [R for Data Science: Workflow Projects](https://r4ds.hadley.nz/workflow-scripts.html) --- working directory and project organisation
- [renv documentation](https://rstudio.github.io/renv/) --- package management for reproducibility
- [here package](https://here.r-lib.org/) --- portable file paths
- [Quarto Guide](https://quarto.org/docs/guide/) --- full Quarto documentation
- [Happy Git with R](https://happygitwithr.com/) --- Jenny Bryan's guide to Git and GitHub for R users
