---
title: "Sampling, Estimation, and Confidence Intervals"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(survival)
library(boot)
```

## When Do You Use This? {#sec-sampling}

::: {.callout-tip}

You have measured something in a sample of patients, animals, or cells,
and you want to say something about the broader population they came
from. Estimation gives you a best guess for the population value, and a
confidence interval tells you how precise that guess is. For example:
you measure albumin in 418 patients with a liver disease and want to
estimate the mean albumin in *all* patients with that disease, not just
the ones in your clinic. Or you run a trial and want to report what
proportion of patients responded, with an honest statement of how
uncertain that proportion is.

This session covers the concepts that underpin almost everything else in
this course. When you later read a t-test result or a regression
coefficient, what you are really reading is an estimate with a
confidence interval. Understanding this session well will make every
session that follows easier to interpret.
:::

## Learning Objectives

After completing this session you will be able to:

- Explain what a sampling distribution and standard error are, and why
  they matter
- State the Central Limit Theorem and understand why it makes
  confidence intervals valid even for skewed data
- Compute confidence intervals for means and proportions in R
- Interpret a 95% confidence interval correctly --- and avoid the most
  common misinterpretation
- Use bootstrap resampling when distributional assumptions are uncertain


## Background: From Sample to Population

### The core problem

You can never measure every patient, mouse, or cell in the world. You
measure a **sample** and use it to estimate something about the
**population** it came from. The key challenge is: your sample is just
one of infinitely many samples you could have drawn. A different clinic,
a different week, a different cohort --- and you would have gotten
slightly different numbers.

Statistics gives you tools to say: "here is my best estimate, and here
is a honest measure of how much it would have wobbled if I had drawn a
different sample."

### A concrete analogy

Before any formulas: think of an election poll. A polling firm surveys
1,000 voters and reports that 54% support Party A (95% CI: 51% to 57%).
They did not survey every voter. They surveyed a sample and used it to
estimate the population proportion --- the share of all voters who
support Party A. The confidence interval is their honest statement about
how much that estimate could have moved if they had called a different
1,000 people.

That is exactly what you are doing in every study. Your sample is the
poll. Your confidence interval is the margin of error. The only
difference is that instead of voting preferences, you are measuring
bilirubin, survival times, or gene expression.

### Three numbers you need to know

**Point estimate** --- Your best single guess at the population value.
For a mean, this is just the sample mean $\bar{x}$.

**Standard error (SE)** --- How much your point estimate would move if
you repeated the study on a fresh sample of the same size. It is not the
spread of your data (that is the standard deviation); it is the spread
of your *estimate*:

$$SE(\bar{x}) = \frac{s}{\sqrt{n}}$$

Two things follow directly from this formula. First, larger samples give
smaller SEs --- your estimate gets more precise as $n$ grows. Second,
more variable data (larger $s$) gives larger SEs --- noisier
measurements are harder to estimate precisely.

**Confidence interval (CI)** --- A range built around the point estimate
so that, across many repetitions of the study, the specified fraction of
intervals would contain the true population value:

$$95\% \text{ CI for } \mu \approx \bar{x} \pm 1.96 \times SE(\bar{x})$$

::: {.callout-warning}
### The most common CI misinterpretation

"There is a 95% probability that the true mean lies in this interval."

This is wrong. The true mean is a fixed (unknown) number --- it either
is in your interval or it is not. The 95% refers to the *procedure*: if
you repeated your study 100 times, approximately 95 of the resulting
intervals would contain the true mean.

Think of it as a fishing net, not a probability. The fish (true mean)
is somewhere in the lake. Your CI is a single throw of the net. A
"95% CI" means your net-throwing technique, if used repeatedly, catches
the fish 95% of the time. On any single throw, you either caught it or
you did not.
:::

### The Central Limit Theorem: why any of this works

Here is the remarkable fact at the heart of statistics: even if
individual measurements are skewed, heavy-tailed, or oddly distributed,
the *sampling distribution of the mean* becomes approximately normal as
sample size increases. This is the **Central Limit Theorem (CLT)**.

It is worth seeing rather than just reading about. Run this simulation
and look at what happens:

```{r clt-simulation}
#| eval: true

set.seed(42)

# We start with a strongly right-skewed distribution (exponential).
# Individual values look nothing like a normal distribution.
# But what does the distribution of *sample means* look like?

sim_means <- tibble(
  n_10  = replicate(2000, mean(rexp(10,  rate = 0.5))),
  n_30  = replicate(2000, mean(rexp(30,  rate = 0.5))),
  n_100 = replicate(2000, mean(rexp(100, rate = 0.5)))
) %>%
  pivot_longer(everything(), names_to = "sample_size", values_to = "mean") %>%
  mutate(sample_size = factor(sample_size,
                              levels = c("n_10", "n_30", "n_100"),
                              labels = c("n = 10", "n = 30", "n = 100")))

ggplot(sim_means, aes(x = mean)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", alpha = 0.8) +
  facet_wrap(~ sample_size, scales = "free_y") +
  labs(
    title = "Sampling distribution of the mean: skewed data becoming normal",
    subtitle = "Each histogram shows 2,000 sample means drawn from a right-skewed distribution",
    x = "Sample mean", y = "Count"
  ) +
  theme_bw()
```

Each histogram shows 2,000 means, each computed from a fresh random
sample of exponential (right-skewed) data. At $n = 10$ the distribution
of means is already roughly bell-shaped. By $n = 30$ it is nearly
normal. By $n = 100$ it is indistinguishable from a normal distribution,
even though every single raw data point came from a skewed distribution.

This is why t-tests and confidence intervals work for most biomedical
data: you do not need your raw measurements to be normal. You need
your *sample size* to be large enough for the CLT to apply. As a rough
guide, $n \geq 30$ is sufficient for mildly skewed data; heavier skew
requires larger samples.


## Example 1: Clinical Data (PBC)

The `survival::pbc` dataset contains data from a clinical trial of
418 patients with primary biliary cirrhosis (PBC), a chronic liver
disease. We will use it to estimate key population quantities.

```{r pbc-setup}
pbc <- survival::pbc %>%
  as_tibble() %>%
  janitor::clean_names() %>%
  mutate(
    sex = factor(sex, levels = c("f", "m"), labels = c("Female", "Male")),
    trt = factor(trt, levels = c(1, 2), labels = c("D-penicillamine", "Placebo"))
  )
```

### Estimating mean albumin with a 95% CI

Serum albumin is a marker of liver function. Normal levels in healthy
adults are roughly 3.5--5.0 g/dL. Before running any test, it is
reasonable to expect that PBC patients --- who have impaired liver
function --- would run below the lower end of this range. Let's
estimate their mean albumin and quantify our uncertainty:

```{r albumin-ci}
albumin_est <- pbc %>%
  filter(!is.na(albumin)) %>%
  summarise(
    n        = n(),
    mean     = mean(albumin),
    sd       = sd(albumin),
    se       = sd / sqrt(n),
    ci_lower = mean - qt(0.975, df = n - 1) * se,
    ci_upper = mean + qt(0.975, df = n - 1) * se
  )
albumin_est
```

::: {.callout-note}
### Reading the Output

| Column | What it means |
|---|---|
| **n** | Sample size. More observations → smaller SE → narrower CI. |
| **mean** | The point estimate: our best guess at the population mean. |
| **sd** | Spread of individual albumin values. This is *not* the SE. |
| **se** | How much our sample mean would vary across repeated studies. Equals sd / √n. |
| **ci_lower / ci_upper** | The 95% confidence interval. The formula is mean ± qt(0.975, df=n−1) × se. |

**In plain English:** We estimate that the mean albumin in PBC patients
is `r round(albumin_est$mean, 2)` g/dL (95% CI:
`r round(albumin_est$ci_lower, 2)` to
`r round(albumin_est$ci_upper, 2)` g/dL). This is below the lower limit
of the normal range (3.5 g/dL), consistent with impaired liver
synthesis.
:::

### The fast way: using `t.test()`

You do not need to write the formula manually. `t.test()` on a single
vector gives you the same CI:

```{r albumin-ttest}
t.test(pbc$albumin)
```

The default null hypothesis ($\mu = 0$) is irrelevant here --- we are
only using the function to get the confidence interval. You can ignore
the p-value in this context.

### Confidence interval for a proportion

What proportion of trial participants are female? Use `prop.test()`,
which gives a Wilson interval rather than the simple Wald formula.
The Wilson interval is more reliable when the proportion is near 0 or 1.

```{r prop-ci}
n_female <- sum(pbc$sex == "Female", na.rm = TRUE)
n_total  <- sum(!is.na(pbc$sex))

prop.test(x = n_female, n = n_total, conf.level = 0.95)
```

### Comparing estimates across groups

Now let's compute separate estimates for each treatment group. This is
the first step before any formal test: look at the estimates and their
intervals side by side.

```{r group-ci}
group_est <- pbc %>%
  filter(!is.na(trt), !is.na(albumin)) %>%
  group_by(trt) %>%
  summarise(
    n        = n(),
    mean     = mean(albumin),
    se       = sd(albumin) / sqrt(n),
    ci_lower = mean - qt(0.975, df = n - 1) * se,
    ci_upper = mean + qt(0.975, df = n - 1) * se
  )
group_est
```

```{r ci-plot}
group_est %>%
  ggplot(aes(x = trt, y = mean, ymin = ci_lower, ymax = ci_upper)) +
  geom_point(size = 4, color = "steelblue") +
  geom_errorbar(width = 0.1, color = "steelblue", linewidth = 1) +
  labs(
    title = "Mean serum albumin by treatment group",
    subtitle = "Points = mean; error bars = 95% CI",
    x = NULL,
    y = "Albumin (g/dL)"
  ) +
  theme_bw()
```

The two CIs overlap substantially --- a visual hint that the groups
probably do not differ on albumin. But overlapping CIs are not a
formal test. For a definitive answer you need a two-sample t-test,
which is the subject of the next session.


## Example 2: Small Samples and Bootstrap CIs

The formula-based CI assumes the sampling distribution of the mean is
normal. The CLT guarantees this for large samples, but with small
samples from skewed distributions that assumption can break down. The
**bootstrap** is an alternative that makes no distributional
assumptions at all.

The idea is simple: if you had infinite money you could repeat your
study thousands of times and directly observe the sampling distribution
of your estimate. You do not have infinite money, but you do have your
data. The bootstrap resamples *your data* (with replacement) thousands
of times, treating it as a stand-in for the population. The spread of
the resulting estimates becomes your confidence interval.

```{r mouse-data}
set.seed(2024)
mice <- tibble(
  group  = rep(c("Control", "Treatment"), each = 20),
  weight = c(rnorm(20, mean = 28, sd = 3),
             rnorm(20, mean = 25, sd = 3))
)

mice %>%
  group_by(group) %>%
  summarise(
    n        = n(),
    mean     = mean(weight),
    sd       = sd(weight),
    se       = sd / sqrt(n),
    ci_lower = mean - qt(0.975, df = n - 1) * se,
    ci_upper = mean + qt(0.975, df = n - 1) * se
  )
```

```{r bootstrap}
ctrl_weights <- mice$weight[mice$group == "Control"]

boot_mean <- boot(
  data      = ctrl_weights,
  statistic = function(x, i) mean(x[i]),
  R         = 1000
)

boot.ci(boot_mean, type = "perc")
```

::: {.callout-note}
### When to use the bootstrap

The bootstrap is worth reaching for when:

- Your sample is small (n < 30) *and* your data are clearly skewed
- You are estimating something other than a mean (a median, a
  correlation, a ratio) where no standard formula exists
- You want a sanity check on a formula-based CI

For means from reasonably sized samples, the t-interval and the
bootstrap give nearly identical results. If they diverge substantially,
that is a signal your data are unusual enough to think carefully about.
:::


## What Can Go Wrong

::: {.callout-warning}
**Confusing standard deviation with standard error**

The SD describes the spread of individual measurements in your sample.
The SE describes how much your *sample mean* would vary across repeated
studies. They answer different questions. A large SD can coexist with a
very small SE if your sample is large. Always be explicit: report SD
when describing the sample, SE (or CI) when describing the precision of
an estimate.

**Using 1.96 instead of a t-quantile for small samples**

The formula $\bar{x} \pm 1.96 \times SE$ uses the normal distribution
quantile, which is only valid for large samples. For $n < 100$ or so,
use `qt(0.975, df = n - 1)`. The `t.test()` function does this
automatically, which is one reason to prefer it over manual calculation.

**Ignoring missing data**

`mean(x)` returns `NA` if any values are missing. Use `na.rm = TRUE`
or filter explicitly. More importantly, ask *why* values are missing.
If sicker patients are more likely to have missing lab values, your
estimate of the mean will be biased upward even after removing the
missing observations.

**Treating overlapping CIs as evidence of no difference**

Non-overlapping 95% CIs do imply a statistically significant difference
(roughly p < 0.05). But overlapping CIs do not imply no difference. The
CI for each group mean is not the same as the CI for their *difference*.
A visual overlap of ~25% of the CI length is roughly consistent with
p = 0.05. For a definitive test, use a two-sample procedure.
:::


## Exercises

### Exercise 1 (Guided): CI for Bilirubin

Using `survival::pbc`:

1. Compute the sample mean, SD, SE, and 95% CI for serum bilirubin
   (`bili`).
2. Bilirubin is right-skewed. Repeat using `log(bili)`. Report the
   geometric mean by back-transforming with `exp()`.
3. Compare the raw mean and the geometric mean. Which better represents
   the "typical" patient, and why?

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r ex1-solution}
# Raw bilirubin
pbc %>%
  filter(!is.na(bili)) %>%
  summarise(
    n        = n(),
    mean     = mean(bili),
    se       = sd(bili) / sqrt(n),
    ci_lower = mean - qt(0.975, df = n - 1) * se,
    ci_upper = mean + qt(0.975, df = n - 1) * se
  )

# Log-transformed
log_est <- pbc %>%
  filter(!is.na(bili)) %>%
  summarise(
    n        = n(),
    log_mean = mean(log(bili)),
    se       = sd(log(bili)) / sqrt(n),
    ci_lower = log_mean - qt(0.975, df = n - 1) * se,
    ci_upper = log_mean + qt(0.975, df = n - 1) * se
  )

# Back-transform to original scale (geometric mean and CI)
log_est %>%
  mutate(across(c(log_mean, ci_lower, ci_upper), exp))
```

The geometric mean is lower than the arithmetic mean because bilirubin
is right-skewed --- a small number of patients have very high values
that pull the arithmetic mean upward. The geometric mean better
represents the "typical" patient in a skewed distribution and is the
correct summary to report alongside a log-transformed analysis.
:::

### Exercise 2 (Semi-guided): Bootstrap vs t-interval

Simulate 15 observations from a skewed distribution:

```{r ex2-setup}
#| eval: false
set.seed(2024)
skewed_data <- rexp(15, rate = 0.3)  # right-skewed
```

1. Compute the t-interval for the mean using the formula, and verify it
   matches `t.test()`.
2. Compute a bootstrap percentile CI using `boot()`.
3. Are the two intervals similar? What would happen if the distribution
   were more skewed?

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r ex2-solution}
set.seed(2024)
skewed_data <- rexp(15, rate = 0.3)

# t-interval
n <- length(skewed_data)
m <- mean(skewed_data)
s <- sd(skewed_data)
ci_t <- c(m - qt(0.975, df = n - 1) * s / sqrt(n),
          m + qt(0.975, df = n - 1) * s / sqrt(n))
ci_t
t.test(skewed_data)$conf.int

# bootstrap
b <- boot(skewed_data, function(x, i) mean(x[i]), R = 2000)
boot.ci(b, type = "perc")$percent[4:5]
```

With n = 15 and moderate skew, the intervals are similar but not
identical. The bootstrap CI will tend to be asymmetric when data are
skewed, because the resampling captures the asymmetry directly. With
heavier skew or a smaller sample, the divergence grows --- and in those
cases you should trust the bootstrap.
:::

### Exercise 3 (Open-ended)

Using a dataset from your own research, compute 95% confidence intervals
for the key outcome variable, separately by group. Produce an error-bar
plot. Write a one-paragraph interpretation as you would for a Results
section in a journal article, reporting the point estimate, CI, and
sample size for each group.


## Comprehension Check

1. What is the difference between a standard deviation and a standard
   error? If you double the sample size, what happens to the SE?
2. A study reports: "mean albumin 3.5 g/dL (95% CI: 3.3 to 3.7)."
   What does this mean in plain language?
3. Why does the Central Limit Theorem allow confidence intervals to work
   even when the raw data are not normally distributed?
4. You have 10 mice with right-skewed weight measurements. Should you
   use a t-interval or a bootstrap CI? Why?
5. Two groups have CIs of [10.0, 12.0] and [11.5, 13.5]. They overlap.
   Can you conclude there is no statistically significant difference?

::: {.callout-note collapse="true"}
### Answers

1. The SD measures the spread of individual observations in your sample.
   The SE measures how much your *sample mean* would vary if you repeated
   the study with a new sample of the same size. SE = SD / √n. Doubling
   the sample size reduces the SE by a factor of √2 ≈ 1.41 (not by 2).

2. The best estimate of the population mean albumin is 3.5 g/dL. If
   this study were repeated many times with samples of the same size,
   95% of the resulting confidence intervals would contain the true
   population mean. It does *not* mean there is a 95% probability the
   true mean is in this specific interval.

3. The CLT states that the sampling distribution of the mean is
   approximately normal for large enough samples, regardless of the
   shape of the original data distribution. This means the
   normal-distribution-based formula for CIs gives valid results even
   for skewed or non-normal measurements, provided n is large enough
   (roughly ≥ 30 for mildly skewed data).

4. With only 10 mice and skewed data, the sample size is too small for
   the CLT to reliably normalise the sampling distribution. Use a
   bootstrap CI, which makes no distributional assumption and directly
   reflects the skew in your data.

5. No. Overlapping CIs do not imply no statistically significant
   difference. The 95% CI for each group mean is not the same as the
   95% CI for their *difference*. Two groups can have overlapping
   individual CIs and still differ significantly on a two-sample test.
   Always run the formal test rather than judging from a visual overlap.
:::


## How to Report

::: {.callout-note}
### Reporting Estimates and Confidence Intervals

**In a Methods section:**

> "Continuous variables are summarised as mean (95% confidence
> interval). Bootstrap confidence intervals (1,000 resamples) were used
> for [variable] because the distribution was markedly skewed and sample
> size was small."

**In a Results section:**

> "The mean [outcome] was X.XX g/dL (95% CI: L.LL to U.UU; n = N)."
>
> "The proportion of [characteristic] was XX.X% (95% CI: L.L% to
> U.U%)."

**Always include:**

- The point estimate and its 95% CI, not just the estimate alone
- The sample size $n$, so readers can judge precision
- Whether the CI is a t-based interval, Wilson interval (for
  proportions), or bootstrap interval

**Common reporting errors:**

- Writing "mean ± SD" when you mean "mean (95% CI)" --- they answer
  different questions. SD describes your sample; CI describes your
  estimate.
- Saying "the true value lies in this interval" --- a 95% CI is a
  statement about the procedure, not about this specific interval.
- Reporting a CI without the point estimate, or vice versa.
:::


## Further Reading

- @Bland-2015 --- *An Introduction to Medical Statistics*: the standard
  reference for clinical researchers, with chapters on estimation and
  CIs
- [R for Data Science](https://r4ds.hadley.nz) --- Wickham & Grolemund:
  excellent coverage of data wrangling and visualisation in R
- @Efron-1993 --- *An Introduction to the Bootstrap*: the
  original book on bootstrap methods, readable without heavy mathematics
