---
title: "Before You Begin: Setup"
---

```{r setup}
#| include: false
#| eval: false
```

This page tells you everything you need to do **once**, before starting the course. Setting up R, RStudio, and the core packages takes about **20-35 minutes**, most of it spent waiting for installations to finish (Step 2 alone takes 10-20 minutes, mostly unattended). Steps 3 and 4 are optional and only needed for specific sessions; add ~10-15 minutes if you need them. After that, each session just works.

::: {.callout-note}
### Only here for the statistics, not learning R?
If your goal is to understand statistical concepts and methods (not to learn R programming), you can **skip Parts 1 and 2** and start directly at **Part 3: Inference**. The statistics begins there. You can still follow the Examples in each session by reading the code output without running it yourself, but you will still need R and the packages installed to run anything interactively.
:::

::: {.callout-tip}
### How to Use the Code on This Page

Every code block below is a command to copy into your RStudio
**Console**, not code that runs automatically when this page is built
(that is why each block is marked `eval: false`). Work through the
steps in order from top to bottom - each one builds on the last - and
wait for each install to finish (the `>` prompt returns) before
starting the next.
:::


## Step 1: Install R and RStudio {#sec-setup}

*Estimated time: ~5-10 minutes (Setup)*

You need two things on your computer:

| Software | What it is | Download |
|---|---|---|
| **R** (version = 4.2) | The statistical computing language | [cran.r-project.org](https://cran.r-project.org) |
| **RStudio Desktop** | The editor you will use to write and run R code | [posit.co/download/rstudio-desktop](https://posit.co/download/rstudio-desktop/) |

Install R first, then RStudio. On Windows, accept all defaults during installation.

To check your R version, run this in the RStudio Console:

```{r}
#| eval: false
R.version.string
```


## Step 1b: Install Rtools (Windows Only) {.unnumbered}

*Estimated time: ~5 minutes (Setup, Windows only)*

::: {.callout-important}
### Windows users must do this before installing packages
Many R packages compile C or Fortran code during installation. Without **Rtools**, Windows will silently install outdated pre-compiled versions of packages, which can cause version conflicts that break Quarto rendering later.

1. Go to [cran.r-project.org/bin/windows/Rtools/](https://cran.r-project.org/bin/windows/Rtools/)
2. Download and install the version that matches your R version (e.g., Rtools44 for R 4.4, Rtools43 for R 4.3)
3. Accept all defaults (no changes needed)
4. Restart RStudio after installation

To confirm Rtools is active, run this in the Console; you should see `TRUE`:
```r
pkgbuild::has_build_tools(debug = TRUE)
```
:::


## Step 2: Install Course Packages (CRAN)

*Estimated time: ~10-20 minutes (Setup, mostly unattended)*

Copy and paste the block below into your **RStudio Console** and press Enter. This installs every package used across all sessions in one go. You only need to do this once.

::: {.callout-warning}
### This will take 10-20 minutes (that is normal)
Installing 25+ packages for the first time takes time. R will print a lot of output as it downloads and compiles each package. **Do not close RStudio and do not interrupt the process.** Just wait until you see the `>` prompt return in the Console. Go make a coffee.
:::

```{r}
#| eval: false
install.packages(c(
  # Quarto rendering essentials (must have current versions)
  "knitr",
  "rmarkdown",
  "xfun",

  # Core data wrangling and visualization (used in every session)
  "tidyverse",
  "patchwork",

  # Exploration and summary tables
  "skimr",
  "janitor",
  "gtsummary",
  "GGally",
  "nhanesA",

  # Data import
  "readxl",
  "palmerpenguins",

  # Missing data
  "naniar",
  "mice",

  # Survival analysis
  "survival",
  "ggsurvfit",

  # Regression and modeling
  "broom",
  "broom.mixed",
  "car",
  "lme4",
  "glmnet",
  "parsnip",
  "performance",
  "pROC",
  "AER",

  # Causal inference
  "dagitty",
  "ggdag",

  # Inference and testing
  "pwr",

  # Reproducibility
  "reprex"
))
```

```{r}
#| eval: false
# Install Bioconductor packages used in some sessions
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
## (Omics session removed)
## Bioconductor note: advanced visualization only
```

::: {.callout-tip}
### If a package fails to install
Run `install.packages("packagename")` individually to see the specific error message. On Windows, restart RStudio and try again; some packages need RTools. On macOS, you may need to install Xcode Command Line Tools first: `xcode-select --install` in the Terminal.
:::


## Step 3: Install Bioconductor Packages

*Estimated time: ~5-10 minutes (Setup, optional)*

The Advanced Visualization session optionally uses a Bioconductor package. Install it with:

```{r}
#| eval: false
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install(c(
  "GEOquery"        # fetching GEO datasets (advanced-visualization session)
))
```

::: {.callout-warning}
### When BiocManager asks about updates, press `a`
During installation you may see:
```
Update all/some/none? [a/s/n]:
```
**Type `a` and press Enter.** This updates all dependent packages to compatible versions. Pressing `n` can leave you with mismatched package versions that cause errors later.
:::

::: {.callout-note}
Bioconductor packages can take 5-10 minutes to install. If you are not doing the Advanced Visualization session, you can skip this step.
:::


## Step 4: Install TwoSampleMR (GitHub)

*Estimated time: ~5 minutes (Setup, optional)*

The Mendelian Randomization session uses a package hosted on GitHub:

```{r}
#| eval: false
if (!requireNamespace("remotes", quietly = TRUE))
  install.packages("remotes")

remotes::install_github("MRCIEU/TwoSampleMR")
```

::: {.callout-note}
This requires an internet connection. If you are not doing the Mendelian Randomization session, you can skip this step.
:::


## Step 5: Verify Your Setup

*Estimated time: ~2 minutes (Setup)*

Run this block to confirm the core packages are installed. You should see no errors.

```{r}
#| eval: false
core_pkgs <- c(
  "knitr", "rmarkdown", "xfun",
  "tidyverse", "patchwork", "skimr", "janitor", "gtsummary", "broom",
  "survival", "ggsurvfit", "lme4", "broom.mixed", "car",
  "glmnet", "performance", "pROC", "mice", "naniar",
  "pwr", "dagitty", "ggdag", "GGally", "nhanesA",
  "palmerpenguins", "AER", "reprex"
)

missing <- core_pkgs[!sapply(core_pkgs, requireNamespace, quietly = TRUE)]

if (length(missing) == 0) {
  message("All core packages are installed. You are ready to start.")
} else {
  message("Missing packages: ", paste(missing, collapse = ", "))
  message("Run: install.packages(c(", paste0('"', missing, '"', collapse = ", "), "))")
}
```


## Quick Reference: Which Session Uses Which Package

| Package | Sessions |
|---|---|
| `tidyverse` | All sessions |
| `survival`, `broom` | Most Part 3-5 sessions |
| `janitor` | All Part 2--5 sessions (clean column names) |
| `gtsummary`, `skimr` | Descriptive Statistics, Explore Clinical Data |
| `ggsurvfit` | Survival Analysis |
| `lme4`, `broom.mixed`, `performance` | Mixed Models |
| `mice`, `naniar` | Data Imputation; Tidy Data |
| `glmnet` | Model Building and Diagnostics |
| `pROC` | Logistic Regression |
| `car` | Multiple Regression; Model Diagnostics |
| `GGally` | Correlation and Association |
| `dagitty`, `ggdag` | Causal Inference |
| `AER` | Mendelian Randomization |
| `TwoSampleMR` | Mendelian Randomization |
| `pwr` | Power and Sample Size |
| `GEOquery` | Advanced Visualization |
| `nhanesA` | Descriptive Statistics; Tidy Data; Explore Clinical Data |
| `palmerpenguins` | Multiple sessions (practice data) |
| `knitr`, `rmarkdown`, `xfun` | All sessions (Quarto rendering) |
| `reprex` | Getting Help |
