# Before You Begin: Setup {#sec-setup .unnumbered}

```{r setup}
#| include: false
#| eval: false
```

This page tells you everything you need to do **once**, before starting the course. The setup takes about 10–15 minutes. After that, each session just works.

---

## Step 1: Install R and RStudio

You need two things on your computer:

| Software | What it is | Download |
|---|---|---|
| **R** (version ≥ 4.2) | The statistical computing language | [cran.r-project.org](https://cran.r-project.org) |
| **RStudio Desktop** | The editor you will use to write and run R code | [posit.co/download/rstudio-desktop](https://posit.co/download/rstudio-desktop/) |

Install R first, then RStudio. On Windows, accept all defaults during installation.

To check your R version, run this in the RStudio Console:

```{r}
#| eval: false
R.version.string
```

---

## Step 2: Install Course Packages (CRAN)

Copy and paste the block below into your **RStudio Console** and press Enter. This installs every package used across all sessions in one go. It will take a few minutes — you only need to do this once.

```{r}
#| eval: false
install.packages(c(
  # Core data wrangling and visualization (used in every session)
  "tidyverse",
  "patchwork",
  "plotly",

  # Exploration and summary tables
  "skimr",
  "gtsummary",
  "GGally",
  "nhanesA",

  # Data import
  "readxl",
  "palmerpenguins",

  # Missing data
  "naniar",
  "mice",

  # Survival analysis
  "survival",
  "ggsurvfit",

  # Regression and modeling
  "broom",
  "broom.mixed",
  "car",
  "lme4",
  "glmnet",
  "performance",
  "pROC",
  "AER",

  # Causal inference
  "dagitty",
  "ggdag",

  # Inference and testing
  "pwr",

  # Reproducibility
  "reprex"
))
```

::: {.callout-tip}
### If a package fails to install
Run `install.packages("packagename")` individually to see the specific error message. On Windows, restart RStudio and try again — some packages need RTools. On macOS, you may need to install Xcode Command Line Tools first: `xcode-select --install` in the Terminal.
:::

---

## Step 3: Install Bioconductor Packages

Some sessions in Part 5 (Omics Statistics, Advanced Visualization) use packages from Bioconductor. Install them with:

```{r}
#| eval: false
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install(c(
  "limma",          # differential expression (omics-statistics session)
  "clusterProfiler", # pathway analysis (omics-statistics session)
  "org.Mm.eg.db",   # mouse gene annotations (omics-statistics session)
  "GEOquery"        # fetching GEO datasets (advanced-visualization session)
))
```

::: {.callout-note}
Bioconductor packages can take 5–10 minutes to install. If you are not doing the Omics Statistics or Advanced Visualization sessions, you can skip this step.
:::

---

## Step 4: Install TwoSampleMR (GitHub)

The Mendelian Randomization session uses a package hosted on GitHub:

```{r}
#| eval: false
if (!requireNamespace("remotes", quietly = TRUE))
  install.packages("remotes")

remotes::install_github("MRCIEU/TwoSampleMR")
```

::: {.callout-note}
This requires an internet connection. If you are not doing the Mendelian Randomization session, you can skip this step.
:::

---

## Step 5: Verify Your Setup

Run this block to confirm the core packages are installed. You should see no errors.

```{r}
#| eval: false
core_pkgs <- c(
  "tidyverse", "patchwork", "skimr", "gtsummary", "broom",
  "survival", "ggsurvfit", "lme4", "broom.mixed", "car",
  "glmnet", "performance", "pROC", "mice", "naniar",
  "pwr", "dagitty", "ggdag", "GGally", "nhanesA",
  "palmerpenguins", "AER", "reprex", "plotly"
)

missing <- core_pkgs[!sapply(core_pkgs, requireNamespace, quietly = TRUE)]

if (length(missing) == 0) {
  message("All core packages are installed. You are ready to start.")
} else {
  message("Missing packages: ", paste(missing, collapse = ", "))
  message("Run: install.packages(c(", paste0('"', missing, '"', collapse = ", "), "))")
}
```

---

## Quick Reference: Which Session Uses Which Package

| Package | Sessions |
|---|---|
| `tidyverse` | All sessions |
| `survival`, `broom` | Most Part 3–5 sessions |
| `gtsummary`, `skimr` | Descriptive Statistics, Explore Clinical Data |
| `ggsurvfit` | Survival Analysis |
| `lme4`, `broom.mixed`, `performance` | Mixed Models |
| `mice`, `naniar` | Data Imputation; Tidy Data |
| `glmnet` | Model Building and Diagnostics |
| `pROC` | Logistic Regression |
| `car` | Multiple Regression; Model Diagnostics |
| `GGally` | Correlation and Association |
| `dagitty`, `ggdag` | Causal Inference |
| `AER` | Mendelian Randomization |
| `TwoSampleMR` | Mendelian Randomization |
| `pwr` | Power and Sample Size |
| `limma`, `clusterProfiler`, `org.Mm.eg.db` | Omics Statistics |
| `GEOquery` | Advanced Visualization |
| `nhanesA` | Descriptive Statistics; Tidy Data; Explore Clinical Data |
| `palmerpenguins` | Multiple sessions (practice data) |
| `plotly` | Vectors (interactive plots) |
| `reprex` | Getting Help |
