# Survival and Time-to-Event Analysis {#sec-survival}

```{r setup}
#| include: false
#| eval: true
source(here::here("R/functions.R"))
source(here::here("R/load-packages.R"))
```

## Introduction

This session covers survival analysis concepts such as Kaplan-Meier
estimation and time-to-event modeling, with an emphasis on reproducible
research practices.

## Learning outcomes

- Describe survival and censored data.
- Fit Kaplan-Meier curves using R.
- Interpret survival summaries in clinical research.

## Methods covered in this session

- Kaplan-Meier estimation for survival probability.
- Stratified survival curves and group comparisons.
- Log-rank tests for comparing survival curves.
- Cox proportional hazards regression for time-to-event modeling.
- Interpretation of hazard ratios, confidence intervals, and p-values.

## Kaplan–Meier estimation and log-rank test

We use the `lung` dataset from the `survival` package for examples.

```{r km-logrank}
library(survival)

lung2 <- survival::lung %>% as_tibble() %>%
	mutate(status_event = (status == 2)) # TRUE when event (death) occurred

# Kaplan-Meier by sex
km_sex <- survfit(Surv(time, status_event) ~ sex, data = lung2)
summary(km_sex)

# Plot the KM curves (base plot for simplicity)
plot(km_sex, col = c("blue", "red"), xlab = "Days", ylab = "Survival probability")
legend("topright", legend = c("Male","Female"), col = c("blue","red"), lty = 1)

# Log-rank test
survdiff(Surv(time, status_event) ~ sex, data = lung2)
```

### How to explain KM and log-rank to a clinician

- **KM:** "Kaplan–Meier curves estimate the proportion of patients surviving beyond each time point, accounting for censoring."  
- **Log-rank:** "The log-rank test assesses whether survival curves differ across groups; report the p-value and consider reporting median survival times with CIs if available."

## Cox proportional hazards regression

Cox regression estimates hazard ratios (HR) and adjusts for covariates.

```{r cox-example}
cox1 <- coxph(Surv(time, status_event) ~ sex + age + ph.ecog, data = lung2)
summary(cox1)

# Hazard ratios and 95% CI
exp(coef(cox1))
exp(confint.default(cox1))
```

### How to explain Cox results to a clinician

- **Report:** the hazard ratio with 95% CI and an interpretation in clinical terms (e.g., "HR=1.5 means a 50% higher instantaneous risk of event").  
- **Context:** note whether proportional hazards assumptions were checked and how adjustments change the effect estimates.

## Short exercise

- Plot Kaplan–Meier curves by sex using `lung`; run a log-rank test and fit a Cox model adjusting for age. Interpret the hazard ratio for `sex` in one sentence.

**Hint:** use `survfit()`, `survdiff()`, and `coxph()`; convert `status` to an event indicator with `status==2` for `lung`.

**Exercises & Further Materials**

- Reproducible survival exercises and example scripts are provided inline or in `scripts/`; consult `data_sources.md` for dataset load snippets and recommended public survival datasets.

::: {.callout-tip collapse="true" title="Survival Exercises — Notes"}
- Exercise files present worked code; no separate `Solution*.pdf` files are included in this course material. Run the Rmds to view outputs and answers.
:::

## Further reading

- Survival analysis and Cox models: @Matloff-2011; see CRAN survival package documentation.
- Practical survival examples are documented in course materials and supporting scripts.

## Resources

- Survival exercises and worked examples were consolidated into the course sessions and supporting scripts; see `data_sources.md` for suggested datasets and scripts.
