---
title: "T-tests and Group Comparisons"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(survival)
library(broom)
```

::: {.callout-note}
### Session at a Glance

**Total core time: about 115 minutes (~2 hours).**

| Section | Time | Type |
|---|---|---|
| When Do You Use This? + Learning Objectives | 5 min | Reading |
| Background | 15 min | Reading |
| Example 1: Animal and Lab Data | 20 min | Worked example |
| Example 2: Clinical Data (PBC) | 15 min | Worked example |
| Checking Assumptions: QQ-plots | 10 min | Worked example |
| What Can Go Wrong | 10 min | Reading |
| Exercises 1 & 2 | 30 min | Practice |
| Comprehension Check | 10 min | Self-test |

In a taught course, this session fits one teaching block. For
self-paced study, it splits naturally into two sittings, e.g.
Background and Example 1 first, the rest later. Exercise 3 is
open-ended and works best as a follow-up using your own data.
:::

::: {.callout-tip}
### How to Use the Code in This Session

Every code chunk below is written so you can copy it into your R
console (or an R script) and run it yourself, in the order shown.
Where you see a "Run It Yourself" box, stop and run the code above it
before reading the explanation that follows.
:::

## When Do You Use This? {#sec-t-tests}

::: {.callout-tip}
You measured haemoglobin before and after treatment in 30 patients. Or
you measured body weight in 20 control mice and 20 treated mice. Or you
want to know whether the mean platelet count in your patient group is
below the clinical reference value. In each case, a t-test answers the
same question: **is this difference bigger than what I would expect from
sampling variation alone?**

This session is the first place you will apply the framework from
[Hypothesis Testing](hypothesis-testing.qmd). If hypothesis testing is
the logic, t-tests are one of its most common tools.
:::

::: {.callout-note collapse="true"}
### Teacher Note

Open by asking the room for a "before vs after" or "treated vs control"
comparison from their own work. Use one of those examples, rather than
the textbook ones, to motivate the one-sample / two-sample / paired
distinction in the Background section before any formula appears.
:::

## Learning Objectives

After completing this session you will be able to:

- Choose the correct t-test variant for your study design (one-sample,
  two-sample, paired)
- Understand why Welch's t-test should be your default
- Check the normality assumption and know when it matters
- Run all three variants in R and read the output line by line
- Report t-test results correctly in a manuscript


## Background: The One Idea Behind All Three Tests

*Estimated time: ~15 minutes (reading)*

::: {.callout-note collapse="true"}
### Teacher Note

The signal-to-noise framing below is the single most important idea in
this session, arguably in the whole course. A useful board exercise:
write $t = \text{signal} / \text{noise}$ and ask what happens to $t$ if
(a) the sample size doubles, (b) the measurement becomes noisier, (c)
the true effect gets bigger. Work through all three before moving on to
which test variant to use; the variant choice is comparatively
mechanical once this idea is solid.
:::

Before the three variants, here is the single idea that ties them all
together. A t-test answers one question: **is the difference I measured
bigger than the noise?** It takes the difference between means (the
*signal*) and divides by the standard error (the *noise*: how much
that difference would vary if you ran the study again on a new sample).
That ratio is the **t-statistic**:

$$t = \frac{\text{difference between means}}{\text{standard error of that difference}} = \frac{\text{signal}}{\text{noise}}$$

A large $|t|$ means the difference stands out against background
variability: chance is an unlikely explanation. A small $|t|$ means
the difference could easily be noise. Everything below (which variant
to pick, Welch versus Student, checking normality) is just this one
idea adapted to your specific study design.

### Choosing the right variant

Ask yourself two questions before running anything:

1. **Am I comparing to a fixed value, or to another group?**
   If comparing to a fixed value (a clinical threshold, a known
   reference), use a **one-sample** t-test. If comparing two groups,
   continue to question 2.

2. **Are the two measurements on the same subjects, or on different
   subjects?**
   If the same subjects appear in both groups (before/after, matched
   pairs, crossover trial), use a **paired** t-test. If different
   subjects are in each group, use a **two-sample** t-test.

| Situation | Test | R syntax |
|---|---|---|
| Compare sample mean to a known value | One-sample | `t.test(x, mu = value)` |
| Compare means of two independent groups | Two-sample (Welch's) | `t.test(y ~ group, data = df)` |
| Compare two measurements on the same subjects | Paired | `t.test(x, y, paired = TRUE)` |

### Welch's vs Student's: always use Welch's

::: {.callout-note collapse="true"}
### Teacher Note

A common question here is "then why does Student's t-test still get
taught at all?" The short, honest answer is history, plus a handful of
specialised designs that assume equal variances by construction. For
everything in this course, Welch's is the right default; treat this as
a one-minute aside, not a detour.
:::

Welch's t-test (R default: `var.equal = FALSE`) does not assume the two
groups have equal variances. Student's t-test does. Since you usually
cannot know in advance whether variances are equal, Welch's is safer.
The cost in power is negligible when variances *are* equal, but the
cost of using Student's when they are not is inflated Type I error.
There is no good reason to use `var.equal = TRUE`.

### Normality: what the assumption actually means

The t-test does not require your raw data to be normally distributed.
It requires the *sampling distribution of the mean* to be normal. By
the Central Limit Theorem, this holds automatically for $n \geq 30$
even with moderately skewed data. For small samples ($n < 30$) with
obviously non-normal distributions, check a QQ-plot. If the data are
heavily skewed or have extreme outliers, use a nonparametric test
(see [Nonparametric Tests](nonparametric-tests.qmd)).

::: {.callout-important}
### The one thing to remember

A t-test is **signal ÷ noise**. A small p-value means your signal is
loud relative to the noise: it does not tell you the difference is
large or clinically important. Always read the p-value alongside the
effect size (the mean difference) and its 95% CI.
:::


## Example 1: Animal and Lab Data

*Estimated time: ~20 minutes (worked example)*

::: {.callout-note collapse="true"}
### Teacher Note

Live-session idea: before showing the boxplot below, ask students to
predict which delivery method will show higher tooth growth, and by
roughly how much. Reveal the plot, then run the test. The plot makes a
difference look obvious; the test result (p = 0.18) tells a more
cautious story. That tension, a visually convincing plot paired with an
inconclusive test, is the core lesson of this example and is worth
dwelling on rather than rushing past.
:::

### Two-sample: does supplement type affect tooth growth?

The `ToothGrowth` dataset records the length of odontoblast cells in
60 guinea pigs given Vitamin C either as ascorbic acid (VC) or orange
juice (OJ) at three doses. At 1 mg/day, does the delivery method
change tooth-cell growth?

Before running any test, look at the data. As you study the plot below,
focus on one thing: does the confidence interval for the difference
between groups cross zero? That is the whole story.

```{r toothgrowth-setup}
#| eval: true

tg <- ToothGrowth %>%
  as_tibble() %>%
  mutate(
    supp = factor(supp, levels = c("VC", "OJ"),
                  labels = c("Ascorbic acid", "Orange juice")),
    dose = factor(dose)
  )

tg_1mg <- tg %>% filter(dose == "1")

tg_1mg %>%
  ggplot(aes(x = supp, y = len, fill = supp)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.6, size = 2.5) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(
    title = "Tooth length by supplement type (1.0 mg/day)",
    x     = NULL,
    y     = "Odontoblast length (mm)"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

Orange juice looks higher, but with only 10 animals per group the
overlap is substantial. Let the test decide:

```{r toothgrowth-test}
#| eval: true

res_tg <- t.test(len ~ supp, data = tg_1mg, var.equal = FALSE)
tidy(res_tg)
```

::: {.callout-tip}
### Run It Yourself

Execute the two chunks above. Your plot should show orange juice (OJ)
somewhat higher than ascorbic acid (VC), with overlapping spread between
the two groups. Your table should match: `t ≈ -1.41`, `p.value ≈ 0.18`,
`conf.low ≈ -8.78`, `conf.high ≈ 1.78`. Before reading the table below,
decide for yourself: based on the plot and these numbers, would you say
the delivery method matters at this dose?
:::

::: {.callout-note}
### Reading the Output Line by Line

| Column | What it means |
|---|---|
| **statistic (t = −1.41)** | The means differ by 1.41 standard errors. Negative because VC < OJ. Close to zero: the gap could easily be noise. |
| **parameter (df = 15.36)** | Welch's degrees of freedom. Not a whole number: this is expected and correct. |
| **p.value (0.18)** | If there were truly no difference, we would see a gap this large 18% of the time from chance alone. Not significant. |
| **conf.low / conf.high (−8.78 to 1.78)** | The true difference could range from OJ being 8.78 mm better to VC being 1.78 mm better. The CI contains zero. |
| **estimate1 / estimate2** | OJ mean = 22.70 mm; VC mean = 16.77 mm. The raw difference is 5.93 mm: always report this, not just p. |

The difference is not significant at 1 mg/day. At 0.5 mg/day it is:
try it in Exercise 1.
:::

### Paired: do two sleep drugs differ?

The `sleep` dataset records *additional hours of sleep* for 10 patients
under each of two drugs. Each patient received **both** drugs, making
this a paired design. The key insight: some people simply sleep longer
than others regardless of the drug. A two-sample test would treat that
between-person variation as noise. The paired test removes it by
focusing on each patient's *within-person* difference.

```{r sleep-plot}
#| eval: true

sleepdat <- sleep %>%
  as_tibble() %>%
  rename(drug = group, extra_hours = extra) %>%
  mutate(drug = factor(drug, labels = c("Drug 1", "Drug 2")))

sleepdat %>%
  ggplot(aes(x = drug, y = extra_hours)) +
  geom_line(aes(group = ID), alpha = 0.4, colour = "grey50") +
  geom_point(aes(colour = drug), size = 3) +
  scale_colour_manual(values = c("steelblue", "tomato")) +
  labs(
    title = "Additional sleep hours per patient under each drug (n = 10)",
    x     = NULL,
    y     = "Additional hours of sleep"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

The connecting lines show each patient's trajectory between drugs. Most
lines slope upward from Drug 1 to Drug 2. The paired test uses exactly
this information, the consistent direction of change across patients.

```{r sleep-test}
#| eval: true

sleep_wide <- sleepdat %>%
  tidyr::pivot_wider(
    names_from  = drug,
    values_from = extra_hours,
    id_cols     = ID
  )

res_sleep <- t.test(sleep_wide[["Drug 2"]], sleep_wide[["Drug 1"]],
                    paired = TRUE)
tidy(res_sleep)
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. You should get `estimate ≈ 1.58`, `statistic ≈
4.06`, `p.value ≈ 0.0028`. Compare this p-value to the two-sample result
from the previous example (p ≈ 0.18). Same kind of test, a much smaller
p-value: the next chunk shows why.
:::

To see exactly why using the paired test matters, run both on the same
data and compare:

```{r sleep-compare}
#| eval: true

bind_rows(
  tidy(t.test(extra_hours ~ drug, data = sleepdat, var.equal = FALSE)) %>%
    mutate(test = "Two-sample (wrong for this design)"),
  tidy(res_sleep) %>%
    mutate(test = "Paired (correct)")
) %>%
  select(test, estimate, conf.low, conf.high, p.value)
```

The paired test gives a much smaller p-value because it strips out the
between-patient variability that inflates the noise estimate in the
two-sample test. The same data, the same effect size, but the right
design choice makes the difference detectable.


## Example 2: Clinical Data (PBC)

*Estimated time: ~15 minutes (worked example)*

::: {.callout-note collapse="true"}
### Teacher Note

This example connects back to the Hypothesis Testing session's
discussion of baseline comparability in randomised trials. If the group
has covered that session, ask them to recall why checking baseline
balance matters before running the two-sample test below.
:::

### One-sample: is mean albumin different from the clinical threshold?

The lower limit of normal serum albumin is 3.5 g/dL. PBC is a
progressive liver disease that can impair the liver's ability to
synthesise albumin, so it is natural to ask: in this cohort, does the
average patient sit below the normal range?

```{r pbc-setup}
#| eval: true

pbc <- survival::pbc %>%
  as_tibble() %>%
  janitor::clean_names() %>%
  mutate(
    sex = factor(sex, levels = c("f", "m"), labels = c("Female", "Male")),
    trt = factor(trt, levels = c(1, 2), labels = c("D-penicillamine", "Placebo"))
  )

tidy(t.test(pbc$albumin, mu = 3.5))
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. You should get `estimate ≈ 3.50`, `statistic ≈
-0.12`, `p.value ≈ 0.90`, with the 95% CI (3.46 to 3.54) sitting almost
exactly on 3.5. Notice how close `estimate` is to `mu`: this is what a
genuinely null result looks like.
:::

Read across the row: `estimate` is the sample mean (3.50 g/dL);
`statistic` is the t-value (−0.12); `p.value` (0.90) is how often a
sample this far from 3.5 would arise by chance if the true population
mean really were 3.5. The 95% CI (3.46 to 3.54) sits almost exactly on
3.5: this is a precise null result. Across the whole cohort, mean
albumin is indistinguishable from the lower limit of normal.

That does **not** mean PBC has no effect on albumin. PBC ranges from
early, asymptomatic disease to advanced cirrhosis, and this cohort spans
that whole spectrum. A cohort-wide average mixes patients whose albumin
is still normal with patients whose albumin is severely depleted, and
the two can cancel out. The next section asks a more useful question:
does albumin differ between groups defined by something other than
disease status, namely treatment arm at baseline.

(`broom::tidy()` packs the output into a data frame rather than the
default printed block. Both show the same numbers; `tidy()` is easier
to work with in reports and pipelines.)

### Two-sample: does albumin differ between treatment groups?

We already know from [Hypothesis Testing](hypothesis-testing.qmd)
that baseline albumin should not differ between randomised groups. Here
we confirm it formally:

```{r pbc-two}
#| eval: true

pbc_rand <- pbc %>% filter(!is.na(trt), !is.na(albumin))

pbc_rand %>%
  ggplot(aes(x = trt, y = albumin, fill = trt)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1.5) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(
    title = "Serum albumin by treatment group (baseline)",
    x     = NULL,
    y     = "Albumin (g/dL)"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

```{r pbc-two-test}
#| eval: true

tidy(t.test(albumin ~ trt, data = pbc_rand, var.equal = FALSE))
```

::: {.callout-tip}
### Run It Yourself

Run the chunk above. You should get `p.value ≈ 0.87` and a confidence
interval for the difference that comfortably contains zero (about −0.10
to 0.09 g/dL). This is the result you want to see when checking baseline
balance: no evidence of a difference.
:::

::: {.callout-tip}
### Putting the Two Results Together

> "Mean serum albumin across the cohort was 3.50 g/dL (95% CI:
> 3.46--3.54), not significantly different from the lower limit of
> normal (3.5 g/dL; one-sample t-test: *t*(417) = −0.12, p = 0.90).
> There was no significant difference in baseline albumin between the
> D-penicillamine and placebo groups (two-sample t-test: p = 0.87),
> confirming comparable randomisation."

Two tests, two null results, but they answer different questions. The
first tells you the cohort average sits right at the clinical threshold,
which, given the spread of disease severity in PBC, is unsurprising and
not very informative on its own. The second tells you something
genuinely useful for trial design: randomisation balanced albumin
between arms at baseline, so a later difference between groups is
unlikely to be explained by a baseline imbalance in this variable.
:::


## Checking Assumptions: QQ-plots

*Estimated time: ~10 minutes (worked example)*

::: {.callout-note collapse="true"}
### Teacher Note

Students often expect QQ-plot points to fall in a perfectly straight
line and worry when $n = 10$ produces visible wiggle. After showing the
plot below, ask: "if I simulated 10 points from a true normal
distribution and made this plot 20 times, would they all look
identical?" Running `qqnorm(rnorm(10))` a few times live is a quick,
memorable way to calibrate expectations for small-sample QQ-plots.
:::

For small samples ($n < 30$), use a QQ-plot to check approximate
normality. `ToothGrowth` has only 10 animals per group at each dose,
exactly the situation where checking matters.

A QQ-plot compares the quantiles of your data against the quantiles of a
theoretical normal distribution. If the points fall roughly along the
diagonal line, normality is a reasonable assumption. What to watch for:

- **Heavy tails**: points curve away from the line at both ends,
  suggesting a distribution with more extreme values than normal
- **Right skew**: points curve upward at the top right
- **Outliers**: one or two points far off the line

```{r qq-plot}
tg_1mg %>%
  ggplot(aes(sample = len)) +
  stat_qq() +
  stat_qq_line(colour = "red") +
  facet_wrap(~supp) +
  labs(title = "QQ-plot: tooth length by supplement (1.0 mg/day)") +
  theme_bw()
```

With $n = 10$, points roughly following the line are sufficient:
you cannot expect perfection from ten observations. If you see a
systematic curve or extreme departures, consider a Wilcoxon test
([Nonparametric Tests](nonparametric-tests.qmd)).


## What Can Go Wrong

*Estimated time: ~10 minutes (reading)*

::: {.callout-note collapse="true"}
### Teacher Note

The "Common misinterpretations" callout below works well as a quick
true/false round: read each statement aloud, ask for a show of hands
before revealing why it is wrong. This surfaces misconceptions the group
already holds rather than just presenting them.
:::

**Using paired when data are independent, or vice versa**

The paired t-test requires a genuine one-to-one link between
observations. If you apply it to two independent groups, the degrees of
freedom will be wrong and the p-value meaningless. If you apply a
two-sample test to paired data, you throw away the within-subject
information and lose statistical power. The design, not the software,
determines which test is correct.

**Relying on Shapiro-Wilk to decide whether to use a t-test**

`shapiro.test()` is the wrong tool for this decision. With small
samples (where normality matters most), it has low power and will miss
real non-normality. With large samples (where normality matters least,
because the CLT takes over), it will flag trivially small departures as
"significant." Use a QQ-plot and think about whether your outcome could
plausibly come from a normal distribution.

**Reporting only the p-value**

"p = 0.03" tells the reader almost nothing. How big is the difference?
In what direction? How precise is the estimate? Always report the mean
difference, SD or SE for each group, and the 95% CI.

**Forgetting that the default is Welch's, not Student's**

`t.test()` in R uses `var.equal = FALSE` by default, which gives
Welch's test. This is the right choice. Do not override it with
`var.equal = TRUE` unless you have a genuine scientific reason to assume
equal variances, and there is rarely such a reason in biomedical
research.

::: {.callout-warning}
### Common misinterpretations

**"p = 0.04 means the difference is large"**
With 400 patients, even a 0.05 g/dL difference (clinically trivial) can
produce p = 0.04. Always report the mean difference and CI.

**"p = 0.07 means the groups are the same"**
It means the data are consistent with no difference at $\alpha = 0.05$.
With 20 patients, it may simply mean the study was underpowered. Never
conclude "no difference" from a non-significant p-value alone.

**"I used a one-sided test to get a smaller p-value"**
A one-sided test is only valid if you pre-specified the direction before
collecting data. Using it because the data happened to go the right way
is p-hacking.

**"The paired t-test always gives a smaller p-value"**
Only when the within-subject measurements are correlated. If your data
are genuinely independent, the paired test is simply wrong.
:::


## Exercises

::: {.callout-note collapse="true"}
### Teacher Note

In a live session, have students attempt Exercise 1 in pairs (~10
minutes) before revealing the solution, then work Exercise 2
individually. Exercise 3 needs each student's own data, so it works best
as a take-home or end-of-day task rather than something to complete in
the room.
:::

### Exercise 1 (Guided): Two-Sample Test at a Different Dose

*Estimated time: ~15 minutes*

Using `ToothGrowth`, test whether tooth length differs by supplement
type at the **0.5 mg/day** dose (where the difference looks larger than
at 1 mg/day).

1. Filter to dose = 0.5, plot, then run a Welch's two-sample t-test.
2. Report the result: t-statistic, df, p-value, 95% CI, and group
   means.
3. How does the conclusion change compared to the 1 mg/day result?

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r ex1}
tg_0.5mg <- tg %>% filter(dose == "0.5")

tg_0.5mg %>%
  ggplot(aes(x = supp, y = len, fill = supp)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.6, size = 2.5) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(title = "Tooth length by supplement (0.5 mg/day)",
       x = NULL, y = "Odontoblast length (mm)") +
  theme_bw() + theme(legend.position = "none")

tidy(t.test(len ~ supp, data = tg_0.5mg, var.equal = FALSE))
```

At 0.5 mg/day the difference is statistically significant and the CI
excludes zero. The supplement delivery method matters more at lower
doses. This illustrates a dose-interaction that would need to be
explored with ANOVA if you want to formally test whether the supplement
effect changes across doses.
:::

### Exercise 2 (Semi-guided): One-Sample Test for Bilirubin

*Estimated time: ~15 minutes*

The upper limit of normal for bilirubin is approximately 1.2 mg/dL.

1. Plot the bilirubin distribution in `survival::pbc`. Is it symmetric?
2. Log-transform bilirubin. Test whether mean `log(bili)` equals
   `log(1.2)`.
3. Report the geometric mean (back-transform with `exp()`), the 95%
   CI, t-statistic, df, and p-value. Write one sentence of
   interpretation.

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r ex2}
ggplot(pbc, aes(x = bili)) +
  geom_histogram(bins = 30, fill = "steelblue", colour = "white") +
  labs(x = "Bilirubin (mg/dL)", y = "Count") +
  theme_bw()

res <- t.test(log(pbc$bili), mu = log(1.2))
tidy(res)

exp(res$conf.int)  # geometric mean CI on original scale
```

Bilirubin is strongly right-skewed, so the log scale is appropriate.
The geometric mean bilirubin in PBC patients is well above 1.2 mg/dL
and the difference is highly significant (p < 0.001), consistent with
cholestatic liver disease.
:::

### Exercise 3 (Open-ended)

*Estimated time: 15--30 minutes, depending on your data*

Using a dataset from your own research, identify a comparison where a
t-test is appropriate. Decide which variant (one-sample, two-sample, or
paired) fits your design. Run the test, check assumptions with a
QQ-plot, and write a complete Methods and Results paragraph as you
would for a journal submission.


## Comprehension Check

*Estimated time: ~10 minutes*

1. What is the difference between a one-sample, two-sample, and paired
   t-test? Give a concrete biomedical example for each.
2. Why is Welch's t-test preferred over Student's t-test in practice?
3. A QQ-plot shows points deviating from the diagonal at the upper tail.
   The sample size is n = 12. What should you do?
4. You run a two-sample t-test and get p = 0.04, 95% CI for the
   difference: (0.01 to 2.1 g/dL). Is this a convincing, informative
   result? What would you report?
5. A colleague says: "I ran Shapiro-Wilk and got p = 0.06, so the data
   are normal enough for a t-test." What is wrong with this reasoning?

::: {.callout-note collapse="true"}
### Answers

1. **One-sample:** compare a sample mean to a fixed reference value
   (e.g., "Is mean haemoglobin in my patients below 12 g/dL?").
   **Two-sample:** compare means of two independent groups (e.g., "Does
   albumin differ between treated and untreated mice?").
   **Paired:** compare two measurements on the same subjects (e.g.,
   "Did blood pressure change from baseline to 6 weeks in the same
   patients?").

2. Welch's test does not assume equal group variances, making it more
   robust. The cost in statistical power when variances are equal is
   negligible. Using Student's test when variances differ inflates the
   Type I error rate. Welch's is the R default and should stay that way.

3. With n = 12 and visible non-normality, the sample is too small for
   the CLT to rescue you. Use a nonparametric Wilcoxon rank-sum test
   instead.

4. The result is statistically significant (p = 0.04) and the CI
   excludes zero, but the CI is very wide: the true difference could
   be as small as 0.01 g/dL (negligible) or as large as 2.1 g/dL
   (potentially important). A barely significant result with a wide CI
   tells you almost nothing about the size of the effect. Report both
   the p-value and the CI, and note that the estimate is imprecise.

5. Shapiro-Wilk with p = 0.06 does not mean the data are normal: it
   means non-normality was not detected at $\alpha = 0.05$. With
   n = 12 the test has very low power; it would miss moderate non-
   normality routinely. Use a QQ-plot and domain knowledge to assess
   whether the normal distribution is a plausible model for your data.
:::


## How to Report

::: {.callout-note}
### Reporting t-test Results

**In a Methods section:**

> "Differences in [outcome] between [group A] and [group B] were
> compared using an independent-samples Welch's t-test. A paired
> t-test was used for [paired outcome] because measurements were taken
> on the same subjects at two time points."

**In a Results section:**

> "[Outcome] was higher in [group A] (mean X.XX, SD Y.YY) than in
> [group B] (mean X.XX, SD Y.YY; *t*(df) = X.XX, p = .YYY, 95% CI
> for difference: L.LL to U.UU)."

**Always include:**

- Means and SDs for both groups
- Test statistic (*t*) and degrees of freedom (df)
- Exact p-value (write p < 0.001 when very small; never p = 0.000)
- 95% CI for the mean difference
- Whether the test was one- or two-sided (two-sided is standard)

**Common reporting errors:**

- Reporting only "p < 0.05" without the actual value or the effect size
- Omitting the 95% CI for the difference
- Writing "mean ± SEM" in the text but calling it "mean ± SD"
:::


## Further Reading

- @Welch-1947: the original Welch t-test paper
- @Bland-2015: *An Introduction to Medical Statistics*, chapter on
  t-tests; written for clinicians
- `?t.test` in R: the built-in help page covers all options clearly
