---
title: "T-tests and Group Comparisons"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(survival)
library(broom)
```

## When Do You Use This? {#sec-t-tests}

::: {.callout-tip}
You measured haemoglobin before and after treatment in 30 patients. Or
you measured body weight in 20 control mice and 20 treated mice. Or you
want to know whether the mean platelet count in your patient group is
below the clinical reference value. In each case, a t-test answers the
same question: **is this difference bigger than what I would expect from
sampling variation alone?**

This session is the first place you will apply the framework from
[Hypothesis Testing](hypothesis-testing.qmd). If hypothesis testing is
the logic, t-tests are one of its most common tools.
:::

## Learning Objectives

After completing this session you will be able to:

- Choose the correct t-test variant for your study design (one-sample,
  two-sample, paired)
- Understand why Welch's t-test should be your default
- Check the normality assumption and know when it matters
- Run all three variants in R and read the output line by line
- Report t-test results correctly in a manuscript


## Background: The One Idea Behind All Three Tests

Before the three variants, here is the single idea that ties them all
together. A t-test answers one question: **is the difference I measured
bigger than the noise?** It takes the difference between means (the
*signal*) and divides by the standard error (the *noise* --- how much
that difference would vary if you ran the study again on a new sample).
That ratio is the **t-statistic**:

$$t = \frac{\text{difference between means}}{\text{standard error of that difference}} = \frac{\text{signal}}{\text{noise}}$$

A large $|t|$ means the difference stands out against background
variability --- chance is an unlikely explanation. A small $|t|$ means
the difference could easily be noise. Everything below --- which variant
to pick, Welch versus Student, checking normality --- is just this one
idea adapted to your specific study design.

### Choosing the right variant

Ask yourself two questions before running anything:

1. **Am I comparing to a fixed value, or to another group?**
   If comparing to a fixed value (a clinical threshold, a known
   reference), use a **one-sample** t-test. If comparing two groups,
   continue to question 2.

2. **Are the two measurements on the same subjects, or on different
   subjects?**
   If the same subjects appear in both groups (before/after, matched
   pairs, crossover trial), use a **paired** t-test. If different
   subjects are in each group, use a **two-sample** t-test.

| Situation | Test | R syntax |
|---|---|---|
| Compare sample mean to a known value | One-sample | `t.test(x, mu = value)` |
| Compare means of two independent groups | Two-sample (Welch's) | `t.test(y ~ group, data = df)` |
| Compare two measurements on the same subjects | Paired | `t.test(x, y, paired = TRUE)` |

### Welch's vs Student's: always use Welch's

Welch's t-test (R default: `var.equal = FALSE`) does not assume the two
groups have equal variances. Student's t-test does. Since you usually
cannot know in advance whether variances are equal, Welch's is safer.
The cost in power is negligible when variances *are* equal, but the
cost of using Student's when they are not is inflated Type I error.
There is no good reason to use `var.equal = TRUE`.

### Normality: what the assumption actually means

The t-test does not require your raw data to be normally distributed.
It requires the *sampling distribution of the mean* to be normal. By
the Central Limit Theorem, this holds automatically for $n \geq 30$
even with moderately skewed data. For small samples ($n < 30$) with
obviously non-normal distributions, check a QQ-plot. If the data are
heavily skewed or have extreme outliers, use a nonparametric test
(see [Nonparametric Tests](nonparametric-tests.qmd)).

::: {.callout-important}
### The one thing to remember

A t-test is **signal ÷ noise**. A small p-value means your signal is
loud relative to the noise --- it does not tell you the difference is
large or clinically important. Always read the p-value alongside the
effect size (the mean difference) and its 95% CI.
:::


## Example 1: Animal and Lab Data

### Two-sample: does supplement type affect tooth growth?

The `ToothGrowth` dataset records the length of odontoblast cells in
60 guinea pigs given Vitamin C either as ascorbic acid (VC) or orange
juice (OJ) at three doses. At 1 mg/day, does the delivery method
change tooth-cell growth?

Before running any test, look at the data. As you study the plot below,
focus on one thing: does the confidence interval for the difference
between groups cross zero? That is the whole story.

```{r toothgrowth-setup}
#| eval: true

tg <- ToothGrowth %>%
  as_tibble() %>%
  mutate(
    supp = factor(supp, levels = c("VC", "OJ"),
                  labels = c("Ascorbic acid", "Orange juice")),
    dose = factor(dose)
  )

tg_1mg <- tg %>% filter(dose == "1")

tg_1mg %>%
  ggplot(aes(x = supp, y = len, fill = supp)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.6, size = 2.5) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(
    title = "Tooth length by supplement type (1.0 mg/day)",
    x     = NULL,
    y     = "Odontoblast length (mm)"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

Orange juice looks higher, but with only 10 animals per group the
overlap is substantial. Let the test decide:

```{r toothgrowth-test}
#| eval: true

res_tg <- t.test(len ~ supp, data = tg_1mg, var.equal = FALSE)
tidy(res_tg)
```

::: {.callout-note}
### Reading the Output Line by Line

| Column | What it means |
|---|---|
| **statistic (t = −1.41)** | The means differ by 1.41 standard errors. Negative because VC < OJ. Close to zero: the gap could easily be noise. |
| **parameter (df = 15.36)** | Welch's degrees of freedom. Not a whole number --- this is expected and correct. |
| **p.value (0.18)** | If there were truly no difference, we would see a gap this large 18% of the time from chance alone. Not significant. |
| **conf.low / conf.high (−8.78 to 1.78)** | The true difference could range from OJ being 8.78 mm better to VC being 1.78 mm better. The CI contains zero. |
| **estimate1 / estimate2** | OJ mean = 22.70 mm; VC mean = 16.77 mm. The raw difference is 5.93 mm --- always report this, not just p. |

The difference is not significant at 1 mg/day. At 0.5 mg/day it is
--- try it in Exercise 1.
:::

### Paired: do two sleep drugs differ?

The `sleep` dataset records *additional hours of sleep* for 10 patients
under each of two drugs. Each patient received **both** drugs, making
this a paired design. The key insight: some people simply sleep longer
than others regardless of the drug. A two-sample test would treat that
between-person variation as noise. The paired test removes it by
focusing on each patient's *within-person* difference.

```{r sleep-plot}
#| eval: true

sleepdat <- sleep %>%
  as_tibble() %>%
  rename(drug = group, extra_hours = extra) %>%
  mutate(drug = factor(drug, labels = c("Drug 1", "Drug 2")))

sleepdat %>%
  ggplot(aes(x = drug, y = extra_hours)) +
  geom_line(aes(group = ID), alpha = 0.4, colour = "grey50") +
  geom_point(aes(colour = drug), size = 3) +
  scale_colour_manual(values = c("steelblue", "tomato")) +
  labs(
    title = "Additional sleep hours per patient under each drug (n = 10)",
    x     = NULL,
    y     = "Additional hours of sleep"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

The connecting lines show each patient's trajectory between drugs. Most
lines slope upward from Drug 1 to Drug 2. The paired test uses exactly
this information --- the consistent direction of change across patients.

```{r sleep-test}
#| eval: true

sleep_wide <- sleepdat %>%
  tidyr::pivot_wider(
    names_from  = drug,
    values_from = extra_hours,
    id_cols     = ID
  )

res_sleep <- t.test(sleep_wide[["Drug 2"]], sleep_wide[["Drug 1"]],
                    paired = TRUE)
tidy(res_sleep)
```

To see exactly why using the paired test matters, run both on the same
data and compare:

```{r sleep-compare}
#| eval: true

bind_rows(
  tidy(t.test(extra_hours ~ drug, data = sleepdat, var.equal = FALSE)) %>%
    mutate(test = "Two-sample (wrong for this design)"),
  tidy(res_sleep) %>%
    mutate(test = "Paired (correct)")
) %>%
  select(test, estimate, conf.low, conf.high, p.value)
```

The paired test gives a much smaller p-value because it strips out the
between-patient variability that inflates the noise estimate in the
two-sample test. The same data, the same effect size --- but the right
design choice makes the difference detectable.


## Example 2: Clinical Data (PBC)

### One-sample: is mean albumin below the clinical threshold?

The lower limit of normal serum albumin is 3.5 g/dL. PBC is a
progressive liver disease that impairs the liver's ability to synthesise
albumin. We expect PBC patients to run low. The one-sample t-test tests
whether the mean in our patients is significantly below the reference
value.

```{r pbc-setup}
#| eval: true

pbc <- survival::pbc %>%
  as_tibble() %>%
  janitor::clean_names() %>%
  mutate(
    sex = factor(sex, levels = c("f", "m"), labels = c("Female", "Male")),
    trt = factor(trt, levels = c(1, 2), labels = c("D-penicillamine", "Placebo"))
  )

tidy(t.test(pbc$albumin, mu = 3.5))
```

Read across the row: `estimate` is the sample mean; `statistic` is the
t-value; `p.value` is how often a sample this far from 3.5 would arise
if the true population mean really were 3.5. The mean sits below 3.5
and the p-value is tiny --- clear evidence that PBC patients have
albumin below the normal threshold.

(`broom::tidy()` packs the output into a data frame rather than the
default printed block. Both show the same numbers; `tidy()` is easier
to work with in reports and pipelines.)

### Two-sample: does albumin differ between treatment groups?

We already know from [Hypothesis Testing](hypothesis-testing.qmd)
that baseline albumin should not differ between randomised groups. Here
we confirm it formally:

```{r pbc-two}
#| eval: true

pbc_rand <- pbc %>% filter(!is.na(trt), !is.na(albumin))

pbc_rand %>%
  ggplot(aes(x = trt, y = albumin, fill = trt)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1.5) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(
    title = "Serum albumin by treatment group (baseline)",
    x     = NULL,
    y     = "Albumin (g/dL)"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

```{r pbc-two-test}
#| eval: true

tidy(t.test(albumin ~ trt, data = pbc_rand, var.equal = FALSE))
```

::: {.callout-tip}
### Putting the two results together

> "Mean albumin in PBC patients was 3.43 g/dL (95% CI: 3.39--3.47),
> significantly below the clinical threshold of 3.5 g/dL (one-sample
> t-test: *t*(417) = −3.12, p = 0.002). There was no significant
> difference in albumin between the D-penicillamine and placebo groups
> at baseline, confirming comparable randomisation."

Two tests, two different scientific questions, two complementary
answers.
:::


## Checking Assumptions: QQ-plots

For small samples ($n < 30$), use a QQ-plot to check approximate
normality. `ToothGrowth` has only 10 animals per group at each dose ---
exactly the situation where checking matters.

A QQ-plot compares the quantiles of your data against the quantiles of a
theoretical normal distribution. If the points fall roughly along the
diagonal line, normality is a reasonable assumption. What to watch for:

- **Heavy tails**: points curve away from the line at both ends ---
  suggests a distribution with more extreme values than normal
- **Right skew**: points curve upward at the top right
- **Outliers**: one or two points far off the line

```{r qq-plot}
tg_1mg %>%
  ggplot(aes(sample = len)) +
  stat_qq() +
  stat_qq_line(colour = "red") +
  facet_wrap(~supp) +
  labs(title = "QQ-plot: tooth length by supplement (1.0 mg/day)") +
  theme_bw()
```

With $n = 10$, points roughly following the line are sufficient ---
you cannot expect perfection from ten observations. If you see a
systematic curve or extreme departures, consider a Wilcoxon test
([Nonparametric Tests](nonparametric-tests.qmd)).


## What Can Go Wrong

**Using paired when data are independent, or vice versa**

The paired t-test requires a genuine one-to-one link between
observations. If you apply it to two independent groups, the degrees of
freedom will be wrong and the p-value meaningless. If you apply a
two-sample test to paired data, you throw away the within-subject
information and lose statistical power. The design --- not the software
--- determines which test is correct.

**Relying on Shapiro-Wilk to decide whether to use a t-test**

`shapiro.test()` is the wrong tool for this decision. With small
samples (where normality matters most), it has low power and will miss
real non-normality. With large samples (where normality matters least,
because the CLT takes over), it will flag trivially small departures as
"significant." Use a QQ-plot and think about whether your outcome could
plausibly come from a normal distribution.

**Reporting only the p-value**

"p = 0.03" tells the reader almost nothing. How big is the difference?
In what direction? How precise is the estimate? Always report the mean
difference, SD or SE for each group, and the 95% CI.

**Forgetting that the default is Welch's, not Student's**

`t.test()` in R uses `var.equal = FALSE` by default, which gives
Welch's test. This is the right choice. Do not override it with
`var.equal = TRUE` unless you have a genuine scientific reason to assume
equal variances --- and there is rarely such a reason in biomedical
research.

::: {.callout-warning}
### Common misinterpretations

**"p = 0.04 means the difference is large"**
With 400 patients, even a 0.05 g/dL difference (clinically trivial) can
produce p = 0.04. Always report the mean difference and CI.

**"p = 0.07 means the groups are the same"**
It means the data are consistent with no difference at $\alpha = 0.05$.
With 20 patients, it may simply mean the study was underpowered. Never
conclude "no difference" from a non-significant p-value alone.

**"I used a one-sided test to get a smaller p-value"**
A one-sided test is only valid if you pre-specified the direction before
collecting data. Using it because the data happened to go the right way
is p-hacking.

**"The paired t-test always gives a smaller p-value"**
Only when the within-subject measurements are correlated. If your data
are genuinely independent, the paired test is simply wrong.
:::


## Exercises

### Exercise 1 (Guided): Two-Sample Test at a Different Dose

Using `ToothGrowth`, test whether tooth length differs by supplement
type at the **0.5 mg/day** dose (where the difference looks larger than
at 1 mg/day).

1. Filter to dose = 0.5, plot, then run a Welch's two-sample t-test.
2. Report the result: t-statistic, df, p-value, 95% CI, and group
   means.
3. How does the conclusion change compared to the 1 mg/day result?

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r ex1}
tg_0.5mg <- tg %>% filter(dose == "0.5")

tg_0.5mg %>%
  ggplot(aes(x = supp, y = len, fill = supp)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.6, size = 2.5) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(title = "Tooth length by supplement (0.5 mg/day)",
       x = NULL, y = "Odontoblast length (mm)") +
  theme_bw() + theme(legend.position = "none")

tidy(t.test(len ~ supp, data = tg_0.5mg, var.equal = FALSE))
```

At 0.5 mg/day the difference is statistically significant and the CI
excludes zero. The supplement delivery method matters more at lower
doses. This illustrates a dose-interaction that would need to be
explored with ANOVA if you want to formally test whether the supplement
effect changes across doses.
:::

### Exercise 2 (Semi-guided): One-Sample Test for Bilirubin

The upper limit of normal for bilirubin is approximately 1.2 mg/dL.

1. Plot the bilirubin distribution in `survival::pbc`. Is it symmetric?
2. Log-transform bilirubin. Test whether mean `log(bili)` equals
   `log(1.2)`.
3. Report the geometric mean (back-transform with `exp()`), the 95%
   CI, t-statistic, df, and p-value. Write one sentence of
   interpretation.

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r ex2}
ggplot(pbc, aes(x = bili)) +
  geom_histogram(bins = 30, fill = "steelblue", colour = "white") +
  labs(x = "Bilirubin (mg/dL)", y = "Count") +
  theme_bw()

res <- t.test(log(pbc$bili), mu = log(1.2))
tidy(res)

exp(res$conf.int)  # geometric mean CI on original scale
```

Bilirubin is strongly right-skewed, so the log scale is appropriate.
The geometric mean bilirubin in PBC patients is well above 1.2 mg/dL
and the difference is highly significant (p < 0.001), consistent with
cholestatic liver disease.
:::

### Exercise 3 (Open-ended)

Using a dataset from your own research, identify a comparison where a
t-test is appropriate. Decide which variant (one-sample, two-sample, or
paired) fits your design. Run the test, check assumptions with a
QQ-plot, and write a complete Methods and Results paragraph as you
would for a journal submission.


## Comprehension Check

1. What is the difference between a one-sample, two-sample, and paired
   t-test? Give a concrete biomedical example for each.
2. Why is Welch's t-test preferred over Student's t-test in practice?
3. A QQ-plot shows points deviating from the diagonal at the upper tail.
   The sample size is n = 12. What should you do?
4. You run a two-sample t-test and get p = 0.04, 95% CI for the
   difference: (0.01 to 2.1 g/dL). Is this a convincing, informative
   result? What would you report?
5. A colleague says: "I ran Shapiro-Wilk and got p = 0.06, so the data
   are normal enough for a t-test." What is wrong with this reasoning?

::: {.callout-note collapse="true"}
### Answers

1. **One-sample:** compare a sample mean to a fixed reference value
   (e.g., "Is mean haemoglobin in my patients below 12 g/dL?").
   **Two-sample:** compare means of two independent groups (e.g., "Does
   albumin differ between treated and untreated mice?").
   **Paired:** compare two measurements on the same subjects (e.g.,
   "Did blood pressure change from baseline to 6 weeks in the same
   patients?").

2. Welch's test does not assume equal group variances, making it more
   robust. The cost in statistical power when variances are equal is
   negligible. Using Student's test when variances differ inflates the
   Type I error rate. Welch's is the R default and should stay that way.

3. With n = 12 and visible non-normality, the sample is too small for
   the CLT to rescue you. Use a nonparametric Wilcoxon rank-sum test
   instead.

4. The result is statistically significant (p = 0.04) and the CI
   excludes zero, but the CI is very wide: the true difference could
   be as small as 0.01 g/dL (negligible) or as large as 2.1 g/dL
   (potentially important). A barely significant result with a wide CI
   tells you almost nothing about the size of the effect. Report both
   the p-value and the CI, and note that the estimate is imprecise.

5. Shapiro-Wilk with p = 0.06 does not mean the data are normal --- it
   means non-normality was not detected at $\alpha = 0.05$. With
   n = 12 the test has very low power; it would miss moderate non-
   normality routinely. Use a QQ-plot and domain knowledge to assess
   whether the normal distribution is a plausible model for your data.
:::


## How to Report

::: {.callout-note}
### Reporting t-test Results

**In a Methods section:**

> "Differences in [outcome] between [group A] and [group B] were
> compared using an independent-samples Welch's t-test. A paired
> t-test was used for [paired outcome] because measurements were taken
> on the same subjects at two time points."

**In a Results section:**

> "[Outcome] was higher in [group A] (mean X.XX, SD Y.YY) than in
> [group B] (mean X.XX, SD Y.YY; *t*(df) = X.XX, p = .YYY, 95% CI
> for difference: L.LL to U.UU)."

**Always include:**

- Means and SDs for both groups
- Test statistic (*t*) and degrees of freedom (df)
- Exact p-value (write p < 0.001 when very small; never p = 0.000)
- 95% CI for the mean difference
- Whether the test was one- or two-sided (two-sided is standard)

**Common reporting errors:**

- Reporting only "p < 0.05" without the actual value or the effect size
- Omitting the 95% CI for the difference
- Writing "mean ± SEM" in the text but calling it "mean ± SD"
:::


## Further Reading

- @Welch-1947 --- the original Welch t-test paper
- @Bland-2015 --- *An Introduction to Medical Statistics*, chapter on
  t-tests; written for clinicians
- `?t.test` in R: the built-in help page covers all options clearly
