# Basic Statistics for Clinical Research {#sec-basic-statistics}

```{r setup}
#| include: false
#| eval: true
source(here::here("R/functions.R"))
source(here::here("R/load-packages.R"))
```

## Introduction

This session covers foundational statistics topics for clinical research
using a tidyverse workflow. It mirrors the basic statistics coverage from
`dev_code_data` while remaining reliable in the current environment.

## Learning outcomes

- Compute descriptive statistics for clinical variables.
- Visualize distributions and compare groups.
- Use transformations and confidence intervals.
- Apply t-tests, ANOVA, correlation, regression, and contingency tests.
- Interpret power and sample-size calculations.

## 1. Load and inspect the PBC dataset

```{r load-pbc}
pbc <- survival::pbc %>%
  as_tibble() %>%
  janitor::clean_names() %>%
  mutate(
    sex = factor(sex, levels = c("f", "m"), labels = c("Female", "Male")),
    stage = factor(stage)
  )

pbc %>%
  glimpse()
```

The `pbc` dataset contains clinical laboratory measurements and disease
stage for patients with primary biliary cirrhosis.

## 2. Descriptive statistics and distributions

```{r descriptive-stats}
pbc %>%
  summarize(
    n_obs = n(),
    mean_bili = mean(bili, na.rm = TRUE),
    sd_bili = sd(bili, na.rm = TRUE),
    median_bili = median(bili, na.rm = TRUE),
    iqr_bili = IQR(bili, na.rm = TRUE)
  )

pbc %>%
  ggplot(aes(x = bili)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 30) +
  labs(
    title = "Distribution of serum bilirubin",
    x = "Bilirubin (mg/dL)",
    y = "Count"
  ) +
  theme_minimal()
```

We can also compare bilirubin by stage.

```{r bili-by-stage}
pbc %>%
  ggplot(aes(x = stage, y = bili, fill = stage)) +
  geom_boxplot(alpha = 0.75) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  labs(
    title = "Bilirubin by disease stage",
    x = "Stage",
    y = "Bilirubin (mg/dL)"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Pastel1")
```

## 3. Transformations and normal approximation

Serum bilirubin is right-skewed, so a log transformation can improve
normal approximation.

```{r log-transform}
pbc <- pbc %>%
  mutate(log_bili = log(bili))

pbc %>%
  summarize(
    mean_log = mean(log_bili, na.rm = TRUE),
    sd_log = sd(log_bili, na.rm = TRUE)
  )

pbc %>%
  ggplot(aes(sample = log_bili)) +
  stat_qq() +
  stat_qq_line(color = "firebrick") +
  labs(
    title = "QQ plot of log-transformed bilirubin",
    x = "Theoretical quantiles",
    y = "Sample quantiles"
  ) +
  theme_minimal()
```

## 4. Confidence intervals and one-sample inference

Estimate the mean log bilirubin and back-transform the interval.

```{r ci-one-sample}
ci_log <- pbc %>%
  summarize(
    estimate = mean(log_bili, na.rm = TRUE),
    se = sd(log_bili, na.rm = TRUE) / sqrt(sum(!is.na(log_bili))),
    df = sum(!is.na(log_bili)) - 1
  ) %>%
  mutate(
    t_crit = qt(0.975, df),
    lower = estimate - t_crit * se,
    upper = estimate + t_crit * se,
    lower_orig = exp(lower),
    upper_orig = exp(upper),
    estimate_orig = exp(estimate)
  )

ci_log
```

This shows a confidence interval on both the log scale and the original scale.

## 5. Two-sample comparison: ToothGrowth

The `ToothGrowth` dataset is useful for comparing two groups with a t-test.

```{r toothgrowth-setup}
tg <- datasets::ToothGrowth %>%
  as_tibble() %>%
  mutate(dose = factor(dose))

tg %>%
  group_by(supp, dose) %>%
  summarize(
    n = n(),
    mean_len = mean(len),
    sd_len = sd(len)
  ) %>%
  ungroup()
```

Compare supplement groups at the lowest dose.

```{r t-test-toothgrowth}
tg_low <- tg %>%
  filter(dose == 0.5)

test_result <- t.test(len ~ supp, data = tg_low)

broom::tidy(test_result)
```

The t-test gives a mean difference, confidence interval, and p-value.

## 6. One-way ANOVA and group comparisons

A one-way ANOVA compares means across three treatment groups.

```{r plantgrowth-aov}
pg <- datasets::PlantGrowth %>%
  as_tibble()

pg %>%
  ggplot(aes(x = group, y = weight)) +
  geom_boxplot(fill = "lightblue") +
  geom_jitter(width = 0.15, alpha = 0.6) +
  labs(
    title = "Plant growth by treatment group",
    x = "Group",
    y = "Plant weight"
  ) +
  theme_minimal()

pg_aov <- aov(weight ~ group, data = pg)
summary(pg_aov)
```

Inspect pairwise differences with Tukey's method.

```{r plantgrowth-tukey}
TukeyHSD(pg_aov)
```

## 7. Correlation and linear regression

We now use `mtcars` to illustrate correlation and regression.

```{r mtcars-regression}
mtcars <- datasets::mtcars %>%
  as_tibble(rownames = "model") %>%
  mutate(am = factor(am, labels = c("Automatic", "Manual")))

mtcars %>%
  summarize(
    cor_wt_mpg = cor(wt, mpg),
    cor_wt_hp = cor(wt, hp)
  )

mtcars %>%
  ggplot(aes(x = wt, y = mpg)) +
  geom_point(color = "darkgreen", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "steelblue") +
  labs(
    title = "Linear relationship between weight and MPG",
    x = "Weight (1000 lbs)",
    y = "Miles per gallon"
  ) +
  theme_minimal()

lm_mpg <- lm(mpg ~ wt, data = mtcars)
broom::tidy(lm_mpg)

## How to explain to a clinician

- **Report:** present the slope with its 95% CI and an example change anchored to familiar units (e.g. "a 0.5 increase in weight is associated with ~X change in mpg, 95% CI [L, U]").  
- **Prediction:** show a predicted value for a realistic patient profile and give the 95% prediction interval to reflect individual variability.

```{r lm-predictions}
# Predictions for representative weights (confidence and prediction intervals)
new_wt <- tibble(wt = seq(min(mtcars$wt), max(mtcars$wt), length.out = 5))
pred_ci <- predict(lm_mpg, newdata = new_wt, interval = "confidence")
pred_pi <- predict(lm_mpg, newdata = new_wt, interval = "prediction")
new_wt_preds <- bind_cols(new_wt, as_tibble(pred_ci))
new_wt_preds
```
```

The regression output includes slope, standard error, and p-value.

## 8. Contingency tables and chi-square tests

Categorical association is a core component of clinical analysis.

```{r categorical-analysis}
pbc %>%
  count(stage, sex) %>%
  pivot_wider(names_from = sex, values_from = n, values_fill = 0)

chisq_test <- pbc %>%
  select(stage, sex) %>%
  table() %>%
  chisq.test()
chisq_test

## Fisher's exact example (small counts)

```{r fisher-example}
# Example 2x2 table with small counts
small_tab <- matrix(c(10, 2, 3, 20), nrow = 2, byrow = TRUE)
dimnames(small_tab) <- list(Treatment = c("A", "B"), Outcome = c("Event", "NoEvent"))
fisher.test(small_tab)
```

## How to explain categorical results to a clinician

- **Chi-square:** report whether there is statistical evidence for an association and, when possible, provide effect-size measures (risk difference, odds ratio).  
- **Fisher's exact:** explain that it is used for small expected counts and report the p-value; always pair with an effect-size or a simple descriptive sentence about the observed proportions.
```

The chi-square test evaluates whether disease stage is associated with sex.

## 9. Power and sample size intuition

Study planning often begins with a power calculation.

```{r power-calculation}
power.t.test(power = 0.8, delta = 0.5, sd = 1)
power.t.test(n = 20, sd = 1, power = 0.8)
```

These calculations show how sample size, effect size, and power interact.

## How to explain power to a clinician

- **Plain language:** "Power tells us how likely the study is to detect a clinically important effect if it exists. Choosing an effect size should be driven by clinical importance, not just statistical convenience."  
- **Actionable statement:** give the required sample size for a specified, clinically meaningful difference and discuss feasibility.

## 10. Exercise

- Compute mean and standard deviation of `bili` by sex in `pbc`.
- Fit a linear model predicting `mpg` from `wt` and `am` in `mtcars`.
- Create a contingency table for `am` and `cyl` and describe whether
  proportions differ by transmission type.

## Key takeaway

Basic statistics for clinical research combines descriptive summary,
comparative testing, and simple modeling. A tidy workflow keeps these
steps reproducible and easy to communicate.

*** End Patch
