---
title: "Importing, Cleaning, and Reshaping Data"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
library(naniar)
library(palmerpenguins)
```

::: {.callout-note}
## Already know this?
If you can import data with `read_csv()`, filter and summarise with `dplyr`, and reshape between wide and long format with `pivot_longer()` / `pivot_wider()`, skip ahead to **Part 2: Describing and Visualizing Data**.
:::

## When Do You Use This? {#sec-tidy-data-wrangling .callout-tip}

> You received a spreadsheet of mouse body weights measured every week for 12 weeks. Column names are `Week1`, `Week2`, ..., `Week12`. Before you can run any analysis, you need to reshape this into a format R can work with --- one row per measurement, not one row per mouse. This session teaches you the entire pipeline: import, inspect, clean, reshape, and prepare for analysis.

## Learning Objectives

After completing this session you will be able to:

- Import `.csv` and `.xlsx` files into R
- Recognize the difference between "wide" and "tidy" (long) data formats
- Use `dplyr` verbs to filter, select, and transform a data frame
- Reshape data between wide and long format with `pivot_longer()` and `pivot_wider()`
- Handle missing values using `naniar`
- Join two data frames with `left_join()`

## Background: What Is Tidy Data?

Most statistical functions in R expect data in **tidy format**: every row is one observation, every column is one variable. This is also called "long" format.

Data from lab instruments, spreadsheets, or electronic health records often arrives in "wide" format --- especially repeated measurements over time, where each time point becomes a separate column.

```{r}
#| echo: false
wide <- tibble(
  mouse_id = c("M01", "M02"),
  Week1    = c(20.1, 21.3),
  Week2    = c(21.5, 22.0),
  Week3    = c(22.8, 23.1)
)

long <- wide %>%
  pivot_longer(cols = starts_with("Week"),
               names_to = "week",
               values_to = "body_weight")

cat("--- Wide format ---\n")
print(wide)
cat("\n--- Tidy (long) format ---\n")
print(long)
```

::: callout-note
The key rule of tidy data: **one row = one observation**. For a mouse measured weekly, each mouse-week combination is a separate observation.
:::


## Example 1: Mouse Body Weight Study

### The Data

We have body weight data for 10 mice (5 control, 5 high-fat diet) measured at Weeks 1, 4, 8, and 12. The data arrives as a wide spreadsheet.

```{r}
set.seed(42)
mouse_wide <- tibble(
  mouse_id = paste0("M", sprintf("%02d", 1:10)),
  group    = rep(c("control", "high_fat"), each = 5),
  Week1    = c(rnorm(5, 23, 1.5), rnorm(5, 23, 1.5)),
  Week4    = c(rnorm(5, 24, 1.5), rnorm(5, 26, 1.5)),
  Week8    = c(rnorm(5, 25, 1.5), rnorm(5, 30, 2.0)),
  Week12   = c(rnorm(5, 25, 1.5), rnorm(5, 35, 2.5))
)

mouse_wide
```

### Importing Data from Files

In real work you would import from a file rather than typing it in.

```{r}
#| eval: false
# CSV file
mouse_data <- read_csv(here::here("data", "mouse_weights.csv"))

# Excel file
library(readxl)
mouse_data <- read_excel(here::here("data", "mouse_weights.xlsx"), sheet = "Sheet1")
```

::: callout-tip
Always use `here::here("data", "myfile.csv")` instead of a full file path like `"C:/Users/me/project/data/myfile.csv"`. The `here` package makes your code work on any computer regardless of where the project folder is located.
:::

### Inspect the Data

```{r}
nrow(mouse_wide)
ncol(mouse_wide)
glimpse(mouse_wide)
```

### Reshape: Wide to Long with `pivot_longer()`

```{r}
mouse_long <- mouse_wide %>%
  pivot_longer(
    cols      = starts_with("Week"),
    names_to  = "week",
    values_to = "body_weight"
  )

mouse_long
```

The data now has **40 rows** (10 mice x 4 time points). This is the format `ggplot2`, `lme4`, and most statistical functions expect.

### Clean: Extract the Week Number

```{r}
mouse_long <- mouse_long %>%
  mutate(week_num = as.numeric(str_remove(week, "Week")))
```

### Filter, Select, and Mutate

```{r}
# Filter: only high-fat mice from Week 8 onwards
hfd_late <- mouse_long %>%
  filter(group == "high_fat", week_num >= 8)

# Select: keep only the columns needed
analysis_cols <- mouse_long %>%
  select(mouse_id, group, week_num, body_weight)

# Mutate: set factor levels and add a derived column
mouse_long <- mouse_long %>%
  mutate(
    group     = factor(group, levels = c("control", "high_fat")),
    weight_kg = body_weight / 1000
  )
```

### Summarise by Group

```{r}
mouse_long %>%
  group_by(group, week_num) %>%
  summarise(
    n           = n(),
    mean_weight = mean(body_weight, na.rm = TRUE),
    sd_weight   = sd(body_weight, na.rm = TRUE),
    .groups     = "drop"
  )
```

### Visualize the Trajectories

```{r}
#| fig-cap: "Mean body weight over time. Error bars show +/-1 SD."

mouse_summary <- mouse_long %>%
  group_by(group, week_num) %>%
  summarise(
    mean_weight = mean(body_weight),
    sd_weight   = sd(body_weight),
    .groups = "drop"
  )

ggplot(mouse_summary, aes(x = week_num, y = mean_weight,
                           color = group, group = group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_weight - sd_weight,
                    ymax = mean_weight + sd_weight),
                width = 0.4) +
  scale_color_manual(values = c("control" = "steelblue", "high_fat" = "tomato")) +
  labs(title = "Body Weight Over Time", x = "Week",
       y = "Body weight (g)", color = "Group") +
  theme_bw()
```


## Example 2: Penguin Data

The `palmerpenguins` dataset contains measurements on 344 penguins from three species. It has realistic missing values and is a useful stand-in for clinical cohort data.

### Load and Inspect

```{r}
data("penguins", package = "palmerpenguins")
glimpse(penguins)
```

### Check for Missing Values

```{r}
colSums(is.na(penguins))
vis_miss(penguins)
```

The `vis_miss()` plot shows the location and proportion of missing data across your dataset. Always do this before any analysis.

### Drop, Rename, and Summarise

```{r}
penguins_clean <- penguins %>%
  drop_na() %>%
  rename(bill_len = bill_length_mm, bill_dep = bill_depth_mm,
         flipper = flipper_length_mm, mass_g = body_mass_g)

penguins_clean %>%
  group_by(species, sex) %>%
  summarise(
    n            = n(),
    mean_mass    = round(mean(mass_g), 1),
    mean_flipper = round(mean(flipper), 1),
    .groups      = "drop"
  )
```

::: callout-warning
`drop_na()` is only appropriate when missing data are few and likely missing at random. In clinical data, missingness is often not random. Use multiple imputation (covered in the Data Imputation session) when missingness may be systematic.
:::

### Reshape: Long to Wide with `pivot_wider()`

```{r}
penguins_clean %>%
  group_by(species, island) %>%
  summarise(mean_mass = round(mean(mass_g), 0), .groups = "drop") %>%
  pivot_wider(names_from = island, values_from = mean_mass)
```


## Joining Data Frames

It is common to have measurements in one file and metadata (group assignments, clinical variables) in another.

```{r}
measurements <- tibble(
  mouse_id     = c("M01", "M02", "M03", "M04", "M05"),
  glucose_mmol = c(5.4, 5.1, 8.9, 9.3, 10.1)
)

metadata <- tibble(
  mouse_id = c("M01", "M02", "M03", "M04", "M05"),
  group    = c("control", "control", "hfd", "hfd", "hfd"),
  sex      = c("F", "M", "F", "M", "F")
)

combined <- left_join(measurements, metadata, by = "mouse_id")
combined
```

| Join type | Keeps rows from... |
|---|---|
| `left_join(x, y)` | All rows from `x`, matched from `y` |
| `inner_join(x, y)` | Only rows present in both |
| `full_join(x, y)` | All rows from both |

::: callout-tip
Use `left_join()` by default. Always check `nrow()` before and after to confirm no rows were unexpectedly dropped or duplicated.
:::


## Where to Find Data Like This {.callout-note}

**Mouse phenotype data:**
[Mouse Phenome Database](https://phenome.jax.org) (phenome.jax.org) --- hundreds of phenotyping studies with body weight, metabolic, and behavioral data. Data downloads as `.csv`.

**Penguin data:**
`palmerpenguins` R package --- `install.packages("palmerpenguins")`. Original data from Gorman et al. (2014).

**Human clinical data:**
NHANES via the `nhanesA` R package:
```{r}
#| eval: false
library(nhanesA)
demo <- nhanes("DEMO_J")   # 2017-2018 demographics cycle
```


## What Can Go Wrong {.callout-warning}

**Column names with spaces or special characters**
Use `janitor::clean_names()` immediately after import: `df <- read_csv("file.csv") %>% clean_names()`

**Silent type coercion**
If a numeric column has a stray `"N/A"` string, `read_csv()` imports it as character. Always check with `glimpse()` and fix with `mutate(col = as.numeric(col))`.

**Row count changes after a join**
If the key is not unique in both tables, `left_join()` can duplicate rows. Run `nrow(result) == nrow(original)` after every join.


## Exercises

### Exercise 1 (Guided): Reshape Clinical Trial Data

```{r}
set.seed(100)
hr_wide <- tibble(
  patient_id = paste0("P", 1:8),
  treatment  = rep(c("drug", "placebo"), each = 4),
  Week0      = round(rnorm(8, 78, 8)),
  Week4      = round(rnorm(8, 74, 8)),
  Week12     = round(rnorm(8, 72, 8))
)
```

1. Reshape to long format. New columns: `"timepoint"` and `"heart_rate"`.
2. Extract the numeric week into a new column `week_num`.
3. Calculate mean heart rate per `treatment` x `week_num`.
4. Plot mean heart rate over time for drug vs placebo.

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r}
hr_long <- hr_wide %>%
  pivot_longer(starts_with("Week"), names_to = "timepoint", values_to = "heart_rate") %>%
  mutate(week_num = as.numeric(str_remove(timepoint, "Week")))

hr_summary <- hr_long %>%
  group_by(treatment, week_num) %>%
  summarise(mean_hr = mean(heart_rate), se = sd(heart_rate)/sqrt(n()), .groups = "drop")

ggplot(hr_summary, aes(x = week_num, y = mean_hr, color = treatment, group = treatment)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_hr - se, ymax = mean_hr + se), width = 0.4) +
  labs(title = "Heart Rate Over Time", x = "Week", y = "Mean heart rate (bpm)") +
  theme_bw()
```
:::

### Exercise 2 (Semi-guided): Clean the Penguins Dataset

1. Remove rows where `sex` is missing.
2. Keep only Adelie and Chinstrap penguins.
3. Create a new variable `mass_kg`.
4. Summarise: mean and SD of body mass by species and sex.

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r}
penguins %>%
  drop_na(sex) %>%
  filter(species %in% c("Adelie", "Chinstrap")) %>%
  mutate(mass_kg = body_mass_g / 1000) %>%
  group_by(species, sex) %>%
  summarise(mean_mass = round(mean(body_mass_g), 1),
            sd_mass = round(sd(body_mass_g), 1),
            n = n(), .groups = "drop")
```
:::

### Exercise 3 (Open-ended)

Take a wide-format dataset from your own research or download one from the Mouse Phenome Database. Import it into R, reshape to long format, and produce one summary table and one plot. Write 2-3 sentences interpreting what you see.


## Comprehension Check

1. A data frame has 5 mice measured at 6 time points in wide format (one row per mouse). After `pivot_longer()`, how many rows will the result have?
2. What does `drop_na()` do, and when should you use it with caution?
3. Which join keeps all rows from the left table and adds matching columns from the right?
4. After importing a CSV, a numeric column shows as `<chr>` in `glimpse()`. What most likely caused this?
5. What is the purpose of `group_by()` followed by `summarise()`?

::: {.callout-note collapse="true"}
### Answers
1. 30 rows (5 x 6 = 30 observations).
2. It removes any row containing at least one NA. Use it with caution in clinical data because patients with missing values may be systematically different from those without --- dropping them can bias results.
3. `left_join(x, y, by = "id")`
4. A non-numeric character is present in that column (e.g., `"N/A"`, `"."`, or a text note). `read_csv()` converts the whole column to character.
5. `group_by()` splits the data into groups silently. `summarise()` computes a summary statistic within each group and returns one row per group.
:::


## How to Report

### Reporting Data Preparation Steps

::: {.callout-note}
**In a methods section:** "Data were prepared for analysis in R. Variables were renamed and recoded as documented in the supplementary materials. [N] observations were excluded due to [reason: e.g., missing outcome variable, outside the study period]. The final analytic dataset comprised [N] observations and [P] variables."

**What to document (and make reproducible):**

- Source of the data (dataset version, access date for public repositories)
- Inclusion and exclusion criteria applied, with counts at each step
- Derived variables and the formulas used to create them
- Any recoding of categorical variables (original label \u2192 new label)
- Software and package versions

**Reproducibility standard:** All data cleaning steps should be in a documented, version-controlled R script or Quarto document so that any analyst can reproduce the final dataset from the raw data.

**Supplementary materials:** Complex cleaning pipelines should be provided as supplementary code or a data dictionary rather than described in full in the main manuscript text.
:::

## Further Reading

- [R for Data Science: Data Tidying](https://r4ds.hadley.nz/data-tidy.html) --- Hadley Wickham's full treatment
- [dplyr cheat sheet](https://rstudio.github.io/cheatsheets/data-transformation.pdf) --- one-page reference
- [naniar documentation](https://naniar.njtierney.com/) --- missing data visualisation
- [Mouse Phenome Database](https://phenome.jax.org) --- public mouse phenotyping datasets
