---
title: "Vectors, Data Frames, and Factors"
---

```{r setup}
#| include: false
#| eval: true
library(tidyverse)
```

::: {.callout-note}
## Already know this?
If you are comfortable working with vectors, data frames, and factors in R — and you understand what `str()`, `class()`, and `as.factor()` do — skip ahead to the next session or go straight to **Part 2: Describing and Visualizing Data**.
:::

## When Do You Use This? {#sec-vectors .callout-tip}

> Everything in R is stored as a vector, a data frame, or a factor. Before you can clean data, run a t-test, or draw a plot, you need to understand these building blocks. This session covers what they are and how to work with them.

## Learning Objectives

After completing this session you will be able to:

- Create and manipulate numeric, character, and logical vectors
- Understand factors and their role in statistical analysis
- Create and work with data frames and tibbles
- Select, filter, and modify columns and rows in a data frame
- Control factor levels and reference groups for regression

Vectors are fundamental data structures in R, which allow you to store
multiple values in a single variable. These values should be of the same
data type (e.g., all numeric, all character, etc.). Vectors are ordered,
meaning that the values have a specific order in the vector.

::: callout-important
Moving forward, please work within the `workbook` file you created
earlier. To strengthen your coding skills and develop finger muscles
memory, we recommend typing out the code inside the code chunks as you
learn and practice R.
:::

### Creating Vectors

You can create a vector using the `c()` function, which stands for
concatenate. Here's how:

```{r}
# Creating a numeric vector
numeric_vector <- c(1, 3, 5, 7, 9)
```

```{r}
# Creating a character vector
character_vector <- c("apple", "banana", "cherry")
```

```{r}
# Creating a logical vector
logical_vector <- c(TRUE, FALSE, TRUE)
```

### Accessing Vector Elements

You can access individual elements of a vector using square brackets
\[\] with the index (position) of the element in the vector that you
want.

```{r}
#Accessing the second element of numeric_vector
numeric_vector[2]
# here we asked R what element is at index 2
# the answer is  3
```

### Manipulating Vectors

There are various ways to manipulate vectors in R. Some of them include:

Adding Elements: You can add elements to a vector using the c()
function.

```{r}
# Adding an element to numeric_vector
numeric_vector <- c(numeric_vector, 11)  # numeric_vector is now c(1, 3, 5, 7, 9, 11)

```

Removing Elements: A common way to remove elements is by using \[ \] and
giving the index of the element in the vector.

```{r}
# Removing the first element from numeric_vector
numeric_vector <- numeric_vector[-1]  # numeric_vector is now c(3, 5, 7, 9, 11)
```

Changing Elements: You can change the value of a vector element by
assigning a new value to it.

```{r}
# Changing the third element of numeric_vector
numeric_vector[3] <- 10  # numeric_vector is now c(2, 3, 10, 5, 6)
```

## Exercise 3

Now it's your turn to practice!

1.  Create a character vector with names of three of your favorite
    fruits.
2.  Access and print the second element of the vector.
3.  Add another favorite fruit to your vector.
4.  Remove the first element from your vector.
5.  Change the last element of your vector to a different food item.
6.  On every step verify the output, is it what you expected?

::: {.callout-tip collapse="true" title="Exercise 3: Click to see solution"}
```{r}

# 1. Create a character vector with names of three of your favorite fruits.
favorite_fruits <- c("Apple", "Banana", "Cherry")

# 2. Access and print the second element of the vector.
second_fruit <- favorite_fruits[2]
print(second_fruit)

# 4. Add another favorite fruit to your vector.
favorite_fruits <- c(favorite_fruits, "Date")
print(favorite_fruits)

# 4. Remove the first element from your vector.
favorite_fruits <- favorite_fruits[-1]
print(favorite_fruits)

# 4. Change the last element of your vector to a different fruit item.

favorite_fruits[3] <- "Elderberry"
print(favorite_fruits)

# Now, the favorite_fruits vector should be: c("Banana", "Cherry", "Elderberry")

```
:::

## Factors and levels

When working with data in R, we often deal with categorical variables,
which are variables that can be divided into different categories, such
as "Yes" or "No", or "Small", "Medium", "Large". In R, we use a special
type of variable called a *factor* to handle these categories. Think of
factors as containers that hold and manage these categories, which we
call *levels*.

### Understanding Factors

Imagine you're sorting fruits into baskets. Each basket is labeled with
the name of a fruit. In R, each basket is like a *level* of a factor.
The factor is the concept of "fruit types". If you have three types of
fruits---apples, bananas, and cherries---then your factor has three
levels: "Apple", "Banana", and "Cherry".

**Why Use Factors?**

-   **Organization for Analysis:** Just as baskets help us organize
    fruits, factors help organize our data, making it easier to analyze.
-   **Clear Categories:** Even if we didn't collect any cherries, our
    "fruit type" factor still knows cherries exist as a category,
    keeping our data consistent.
-   **Order Matters:** We can order our factors (like saying small,
    medium, and large sizes), which is important for analysis that
    depends on ranking.
-   **Visualization:** Factors help R know how to group and label data
    in charts, which makes our visualizations clear and accurate.

**Practical Example: Fruit Count Chart with factor levels**

Let's relate this to our fruit count chart below. We categorized the
count of different fruits. In our chart, each fruit type is a level in
our factor. This allows us to make a clear and organized bar chart,
showing us how many of each fruit type we have. Remember, factors are
there to make our data analysis and visualization tasks easier and more
accurate, much like baskets help keep fruits organized in real life!

::: callout-note
Move your cursor over the bars in the chart to view the precise counts
of each fruit type.
:::

```{r fruit-plot-with-factors, echo=FALSE, message=FALSE}
library(ggplot2)

# Example dataset: Fruits and their counts
fruits <- c("Apple", "Banana", "Cherry", "Date", "Elderberry")
counts <- c(23, 15, 19, 10, 5)

# Calculate total count of all fruits
total_count <- sum(counts)

# Create a data frame
fruit_data <- data.frame(Fruit = factor(fruits, levels = fruits), Count = counts)

# Use ggplot2 to create a bar plot
ggplot(fruit_data, aes(x = Fruit, y = Count, fill = Fruit)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set3") +
  labs(x = "Fruit Type", y = "Count", title = "Fruits Count by factors") +
  theme_minimal() +
  theme(legend.position = "none")
```

**Practical Example: Fruit Count Chart without factor levles** In this
example, we didn't specify to R that our data should be treated as
categorical with distinct factor levels. Instead, we've labeled all
entries simply as "All Fruits," akin to placing various fruits into one
basket. This approach makes it challenging to discern the quantity of
each individual fruit type because we're presented with a combined total
rather than a detailed breakdown.

```{r fruit-plot-witout-factors, echo=FALSE, message=FALSE}
# R code to plot all fruits as a single bar, ignoring individual counts

# Example dataset: Fruits and their counts without factor levels
fruits <- rep("Apple Banana Cherry Date Elderberry", 5)
counts <- c(23, 15, 19, 10, 5)

# Calculate the total count
total_count <- sum(counts)

# Create a data frame with a single fruit category
fruit_data_combined <- data.frame(Fruit = fruits, Count = total_count)

# Use ggplot2 to create a bar plot with all fruits combined into a single bar
ggplot(fruit_data_combined, aes(x = Fruit, y = Count, fill = Fruit)) +
  geom_bar(stat = "identity") +
  labs(x = "", y = "Total Count", title = "Fruits count without factors") +
  theme_minimal() +
  theme(legend.position = "none")
```

### Creating and converting Factors

You can create a factor from a vector using the `factor()` function.
Here's how:

```{r}
# Create a vector of fruit names
fruit_vector <- c("Apple", "Banana", "Cherry", "Apple", "Cherry", "Banana", "Banana", "Cherry", "Apple","Apple","Apple")
```

```{r}
fruit_vector #  you can get the value of a vector just by running its name in the console or in a code chunk
```

**Convert the vector to a factor**

```{r}
# In this code, convert the fruit_vector vector into a factor using the factor() function.
fruit_factor <- factor(fruit_vector)
```

```{r}
fruit_factor  
```

**Levels of a Factor** The unique values in a factor are called levels.
You can see the levels of a factor using the levels() function. Get the
levels of factor

```{r}
levels(fruit_factor)  
```

**Getting Summary of Factors**

```{r}
summary(fruit_factor)

```

You can change the number of the elements (fruit names in the
fruit_vector) and then play with it.

::: callout-note
Why get summary of factors? It is because just like you'd want to know
how many of each type of fruit you have in the basket, in statistics,
you often need to sort data into categories and count them. Factors help
you do that in R.
:::

### Ordered Factors

**Understanding Ordered Factors in R** In R, when we talk about data
like T-shirt sizes or class levels, we call them "factors." They are
like labels we put on things that are similar, such as "Small,"
"Medium," "Large" for sizes.

But sometimes, it's not enough to just label these things. We need to
remember that some labels come before others, like how "Small" is
smaller than "Medium." We use something called "ordered factors" to do
this.

**Why Ordered Factors Matter**

Think about lining up for a race. The order matters, right? First,
second, third. If we mix it up, it would be confusing. It's the same
with our data. If we don't use ordered factors when the order is
important, we might get mixed up results.

**How to Make Ordered Factors**

Here's how we make ordered factors in R:

1.  We tell R which labels we have, like "Small," "Medium," "Large."
2.  We tell R to remember the order by saying `TRUE` to "ordered."

**Example with T-Shirt Sizes**

Let's say we're sorting T-shirts by size:

```{r}
sizes <- c("Small", "Medium", "Large", "Medium", "Small")
```

To keep the sizes in order, we make an ordered factor:

```{r}
ordered_sizes <- factor(sizes, ordered = TRUE, levels = c("Small", "Medium", "Large"))
```

Now R knows that "Small" comes before "Medium," and "Medium" comes
before "Large."

**Using Ordered Factors**

When we have our sizes in order, we can do things like see if one size
is bigger than another, or make graphs where "Small" comes before
"Medium," not just wherever they popped up in our list. So, ordered
factors help us keep our data tidy and in line, like students in a
school line-up. It helps us make sense of things that should follow a
certain order, giving us clearer results when we're working with our
information. This is super helpful for making sure we understand our
data the right way.

### Converting Factors

Let's use a fruit basket again as an example to illustrate this concept:

Suppose you have a basket filled with an assortment of fruits: apples,
bananas, and cherries. You decide to organize them by putting a little
sticker on each that says "apple", "banana", or "cherry". In R, this is
like creating a factor for your fruits.

Here's some R code that might represent our fruit basket as a factor:

```{r}
fruits <- c("apple", "banana", "cherry", "apple", "cherry")
fruit_factor <- factor(fruits)
```

Now, let's explore the conversions and understand why and when you might
need them:

1.  **To Character:**

    Sometimes, you want to check the stickers without sorting the
    fruits. This is like converting your factor to a character vector.
    You're not interested in how many apples or cherries you have, just
    what's in the basket.

```{r}
fruit_labels <- as.character(fruit_factor)
```

**Why and when you need this:** You might need to do this when you're
only interested in the names of the fruits for a task like labeling a
shelf, where the count doesn't matter.

2.  **To Numeric:**

On other occasions, you might decide to assign a number to each type of
fruit: apples = 1, bananas = 2, cherries = 4. This is like converting
your factor to a numeric vector. Instead of names, each fruit is
represented by a number.

```{r}
fruit_numbers <- as.numeric(fruit_factor)
```

**Why and when you need this:** This is useful when you're entering your
fruit data into a system that requires numbers instead of names, maybe
for a stocktaking system that tracks inventory with codes instead of
fruit names.

In summary, whether you convert a factor into characters or numbers
depends on what you need at the moment:

-   Use characters when you want to work with the names or labels
    directly.
-   Use numbers when you need to enter data into a numerical system or
    do some sort of calculation where categories are represented by
    numbers.

### Exercise 4

Now, practice what you've learned!

1.  Create a factor variable with your favorite colors.
2.  Find out the levels of your factor variable.
3.  Create an ordered factor with levels "Low", "Medium", "High".
4.  Get a summary of your ordered factor.
5.  Convert the factor back to a character vector.

::: {.callout-tip collapse="true" title="Exercise 4: Click to see solution"}
```{r}

# 1. Create a factor variable with your favorite colors.
favorite_colors <- c("Red", "Blue", "Green", "Blue", "Red")
color_factor <- factor(favorite_colors)

# 2. Find out the levels of your factor variable.
color_levels <- levels(color_factor)
color_levels  # Output: "Blue" "Green" "Red"

# 5. Create an ordered factor with levels "Low", "Medium", "High".
ordered_vector <- c("Low", "Medium", "High", "Medium", "Low")
ordered_factor <- factor(ordered_vector, ordered = TRUE, levels = c("Low", "Medium", "High"))

# 5. Get a summary of your ordered factor.
ordered_summary <- summary(ordered_factor)
ordered_summary  # Output: a table showing the counts of each level

# 5. Convert the factor back to a character vector.
char_vector <- as.character(ordered_factor)
char_vector  # Output: "Low" "Medium" "High" "Medium" "Low"

```
:::


## Data Frames and Tibbles

A vector stores a single column of values. A **data frame** stores many columns together — like a spreadsheet where every column can have a different type. This is how you will actually store and work with your research data in R.

A **tibble** is a modern version of a data frame from the `tidyverse`. It behaves the same way but has slightly better defaults (it never converts character columns to factors automatically, and it prints more neatly).

### Creating a Data Frame

```{r}
# A simple data frame with three columns
mouse_data <- data.frame(
  mouse_id   = c("M01", "M02", "M03", "M04", "M05"),
  group      = c("control", "control", "treatment", "treatment", "treatment"),
  body_weight = c(24.1, 23.8, 28.5, 27.2, 29.0)
)

mouse_data
```

### Viewing the Structure

```{r}
# nrow and ncol tell you the dimensions
nrow(mouse_data)
ncol(mouse_data)

# str() shows column types — very useful for diagnosing problems
str(mouse_data)

# head() shows the first 6 rows (useful for large datasets)
head(mouse_data)
```

### Accessing Columns

There are three ways to get a single column. All three give the same result.

```{r}
# Dollar sign: the most common approach
mouse_data$body_weight

# Double brackets with column name
mouse_data[["body_weight"]]

# Double brackets with column index number
mouse_data[[3]]
```

### Accessing Rows and Cells

Data frames use `[row, column]` indexing.

```{r}
# Row 2, all columns
mouse_data[2, ]

# All rows, column 3
mouse_data[, 3]

# Row 2, column 3
mouse_data[2, 3]
```

### Creating a Tibble

```{r}
library(tibble)

# tibble() works exactly like data.frame() but prints more neatly
mouse_tibble <- tibble(
  mouse_id    = c("M01", "M02", "M03", "M04", "M05"),
  group       = c("control", "control", "treatment", "treatment", "treatment"),
  body_weight = c(24.1, 23.8, 28.5, 27.2, 29.0)
)

mouse_tibble
```

::: callout-note
When you import data with `read_csv()` from the `readr` package, it returns a tibble automatically. When you import with `read.csv()` (base R), it returns a data frame. Either works, but tibbles are generally easier to work with.
:::

### Adding and Modifying Columns

```{r}
# Add a new column using the dollar sign
mouse_data$weight_kg <- mouse_data$body_weight / 1000

# Modify an existing column
mouse_data$body_weight <- round(mouse_data$body_weight, 0)

mouse_data
```


## Controlling Factor Levels for Statistical Analysis

When you use a factor in a regression model or an ANOVA, R automatically treats the first level alphabetically as the **reference group**. This affects how coefficients are interpreted. Controlling the level order is an important practical skill.

### Why Level Order Matters

```{r}
# R chooses "control" as reference automatically (alphabetical)
group_factor <- factor(c("treatment", "control", "treatment", "control"))
levels(group_factor)  # "control" comes first = reference group

# In a regression, the coefficient for "treatment" is relative to "control"
# This is what you want — but it happens to be correct here by accident
```

### Setting the Reference Level with `relevel()`

```{r}
# Explicitly set "control" as the reference group
group_factor <- relevel(group_factor, ref = "control")
levels(group_factor)
```

### Using `forcats` for More Control

The `forcats` package (part of the `tidyverse`) gives you powerful tools for reordering factor levels. This is important for plots (which group appears first) and for regression (which group is the reference).

```{r}
library(forcats)

# fct_relevel: move specific levels to the front
diet_group <- factor(c("high_fat", "low_fat", "control", "high_fat", "control"))
diet_reordered <- fct_relevel(diet_group, "control", "low_fat", "high_fat")
levels(diet_reordered)

# fct_recode: rename levels
diet_renamed <- fct_recode(diet_group,
  "Control diet"   = "control",
  "Low-fat diet"   = "low_fat",
  "High-fat diet"  = "high_fat"
)
levels(diet_renamed)

# fct_infreq: order by frequency (most common first) — useful for bar charts
diet_by_freq <- fct_infreq(diet_group)
levels(diet_by_freq)
```

::: callout-tip
**Rule of thumb:** Always check `levels()` on your factor columns before running a regression. Confirm that the reference group is what you intend it to be. If it is not, use `fct_relevel()` or `relevel()` to fix it.
:::


## Exercise 5

Practice building and working with a data frame that represents a real experiment.

**Scenario:** You ran an experiment with 8 mice in three groups: control, low-dose treatment, and high-dose treatment. You measured body weight and blood glucose.

1. Create a data frame with these columns: `mouse_id` (M01-M08), `group` (assign mice to three groups as you like), `body_weight` (make up realistic values in grams, roughly 22-32g), `glucose` (roughly 5-12 mmol/L).
2. Convert `group` to a factor with levels ordered: `"control"`, `"low_dose"`, `"high_dose"`.
3. Print the levels of the `group` factor. Is the reference group correct?
4. Calculate the mean body weight for each group using `tapply()`:  `tapply(df$body_weight, df$group, mean)`.
5. Add a new column called `glucose_category` that is `"normal"` if glucose < 7 and `"elevated"` if glucose >= 7.

::: {.callout-tip collapse="true"}
### Exercise 5: Solution

```{r}
# 1. Create the data frame
mice <- data.frame(
  mouse_id    = paste0("M0", 1:8),
  group       = c("control", "control", "control",
                  "low_dose", "low_dose",
                  "high_dose", "high_dose", "high_dose"),
  body_weight = c(23.5, 24.1, 22.8, 26.3, 27.0, 29.5, 30.1, 28.8),
  glucose     = c(5.2, 5.8, 5.5, 7.1, 6.9, 9.3, 10.2, 8.7)
)

# 2. Convert group to an ordered factor
mice$group <- factor(mice$group,
                     levels = c("control", "low_dose", "high_dose"))

# 3. Check levels
levels(mice$group)  # "control" is the reference — correct

# 4. Mean body weight per group
tapply(mice$body_weight, mice$group, mean)

# 5. Add glucose category
mice$glucose_category <- ifelse(mice$glucose < 7, "normal", "elevated")
mice
```
:::


## Common Mistakes

::: {.callout-warning}
### What Can Go Wrong

**Type coercion in vectors.** Mixing types in a single vector silently converts everything to the most general type: `c(1, 2, "three")` produces `c("1", "2", "three")` — all characters. Downstream arithmetic will then fail with confusing errors. Check with `class()` if a numeric column behaves strangely.

**Factor levels in the wrong order.** R assigns factor levels alphabetically by default. In a regression, the first level is the reference. `"control"` becomes the reference only if it sorts first alphabetically — `"A"` would be the reference if your groups are `"A"`, `"B"`, `"control"`. Always set levels explicitly with `factor(..., levels = c("control", "low_dose", "high_dose"))`.

**`[[` vs `[` indexing.** `df[, "col"]` returns a data frame; `df[["col"]]` (or `df$col`) returns the vector. Functions like `t.test()` expect a vector, not a one-column data frame. If you get "non-numeric argument" errors, this is often the cause.

**`NA` propagation.** Arithmetic on a vector containing `NA` returns `NA`: `mean(c(1, 2, NA))` returns `NA`. Pass `na.rm = TRUE` to summary functions, or remove `NA` values deliberately with `na.omit()`.

**Implicit recoding when coercing factors to numeric.** `as.numeric(factor(c("10", "20", "30")))` returns `1 2 3` — the level indices, not the original values. Use `as.numeric(as.character(my_factor))` to recover the original numbers.
:::

## Exercises

### Exercise 1 (Guided): Exploring `ToothGrowth`

`ToothGrowth` is built into R. This exercise practises the core skills: inspecting structure, checking types, filtering, and adding columns.

1. Run `str(ToothGrowth)`. What are the column names and types?
2. Convert the `supp` column to a factor with levels `c("VC", "OJ")` and labels `c("Ascorbic acid", "Orange juice")`. Check with `levels()`.
3. Convert `dose` to an ordered factor with levels `0.5 < 1.0 < 2.0`.
4. Use `[` indexing to extract all rows where `dose == 2.0` and `supp == "OJ"`.
5. Add a column `len_cm` that converts `len` from micrometres to centimetres (divide by 10 000).

::: {.callout-tip collapse="true"}
### Exercise 1: Solution

```{r}
# 1. Structure
str(ToothGrowth)

# 2. Factor with labels
tg <- ToothGrowth
tg$supp <- factor(tg$supp,
                  levels = c("VC", "OJ"),
                  labels = c("Ascorbic acid", "Orange juice"))
levels(tg$supp)

# 3. Ordered dose factor
tg$dose <- factor(tg$dose, levels = c(0.5, 1.0, 2.0), ordered = TRUE)
levels(tg$dose)

# 4. Subset with [
tg[tg$dose == "2" & tg$supp == "Orange juice", ]

# 5. New column
tg$len_cm <- tg$len / 10000
head(tg)
```
:::

### Exercise 2 (Semi-guided): Factor Manipulation with `PlantGrowth`

`PlantGrowth` has a `group` column with three levels: `ctrl`, `trt1`, `trt2`.

1. Load `PlantGrowth` and inspect its structure with `str()`.
2. What are the current levels of `group`? Which is the reference if you used this in a regression right now?
3. Reorder the levels so `ctrl` is first, then `trt1`, then `trt2`.
4. Use `fct_recode()` from `forcats` to rename `ctrl` → `"Control"`, `trt1` → `"Treatment 1"`, `trt2` → `"Treatment 2"`.
5. Calculate the mean and SD of `weight` for each group using `tapply()`.

::: {.callout-tip collapse="true"}
### Exercise 2: Solution

```{r}
library(forcats)

pg <- PlantGrowth

# 1-2. Inspect
str(pg)
levels(pg$group)  # "ctrl" "trt1" "trt2" — ctrl is reference (first level)

# 3. Reorder (already correct, but explicit is better)
pg$group <- factor(pg$group, levels = c("ctrl", "trt1", "trt2"))

# 4. Rename levels
pg$group <- fct_recode(pg$group,
  "Control"     = "ctrl",
  "Treatment 1" = "trt1",
  "Treatment 2" = "trt2"
)
levels(pg$group)

# 5. Summary stats
tapply(pg$weight, pg$group, mean)
tapply(pg$weight, pg$group, sd)
```
:::

### Exercise 3 (Open-ended)

Using a dataset from your own research (or any built-in R dataset), complete the following:

1. Identify one categorical column. Convert it to a factor and set the levels in a meaningful order (e.g., control before treatment, smallest dose before largest).
2. Identify one numeric column. Check for `NA` values with `sum(is.na())` and compute the mean with `na.rm = TRUE`.
3. Create a new derived column (e.g., a ratio, a log transform, or a binary category based on a threshold).
4. Write two sentences describing what the dataset contains and what you found from the structure inspection — as you would in a Data section of a manuscript.


## Comprehension Check

1. You create a vector with `c(1, 2, "three")`. What type is the result, and why?
2. What does `levels(my_factor)` return?
3. You have a column `treatment` with values `"A"`, `"B"`, `"C"`. In a regression, R uses `"A"` as the reference. How do you change the reference to `"B"`?
4. What is the difference between a data frame and a tibble?
5. You want the third row and second column of a data frame called `df`. What code do you write?

::: {.callout-note collapse="true"}
### Answers
1. Character. In R, if you mix types in a vector, everything is converted to the most general type. Numbers become character strings here (`"1"`, `"2"`, `"three"`).
2. The unique categories (levels) of the factor, in the order R will use them.
3. `treatment <- relevel(treatment, ref = "B")` or `treatment <- fct_relevel(treatment, "B")` using `forcats`.
4. A tibble is a modern version of a data frame from the `tidyverse`. It prints more neatly and never silently converts string columns to factors. For statistical analysis, they behave identically.
5. `df[3, 2]`
:::

## Further Reading

- [Tibbles in R for Data Science](https://r4ds.hadley.nz/tibbles.html) — Hadley Wickham's explanation of tibbles
- [forcats documentation](https://forcats.tidyverse.org/) — full guide to factor manipulation
- [R Data Structures cheat sheet](https://rstudio.github.io/cheatsheets/base-r.pdf) — quick reference for vectors, matrices, lists, and data frames

